### 2017-01-29 writeups/content.tex
Writeup on concurrency model added.

\section{Introduction}

In this document, I describe the concurrency model of dali, and ways
to realize it on single and distributed machines.

\section{The Branch Concurrency Model}

\begin{figure}[ht]

\includegraphics[scale=0.75]{Figures/sync-1}

\caption{The branch concurrency model with a simple two-level
hierarchy}
\label{fig:sync-1}
\end{figure}

A DaLi program consists of multiple concurrent threads operating on a
persistent shared state. The OCaml type of the shared state
distinguishes it from the thread-local state. Unlike imperative
programs, where the there exists a single copy of the shared state
that is \emph{updated} by concurrent threads, DaLi programs manage
shared state in a fashion similar to the version control software,
i.e., by \emph{forking} off new (concurrent) versions and
\emph{merging} them back. We refer to a totally-ordered sequence of
versions as a \emph{branch}. The program starts execution in its main
thread with an initial version of the shared state, which is the first
version of the \emph{master} branch. Certain operations performed by
the program (described later) result in the creation of later versions
on the master branch. Forking off new threads by the program results
in forking off new versions of the state along the new branches; each
thread is associated a new branch. In Fig.~\ref{fig:sync-1} for
example, main thread ($M$) operating on the master branch ($m$)
creates two new threads - $T_1$ and $T_2$, which fork off two new
branches - $t_1$ and $t_2$, respectively. The first versions on both
the branches, i.e., $t_1(1)$ and $t_2(1)$, denote the same state (same
as the forked version $m(1)$ on the master), yet the versions themselves
are deemed distinct and concurrent. Formally, $t_1(1) \neq t_2(1) \neq
m(1)$, but $\denot{t_1(1)} = \denot{t_2(1)} = \denot{m(1)}$.
% Note that the thread fork operation is blocking; the caller thread
% blocks until all the child threads finish execution and return.

The concurrency model allows threads to periodically synchronize their
branches with the master branch via the \C{sync\_next\_version}
operation. The operation is so named because it synchronizes the
immediate next version on the caller thread's branch with a version on
the master that is later than (or same as) the last master version
known to the branch. Concurrent threads that only serve read requests
are expected to periodically sync with the master to contain the
staleness of the data served. However, threads may also wish to
write new data and make it available globally. For this purpose,
\C{sync\_next\_version} allows the caller to (optionally) pass an
argument proposing the next version of the state. If there haven't
been any later versions on the master since the last known version,
the proposal is accepted and published as the next version on the
branch, and also the master. If there have been later master versions,
then latest such version is merged with the proposed version and
published. For this reason, applications are expected to implement a
merge function for the type of the state. In Fig.~\ref{fig:sync-1},
thread $T_2$ proposes a new version (the blue dot), which is merged
with the version $m(2)$ of the master (merge highlighted in the green
background), and the result is published as the version $t_2(2)$ on
the $t_2$ branch and $m(3)$ on the master branch. It may sometimes
transpire that the version proposed by the thread may be in conflict
with the latest version on the master, in which case the merge
function is allowed to return an error value and fail. This scenario
is captured in Fig.~\ref{fig:sync-1} in the red background. The
failure of merge results in the thread's proposal not being accepted,
and master's version being published as the next version on the
branch. In Fig.~\ref{fig:sync-1}, the failure of merge in $T_1$
results in master's $m(3)$ getting published as $t_1(2)$ on $t_1$.
Note that this version denotes the same state as present on the $t_2$
branch, which incorporates $T_2$'s proposal, thus demonstrating the
information flow among the branches.

The threads can access the lastest version on their branches via the
\C{get\_latest\_version} operation. If the last merge operation is a
success, then this operation returns the merged version, allowing the
thread to continue its computation. If the last merge operation
failed, then it allows the thread to read the latest version of the 
master and revise its proposal for the next \C{sync} operation.

The behavior of \C{sync\_next\_version} and \C{get\_latest\_version}
is more-or-less same in the main thread. While \C{sync} without a
propoisal is effectively a \C{nop} in the main thread, a \C{sync} with
a proposal tries to merge the proposal with the lastest master
version, which happens to be the local version for the main thread. As
with other threads, \C{sync} fails if the proposed version is
incompatible with the later version. There is however a slight
difference in behavior of the \C{get\_latest\_version} in the main
thread; unlike in other threads, successive calls to
\C{get\_latest\_version} in the main thread may return different
results even if they are not interleaved by a \C{sync} operation. This
happens if one of the child threads syncs its branch with the master
inbetween the two calls.

\subsection{Generalizing to branch hierarchies of any depth}

\begin{figure}[t]
\centering
\includegraphics[scale=0.75]{Figures/sync-2}

\caption{The branch concurrency model generalized to hierarchies of
any depth}
\label{fig:sync-2}
\end{figure}

Observe that the branch concurrency model desribed above confines
branching to a 2-level hierarchy with the master branch at the top and
forked branches at the bottom. The model does not define the behavior
of the version/thread fork operation originating from one of the child
threads. We now generalize the model for branch hierarchies of any
depth, while also eliminating the slight asymmetry between the main
thread and the child threads described above. We achieve this
generalization by redefining the \C{sync\_next\_version} operation as
a two-step process that first merges the proposed version with the
latest version on the current branch, and then merges the result with
the latest version on the parent branch (note: not necessarily the
master branch). The additional first step is now necessary because the
latest version on the current branch may not be same as the one that
was last read through the \C{get\_latest\_version}; a new version may
have since been committed by one of the child threads of the current
thread. The first step merges the proposed version with this version.
The second stepe merges the result of the first step with the lastest
version on the master, and publishes the merge result on both the
branches.

In Fig.~\ref{fig:sync-2}, thread $T_1$ forks a child $T_2$, which
forks $T_3$. The initial versions on the corresponding branches are
$t_1(1)$, $t_2(1)$, and $t_3(1)$, respectively. Thread $T_2$ starts a
local computation on its initial version copied into its heap (the red
dot). At some point, $T_3$ publishes a later version ($t_2(2)$) to its
parent. A little while later, $T_2$ concludes its local computation
and proposes a candidate for the new version on $t_2$ (the blue dot;
the dotted line connecting the red and blue dots denotes data
dependency). This candidate is merged with the latest local version
($t_2(2)$), and also the latest parent version ($t_1(2)$), and the
merge result\footnote{Convergence requires merge to be commutative, so
the order of merging doesn't matter.} is published as $t_2(3)$ on the
local branch and $t_1(3)$ on the parent branch. Subsequence
\C{get\_latest\_version} returns these versions on both the branches.

Observe that the behavior of \C{sync\_next\_version} and
\C{get\_latest\_version} is now uniform across all the threads. The
first and second steps (resp.) of \C{sync\_next\_version} are
effectively nop for the root and leaves of the thread hierarchy,
because the master branch doesn't have a parent and the bottom-most
branch doesn't have a child.
the main thread doesn't have a parent,

\subsection{Blocking sync only when necessary}

One problem with the previous model is that the call to
\C{sync\_later\_version} always blocks until the merge operation
succeeds and the result is published on both the branches. This is
unnecessary if the merge operation is always guaranteed to succeed (for
e.g., in the case of a mergeable queue). A non-blocking \C{sync} is
more appropriate for such cases. Selective non-blocking behavior for
can be accommodated in the semantics of \C{sync} as following: Along
with the proposal for the later version, \C{sync} also accepts the
current continuation of the caller. If there is a possiblity that any
of the two merge operations might fail, then \C{sync} waits until it
obtains the result of merges before calling the continuation with an
appropriate status.  Otherwise, it calls the current continuation
immediately, reporting success status. Some subsequent (but not
necessarily immediate) call to \C{get\_latest\_version} is guaranteed
to return the merged version\footnote{This guarantee is conditional on
the EC guarantee of the underlying machine and the commutativity
property of the merge function.}.

But how does the \C{sync} know if the merge might return conflict? By
observing the diff (\C{git diff}) of the proposed and the latest local
versions w.r.t to the last known version of the parent. For example,
let us consider an application whose shared state contains three data
items - \C{x}, \C{y}, and \C{z}, which are indepedently mergeable. Suppose
that the type of the merge function for \C{z} indicates a possibility of
conflict (\C{type 'a result = Ok of 'a | Conflict}), whereas the merge
functions for \C{x} and \C{y} do not admit that possibility. Let the last
know parent version be \C{\{x=a; y=b; z=c\}}, the latest local version
be \C{\{x=a'; y=b'; z=c\}}, and the proposed version be \C{\{x=a''; y=b;
z=c\}}. It is clear that the merge function on \C{z} will not be called
while merging the versions, hence there is not possibility that the
merge fails. Consequencely, \C{sync} can be non-blocking in this case.
On the other hand, if \C{z} is assigned \C{c'} in the proposed version,
then the merge on \C{z} will be called, which might fail. Therefore,
\C{sync} needs to block.

The banality of selective blocking optimization belies its true
importance, for the selectively blocking \C{sync} is what lets DaLi
implement a highly available replicated data store with on-demand
strong consistency, without requiring programmers to explicitly reason
about the consistency.

\section{Instantiations}

DaLi's concurrency model is general enough that it can be instantiated
in system-specific ways.

\subsection{Single-node database}

ToDo.

\subsection{Replicated data store}

ToDo.

### 2017-02-12 paper/dali.tex
Added paper.

%% For double-blind review submission
\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format should change 'acmlarge' to
%% 'sigplan,10pt'.

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\makeatletter\if@ACM@journal\makeatother
%% Journal information (used by PACMPL format)
%% Supplied to authors by publisher for camera-ready submission
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{1}
\acmArticle{1}
\acmYear{2017}
\acmMonth{1}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\else\makeatother
%% Conference information (used by SIGPLAN proceedings format)
%% Supplied to authors by publisher for camera-ready submission
\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
\acmYear{2017}
\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\fi

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission
\setcopyright{none}             %% For review submission
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

\begin{document}

%% Title information
\title[DaLi]{DaLi: Database as a Library}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{First1 Last1}
\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position1}
  \department{Department1}              %% \department is recommended
  \institution{Institution1}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{City1}
  \state{State1}
  \postcode{Post-Code1}
  \country{Country1}
}
\email{first1.last1@inst1.edu}          %% \email is recommended

%% Author with two affiliations and emails.
\author{First2 Last2}
\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position2a}
  \department{Department2a}             %% \department is recommended
  \institution{Institution2a}           %% \institution is required
  \streetaddress{Street2a Address2a}
  \city{City2a}
  \state{State2a}
  \postcode{Post-Code2a}
  \country{Country2a}
}
\email{first2.last2@inst2a.com}         %% \email is recommended
\affiliation{
  \position{Position2b}
  \department{Department2b}             %% \department is recommended
  \institution{Institution2b}           %% \institution is required
  \streetaddress{Street3b Address2b}
  \city{City2b}
  \state{State2b}
  \postcode{Post-Code2b}
  \country{Country2b}
}
\email{first2.last2@inst2b.org}         %% \email is recommended

%% Paper note
%% The \thanks command may be used to create a "paper note" ---
%% similar to a title note or an author note, but not explicitly
%% associated with a particular element.  It will appear immediately
%% above the permission/copyright statement.
%\thanks{with paper note}                %% \thanks is optional
                                        %% can be repeated if necesary
                                        %% contents suppressed with 'anonymous'

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Text of abstract \ldots.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords is optional

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\input{intro}

%% Acknowledgments
% \begin{acks}                            %% acks environment is optional
%                                         %% contents suppressed with 'anonymous'
%   %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
%   %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
%   %% acknowledge financial support and will be used by metadata
%   %% extraction tools.
%   This material is based upon work supported by the
%   \grantsponsor{GS100000001}{National Science
%     Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
%   No.~\grantnum{GS100000001}{nnnnnnn} and Grant
%   No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
%   conclusions or recommendations expressed in this material are those
%   of the author and do not necessarily reflect the views of the
%   National Science Foundation.
% \end{acks}

%% Bibliography
%\bibliography{bibfile}

%% Appendix
\appendix
\section{Appendix}

Text of appendix \ldots

\end{document}

### 2017-02-12 paper/intro.tex
Added paper.

\section{Introduction}

Distributed programs are challenging to write because they must juggle
two distinct mechanisms.  The first deals with application logic - how
do we define computation to be robust in the presence of disconnected
non-coherent communication among distributed threads of control?  The
second deals with system concerns - how do define notions of
atomicity, distribution, persistence, and replication?  Because of the
subtleties in dealing with the latter, especially when high
availability (low response latency) is desired, systems typically
separate these mechanisms into separate tools or services (e.g., CRDTs
have a well-defined semantics in an eventually consistent system that
may experience network partitions, while database and related storage
systems provide support for atomic transactions, persistence,
replication.  Unfortunately, because these different mechanisms have
vastly different semantics and goals, program structure becomes
complicated and muddled.  More importantly, it becomes problematic to
relate properties relevant to applications to the guarantees provided
by these lower-level services.  In this paper, we propose a radically
different vision of a distributed application that is predicated on
defining both mechanisms within the same language framework, with
those aspects not directly related to coherence and consistency
encapsulated within a library.  This not only simplifies program
reasoning, but leads to significant expressivity gains since
language-level types and abstractions are directly supported by the
library.   Realizing this vision leads to a radically different
programming model, one which realizes a monadic functional version
control system centered around data, rather than file coherence.  We
have implemented an end-to-end instantiation of this idea in OCaml
called DaLi and demonstrate how efficient distributed applications can
be written and deployed without having to deal with \emph{any}
system-centric features.

### 2017-02-17 paper/semantics.tex
OpSem added to the paper.

\section{Operational Semantics}

\input{opsem}

\begin{definition} [\bfseries Semantic Successor]
Version $v_2$ is a semantic successor of version $v_1$ under a history
$H$ (written $\under{H}{v_2 \semsucceq v_1}$) if and only if:
\begin{itemize}
  \item There exists a branch $b$ in $H$ (i.e., $\exists(t\in
  dom(H)).\,H(t) = b$) in which $v_2$ immediately succeeds
  $v_1$, but $v_2$ is not a result of a merge operation (i.e.,
  $v_2$'s tag is not \C{MERGE}), or
  \item There exists a branch $b$ in $H$ that contains
  $(v_2, \C{MERGE}\;(v_1,f_1)::b')$, for some $f_1$ and $b'$, or
  \item $v_2$ is a transitive semantic successor of $v_1$, i.e., 
  $\exists v.~ \under{H}{v \semsucceq v_1} \conj \under{H}{v_2
  \semsucceq v}$ 
\end{itemize}
\end{definition}

\begin{definition} [\bfseries Successor]
Version $v_2$ is a syntactic successor (or simply, successor) of
version $v_1$ under a history $H$ (written $\under{H}{v_2 \succeq
v_1}$) if and only if:
\begin{itemize}
  \item $v_2$ is a semantic successor of $v_1$, or
  \item $v_2$ is a result of a merge, and $v_2$ immediately succeeds
  $v_1$ in some branch $b$ in $H$, or
  \item $v_2$ is a transitive successor of $v_1$.
\end{itemize}
\end{definition}

### 2017-02-17 paper/semantics.tex
Last version that is made for Queue.

@@
 
-\begin{definition} [\bfseries Semantic Successor]
-Version $v_2$ is a semantic successor of version $v_1$ under a history
-$H$ (written $\under{H}{v_2 \semsucceq v_1}$) if and only if:
+\begin{definition} [\bfseries Semantic Ancestor]
+Version $v_1$ is a semantic ancestor of version $v_2$ under a history
+$H$ (written $\under{H}{v_1 \sempreceq v_2}$) if and only if:
 \begin{itemize}
@@
   $v_2$'s tag is not \C{MERGE}), or
+  \item There exists a branch $b$ in $H$, such that $b = (v_2, 
+  \C{FORK}\; (v_1,f_1)::b_1)::b'$, for some $f_1$, $b_1$ and $b'$, or
   \item There exists a branch $b$ in $H$ that contains
-  $(v_2, \C{MERGE}\;(v_1,f_1)::b')$, for some $f_1$ and $b'$, or
-  \item $v_2$ is a transitive semantic successor of $v_1$, i.e., 
-  $\exists v.~ \under{H}{v \semsucceq v_1} \conj \under{H}{v_2
-  \semsucceq v}$ 
+  $(v_2, \C{MERGE}\;(v_1,f_1)::b_1)$, for some $f_1$ and $b_1$, or
+  \item $v_1 = v_2$, or $v_1$ is transitively a semantic ancestor of
+  $v_2$, i.e., $\exists v.~ \under{H}{v_1 \sempreceq v} \conj
+  \under{H}{v \sempreceq v}$ 
 \end{itemize}
@@
 
-\begin{definition} [\bfseries Successor]
-Version $v_2$ is a syntactic successor (or simply, successor) of
-version $v_1$ under a history $H$ (written $\under{H}{v_2 \succeq
-v_1}$) if and only if:
+\begin{definition} [\bfseries Least Common Semantic Ancestor]
+Version $v$ is said to be a common semantic ancestor of versions $v_1$
+and $v_2$ under $H$ iff $\under{H}{v \sempreceq v_1}$ and $\under{H}{v
+\sempreceq v_2}$. It is said to be the least common semantic ancestor
+(LCSA) of $v_1$ and $v_2$, iff there does not exist a $v'$ such that
+$\under{H}{v' \sempreceq v_1}$ and $\under{H}{v' \sempreceq v_2}$ and
+$\under{H}{v \preceq v'}$.
+\end{definition}
+
+\begin{lemma}
+Semantic ancestor relation is a semi-lattice. For any two versions
+$v_1$ and $v_2$, there exists a unique LCSA, which is their join ($v_1
+\sqcup v_2$).
+\end{lemma}
+
+\begin{definition} [\bfseries Semantic Origin]
+Version $v_o$ is said to be the semantic origin of a branch $b$ under
+a history $H$ iff:
 \begin{itemize}
-  \item $v_2$ is a semantic successor of $v_1$, or
+  \item $v_o$ lies outside branch $b$, and
+  \item $v_o$ is a semantic ancestor of the latest version ($v$) of $b$
+  under $H$, and
+  \item $v_o$ is the least such version, i.e., every other semantic
+  ancestor ($v_o'$) of $v$ such that $v_o' \not\in b$ is also a
+  semantic ancestor of $v_o$.
+\end{itemize}
+\end{definition}
+
+\begin{definition} [\bfseries Mergeability]
+Versions $v_1$ and $v_2$ are mergeable under $H$ iff their least
+common ancestor is also their semantic ancestor. Branches $b_1$ and
+$b_2$ are mergeable under $H$ iff their latest versions are mergeable
+under $H$.
+\end{definition}
+
+\begin{definition} [\bfseries Ancestor]
+Version $v_1$ is a syntactic ancestor (or simply, ancestor) of
+version $v_2$ under a history $H$ (written $\under{H}{v_1 \preceq
+v_2}$) if and only if:
+\begin{itemize}
+  \item $v_1$ is a semantic ancestor of $v_2$, or
   \item $v_2$ is a result of a merge, and $v_2$ immediately succeeds
   $v_1$ in some branch $b$ in $H$, or
-  \item $v_2$ is a transitive successor of $v_1$.
+  \item $v_1 = v_2$ or $v_1$ is transitively an ancestor of $v_2$.
 \end{itemize}
 \end{definition}
+
+\begin{definition} [\bfseries Least Common Ancestor]
+Version $v$ is said to be a common ancestor of versions $v_1$ and
+$v_2$ under $H$ iff $\under{H}{v \preceq v_1}$ and $\under{H}{v
+\preceq v_2}$. It is said to be the least common ancestor (LCA) of
+$v_1$ and $v_2$, iff there does not exist a $v'$ such that
+$\under{H}{v' \preceq v_1}$ and $\under{H}{v' \preceq v_2}$ and
+$\under{H}{v \preceq v'}$.
+\end{definition}
+
+\begin{theorem} [\bfseries Unique LCA]
+Every pair of versions $v_1$ and $v_2$ in the history $H$ generated by
+the abstract machine of Fig.~\ref{fig:opsem} have a unique least
+common ancestor. More precisely, if every pair of versions in the
+initial history $H$ have a unique LCA, and if the abstract machine
+takes $H$ to $H'$, then every pair of versions in the final history
+$H'$ also have a unique LCA.
+\end{theorem}
+\begin{proof}
+By case analysis on the step relation. The \rulelabel{E-Run} case is
+trivial. The \rulelabel{E-Push} case creates a new version ($v$) with
+\C{PUSH} tag at the end of a branch. Let $v_1$ be the previous version
+on the branch, and let $v_2$ be any other version. The premise is that
+$v_1$ and $v_2$ have a unique LCA $v_0$. Observe that if $v_0 \neq
+v_1$, then LCA of $v$ and $v_2$ must also be an LCA of $v_1$ and
+$v_2$, hence LCA of $v$ and $v_2$ is unique. If $v_0 = v_1$, then LCA
+of $v$ and $v_2$ is $v_1$, hence unique. Thus, uniqueness of LCA is
+preserved. The reasoning for \rulelabel{E-Fork} is similar. The 
+\rulelabel{E-Pull} case is trivial. The \rulelabel{E-Pull-Wait} case
+is where the merge happens. 
+\end{proof}

### 2017-02-18 paper/semantics.tex
Locus definition

--- 2017-02-17 paper/semantics.tex
@@
 
-\begin{definition} [\bfseries Semantic Successor]
-Version $v_2$ is a semantic successor of version $v_1$ under a history
-$H$ (written $\under{H}{v_2 \semsucceq v_1}$) if and only if:
+\begin{definition} [\bfseries Ancestor]
+Version $v_1$ is a ancestor of version $v_2$ under a history
+$H$ (written $\under{H}{v_1 \preceq v_2}$) if and only if:
 \begin{itemize}
@@
   dom(H)).\,H(t) = b$) in which $v_2$ immediately succeeds
-  $v_1$, but $v_2$ is not a result of a merge operation (i.e.,
-  $v_2$'s tag is not \C{MERGE}), or
+  $v_1$, or
+  \item There exists a branch $b$ in $H$, such that $b = (v_2, 
+  \C{FORK}\; (v_1,f_1)::b_1)::b'$, for some $f_1$, $b_1$ and $b'$, or
   \item There exists a branch $b$ in $H$ that contains
-  $(v_2, \C{MERGE}\;(v_1,f_1)::b')$, for some $f_1$ and $b'$, or
-  \item $v_2$ is a transitive semantic successor of $v_1$, i.e., 
-  $\exists v.~ \under{H}{v \semsucceq v_1} \conj \under{H}{v_2
-  \semsucceq v}$ 
+  $(v_2, \C{MERGE}\;(v_1,f_1)::b_1)$, for some $f_1$ and $b_1$, or
+  \item $v_1 = v_2$, or $v_1$ is transitively a ancestor of
+  $v_2$, i.e., $\exists v.~ \under{H}{v_1 \preceq v} \conj
+  \under{H}{v \preceq v}$ 
 \end{itemize}
@@
 
-\begin{definition} [\bfseries Successor]
-Version $v_2$ is a syntactic successor (or simply, successor) of
-version $v_1$ under a history $H$ (written $\under{H}{v_2 \succeq
-v_1}$) if and only if:
-\begin{itemize}
-  \item $v_2$ is a semantic successor of $v_1$, or
-  \item $v_2$ is a result of a merge, and $v_2$ immediately succeeds
-  $v_1$ in some branch $b$ in $H$, or
-  \item $v_2$ is a transitive successor of $v_1$.
-\end{itemize}
+\begin{definition} [\bfseries Least Common Ancestor]
+Version $v$ is said to be a common ancestor of versions $v_1$
+and $v_2$ under $H$ iff $\under{H}{v \preceq v_1}$ and $\under{H}{v
+\preceq v_2}$. It is said to be the least common ancestor
+(LCA) of $v_1$ and $v_2$, iff there does not exist a $v'$ such that
+$\under{H}{v' \preceq v_1}$ and $\under{H}{v' \preceq v_2}$ and
+$\under{H}{v \preceq v'}$.
 \end{definition}
+
+\begin{definition} [\bfseries External Locus]
+Given a branch $b$ and a version $v\in b$, a version $v_o \not\in b$
+is an external locus of $v$ iff it is a least ancestor of $v_o$
+outside of $b$. That is, $\under{H}{v_o \preceq v}$, and there does
+not exist a $v_o' \not\in b$ such that $\under{H}{v_o' \preceq v}$ and
+$\under{H}{v_o \preceq v_o'}$. We lift the definition to the level of
+branches; $v_o$ is an external locus of a branch $b$ iff it is an
+external locus of its latest version.
+\end{definition}
+
+Note that all branches (except $H(t_{\top})$) begin with a \C{FORK}
+tag, hence every version has at least one external locus. We later
+prove that in the histories generated by the rules in
+Fig.~\ref{fig:opsem}, each version (hence, each branch) has exactly
+one external locus.
+
+\begin{definition} [\bfseries Mergeability]
+Given a history $H$, a version $v_1$ and a version $v_2$ that is not
+an ancestor of $v_1$ under $H$, $v_2$ is mergeable into $v_1$ (denoted
+$\under{H}{v_2 \mbleto v_1}$) iff $v_1$'s external locus is an
+ancestor of $v_2$.
+\end{definition}
+
+\begin{lemma} [\bfseries Unique External Locus]
+Every version $v$ (except the \C{INIT} version) in the history $H$
+generated by the rules in Fig.~\ref{fig:opsem} has a unique external
+locus.
+\end{lemma}
+\begin{proof}
+By case analysis on the reduction relation. In each case, we assume
+that the unique external locus property holds for the initial history,
+and prove that it holds for the final history.
+
+In \rulelabel{E-Run} case, initial version is the external locus of
+the new version on the new branch $H(t)$. In \rulelabel{E-Fork} case,
+which creates the branch $H(t')$ by forking from $H(t)$, the latest
+version on $H(t)$ is the external locus of the first version on
+$H(t)$. In \rulelabel{E-Push} case, the version on $H(t)$ has the same
+external locus as the previous latest version, which, premise says, is
+unique. No new version is created in \rulelabel{E-Pull}. In
+\rulelabel{E-Pull-Wait}, new version is created by merging the latest
+version ($v'$) on $H(t')$ into the latest version $v$ on $H(t)$.
+Premise says that $v$ has a unique external locus (call it $v_o$). The
+new version $v_m$ inherits $v_o$ as an external locus, and also has
+$v'$ as another external locus. However, since $\under{H}{v' \mbleto
+v}$, we have $\under{H}{v_o \preceq v'}$. Hence $v'$ is the only
+external locus.
+\qed
+\end{proof}
+
+\begin{theorem} [\bfseries Unique LCA]
+Every pair of versions $v_1$ and $v_2$ in the history $H$ generated by
+the rules in Fig.~\ref{fig:opsem} have a unique least
+common ancestor. 
+\end{theorem}
+\begin{proof}
+By case analysis on the step relation. In each case, we assume that
+the unique LCA property holds for the initial history, and prove that
+it holds for the final history.
+
+The \rulelabel{E-Run} case is trivial. The \rulelabel{E-Push} case
+creates a new version ($v$) with \C{PUSH} tag at the end of a branch.
+Let $v_1$ be the previous version on the branch, and let $v_2$ be any
+other version. The premise is that $v_1$ and $v_2$ have a unique LCA
+$v_0$. Observe that if $v_0 \neq v_1$, then LCA of $v$ and $v_2$ must
+also be an LCA of $v_1$ and $v_2$, hence LCA of $v$ and $v_2$ is
+unique. If $v_0 = v_1$, then LCA of $v$ and $v_2$ is $v_1$, hence
+unique. Thus, uniqueness of LCA is preserved. The reasoning for
+\rulelabel{E-Fork} is similar. The \rulelabel{E-Pull} case is trivial.
+
+The \rulelabel{E-Pull-Wait} case is where the merge happens. 
+\end{proof}

### 2017-02-18 paper/semantics.tex
Unique LCA theorem added.

@@
 
-\begin{definition} [\bfseries External Locus]
-Given a branch $b$ and a version $v\in b$, a version $v_o \not\in b$
-is an external locus of $v$ iff it is a least ancestor of $v_o$
-outside of $b$. That is, $\under{H}{v_o \preceq v}$, and there does
-not exist a $v_o' \not\in b$ such that $\under{H}{v_o' \preceq v}$ and
-$\under{H}{v_o \preceq v_o'}$. We lift the definition to the level of
-branches; $v_o$ is an external locus of a branch $b$ iff it is an
-external locus of its latest version.
+\begin{definition} [\bfseries Internal and External Ancestors]
+Given a branch $b$ and a version $v\in b$, an internal ancestor of $v$
+is an ancestor from the same branch $b$. An external ancestor of $v$
+is an ancestor from a different branch $b'\neq b$. 
+\end{definition}
+
+\begin{definition} [\bfseries Least Internal and External Ancestors]
+Given a branch $b$ and a version $v\in b$, least internal ancestor is
+the immediate ancestor version ($v_i$) of $v$ on $b$. Least external
+ancestor ($v_o$) is an external ancestor of $v$ that is not an
+ancestor of any other external ancestor of $v$. That is,
+$\under{H}{v_o \preceq v}$, and there does not exist a $v_o' \not\in
+b$ such that $\under{H}{v_o' \preceq v}$ and $\under{H}{v_o \preceq
+v_o'}$. 
+% We lift the definition to the level of branches; $v_o$ is an
+% external locus of a branch $b$ iff it is an external locus of its
+% latest version.
 \end{definition}
@@
 Note that all branches (except $H(t_{\top})$) begin with a \C{FORK}
-tag, hence every version has at least one external locus. We later
-prove that in the histories generated by the rules in
-Fig.~\ref{fig:opsem}, each version (hence, each branch) has exactly
-one external locus.
+tag, hence every version has at least one least external ancestor. We
+later prove that in the histories generated by the rules in
+Fig.~\ref{fig:opsem}, each version has exactly one least external
+ancestor. We call the unique least external ancestor of a version as
+its \emph{external locus}. Least internal ancestor, by definition, is
+unique, which we sometimes call \emph{internal locus}.
 
@@
 
-In \rulelabel{E-Run} case, initial version is the external locus of
+In \rulelabel{E-Run} case, initial version is the unique external locus of
 the new version on the new branch $H(t)$. In \rulelabel{E-Fork} case,
@@
 Premise says that $v$ has a unique external locus (call it $v_o$). The
-new version $v_m$ inherits $v_o$ as an external locus, and also has
-$v'$ as another external locus. However, since $\under{H}{v' \mbleto
-v}$, we have $\under{H}{v_o \preceq v'}$. Hence $v'$ is the only
-external locus.
+new version $v_m$ inherits $v_o$ as a least external ancestor, and
+also has $v'$ as another least external ancestor. However, since
+$\under{H}{v' \mbleto v}$, we have $\under{H}{v_o \preceq v'}$, we
+have $v'$ as the the external locus.
 \qed
 \end{proof}
+
+%% If we are allowed to merge earlier versions on branches, then we
+%% have a possibility of utmost two LCAs.
 
@@
 other version. The premise is that $v_1$ and $v_2$ have a unique LCA
-$v_0$. Observe that if $v_0 \neq v_1$, then LCA of $v$ and $v_2$ must
-also be an LCA of $v_1$ and $v_2$, hence LCA of $v$ and $v_2$ is
-unique. If $v_0 = v_1$, then LCA of $v$ and $v_2$ is $v_1$, hence
-unique. Thus, uniqueness of LCA is preserved. The reasoning for
+$v_0$. Observe that $v$'s internal locus is $v_1$ and its external
+locus is same as that of $v_1$. Hence, all ancestors of $v_1$
+(including itself) are also ancestors of $v$. Moreover, $v$ has no new
+ancestors except itself, which is the latest version being added.
+Hence, any common ancestor of $v_1$ and $v_2$ is also a common
+ancestor of $v$ and $v_2$, while there cannot be new common ancestors.
+Hence LCA of $v$ and $v_2$ is unique. The reasoning for
 \rulelabel{E-Fork} is similar. The \rulelabel{E-Pull} case is trivial.
 
-The \rulelabel{E-Pull-Wait} case is where the merge happens. 
+The \rulelabel{E-Pull-Wait} case is where the merge happens. Version
+$v_m$ is merging into version $v$ to produce the next version $v_m$.
+Let $v_2$ be any version. From premise, we know that $v$ and $v_2$
+have a unique LCA (call it $v_0$), and $v'$ and $v_2$ have a unique
+LCA (call it $v_1$). But, we know that $v_2$ has a unique external
+locus, which leads to the following cases:
+\begin{itemize}
+  \item Both $v_0$ and $v_1$ are external ancestors of $v_2$. Hence,
+  $\under{H}{v_0 \preceq v_1} \disj \under{H}{v_1 \preceq v_0}$.
+  Since\under{H}{v_0 \preceq v_m} \disj \under{H}{v_1 \preceq v_m}, 
+  $v_2$ and $v_m$ have a unique LCA.
+
+  \item $v_0$ is an internal ancestor of $v_2$, hence an external
+  ancestor of $v$, hence an external ancestor of $v_m$. $v_1$ is
+  an ancestor of $v'$, hence also an external ancestor of $v_m$. 
+\end{itemize}
+
+As proved
+before, the merged version ($v_m$) has a unique external locus, which
+is the merging version ($v'$).
 \end{proof}

### 2017-02-19 paper/intro.tex
Intro points added.

@@
 system-centric features.
+
+Points:
+\begin{itemize}
+  \item Language is often considered the pivot of the thought process.
+  Unsurprisingly, programmers are conditioned to reason in terms of
+  the language-level abstractions. However, reality of underlying
+  systems often force programmers to think in terms of low-level
+  abstractions that are counter-intuitive from the language
+  standpoint. Even functional programmers are often exposed to this
+  rather unfortunate state of affairs.
+
+  \item Consider, for example, an OCaml programmer who wishes to
+  manage an ordered set of elements in a binary tree, while using a
+  hashtable to quickly obtain references into the tree. While the
+  programmer enjoys high-level abstractions and safety guarantees of
+  OCaml insofar as the program only accesses in-memory data, she has
+  to let go of this convenience if the application requires data
+  persistence. She has to now re-engineer her application around the
+  data model enforced by the underlying storage engine (e.g.,
+  relational data model, or key-value data model), complicating
+  programming, and potentially compormising the language safety
+  guarantees in the process. For instance, while OCaml guarantees the
+  safety of all references, references between tables in a database
+  are unsafe, and have been known to lead to bugs.
+
+  \item Add replication to persistence, and the problems grow
+  manifold. The impossibility of obtaining strong consistency
+  guarantees without compromising scalability/availability under
+  decentralized replication means that applications can now exhibit
+  anomalous behaviors that are often incomprehensible to the
+  programmer. The standard artifacts of replication made available to
+  the programmer, namely mutable key-value pairs, expose the underlying
+  system reality in its full ugliness. While the programmer may manage
+  to reason about the contents of a key-value cell with much effort,
+  the reasoningly unfortunately does not compose; each key-value pair
+  can be manipulated independently, which makes it almost impossible
+  to obtain guarantees over a collection of data items in the same
+  vein as a high-level data structure (e.g., a binary search tree).
+  CRDTs are an attempt to provide high-level semantics of replication
+  for certain useful data types, however CRDTs do not come with a
+  an associated replicated storage system and a programming framework,
+  and programmers are again faced with the task of bridging the gap
+  between the high-level semantics of the CRDT to the abstractions and
+  guarantees offered by the underlying storage system. Moreover,
+  standard data structures need significant re-engineering to be
+  considered CRDTs (e.g., functions need to return \emph{effects} all
+  of which need to commute), and programmers receive little or no tool
+  support in this process. Moreover, CRDTs famously do not compose (in
+  the sense that the invariants over the composition cannot be
+  preserved), requiring programmers to resort to extraneous
+  mechanisms with system-specific properties, such as transactions.
+
+  \item In this paper we describe a library called \name that adresses
+  the aforementioned problems. \name lets programmers obtain add
+  persistence and fully-decentralized replication to their
+  applications without ever having to leave the comforts of high-level
+  reasoning in OCaml. Thus, in keeping with the long tradition of
+  Functional Programming, programmers are \emph{completely} shielded
+  from the reality of the underlying system implementation. This
+  elevation from the low-level reasoning to the high-level reasoning
+  concretely manifests as reasoning in terms of \emph{mergeable} data
+  types instead of \emph{replicated} data types; replication being a
+  system-level concern.
+
+\end{itemize}

### 2017-02-19 paper/intro.tex
Intro expanded.

@@
 
-  \item In this paper we describe a library called \name that adresses
-  the aforementioned problems. \name lets programmers obtain add
+  \item In this paper we describe an OCaml library called \name that
+  adresses the aforementioned problems. \name lets programmers add
   persistence and fully-decentralized replication to their
@@
   Functional Programming, programmers are \emph{completely} shielded
-  from the reality of the underlying system implementation. This
-  elevation from the low-level reasoning to the high-level reasoning
-  concretely manifests as reasoning in terms of \emph{mergeable} data
-  types instead of \emph{replicated} data types; replication being a
-  system-level concern.
+  from the complexities of the underlying system model. This elevation
+  from the low-level reasoning to the high-level reasoning concretely
+  manifests as reasoning in terms of \emph{mergeable} data types
+  instead of \emph{replicated} data types; replication being a
+  system-level concern. A mergeable data type library is a standard
+  OCaml data type library equipped with a \C{merge} function that
+  describes how to merge two concurrent versions of the data type that
+  originated from a single common ancestor. Thus, any OCaml data type
+  for which a meaningful merge function can be written is a mergeable
+  type. We give multiple examples of mergeable types useful in
+  practice, present a simple theory that guides programmers in writing
+  merge functions, and describe the tool support available to them in
+  this process. \name also comes with a novel programming model that
+  realizes a monadic functional version control system centered around
+  data, rather than file coherence. The \name monad lets programmers
+  write and compose concurrent computations around multiple versions
+  of a mergeable data type. The computation makes progress by forking
+  concurrent versions of the existing versions, creating later
+  versions of the existing versions along a \emph{branch}, or by
+  merging existing versions across the branches. However, as we shall
+  see, unconstrained branching may yield a branching structure that
+  becomes unmergeable, thus leading to \emph{stuck} states. An
+  important property of the \name library is that it suitably
+  constrains the branching structure to avoid stuck states without
+  introducing additional coordination that compromises the
+  decentralized nature of the computation. Futhermore, \name
+  profitably exploits the provenance information available via the
+  branching structure to make the application resilient to network
+  partitions. As a result, OCaml programmers get the benefits of
+  persistence and highly available replication, while they continue to
+  enjoy the comforts of high-level reasoning and the familiarity of
+  the standard library data types.
+
+  Our contributions are summarized below:
+  \begin{itemize}
+    \item We formally introduce the concept of a mergeable data type,
+    as an alternative to a replicated data type, to admit high-level
+    reasoning about replicated data in functional programs.
+
+    \item We describe the \name library that transparently adds
+    persistence and replication features to a mergeable type. \name
+    comes with a programming model that brings to bear the power and
+    flexibility of a version control system on administration of the
+    replicated data. \name however hides the complexity of version
+    control behind a monad, and exposes its functionality via a simple
+    API that lets programmers define and compose concurrent OCaml
+    computations around replicated data.
+
+    \item We present a formalization of the \name programming model,
+    and prove results that guarantee eventual convergence of
+    concurrent versions, and progress (stuck-freeness) guarantees.
+
+    \item We describe an implementation of \name atop the Irmin store,
+    and describe multiple examples, and two case studies that
+    attest to the practical utility of \name.
+  \end{itemize}
 

### 2017-02-20 paper/dali.tex
OpSem added to the paper.
Last version that is made for Queue.
Locus definition
Motivation.

@@
 %% Some recommended packages.
-\usepackage{booktabs}   %% For formal tables:
-                        %% http://ctan.org/pkg/booktabs
-\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
-                        %% http://ctan.org/pkg/subcaption
+\usepackage{mathtools}
+\usepackage{mathpartir}
+\usepackage{fullpage}
+\usepackage{ifpdf}
+\usepackage{graphicx}
+%\usepackage[usenames,dvipsnames]{color}
+\usepackage{subcaption}
+\usepackage{stmaryrd}
+%\usepackage[numbers]{natbib}
+\usepackage{amsthm}
+\usepackage{listings}          % format code
+\usepackage{xspace}
+
+% Math mode
+%-----------
+\newenvironment{nop}{}{}
+\newenvironment{smathpar}{
+\begin{nop}\small\begin{mathpar}}{
+\end{mathpar}\end{nop}\ignorespacesafterend}
+
+% Name
+%-----
+\newcommand{\name}{{\sf DaLi}\xspace}
+
+% Formatting
+%---------
+\newcommand{\C}[1]{\code{#1}}
+\newcommand{\tuplee}[1]{\langle #1 \rangle}
+\newcommand*{\rom}[1]{\expandafter\romannumeral #1}
+
+% Formatting commands
+% -------------------
+\newcommand{\code}[1]{\,{\tt #1}\,}
+\newcommand{\spc}[0]{\quad}
+\newcommand{\ALT}{~\mid~}
+\newcommand{\rel}[1]{{R}_{\mathit{#1}}}
+\newcommand{\conj}{~\wedge~}
+\newcommand{\disj}{~\vee~}
+\newcommand{\rulelabel}[1]{\textrm{\sc {#1}}}
+\newcommand{\ilrulelabel}[1]{{\sc #1}}
+\newcommand{\RULE}[2]{\frac{\begin{array}{c}#1\end{array}}
+                           {\begin{array}{c}#2\end{array}}}
+\newcommand{\denot}[1]{\llbracket #1 \rrbracket}
+\newcommand{\bind}{>\!\!>\!\!=}
+\newcommand{\coloneqq}{::=}
+\newcommand{\stepsto}{\longrightarrow}
+\newcommand{\return}[1]{\C{return}\;#1}
+\newcommand{\run}[2]{\C{run}\;#1\;#2}
+\newcommand{\fork}[1]{\C{fork}\;#1}
+\newcommand{\push}[1]{\C{push}\;#1}
+\newcommand{\pull}{\C{pull}}
+\newcommand{\semsucceq}{\succeq_{\circ}}
+\newcommand{\sempreceq}{\preceq_{\circ}}
+\newcommand{\under}[2]{#1\,\vdash\,#2}
+\newcommand{\mbleto}{\rightsquigarrow_m}
+
+% New colors
+%------------
+\definecolor{Bittersweet}{rgb}{1.0, 0.44, 0.37}
+
+% Listings Code
+%---------------
+\newcommand{\lstml}{
+\lstset{ %
+language=ML, % choose the language of the code
+basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
+keywordstyle=\color{Bittersweet},
+% numbers=left,                   % where to put the line-numbers
+numberstyle=\tiny,      % the size of the fonts that are used for the line-numbers
+stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
+numbersep=5pt,                  % how far the line-numbers are from the code
+showspaces=false,               % show spaces adding particular underscores
+showstringspaces=false,         % underline spaces within strings
+showtabs=false,                 % show tabs within strings adding particular underscores
+% frame=single,                   % adds a frame around the code
+tabsize=2,                      % sets default tabsize to 2 spaces
+captionpos=b,                   % sets the caption-position to bottom
+breaklines=true,                % sets automatic line breaking
+breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
+commentstyle=\itshape\color{MidnightBlue},
+%escapeinside={\%*}{*)},         % if you want to add a comment within your code
+morekeywords={module, @@deriving, not, : , /\\}
+}}
+\lstnewenvironment{ocaml}
+    { % \centering
+			\lstml
+      \lstset{}%
+      \csname lst@setfirstlabel\endcsname}
+    { %\centering
+      \csname lst@savefirstlabel\endcsname}
 
@@
 
+\input{motivation}
+
+\input{semantics}
+
 %% Acknowledgments

### 2017-02-20 paper/motivation.tex
Motivation.

\section{Motivation}

In this section, we motivate mergeable data types and \name
programming model via a series of examples.

\subsection{Counter}

\begin{figure}

\begin{subfigure}[b]{0.4\textwidth}
  \begin{ocaml}
    module Counter: sig
      type t
      val add: int -> t -> t
      val mult: int -> t -> t
      val read: t -> int
    end = struct
      type t = int
      let add x v = v + (abs x)
      let mult x v = v * (abs x)
      let read v = v
    end
  \end{ocaml}

\caption{\C{Counter} library in OCaml}
\label{fig:counter-adt}
\end{subfigure}
\begin{subfigure}[b]{0.56\textwidth}
  \begin{ocaml}
    module Counter: sig
      type t
      type eff = t -> t
      val add: int -> t-> eff
      val mult: int -> t -> eff
      val read: t -> int
    end = struct
      type t = int
      type eff = t -> t
      let add x v = fun v' -> v' + (abs x)
      let mult x v = 
        fun v' -> v' + v * (abs x - 1) 
      let read v = v
    end
  \end{ocaml}

\caption{\C{Counter} library re-engineered for op-based replication}
\label{fig:counter-rdt}
\end{subfigure}
% \begin{minipage}{0.62\textwidth}
%   \begin{ocaml}
%     module Counter = struct
%       include Counter
%       type t = counter [@@deriving persistence]
%       let merge old v1 v2 = old + (v1 - old) 
%                                 + (v2 - old)
%     end
%   \end{ocaml}
% \end{minipage}

% \caption{\C{Counter} library in OCaml equipped with a \C{merge}
% function. The \C{deriving} syntax~\cite{ppx-deriving} is processed by
% \name's meta-programming library to automatically derive a persistent
% variant of the \C{Counter}}

% \label{fig:counter}
\end{figure}

We will start with a simple example of a counter that nonetheless
demonstrates the benefits of our approach. An OCaml implementation of
the \C{Counter} library is shown in Fig.~\ref{fig:counter-adt}.
\C{Counter} supports two (update) operations - \C{add} and \C{mult} -
that let a non-negative integer value to be added to the counter, and
multiply the counter, respectively. Observe that the library is
written in an idiomatic style in OCaml; any OCaml programmer faced
with the task of implementing the counting functionality would
presumably write the same.

The counter implementation of Fig.~\ref{fig:counter-adt} serves the
application well as long as it is used in-memory on a single machine.
However, as the application grows in complexity and scales beyond a
single machine, its state may also need to be replicated and
persisted. Unfortunately, replication doesn't come for free, and
existing techniques require significant re-engineering of the
application and its libraries to support replication. A method often
adopted~\cite{crdts,gotsman-popl16} is to re-define the operations to
return \emph{effects} instead of values.  For instance, a \C{Counter}
operation returns a function of type $\C{Counter.t} \rightarrow
\C{Counter.t}$, which is then applied to the counter values at all
replicas to realize its effect.  Fig.~\ref{fig:counter-rdt} shows the
\C{Counter} library with operations re-defined to returns effects.
Such re-engineering has to be done carefully, lest an operation may
may end up causing an unintended effect at a remote replica. For
instance, while the effect of a \C{mult x v} operation may intuitively
seem to be \C{fun v' -> (abs x) * v'}, it is incorrect for it affects
different replicas differently, depending on the local counter value.
Fig.~\ref{fig:counter-rdt} therefore defines the result of \C{mult x
v} as an effect that increments the counter value by \C{v * (abs x -
1)} uniformly at all replicas. Likewise, accidentally defining the
effect of an \C{add x v} operation as \C{fun v' -> (abs x) + v}
results in replacing the counter value at a remote replica with the
one at origin replica, thereby losing updates at the remote replica.
Thus, reasoning about the correctness of effects often involves
appeals to system-specific artifacts, such as replicas, which don't
normally manifest in the program.  While checking the effects for
commutativity may help catch some of these oversights, commutativity
is a non-local property, and automatic verification is difficult in
general. Besides the possibility of new bugs, re-engineering libraries
to return effects has obvious software engineering problems in the
sense that it breaks backwards compatibility with the existing code.

### 2017-02-20 paper/motivation.tex
Counter eg modified.

@@
       type t
-      type eff = t -> t
+      type eff 
       val add: int -> t-> eff
       val mult: int -> t -> eff
+      val apply: eff -> t -> t
       val read: t -> int
@@
       type t = int
-      type eff = t -> t
-      let add x v = fun v' -> v' + (abs x)
-      let mult x v = 
-        fun v' -> v' + v * (abs x - 1) 
+      type eff = Add of int
+      let add x v = Add (abs x)
+      let mult x v = Add (v * (abs x - 1))
+      let apply (Add x) v = x + v
       let read v = v
@@
 
-\caption{\C{Counter} library re-engineered for op-based replication}
+\caption{\C{Counter} library re-engineered for effect-based replication}
 \label{fig:counter-rdt}
@@
 application and its libraries to support replication. A method often
-adopted~\cite{crdts,gotsman-popl16} is to re-define the operations to
-return \emph{effects} instead of values.  For instance, a \C{Counter}
-operation returns a function of type $\C{Counter.t} \rightarrow
-\C{Counter.t}$, which is then applied to the counter values at all
-replicas to realize its effect.  Fig.~\ref{fig:counter-rdt} shows the
+adopted~\cite{crdts, pldi15, gotsman-popl16} is to re-define the operations to
+return \emph{effects} instead of values.  An \emph{effect} is a tag that
+identifies the operation to be applied uniformly at all
+replicas to incorporate the effects of the original operaton. For
+instance, \C{Counter.add x v} operation returns \C{Add x} effect,
+which, when applied at a replica (see \C{apply}), adds \C{x} to the 
+local counter value.  Fig.~\ref{fig:counter-rdt} shows the
 \C{Counter} library with operations re-defined to returns effects.
@@
 may end up causing an unintended effect at a remote replica. For
-instance, while the effect of a \C{mult x v} operation may intuitively
-seem to be \C{fun v' -> (abs x) * v'}, it is incorrect for it affects
-different replicas differently, depending on the local counter value.
-Fig.~\ref{fig:counter-rdt} therefore defines the result of \C{mult x
-v} as an effect that increments the counter value by \C{v * (abs x -
-1)} uniformly at all replicas. Likewise, accidentally defining the
-effect of an \C{add x v} operation as \C{fun v' -> (abs x) + v}
-results in replacing the counter value at a remote replica with the
-one at origin replica, thereby losing updates at the remote replica.
-Thus, reasoning about the correctness of effects often involves
-appeals to system-specific artifacts, such as replicas, which don't
-normally manifest in the program.  While checking the effects for
+instance, one may incorrectly define \C{mult x v} operation as
+generating a \C{Mult x} effect, for it is easy to overlook the fact
+that \C{apply (Mult x)} affects different replicas differently
+depending on the counter value. Thus, reasoning about the correctness
+of effects often involves appeals to system-specific artifacts, such
+as replicas, which don't normally manifest in the program.  While
+checking the effects (or rather their application via \C{apply}) for
 commutativity may help catch some of these oversights, commutativity
-is a non-local property, and automatic verification is difficult in
-general. Besides the possibility of new bugs, re-engineering libraries
-to return effects has obvious software engineering problems in the
-sense that it breaks backwards compatibility with the existing code.
+is a non-local property, and automatically verifying it is difficult.
+Besides introducing the possibility of new bugs, re-engineering
+libraries to return effects has obvious software engineering problems
+in the sense that it breaks backwards compatibility with the existing
+code.
+
+Another downside with effect-generating replicated data type libraries
+is that they don't compose. For instance, let us say that an
+application wants to use two replicated counters, $c_1$ and $c_2$,
+bound by an invariant $c_2 \ge c_1$, which the application duly the
+invariant when updating $c_2$ and $c_1$. Nonetheless, it may still
+witness anamolous states that violate the invariant because updates to
+$c_1$ and $c_2$ are applied independently in any order. For example,
+if a replica increments both the counters by 1 each, the \C{read} at
+another replica, where $(c_1,c_2) = (2,2)$ may witness the increment
+to $c_1$, but not $c_2$, thus violating the invariant. The intention
+to atomically apply both the effects cannot be captured in the absence
+of external mechanisms to compose effects (such as effect-based
+transactions).

### 2017-02-20 paper/motivation.tex
Persistence

@@
 application wants to use two replicated counters, $c_1$ and $c_2$,
-bound by an invariant $c_2 \ge c_1$, which the application duly the
-invariant when updating $c_2$ and $c_1$. Nonetheless, it may still
+bound by an invariant $c_2 \ge c_1$, which the application duly
+enforces when updating $c_2$ and $c_1$. Nonetheless, it may still
 witness anamolous states that violate the invariant because updates to
@@
 of external mechanisms to compose effects (such as effect-based
-transactions).
+transactions~\cite{pldi15}).
+
+Persistence adds another dimension to the problem. Functional data
+structures are often large in-memory linked structures that admit
+functional updates by creating newer versions that share most of their
+structure with previous versions. Sharing is the key to efficiency,
+without which every functional update takes time that is at least
+linear in the size of the data structure. Unfortunately, the
+straightforward way to persist a data structure on disk in a
+machine-agnostic format (e.g., JSON) requires serialization, which is
+linear in the size of the data structure. An alternative is to
+maintain a separate on-disk copy of the data and administer it via a
+database system. The downside is that the programmer is now required
+to reason in terms of a lower-level abstraction (e.g., a relational
+data model) in order to maintain coherence between in-memory and
+on-disk representations. Safety guarantees over the in-memory data do
+not carry over to the disk, which introduces additional complications
+and the possibility of new bugs. Thus, adding persistence to
+applications while preserving high-level safety gurantees offered by
+the language abstraction is a non-trivial problem.

### 2017-02-20 paper/motivation.tex
DaLi counter added.

@@
 the language abstraction is a non-trivial problem.
+
+\name lets programmers add replication and persistence to their
+applications without having to re-engineer their abstractions or
+reason about lower-level artifacts. Here is how you write a mergeable
+counter in \name:
+\begin{figure}[!h]
+  \begin{ocaml}
+    module Counter = struct
+      include Counter
+      type mergeable_t = t [@@deriving persistence]
+      let merge old v1 v2 = old + (v1 - old) + (v2 - old)
+    end
+  \end{ocaml}
+\end{figure}

### 2017-02-21 paper/intro.tex
updates to intro

--- 2017-02-12 paper/intro.tex
@@
 
-Distributed programs are challenging to write because they must juggle
-two distinct mechanisms.  The first deals with application logic - how
-do we define computation to be robust in the presence of disconnected
-non-coherent communication among distributed threads of control?  The
-second deals with system concerns - how do define notions of
-atomicity, distribution, persistence, and replication?  Because of the
-subtleties in dealing with the latter, especially when high
-availability (low response latency) is desired, systems typically
-separate these mechanisms into separate tools or services (e.g., CRDTs
-have a well-defined semantics in an eventually consistent system that
-may experience network partitions, while database and related storage
-systems provide support for atomic transactions, persistence,
-replication.  Unfortunately, because these different mechanisms have
-vastly different semantics and goals, program structure becomes
-complicated and muddled.  More importantly, it becomes problematic to
-relate properties relevant to applications to the guarantees provided
-by these lower-level services.  In this paper, we propose a radically
-different vision of a distributed application that is predicated on
-defining both mechanisms within the same language framework, with
-those aspects not directly related to coherence and consistency
-encapsulated within a library.  This not only simplifies program
-reasoning, but leads to significant expressivity gains since
-language-level types and abstractions are directly supported by the
-library.   Realizing this vision leads to a radically different
-programming model, one which realizes a monadic functional version
-control system centered around data, rather than file coherence.  We
-have implemented an end-to-end instantiation of this idea in OCaml
-called DaLi and demonstrate how efficient distributed applications can
-be written and deployed without having to deal with \emph{any}
-system-centric features.
+Real-world distributed programs are challenging to write and maintain
+because they inevitably juggle and conflate two distinct mechanisms.
+The first concerns the expression of application logic - how do we
+define computations that are robust in the presence of distributed
+communication among concurrently executing threads of control?  The
+second deals with system concerns - how do we express notions of
+visibility, replication, and consistency when the nodes participating
+in such computations may be geographically distributed and the
+networks that connect them unreliable?  To simplify reasoning in such
+complicated environments, models often make strong assumptions on the
+guarantees provided by an implementation (such as
+serializablity~\cite{...} or strong consistency~\cite{...}) that may
+inadvertently mask important albeit unpleasant realities related to
+network partitions~\cite{...}, or which may ignore important system
+architectural features that assume replicated program state~\cite{...}
+to provide high availability and fault tolerance.  On the other hand,
+when these properties are explicitly exposed to the programmer, e.g.,
+by requiring that applications be written in terms of specialized
+distributed data structures~\cite{...} or control
+primitives~\cite{...}, simplicity, composability, and
+ease-of-reasoning suffers.
+
+To illustrate how these kinds of problems manifest in practice,
+consider how we might write a simple counter library (see
+Fig.~\ref{fig:counter-adt}).  A \C{Counter} supports two (update)
+operations - \C{add} and \C{mult} - that lets a non-negative integer
+value be added or multiplied to the counter, resp.  Observe that the
+library is written in an idiomatic style in OCaml, with no special
+reasoning principles needed to realize desired functionality.  As long
+as applications use the library on a single machine, this
+implementation suffices well.  If the library is used in the context
+of a more sophisticated application, say one whose computation is
+distributed among a collection of machines, its state may need to be
+replicated.  Unfortunately, replication doesn't come for free.  Global
+synchronous coordination among all nodes is problematic in distributed
+systems.  Thus, applying an \C{Add} operation on one node may not be
+instanteously witnessed on another, which may in turn be
+simultaneously attempting to perform its own \C{Add} or \C{Mult}
+action.  While synchronizing the activities of all nodes to ensure at
+most one such operation is performed at a time is impractical,
+designating a single node to hold the counter state eschewing
+replicating altogether (as in a client-server configuration), is also
+an undesirable solution, given its sensitivity to network partitions
+and server failures.  But, simply replicating counter state among
+nodes, without having any coordination protocol, is an equally
+infeasible approach.
+
+A method often adopted~\cite{crdts, pldi15, gotsman-popl16} to address
+this tension is to re-define a datatype's operations to return
+\emph{effects} instead of values.  An \emph{effect} is a tag that
+identifies the operation to be applied uniformly at all replicas to
+incorporate the effects of the original
+operaton. Fig.~\ref{fig:counter-rdt} shows the \C{Counter} library
+with operations re-defined to returns effects.  Observe that the
+\C{Counter.add x v} operation now returns an \C{Add x} effect, which,
+when applied at a replica (see \C{apply} in the figure), adds \C{x} to
+the local counter value.  Assuming effects are commutative, they can
+be processed asynchronously using a log local to each replica.  This
+implementation produces an eventually consistent global state in which
+all replicas will eventually contain the same counter value.  While
+effects provide a low-level operational basis for handling replicated
+state, the \emph{ad hoc} nature of the solution confounds desirable
+high-level reasoning principles.  Indeed, the semantic gap between the
+two versions of the counter, one cognizant of replication and the
+other not, breaks backward compatibility.  More significantly, its
+non-trivial construction must be developed in different guises for
+every distributed data structure used by the application.
+\begin{figure}
+\begin{subfigure}[b]{0.4\textwidth}
+  \begin{ocaml}
+    module Counter: sig
+      type t
+      val add: int -> t -> t
+      val mult: int -> t -> t
+      val read: t -> int
+    end = struct
+      type t = int
+      let add x v = v + (abs x)
+      let mult x v = v * (abs x)
+      let read v = v
+    end
+  \end{ocaml}
+\caption{\C{Counter} library in OCaml}
+\label{fig:counter-adt}
+\end{subfigure}
+\begin{subfigure}[b]{0.56\textwidth}
+  \begin{ocaml}
+    module Counter: sig
+      type t
+      type eff 
+      val add: int -> t-> eff
+      val mult: int -> t -> eff
+      val apply: eff -> t -> t
+      val read: t -> int
+    end = struct
+      type t = int
+      type eff = Add of int
+      let add x v = Add (abs x)
+      let mult x v = Add (v * (abs x - 1))
+      let apply (Add x) v = x + v
+      let read v = v
+    end
+  \end{ocaml}
+\caption{\C{Counter} library re-engineered for effect-based replication}
+\label{fig:counter-rdt}
+\end{subfigure}
+\end{figure}
+Lack of composability is another important downside of this approach.
+Consider an application that uses two replicated counters, $c_1$ and
+$c_2$, bound by the invariant $c_2 \ge c_1$, duly enforced by the
+application when updating $c_2$ or $c_1$.  An execution may
+nonetheless witness anamolous states that violate the invariant
+because updates to $c_1$ and $c_2$ are applied independently in any
+order on any replica.  For example, if a replica increments both
+counters, a \C{read} operation performed at another replica may
+witness the increment to $c_1$, but not $c_2$, thus violating the
+invariant. The intention to atomically apply both effects cannot be
+captured in the absence of external mechanisms to compose effects
+(such as effect-based transactions~\cite{pldi15}).
+
+On the other hand, a functional programmer is likely to approach the
+problems very differently.  Rather than thinking about a low-level
+operational variant of distributed state given in terms of explicit
+effects as described above, a declarative interpretation of a
+decentralized computation could be naturally described in terms of a
+version tree whose root contains the initial counter value, and whose
+branches represent different versions maintained by different nodes,
+with the state produced by a counter's operations on a node recorded
+along the node's local branch.  Now, a globally consistent view of a
+counter can be explained simply in terms of \emph{merge} operations of
+different branch tips.  Treating replication as merging yields a
+counter implementation that bears strong similarity to the original
+sequential one:
+  \begin{ocaml}
+    module Counter = struct
+      include Counter
+      type mergeable_t = t [@@deriving persistence]
+      let merge lca v1 v2 = lca + (v1 - lca) + (v2 - lca)
+    end
+  \end{ocaml}
+The role of \C{lca} (least common ancestor) here captures salient
+history - the state resulting from the merge of two versions derived
+from the same ancestor state should not unwittingly duplicate the
+contributions of the ancestor.  This interpretation of a replicated
+datatype is thus given in terms of multi-versions of a program state
+impicitly associated with the different nodes that comprise a
+distributed application and a logical notion of a \emph{master}
+representing some consistent, but not necessarily current, view of
+global state.
+\begin{figure}
+  \begin{center}
+    \includegraphics[scale=0.35]{Figures/dali-counter}
+  \end{center}
+  \caption{\small A multi-version counter can be expressed as a
+    sequence of versions managed by logically-distributed concurrently
+    executing threads.  Local versions produced by these threads
+    periodically merge their state with one another.  In the figure,
+    when the local version of Node 3, labeled 20, merges with the
+    local version of Node 1, whose counter state is 15 at the time, a
+    new counter value is produced.  This value takes into account the
+    previous merged state (10) from which the two versions were
+    derived to yield a new merged state of 25.  Observe that there is
+    no explicit synchronization or coordination among versions.  In
+    the figure, red circles represent state produced through merging,
+    and blue circles represent local counter state produced by
+    applying counter operations on a node-local version.  }
+\end{figure}
+
+The \name programming model realizes a monadic functional version
+control system centered around data, rather than file coherence.  The
+\name monad lets programmers write and compose concurrent computations
+around multiple (implicit) versions of a mergeable datatype.  A
+computation makes progress by forking concurrent versions of existing
+versions, creating later versions along a \emph{branch}, or by merging
+branch tips.  An important soundness property of the model is that
+instances of a replicated datatype whose merge operation is
+commutative with respect to its local version arguments is guaranteed
+to eventually converge to the same value on all replicas.  As a
+result, OCaml programmers get the benefits of highly available
+(low-latency) distributed computation, while continuing to enjoy the
+comforts of high-level reasoning and familiarity of standard library
+datatypes.  Notably, while \name's programming model makes no explitic
+reference to any specific operational manifestation of a distributed
+system (e.g., programmers do not need to worry about replicas and
+their management), it can be nonetheless efficiently realized on
+existing geo-replicated distributed systems.
+
+Our contributions are summarized below:
+
+\begin{itemize}
+    \item We formally introduce the concept of a \emph{mergeable} datatype
+      to admit high-level declarative reasoning about distributed
+      computation and replicated state in functional programs.
+
+    \item We describe the \name library that transparently adds
+      persistence and replication features to a mergeable type. \name
+      comes with a programming model that brings to bear the power and
+      flexibility of a version control system to the administration of
+      replicated data. \name hides the complexity of version control
+      behind a monad, and exposes its functionality via a simple API
+      that lets programmers define and compose concurrent OCaml
+      computations around such data.  Issues related to replication
+      only manifest implicitly in the definition of a merge operation
+      that defines coherence among different instances of replicated
+      state.
+
+    \item We present a formalization of the \name programming model,
+      and establish the comditions under which eventual convergence of
+      concurrent version state can be guaranteed.
+
+    \item We describe an implementation of \name atop the Irmin
+      persistent store~\cite{...}, a content-addressable storage
+      library for OCaml, along with case studies and experimental
+      results to its practical utility.
+\end{itemize}
+
+%% Persistence adds yet another dimension to the problem. Functional data
+%% structures are often large in-memory linked structures that admit
+%% functional updates by creating newer versions that share most of their
+%% structure with previous versions. Sharing is the key to efficiency,
+%% without which every functional update takes time that is at least
+%% linear in the size of the data structure. Unfortunately, the
+%% straightforward way to persist a data structure on disk in a
+%% machine-agnostic format (e.g., JSON) requires serialization, which is
+%% linear in the size of the data structure. An alternative is to
+%% maintain a separate on-disk copy of the data and administer it via a
+%% database system. The downside is that the programmer is now required
+%% to reason in terms of a lower-level abstraction (e.g., a relational
+%% data model) in order to maintain coherence between in-memory and
+%% on-disk representations. Safety guarantees over the in-memory data do
+%% not carry over to the disk, which introduces additional complications
+%% and the possibility of new bugs. Thus, adding persistence to
+%% applications while preserving high-level safety gurantees offered by
+%% the language abstraction is a non-trivial problem.
+
+%% However, as we shall see,
+%% unconstrained branching may yield a branching structure that becomes
+%% unmergeable, thus leading to \emph{stuck} states. An important
+%% property of the \name library is that it guarantees progress even in
+%% the presence of arbitrary merges provided the merge operation is
+%% commutative with respect to its version state arguments.  Furthermore,
+%% \name profitably exploits provenance information available via
+%% branches to make application resilient to network partitions.  
+%% In this paper, we define a semantics of mergeable datatypes that lets
+%% any OCaml program be seamlessly deployed in a distributed environment
+%% with implicit support for replicated and decentralized operation.  The
+%% underlying foundation for merge actions has a natural category
+%% theoretic interpretation that enables us to construct useful mergeable
+%% variants for \emph{any} OCaml type whose states are commutative wtih
+%% respect to the merge operation defie

### 2017-02-21 paper/model.tex
Programming Model.

\section{Programming Model}

\begin{figure}

  \begin{ocaml}
  module type DALI = sig
    type 'a 'b t
    val return : 'b -> 'a 'b t
    val bind : 'a 'b t -> ('b -> 'a 'c t) -> 'a 'c t
    val get_current_version: unit -> 'a 'a t
    val with_init_version_do: 'a -> 'a 'b t -> 'b
    val fork_version : 'a 'b t -> 'a unit t
    val sync_next_version: unit -> ?v:'a -> 'a 'a t
  end
  \end{ocaml}

\label{fig:dali-monad}
\caption{Signature of the \name monad}
\end{figure}

\begin{figure}

  \begin{ocaml}
  module type CANVAS = sig
    type pixel = {r:char; g:char; b:char}
    type tree = 
      | N of pixel
      | B of {tl: tree; tr: tree; bl: tree; br: tree} 
    type t = {max_x:int; max_y:int; canvas:tree} [@@deriving versioned]
    type loc = {x:int; y:int}

    val new_canvas: int -> int -> t
    val set_px: t -> loc -> pixel -> t
    val get_px: t -> loc -> pixel
    val merge: (* lca *)t -> (* v1 *)t -> (* v2 *)t -> t
  end
  \end{ocaml}

\label{fig:canvas-sig}
\caption{\drawsome: a sample \name application}
\end{figure}

In this section, we describe the \name programming model through the
example of a collaborative drawing application we call \drawsome.

Fig.~\ref{canvas-sig} shows the signature of the \drawsome
application. \drawsome represents a free-hand drawing canvas as a tree
of quadrants, where each quadrant is simply a leaf node unless it has
pixels of different colors. Quadrants are expanded into a tree
structures as an when pixels are colored (the color scheme used is
rgb). The representation is thus optimized for sparse canvases, such
as whiteboards. The application supports three simple operations:
creating a new canvas, setting the pixel at a specified coordinate,
and returning the pixel at a given coordinate. The \C{merge} function
and \C{deriving versioned} directive are explained below.

\drawsome lets multiple users collaborate on a canvas that is
conceptually shared among all the collaborators. Under a shared-memory
abstraction, there would be a single copy of the canvas that is
updated concurrently by multiple clients such that from the
perspective of any single client, the canvas would change without any
explicit intervention. \name ascribes functional semantics to sharing
by letting each client work on its own version of the state (a data
structure), and later merging the concurrent versions on-demand. Thus,
the primary artifact of the \name programming model is a versioned
data structure. The library operates on a representation of versioned
data structures optimized for persistence on disk
(Sec.~\ref{sec:persistence}). \name's meta-programming component
automatically synthesizes this representation, along with the
functions that translate between representations, for the data type
definitions marked with the ppx~\cite{ppx} directive \C{deriving
versioned}. Concretely, for the \C{Canvas} module, \name synthesizes
a \C{Canvas.Versioned} module with a type \C{t}, and functions
\C{of\_canvas} and \C{to\_canvas} of types \C{Canvas.t $\rightarrow$
t} and \C{t $\rightarrow$ Canvas.t}, respectively.

\name requires a three-way \C{merge} function to merge the concurrent
versions of a drawing canvas. The three arguments include two
concurrent versions (\C{v1} and \C{v2}), and their least common
ancestor (\C{lca}) - the version from which the two concurrent
versions evolved independently. The merge function can make use of the
pixel values of the common ancestor to merge the pixel values on both
the canvases. For instance, if \C{v1} colors the pixel at $(0,0)$ red,
and \C{v2} colors it green, and the color of the pixel in \C{lca} is
red, then it means that only \C{v2} modified the color. Hence the
pixel is colored green in the merged canvas. On the other hand, if the
pixel is colored white in the original canvas, then it means that both
\C{v1} and \C{v2} have modified the color. In such case, an appopriate
color mixing algorithm can be used to determine the color of pixel.
For instance, the pixel can be colored yellow - an additive
combination of red and green.

The \name programming model lets programmers define and compose
concurrent computations around versioned data structures.
Fig.~\ref{fig:dali-monad} shows the signature of the \name module that
implements the programming model along the lines of the well-known
\C{State} monad~\cite{wadler-monad}. The monad encapsulates a
versioned functional state (\C{'a}) and the type \C{'a 'b t}
represents a monadic computation that returns a \C{'b} result.
Functions \C{return} and \C{bind} have their usual monadic
interpretation. \C{get\_current\_version} is like the \C{State}
monad's \C{get}; it returns the versioned state encapsulated by the
monad. \C{with\_init\_version\_do} runs a monadic computation against
a given initial version and returns the result. \C{fork\_version}
returns a computation that forks a new concurrent version from the
current version, and runs the given monadic computation asynchronously
against the forked version.  The semantics of \C{sync\_next\_version}
(simply called \C{sync}) is somewhat involved: the function accepts a
\emph{proposal} for the next version of the state, where \C{next} is
defined w.r.t the current version, and returns (via a monad) the
actual next version, which becomes the current version for rest of the
computation.  The next version is created by merging the proposal with
a subset of concurrent versions that became available since the last
merge or fork. Thus, \C{sync} effectively lets a computation sync with
the concurrent computations and obtain their latest updates.

\begin{figure}

\caption{\drawsome: a sample collaborative }
\label{fig:dali-canvas}
\end{figure}

### 2017-02-21 paper/model.tex

@@
     val fork_version : 'a 'b t -> 'a unit t
-    val sync_next_version: unit -> ?v:'a -> 'a 'a t
+    val sync ()_next_version: unit -> ?v:'a -> 'a 'a t
   end
@@
 
+\begin{figure}
+\centering
+\includegraphics[scale=0.5]{Figures/canvas-merging}
+
+\caption{Merging concurrent versions (\C{v1} and \C{v2} ) of a drawing
+canvas. \C{lca} is their common ancestor.}
+\label{fig:canvas-merging}
+\end{figure}
+
 \drawsome lets multiple users collaborate on a canvas that is
@@
 pixel values of the common ancestor to merge the pixel values on both
-the canvases. For instance, if \C{v1} colors the pixel at $(0,0)$ red,
-and \C{v2} colors it green, and the color of the pixel in \C{lca} is
-red, then it means that only \C{v2} modified the color. Hence the
+the canvases. For instance, if the color of a pixel in \C{v1} is white,
+and in \C{v2} it is green, and its color in \C{lca} is
+white, then it means that only \C{v2} modified the color. Hence the
 pixel is colored green in the merged canvas. On the other hand, if the
-pixel is colored white in the original canvas, then it means that both
+pixel is red in \C{v1}, then it means that both
 \C{v1} and \C{v2} have modified the color. In such case, an appopriate
-color mixing algorithm can be used to determine the color of pixel.
+color-mixing algorithm can be used to determine the color of pixel.
 For instance, the pixel can be colored yellow - an additive
-combination of red and green.
+combination of red and green. The logic is illustrated in
+Fig.~\ref{fig:canvas-merging}.
 
@@
 returns a computation that forks a new concurrent version from the
-current version, and runs the given monadic computation asynchronously
-against the forked version.  The semantics of \C{sync\_next\_version}
-(simply called \C{sync}) is somewhat involved: the function accepts a
+current version, and runs the given monadic computation async ()hronously
+against the forked version.  The semantics of \C{sync ()\_next\_version}
+(simply called \C{sync ()}) is somewhat involved: the function accepts a
 \emph{proposal} for the next version of the state, where \C{next} is
@@
 a subset of concurrent versions that became available since the last
-merge or fork. Thus, \C{sync} effectively lets a computation sync with
-the concurrent computations and obtain their latest updates.
+merge or fork. Thus, \C{sync ()} effectively lets a computation sync
+() with a subset of concurrent computations and obtain their latest
+updates.
 
 \begin{figure}
+\centering
+\begin{tabular}{l||l||l}
+\begin{ocaml}
+let alice_f : C.t unit t = 
+  get () >>= fun c0 -> 
+  fork bob_f >>= fun () ->
+  let c0' = C.draw_line c0 
+    {x=10;y=20}
+    {x=40;y=20} in
+  sync () ~v:c0' >>= 
+  fun c1 ->
+  let c1' = C.draw_line c1 
+    {x=10;y=40} 
+    {x=40;y=40} in
+  sync () ~v:c1' >>= 
+  fun c2 -> return ()
+\end{ocaml}
+&
+\begin{ocaml}
+let bob_f : C.t unit t = 
+  get () >>= fun c0 -> 
+  fork cheryl_f >>= 
+  fun () ->
+  let c0' = C.draw_line c0 
+    {x=10;y=20} 
+    {x=10;y=40} in
+  sync () ~v:c0' >>= 
+  fun c1 -> sync () >>= 
+  fun c2 -> return ()
+\end{ocaml}
+&
+\begin{ocaml}
+let cheryl_f : C.t unit t = 
+  get () >>= fun c0 -> 
+  let c0' = C.draw_line c0 
+    {x=40;y=20} 
+    {x=40;y=20} in
+  sync () ~v:c0' >>= 
+  fun c1 -> sync () >>= 
+  fun c2 -> return ()
+\end{ocaml}
+\\
+\end{tabular}
+\caption{\drawsome: A collaborative drawing session between Alice,
+Bob, and Cheryl}
+\label{fig:canvas-sessions-code}
+\end{figure}
 
-\caption{\drawsome: a sample collaborative }
-\label{fig:dali-canvas}
+Fig.~\ref{fig:canvas-sessions} shows a collaborative editing session
+between Alice, Bob and Cheryl.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.6]{Figures/canvas-sessions}
+
+\caption{\drawsome: Collaborative drawing session visualized}
+\label{fig:canvas-sessions}
 \end{figure}

### 2017-02-21 paper/model.tex
Model.

@@
   let c0' = C.draw_line c0 
-    {x=10;y=20}
-    {x=40;y=20} in
+    {x=0;y=0}
+    {x=4;y=0} in
   sync () ~v:c0' >>= 
@@
   let c1' = C.draw_line c1 
-    {x=10;y=40} 
-    {x=40;y=40} in
+    {x=0;y=4} 
+    {x=4;y=4} in
   sync () ~v:c1' >>= 
@@
   let c0' = C.draw_line c0 
-    {x=10;y=20} 
-    {x=10;y=40} in
+    {x=0;y=0} 
+    {x=0;y=4} in
   sync () ~v:c0' >>= 
@@
   let c0' = C.draw_line c0 
-    {x=40;y=20} 
-    {x=40;y=20} in
+    {x=4;y=0} 
+    {x=4;y=0} in
   sync () ~v:c0' >>= 
@@
 
-Fig.~\ref{fig:canvas-sessions} shows a collaborative editing session
-between Alice, Bob and Cheryl.
-
 \begin{figure}
@@
 \end{figure}
+
+Fig.~\ref{fig:canvas-sessions-code} demonstrates how a collaborative
+drawing session between Alice, Bob and Cheryl can be composed using
+\name. A possible execution of the session is visualized in
+Fig.~\ref{fig:canvas-sessions}. Assume that the session starts with
+Alice on a $5\times 5$ blank canvas, as shown below:
+\begin{ocaml}
+  module C = Canvas;; open Dali;;
+  with_init_version_do (C.new_blank 5 5) alice_f 
+\end{ocaml}
+Alice starts by reading the current version of the canvas, which is
+blank. She then invites Bob for collaboration by forking a new
+concurrent version for Bob. Bob in turn invites Cheryl for
+collaboration. All three of them start working on the blank canvas.
+Alice draws a red horizontal line from $(0,0)$ (top-left) to $(4,0)$
+(top-right) using \C{C.draw\_line} (definition not shown; can be
+constructed using \C{set\_px}). Meanwhile, Bob draws a green vertical
+line from $(0,0)$ to $(0,4)$, and Cheryl draws a similar line from
+$(4,0)$ to $(4,4)$. All three of them call \C{sync} with their
+respective proposals ($C_0'$). While any partial ordering of
+concurrent \C{syncs} is valid, we consider a linear order where
+Cheryl's \C{sync} happens first, followed by Bob's and then Alice's.
+Cheryl's \C{sync} does not find any concurrent versions, hence
+installs the proposed version ($C_0'$) as the next version on Cheryl's
+branch. Bob's \C{sync} finds Cheryl's $C_0'$ as a concurrent version,
+and merges it with its proposal to produce the next version $C_1$,
+which is installed on Bob's branch.  and returned by \C{sync} (\C{c1}
+in second column of Fig.~\ref{fig:canvas-sessions-code}). The least common
+ancestor (LCA) for this merge is the initial version ($C_0$), and the two
+concurrent versions are Bob's $C_0'$ and Cheryl's $C_0'$. Next,
+Alice's \C{sync} finds Cheryl's $C_0'$ and Bob's $C_1$ as concurrent
+versions, and merges them successively with Alice's proposal. For the
+first merge, the two concurrent versions are Alice's $C_0'$ and
+Cheryl's $C_0'$, and the LCA is the initial version
+($C_0$). The result of this merge is installed as the next version
+($\C{C_0''}$) on Alice's branch. For the next merge, the two
+concurrent versions are Alice's $C_0''$ and Bob's $C_1$ and the LCA is
+Cheryl's $C_0'$\footnote{Thus, LCA of versions on two branches can lie
+outside both the branches.}. The result ($C_1$) becomes the next
+version on Alice's branch, and the return value of Alice's \C{sync}.
+Next, Alice draws a red horizontal line from $(0,4)$ to $(4,4)$,
+proposes this canvas (\C{c1'} in the first column of
+Fig.~\ref{fig:canvas-sessions-code}) as the next version to \C{sync}.
+Since there are no concurrent versions, $C_1'$ becomes the next
+version and the return value of \C{sync}. The subsequent \C{sync}
+operations from Bob and Cheryl propose no new versions, hence simply
+obtain Alice's $C_1'$ as next versions.

### 2017-02-21 paper/model.tex
Model done.

@@
     val fork_version : 'a 'b t -> 'a unit t
-    val sync ()_next_version: unit -> ?v:'a -> 'a 'a t
+    val sync_next_version: unit -> ?v:'a -> 'a 'a t
   end
@@
 returns a computation that forks a new concurrent version from the
-current version, and runs the given monadic computation async ()hronously
-against the forked version.  The semantics of \C{sync ()\_next\_version}
-(simply called \C{sync ()}) is somewhat involved: the function accepts a
-\emph{proposal} for the next version of the state, where \C{next} is
-defined w.r.t the current version, and returns (via a monad) the
-actual next version, which becomes the current version for rest of the
-computation.  The next version is created by merging the proposal with
-a subset of concurrent versions that became available since the last
-merge or fork. Thus, \C{sync ()} effectively lets a computation sync
-() with a subset of concurrent computations and obtain their latest
-updates.
+current version, and runs the given monadic computation asynchronously
+against the forked version.  The semantics of \C{sync\_next\_version}
+(simply called \C{sync}) is somewhat involved: the function accepts
+a \emph{proposal} for the next version of the
+state, where \C{next} is defined w.r.t the current version, and
+returns (via a monad) the actual next version, which becomes the
+current version for rest of the computation.  The next version is
+created by merging the proposal with a subset of concurrent versions
+that became available since the last merge or fork. Thus, \C{sync}
+effectively lets a computation sync with a subset of concurrent
+computations and obtain their latest updates.
 
@@
 obtain Alice's $C_1'$ as next versions.
+
+The \drawsome example demonstrates the utility of mergeable data types
+and the \name programming model in building concurrent applications
+with conceptual sharing of state. This example, however, implicitly
+makes certain assumptions about the model and the underlying system,
+such as the existence of a single least common ancestor (LCA) for any
+pair of versions, the ability to access a previous version on any
+branch, and mergeability of any two concurrent versions. Enforcing
+these guarantees in a fully decentralized distributed setting requires
+addressing non-trivial theoretical challenges described in the next
+section. Unlike the canvas, the process of writing merge functions for
+data types with non-trivial invariants could be quite involved. The
+later sections describe a principled approach to derive a merge
+function for a data type given its invariants. Like \C{Canvas.merge},
+merge functions often need to compare multiple versions of a data type
+for structural equality, which, when performed naively, incurs cost
+linear in the size of versions. The representation of a versioned data
+type in \name is optimized for performing comparisons such that
+structural equality can be determined in constant time. Such
+engineering details, which contribute to the practicality of our
+programming model are also discussed in subsequent sections.

### 2017-02-21 paper/model.tex
Model edits.

@@
 
+\caption{\drawsome: a sample \name application}
 \label{fig:canvas-sig}
-\caption{\drawsome: a sample \name application}
 \end{figure}
@@
 
-Fig.~\ref{canvas-sig} shows the signature of the \drawsome
+Fig.~\ref{fig:canvas-sig} shows the signature of the \drawsome
 application. \drawsome represents a free-hand drawing canvas as a tree
@@
 linear in the size of versions. The representation of a versioned data
-type in \name is optimized for performing comparisons such that
-structural equality can be determined in constant time. Such
+type in \name is optimized for performing comparisons that allows
+structural equality to be determined in constant time. Such
 engineering details, which contribute to the practicality of our

### 2017-02-23 paper/intro-sj.tex
edits

--- 2017-02-21 paper/intro.tex
@@
 consider how we might write a simple counter library (see
-Fig.~\ref{fig:counter-adt}).  A \C{Counter} supports two (update)
-operations - \C{add} and \C{mult} - that lets a non-negative integer
-value be added or multiplied to the counter, resp.  Observe that the
-library is written in an idiomatic style in OCaml, with no special
-reasoning principles needed to realize desired functionality.  As long
-as applications use the library on a single machine, this
-implementation suffices well.  If the library is used in the context
-of a more sophisticated application, say one whose computation is
-distributed among a collection of machines, its state may need to be
-replicated.  Unfortunately, replication doesn't come for free.  Global
-synchronous coordination among all nodes is problematic in distributed
-systems.  Thus, applying an \C{Add} operation on one node may not be
-instanteously witnessed on another, which may in turn be
-simultaneously attempting to perform its own \C{Add} or \C{Mult}
-action.  While synchronizing the activities of all nodes to ensure at
-most one such operation is performed at a time is impractical,
-designating a single node to hold the counter state eschewing
-replicating altogether (as in a client-server configuration), is also
-an undesirable solution, given its sensitivity to network partitions
-and server failures.  But, simply replicating counter state among
-nodes, without having any coordination protocol, is an equally
-infeasible approach.
-
-A method often adopted~\cite{crdts, pldi15, gotsman-popl16} to address
-this tension is to re-define a datatype's operations to return
-\emph{effects} instead of values.  An \emph{effect} is a tag that
-identifies the operation to be applied uniformly at all replicas to
-incorporate the effects of the original
-operaton. Fig.~\ref{fig:counter-rdt} shows the \C{Counter} library
-with operations re-defined to returns effects.  Observe that the
-\C{Counter.add x v} operation now returns an \C{Add x} effect, which,
-when applied at a replica (see \C{apply} in the figure), adds \C{x} to
-the local counter value.  Assuming effects are commutative, they can
-be processed asynchronously using a log local to each replica.  This
-implementation produces an eventually consistent global state in which
-all replicas will eventually contain the same counter value.  While
-effects provide a low-level operational basis for handling replicated
-state, the \emph{ad hoc} nature of the solution confounds desirable
-high-level reasoning principles.  Indeed, the semantic gap between the
-two versions of the counter, one cognizant of replication and the
-other not, breaks backward compatibility.  More significantly, its
-non-trivial construction must be developed in different guises for
-every distributed data structure used by the application.
+Fig.~\ref{fig:counter-adt}).
 \begin{figure}
@@
   \end{ocaml}
+
 \caption{\C{Counter} library in OCaml}
@@
 \end{subfigure}
-\end{figure}
+A \C{Counter} supports two (update) operations - \C{add} and \C{mult}
+- that lets a non-negative integer value be added or multiplied to the
+counter, resp.  Observe that the library is written in an idiomatic
+style in OCaml, with no special reasoning principles needed to realize
+the desired functionality.  As long as applicatons use the library on
+a single machine, this implementation suffices well.  If the library
+is used in the context of a more sophisticated application,
+distributed among a collection of machines, its state may need to be
+replicated.  Unfortunately, replication doesn't come for free.  In
+such a context, global coordination is problematic, and applying an
+\C{Add} operation on one node may not be instanteously witnessed on
+another, which may in turn be simultaneously attempting to perform its
+own \C{Add} or \C{Mult} action.  While synchronizing the activities of
+all nodes to ensure at most one such operation is performed at a time
+is impractical, designating a single node to hold the counter state
+eschewing replicating altogether (as in a client-server
+configuration), is also an undesirable solution, given its sensitivity
+to network partitions and server failures.  But, simply replicating
+counter state among nodes, without having any coordination protocol,
+is an equally infeasible approach.
+
+A method often adopted~\cite{crdts, pldi15, gotsman-popl16} to address
+this tension is to re-define a datatype's operations to return
+\emph{effects} instead of values.  An \emph{effect} is a tag that
+identifies the operation to be applied uniformly at all replicas to
+incorporate the effects of the original
+operaton. Fig.~\ref{fig:counter-rdt} shows the \C{Counter} library
+with operations re-defined to returns effects.  Observe that the
+\C{Counter.add x v} operation now returns an \C{Add x} effect, which,
+when applied at a replica (see \C{apply} in the figure) adds \C{x} to
+the local counter value.  Assuming effects are commutative, they can
+be processed asynchronously using a log local to each replica.  This
+implementation produces an eventually consistent global state in which
+all replicas will eventually contain the same counter value.  While
+effects provide a low-level operational basis for handling replicated
+state, the \emph{ad hoc} nature of the solution confounds desirable
+high-level reasoning principles.  Indeed, the semantic gap between the
+two versions of the counter, one cognizant of replication and the
+other not, breaks backward compatibility.  More significantly, its
+non-trivial construction must be developed in different guises for
+every distributed data structure used by the application.
+
 Lack of composability is another important downside of this approach.
@@
 history - the state resulting from the merge of two versions derived
-from the same ancestor state should not unwittingly duplicate the
+from the same ancestor should not unwittingly duplicate the
 contributions of the ancestor.  This interpretation of a replicated
-datatype is thus given in terms of multi-versions of a program state
+data type is thus given in terms of multi-versions of a program state
 impicitly associated with the different nodes that comprise a
@@
   \begin{center}
-    \includegraphics[scale=0.35]{Figures/dali-counter}
+    \includegraphics[scale=0.4]{Figures/dali-counter}
   \end{center}
   \caption{\small A multi-version counter can be expressed as a
-    sequence of versions managed by logically-distributed concurrently
-    executing threads.  Local versions produced by these threads
-    periodically merge their state with one another.  In the figure,
-    when the local version of Node 3, labeled 20, merges with the
-    local version of Node 1, whose counter state is 15 at the time, a
-    new counter value is produced.  This value takes into account the
-    previous merged state (10) from which the two versions were
-    derived to yield a new merged state of 25.  Observe that there is
-    no explicit synchronization or coordination among versions.  In
-    the figure, red circles represent state produced through merging,
-    and blue circles represent local counter state produced by
-    applying counter operations on a node-local version.  }
+    non-planar graph, consisting of a logical master that defines a
+    notion of a globally consistent state, and a collection of
+    versions, each associated with a different node in a distributed
+    system.  Versions periodically merge their state with the master.
+    In the figure, when the local state of node 2, labeled 1, merges
+    with the master whose counter state is 15, a new counter value is
+    produced.  This value takes into account the global state (10)
+    from which the versions under merge were initiated to yield a new
+    global state of 6.  Observe that there is no synchronization or
+    coordination among versions.  In the figure, red circles represent
+    state produced through coordination with the master, while blue
+    circles represent local counter state produced by applying counter
+    operations on a node-local version.  }
 \end{figure}
 
-The \name programming model realizes a monadic functional version
+The \name programming model thus realizes a monadic functional version
 control system centered around data, rather than file coherence.  The
 \name monad lets programmers write and compose concurrent computations
-around multiple (implicit) versions of a mergeable datatype.  A
-computation makes progress by forking concurrent versions of existing
-versions, creating later versions along a \emph{branch}, or by merging
-branch tips.  An important soundness property of the model is that
-instances of a replicated datatype whose merge operation is
-commutative with respect to its local version arguments is guaranteed
-to eventually converge to the same value on all replicas.  As a
-result, OCaml programmers get the benefits of highly available
-(low-latency) distributed computation, while continuing to enjoy the
-comforts of high-level reasoning and familiarity of standard library
-datatypes.  Notably, while \name's programming model makes no explitic
-reference to any specific operational manifestation of a distributed
-system (e.g., programmers do not need to worry about replicas and
-their management), it can be nonetheless efficiently realized on
-existing geo-replicated distributed systems.
+around multiple versions of a mergeable data type.  Computation makes
+progress by forking concurrent versions of existing versions, creating
+later versions along a \emph{branch}, or by merging existing versions
+across branches. However, as we shall see, unconstrained branching may
+yield a branching structure that becomes unmergeable, thus leading to
+\emph{stuck} states. An important property of the \name library is
+that it suitably constrains branching structure to avoid stuck states
+without introducing additional coordination.  Furthermore, \name
+profitably exploits provenance information available via branches to
+make application resilient to network partitions.  As a result, OCaml
+programmers get the benefits of highly available (low-latency)
+computation, while continuing to enjoy the comforts of high-level
+reasoning and familiarity of the standard library data types.
+
+In this paper, we define a semantics of mergeable datatypes that lets
+any OCaml program be seamlessly deployed in a distributed environment
+with implicit support for replicated and decentralized operation.  The
+underlying foundation for merge actions has a natural category
+theoretic interpretation that enables us to construct useful mergeable
+variants for \emph{any} OCaml type.  Notably, while the resulting
+concurrency semantics makes no reference to any specific operational
+manifestation of a distributed system (e.g., programmers do not need
+to worry about replicas and their management), it can be nonetheless
+efficiently realized on existing geo-replicated distributed systems.
 
@@
 \begin{itemize}
-    \item We formally introduce the concept of a \emph{mergeable} datatype
-      to admit high-level declarative reasoning about distributed
+    \item We formally introduce the concept of a \emph{mergeable} data
+      type to admit high-level declarative reasoning about distributed
       computation and replicated state in functional programs.
@@
       flexibility of a version control system to the administration of
-      replicated data. \name hides the complexity of version control
-      behind a monad, and exposes its functionality via a simple API
-      that lets programmers define and compose concurrent OCaml
-      computations around such data.  Issues related to replication
-      only manifest implicitly in the definition of a merge operation
+      replicated data. \name hides the complexity of version
+      control behind a monad, and exposes its functionality via a
+      simple API that lets programmers define and compose concurrent
+      OCaml computations around replicated data.  Issues related to
+      replication only manifest in the definition of a merge operation
       that defines coherence among different instances of replicated
@@
     \item We present a formalization of the \name programming model,
-      and establish the comditions under which eventual convergence of
-      concurrent version state can be guaranteed.
+      and prove results that guarantee eventual convergence of
+      concurrent versions, and progress (stuck-freeness) guarantees.
 
@@
 %% the language abstraction is a non-trivial problem.
-
-%% However, as we shall see,
-%% unconstrained branching may yield a branching structure that becomes
-%% unmergeable, thus leading to \emph{stuck} states. An important
-%% property of the \name library is that it guarantees progress even in
-%% the presence of arbitrary merges provided the merge operation is
-%% commutative with respect to its version state arguments.  Furthermore,
-%% \name profitably exploits provenance information available via
-%% branches to make application resilient to network partitions.  
-%% In this paper, we define a semantics of mergeable datatypes that lets
-%% any OCaml program be seamlessly deployed in a distributed environment
-%% with implicit support for replicated and decentralized operation.  The
-%% underlying foundation for merge actions has a natural category
-%% theoretic interpretation that enables us to construct useful mergeable
-%% variants for \emph{any} OCaml type whose states are commutative wtih
-%% respect to the merge operation defie

### 2017-02-23 paper/model.tex
edits

@@
 \begin{figure}
-
-  \begin{ocaml}
-  module type DALI = sig
-    type 'a 'b t
-    val return : 'b -> 'a 'b t
-    val bind : 'a 'b t -> ('b -> 'a 'c t) -> 'a 'c t
-    val get_current_version: unit -> 'a 'a t
-    val with_init_version_do: 'a -> 'a 'b t -> 'b
-    val fork_version : 'a 'b t -> 'a unit t
-    val sync_next_version: unit -> ?v:'a -> 'a 'a t
-  end
-  \end{ocaml}
-
-\label{fig:dali-monad}
-\caption{Signature of the \name monad}
-\end{figure}
-
-\begin{figure}
-
+\begin{center}
   \begin{ocaml}
@@
       | B of {tl: tree; tr: tree; bl: tree; br: tree} 
-    type t = {max_x:int; max_y:int; canvas:tree} [@@deriving versioned]
+    type t = {max_x:int; max_y:int; canvas:tree} 
     type loc = {x:int; y:int}
@@
   \end{ocaml}
-
+\end{center}
 \caption{\drawsome: a sample \name application}
@@
 Fig.~\ref{fig:canvas-sig} shows the signature of the \drawsome
-application. \drawsome represents a free-hand drawing canvas as a tree
-of quadrants, where each quadrant is simply a leaf node unless it has
-pixels of different colors. Quadrants are expanded into a tree
-structures as an when pixels are colored (the color scheme used is
-rgb). The representation is thus optimized for sparse canvases, such
-as whiteboards. The application supports three simple operations:
+application. \drawsome represents a free-hand drawing canvas in terms
+of a tree of quadrants.  Each quadrant is simply a leaf node
+containing a single pixel (\C{r},\C{g}, or \C{b}) or a tree, if the
+quadrant contains multiple pixels of different colors. Quadrants are
+expanded into a tree structures as and when pixels are colored.  The
+representation is thus optimized for sparse canvases, such as
+whiteboards. The application supports three simple operations:
 creating a new canvas, setting the pixel at a specified coordinate,
@@
 \drawsome lets multiple users collaborate on a canvas that is
-conceptually shared among all the collaborators. Under a shared-memory
-abstraction, there would be a single copy of the canvas that is
-updated concurrently by multiple clients such that from the
-perspective of any single client, the canvas would change without any
-explicit intervention. \name ascribes functional semantics to sharing
-by letting each client work on its own version of the state (a data
-structure), and later merging the concurrent versions on-demand. Thus,
+conceptually shared among them. Under a shared-memory abstraction,
+there would be a single copy of the canvas that is updated
+concurrently by multiple clients; from the perspective of any single
+client, the canvas could change without any explicit
+intervention. \name ascribes functional semantics to sharing by
+letting each client work on its own version of the state (the tree
+data structure), later merging concurrent versions on-demand.  Thus,
 the primary artifact of the \name programming model is a versioned
-data structure. The library operates on a representation of versioned
-data structures optimized for persistence on disk
-(Sec.~\ref{sec:persistence}). \name's meta-programming component
-automatically synthesizes this representation, along with the
-functions that translate between representations, for the data type
-definitions marked with the ppx~\cite{ppx} directive \C{deriving
-versioned}. Concretely, for the \C{Canvas} module, \name synthesizes
-a \C{Canvas.Versioned} module with a type \C{t}, and functions
-\C{of\_canvas} and \C{to\_canvas} of types \C{Canvas.t $\rightarrow$
-t} and \C{t $\rightarrow$ Canvas.t}, respectively.
+data structure in which different versions are managed by different
+clients.
+
+%% The library operates on a representation of versioned data
+%% structures optimized for persistence on disk
+%% (Sec.~\ref{sec:persistence}).  \name's meta-programming component
+%% automatically synthesizes this representation, along with the
+%% functions that translate between representations, for the data type
+%% definitions marked with the ppx~\cite{ppx} directive \C{deriving
+%%   versioned}. Concretely, for the \C{Canvas} module, \name synthesizes
+%% a \C{Canvas.Versioned} module with a type \C{t}, and functions
+%% \C{of\_canvas} and \C{to\_canvas} of types \C{Canvas.t $\rightarrow$
+%%   t} and \C{t $\rightarrow$ Canvas.t}, respectively.
 
 \name requires a three-way \C{merge} function to merge the concurrent
-versions of a drawing canvas. The three arguments include two
-concurrent versions (\C{v1} and \C{v2}), and their least common
-ancestor (\C{lca}) - the version from which the two concurrent
-versions evolved independently. The merge function can make use of the
-pixel values of the common ancestor to merge the pixel values on both
-the canvases. For instance, if the color of a pixel in \C{v1} is white,
-and in \C{v2} it is green, and its color in \C{lca} is
-white, then it means that only \C{v2} modified the color. Hence the
+versions of a drawing canvas (see Fig.~\ref{fig:merge-canvas}). The
+three arguments include two concurrent versions (\C{v1} and \C{v2}),
+and their least common ancestor (\C{lca}) - the version from which the
+two concurrent versions evolved independently. The merge function can
+make use of the pixel values of the common ancestor to merge the pixel
+values on both the canvases. For instance, if the color of a pixel in
+\C{v1} is white, and in \C{v2} it is green, and its color in \C{lca}
+is white, then it means that only \C{v2} modified the color. Hence the
 pixel is colored green in the merged canvas. On the other hand, if the
-pixel is red in \C{v1}, then it means that both
-\C{v1} and \C{v2} have modified the color. In such case, an appopriate
-color-mixing algorithm can be used to determine the color of pixel.
-For instance, the pixel can be colored yellow - an additive
-combination of red and green. The logic is illustrated in
-Fig.~\ref{fig:canvas-merging}.
+pixel is red in \C{v1}, then it means that both \C{v1} and \C{v2} have
+modified the color. In such case, an appopriate color-mixing algorithm
+can be used to determine the color of pixel.  For instance, the pixel
+can be colored yellow - an additive combination of red and green. The
+logic is illustrated in Fig.~\ref{fig:canvas-merging}.
+\begin{figure}
+\begin{center}
+  \begin{ocaml}
+let color_mix px1 px2 : pixel = 
+let f = Char.code in
+let h x y = Char.chr @@ (x + y)/ 2 in
+let (r1,g1,b1) = (f px1.r, f px1.g, f px1.b) in
+let (r2,g2,b2) = (f px2.r, f px2.g, f px2.b) in
+let (r,g,b) = (h r1 r2, h g1 g2, h b1 b2) in {r=r; g=g; b=b}
+
+let b_of_n px = B {tl_t=N px; tr_t=N px; bl_t=N px; br_t=N px}
+
+let rec merge lca v1 v2 = 
+  if v1=v2 then v1
+  else if v1=lca then v2
+  else if v2=lca then v1
+  else match (lca,v1,v2) with
+    (*
+     * The first three rules isomorphize lca, v1 and v2.
+     *)
+    | (_, B _, N px2) -> merge lca v1 @@ b_of_n px2
+    | (_, N px1, B _) -> merge lca (b_of_n px1) v2
+    | (N px, B _, B _) -> merge (b_of_n px) v1 v2
+    | (B x, B x1, B x2) ->
+        let tl_t' = merge x.tl_t x1.tl_t x2.tl_t in
+        let tr_t' = merge x.tr_t x1.tr_t x2.tr_t in
+        let bl_t' = merge x.bl_t x1.bl_t x2.bl_t in
+        let br_t' = merge x.br_t x1.br_t x2.br_t in
+          B {tl_t=tl_t'; tr_t=tr_t'; bl_t=bl_t'; br_t=br_t'}
+    | (_, N px1, N px2) -> 
+        (* pixels are merged by mixing colors *)
+        let px' = color_mix px1 px2 in N px'
+ \end{ocaml}
+\caption{Merging different versions of a canvas.}
+\label{fig:merge-canvas}
+\end{center}
+\end{figure}
 
@@
 \C{State} monad~\cite{wadler-monad}. The monad encapsulates a
-versioned functional state (\C{'a}) and the type \C{'a 'b t}
+versioned functional state (\C{'a}) and the type \C{('a, 'b) t}
 represents a monadic computation that returns a \C{'b} result.
 Functions \C{return} and \C{bind} have their usual monadic
+\begin{figure}
+\begin{center}
+  \begin{ocaml}
+  module type DALI = sig
+    type ('a, 'b) t
+    val return : 'b -> ('a, 'b) t
+    val bind : ('a, 'b) t -> ('b -> ('a, 'c) t) -> ('a, 'c) t
+    val get_current_version: unit -> ('a, 'a) t
+    val with_init_version_do: 'a -> ('a, 'b) t -> 'b
+    val fork_version : ('a, 'b) t -> 'a unit t
+    val sync_next_version: unit -> ?v:'a -> ('a, 'a) t
+  end
+  \end{ocaml}
+\label{fig:dali-monad}
+\caption{Signature of the \name monad}
+\end{center}
+\end{figure}
 interpretation. \C{get\_current\_version} is like the \C{State}
@@
 current version, and runs the given monadic computation asynchronously
-against the forked version.  The semantics of \C{sync\_next\_version}
-(simply called \C{sync}) is somewhat involved: the function accepts
-a \emph{proposal} for the next version of the
-state, where \C{next} is defined w.r.t the current version, and
-returns (via a monad) the actual next version, which becomes the
-current version for rest of the computation.  The next version is
-created by merging the proposal with a subset of concurrent versions
-that became available since the last merge or fork. Thus, \C{sync}
-effectively lets a computation sync with a subset of concurrent
-computations and obtain their latest updates.
+against the forked version.  \C{sync\_next\_version} (simply called
+\C{sync}) accepts a \emph{proposal} for the next version of the state;
+this proposal is the current local version of the state that reflects
+local modifications not yet witnessed by any other concurrently
+executing computation.  The operation returns (via a monad) the actual
+next version, which becomes the current version for the rest of the
+computation.  This version is created by merging the proposal with a
+subset of concurrent versions that have become available since the
+last merge or fork. Thus, \C{sync} effectively lets a computation sync
+with a subset of concurrent computations to obtain their latest
+updates.
 
@@
 
-\begin{figure}
-\centering
-\includegraphics[scale=0.6]{Figures/canvas-sessions}
-
-\caption{\drawsome: Collaborative drawing session visualized}
-\label{fig:canvas-sessions}
-\end{figure}
-
 Fig.~\ref{fig:canvas-sessions-code} demonstrates how a collaborative
@@
 \begin{ocaml}
-  module C = Canvas;; open Dali;;
+  module C = Canvas;; 
   with_init_version_do (C.new_blank 5 5) alice_f 
@@
 blank. She then invites Bob for collaboration by forking a new
-concurrent version for Bob. Bob in turn invites Cheryl for
+concurrent version for Bob. Bob, in turn, invites Cheryl for
+\begin{figure}
+\centering
+\includegraphics[scale=0.6]{Figures/canvas-sessions}
+\caption{\drawsome: Collaborative drawing session visualized}
+\label{fig:canvas-sessions}
+\end{figure}
 collaboration. All three of them start working on the blank canvas.
 Alice draws a red horizontal line from $(0,0)$ (top-left) to $(4,0)$
-(top-right) using \C{C.draw\_line} (definition not shown; can be
-constructed using \C{set\_px}). Meanwhile, Bob draws a green vertical
-line from $(0,0)$ to $(0,4)$, and Cheryl draws a similar line from
-$(4,0)$ to $(4,4)$. All three of them call \C{sync} with their
-respective proposals ($C_0'$). While any partial ordering of
-concurrent \C{syncs} is valid, we consider a linear order where
-Cheryl's \C{sync} happens first, followed by Bob's and then Alice's.
-Cheryl's \C{sync} does not find any concurrent versions, hence
-installs the proposed version ($C_0'$) as the next version on Cheryl's
-branch. Bob's \C{sync} finds Cheryl's $C_0'$ as a concurrent version,
-and merges it with its proposal to produce the next version $C_1$,
-which is installed on Bob's branch.  and returned by \C{sync} (\C{c1}
-in second column of Fig.~\ref{fig:canvas-sessions-code}). The least common
-ancestor (LCA) for this merge is the initial version ($C_0$), and the two
-concurrent versions are Bob's $C_0'$ and Cheryl's $C_0'$. Next,
+(top-right) using \C{C.draw\_line}.\footnote{Its definitionis not
+  shown, but can be constructed using \C{set\_px}} Meanwhile, Bob
+draws a green vertical line from $(0,0)$ to $(0,4)$, and Cheryl draws
+a similar line from $(4,0)$ to $(4,4)$. All three of them call
+\C{sync} with their respective proposals ($C_0'$). While any partial
+ordering of concurrent \C{syncs} is valid, we consider a linear order
+where Cheryl's \C{sync} happens first, followed by Bob's and then
+Alice's.  Cheryl's \C{sync} does not find any concurrent versions,
+hence installs the proposed version ($C_0'$) as the next version on
+Cheryl's branch. Bob's \C{sync} finds Cheryl's $C_0'$ as a concurrent
+version, and merges it with its proposal to produce the next version
+$C_1$, which is then installed on Bob's branch.  The least common
+ancestor (LCA) for this merge is the initial version ($C_0$), and the
+two concurrent versions are Bob's $C_0'$ and Cheryl's $C_0'$. Next,
 Alice's \C{sync} finds Cheryl's $C_0'$ and Bob's $C_1$ as concurrent
@@
 first merge, the two concurrent versions are Alice's $C_0'$ and
-Cheryl's $C_0'$, and the LCA is the initial version
-($C_0$). The result of this merge is installed as the next version
-($\C{C_0''}$) on Alice's branch. For the next merge, the two
-concurrent versions are Alice's $C_0''$ and Bob's $C_1$ and the LCA is
-Cheryl's $C_0'$\footnote{Thus, LCA of versions on two branches can lie
+Cheryl's $C_0'$, and the LCA is the initial version ($C_0$). The
+result of this merge is installed as the next version ($\C{V_1}$) on
+Alice's branch. For the next merge, the two concurrent versions are
+Alice's $V_1$ and Bob's $C_1$ and the LCA is Cheryl's
+$C_0'$\footnote{Thus, the LCA of versions on two branches can lie
   outside both the branches.}. The result ($C_1$) becomes the next
@@
 Fig.~\ref{fig:canvas-sessions-code}) as the next version to \C{sync}.
-Since there are no concurrent versions, $C_1'$ becomes the next
-version and the return value of \C{sync}. The subsequent \C{sync}
-operations from Bob and Cheryl propose no new versions, hence simply
-obtain Alice's $C_1'$ as next versions.
+Since there are no concurrent versions, $C_1'$ becomes her next
+version. The subsequent \C{sync} operations from Bob and Cheryl
+propose no new versions, hence simply obtain Alice's $C_1'$ as next
+versions.
 
 The \drawsome example demonstrates the utility of mergeable data types
-and the \name programming model in building concurrent applications
-with conceptual sharing of state. This example, however, implicitly
-makes certain assumptions about the model and the underlying system,
-such as the existence of a single least common ancestor (LCA) for any
-pair of versions, the ability to access a previous version on any
-branch, and mergeability of any two concurrent versions. Enforcing
-these guarantees in a fully decentralized distributed setting requires
-addressing non-trivial theoretical challenges described in the next
-section. Unlike the canvas, the process of writing merge functions for
-data types with non-trivial invariants could be quite involved. The
-later sections describe a principled approach to derive a merge
-function for a data type given its invariants. Like \C{Canvas.merge},
-merge functions often need to compare multiple versions of a data type
-for structural equality, which, when performed naively, incurs cost
-linear in the size of versions. The representation of a versioned data
-type in \name is optimized for performing comparisons that allows
-structural equality to be determined in constant time. Such
-engineering details, which contribute to the practicality of our
-programming model are also discussed in subsequent sections.
+in building concurrent applications with conceptual sharing of
+state. This example, however, implicitly makes certain assumptions
+about the model and the underlying system, such as the existence of a
+single least common ancestor (LCA) for any pair of versions, the
+ability to access a previous version on any branch, and mergeability
+of any two concurrent versions. Enforcing these guarantees in a fully
+decentralized distributed setting requires addressing non-trivial
+theoretical challenges described in the next section. Unlike the
+canvas, the process of writing merge functions for data types with
+non-trivial invariants could be quite involved. The following sections
+describe a principled approach to deriving merge functions for any
+data type given its invariants. Like \C{Canvas.merge}, merge functions
+often need to compare multiple versions of a data type for structural
+equality, which, when performed naively, incurs cost linear in the
+size of versions. The representation of a versioned data type in \name
+is optimized for performing comparisons that allows structural
+equality to be determined in constant time. Such engineering details,
+which contribute to the practicality of our programming model are also
+discussed in subsequent sections.

### 2017-02-23 paper/semantics.tex
Intro points added.
opsem.

@@
 
+We formalize our ideas in the context of a lambda calculus ($\lang$)
+shown in Fig.~\ref{fig:opsem}. Expressions of $\lang$ are variables,
+constants, and \name primitives composed using the lambda combinator.
+For brevity, we use short names for \name primitives: \C{run} for
+\C{with\_init\_version\_do}, and \C{fork} for \C{fork\_version}. To
+simplify the technical development, \name's \C{sync\_next\_version}
+operation is broken down into two primitives - \C{push} and \C{pull},
+which can be lambda-composed get the desired effect:
+\begin{smathpar}
+  \C{sync}\;x \;=\; (\lambda y.\pull)\; (\push\,x)
+\end{smathpar}
+The semantics of\C{get\_current\_version} is subsumed by \C{pull},
+hence elided.  Values ($v$) are constants and lambda abstractions.  A
+program ($p$) is a parallel composition of threads, where each thread
+is an expression ($s$) indexed by the corresponding thread identifier
+($t$).
+
+Fig.~\ref{fig:opsem} also shows the syntax of \emph{branches}, which
+are the artifacts of evaluation and only appear during the run-time. A
+branch is a non-empty sequence of tagged values, where the tag
+captures the abstract run-time operation that led to the creation of
+the value. It is implicitly assumed that each value added to a branch
+is uniquely identifiably, hence no two values on a branch are equal.
+The uniqueness assumption is later extended to a collection of
+branches that constitute a branching structure. A real implementation
+meets this assumption by versioning values across the branches. Thus,
+in reality, branches contain \emph{versions} which denote values. The
+semantics, however, doesn't make this distinction, and uses values and
+versions interchangeably.
+
+Small-step operational semantics of $\lang$ is defined via reduction
+relation ($\stepsto$) that relates \emph{program states}. A program
+state ($p;\,H$) consists of a program $p$ and a \emph{branch history}
+$H$ that maps thread Ids to corresponding branches; each thread is
+associated with a branch during the evaluation. Evaluation contexts
+have been defined separately for expressions ($E$) and programs ($P$),
+with the latter subsuming the former. $E$ is defined to evaluate the
+first argument of a \C{run} expression to a value that constitutes the
+initial version (recall that \C{run} models \name's
+\C{with\_init\_version\_do}). Program evaluation context
+non-deterministically picks one of the threads to evaluate. The admin
+rule that relates transitions of holes to transitions of expressions
+and programs is straightforward, hence elided. Rest of the reduction
+rules are presented in Fig.~\ref{fig:opsem}. For brevity, we write
+$H(t\mapsto (v,f))$ to denote the proposition that $H$ maps $t$ to
+$(v,f)$. The notation $H[t \mapsto (v,f)]$ as usual denotes
+the extension of $H$ with the binding $t \mapsto (v,f)$.
+
+Reduction rules let expression evaluation take a step by rewriting the
+expression and suitably updating the branch history ($H$). The
+\rulelabel{E-Run} rule is applicable only when $H$ is empty, i.e.,
+when no prior branching structure exists. The rule rewrites the
+$\C{run}\;v\;s$ expression to $s$, while creating a new branching
+structure with two branches: a \emph{top} branch that has just the
+initial version (tagged with \C{INIT}), and a branch for the current
+thread ($t$) forked-off from the top branch.  The first version on the
+current branch ($H(t)$) denotes the same value ($v$) as the initial
+version on the top branch, although versions themselves are deemed
+distinct. The new version is tagged with a \C{FORK} tag that keeps the
+record of its orgin, namely the \C{fork} operation and the branch from
+which the current branch is forked. The \rulelabel{E-Fork} rule forks
+a new thread with a fresh id ($t'$) and adds it to the thread pool.
+The corresponding branch ($H(t')$) is forked from the parent thread's
+branch ($H(t)$). The semantics of branch forking is same as described
+above. The \C{fork} expression in the parent thread evaluates to
+\C{()}. The \rulelabel{E-Push} rule creates a new version on the
+current branch ($H(t)$) using the pushed value ($v$).
+\rulelabel{E-App} is the standard beta reduction rule.
+
+The semantics non-deterministically chooses \rulelabel{E-Pull} or
+\rulelabel{E-Pull-Wait} rules to reduce a \C{pull} expression. The
+\rulelabel{E-Pull} rule reduces \C{pull} to \C{()}, and returns the
+latest version on the current branch. The \rulelabel{E-Pull-Wait} rule
+can be thought of as a stutter step; it doesn't reduce \C{pull}, but
+updates the branching structure by merging (the latest version of) a
+concurrent branch ($H(t')$) into (the latest version of) the current
+branch ($H(t)$), and extending the current branch with the merged
+version ($v_m$). The new version is tagged with a \C{MERGE} tag that,
+like a \C{FORK} tag, records its origin. The premises of the
+\rulelabel{E-Pull-Wait} serve the important purpose of contraining the
+branching structure by allowing only the legal merges, and
+consequently preserving certain desirable properties of the system;
+we will go into the details shortly. The \rulelabel{E-Pull-Wait} and
+\rulelabel{E-Pull} rules thus let a thread sync with a subset of
+concurrent threads in multiple steps before returning the result of
+the \C{pull}. Since \C{sync} is a composition of \C{push} and
+\C{pull}, its behavior can be explained thus: \C{sync} pushes the
+given value onto the current (local) branch, merges a (possibly empty)
+subset of concurrent branches into the local branch, and returns the
+result.
+
+We will not present a series of definitions and results that let us
+understand the premises of the \rulelabel{E-Pull-Wait} rule, examine
+the merge operation in closer detail, and appreciate the need for
+constraining the branching structure. First, we formalize the
+intuitive notation of the ancestor relationship between versions of a
+legal branching history (i.e., a branching history generated by the
+rules in Fig.~\ref{fig:opsem}):
+
 \begin{definition} [\bfseries Ancestor]
 Version $v_1$ is a ancestor of version $v_2$ under a history
-$H$ (written $\under{H}{v_1 \preceq v_2}$) if and only if:
+$H$ (written $\under{H}{v_1 \preceq v_2}$) if and only if one of the
+following is true:
 \begin{itemize}
@@
   dom(H)).\,H(t) = b$) in which $v_2$ immediately succeeds
-  $v_1$, or
-  \item There exists a branch $b$ in $H$, such that $b = (v_2, 
-  \C{FORK}\; (v_1,f_1)::b_1)::b'$, for some $f_1$, $b_1$ and $b'$, or
+  $v_1$,
+  \item There exists a branch $b$ in $H$ that contains $(v_2, 
+  \C{FORK}\; (v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
   \item There exists a branch $b$ in $H$ that contains
-  $(v_2, \C{MERGE}\;(v_1,f_1)::b_1)$, for some $f_1$ and $b_1$, or
+  $(v_2, \C{MERGE}\;(v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
   \item $v_1 = v_2$, or $v_1$ is transitively a ancestor of
@@
 
+Ancestor relation is therefore a partial order (reflexive, transitive,
+anti-symmetric) with a greatest lower bound (the initial version).
+Thus, for any two versions in a legal history, there exist at least
+one common ancestor. Ancestor relationships among the common ancestors
+let us define the notion of a least common ancestor (LCA):
+
 \begin{definition} [\bfseries Least Common Ancestor]
-Version $v$ is said to be a common ancestor of versions $v_1$
-and $v_2$ under $H$ iff $\under{H}{v \preceq v_1}$ and $\under{H}{v
-\preceq v_2}$. It is said to be the least common ancestor
-(LCA) of $v_1$ and $v_2$, iff there does not exist a $v'$ such that
-$\under{H}{v' \preceq v_1}$ and $\under{H}{v' \preceq v_2}$ and
-$\under{H}{v \preceq v'}$.
+Version $v$ is said to be a common ancestor of versions $v_1$ and
+$v_2$ under a history $H$ if and only if $\under{H}{v \preceq v_1}$
+and $\under{H}{v \preceq v_2}$. It is said to be the least common
+ancestor (LCA) of $v_1$ and $v_2$, iff there does not exist a $v'$
+such that $\under{H}{v' \preceq v_1}$ and $\under{H}{v' \preceq v_2}$
+and $\under{H}{v \preceq v'}$.
 \end{definition}
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.6]{Figures/merge-needs-lca}
+
+\caption{This example of a grow-only counter illustrate why \C{merge}
+needs a least common ancestor, and not just a common ancestor. Both 0
+and 2 are common ancestors of 4 and 7, while 2 is their least common
+ancestor (since $0 \preceq 2$). The result (v) of merging 4 and 7 is
+11 (incorrect) if 0 is used as the common ancestor for merge, and 9
+(correct, because 2+2+3+2 = 9) if 2 is used. }
+\label{fig:merge-needs-lca}
+\end{figure}
+
+\begin{figure}[!t]
+\centering
+\subcaptionbox[] {\small
+  In this example, 1 and 3 have two LCAs (3 and 4) a result of
+  previous merges. The dotted circle denotes a virtual ancestor
+  obtained by merging the two LCAs.
+  \label{fig:criss-cross-lcas}
+} [0.47\columnwidth] {
+  \includegraphics[scale=0.55]{Figures/2-LCAs}
+}
+\hfill
+\subcaptionbox[] {\small
+  In this example, 3 and 6 have two LCAs (2 and 1) despite there not
+  being any previous merges between their respective branches.
+  \label{fig:external-lcas}
+} [0.47\columnwidth] {
+  \includegraphics[scale=0.7]{Figures/2-external-LCAs}
+}
+\caption{Examples where merging versions have more than one LCA}
+\label{fig:many-lcas}
+\end{figure}
+
+When merging two concurrent versions $v_1$ and $v_2$, the common
+ancestor argument for \C{merge} must the LCA of $v_1$ and $v_2$,
+without which \C{merge} may yeild unexpected results. This is
+demonstrated for the grow-only counter in
+Fig.~\ref{fig:merge-needs-lca}, where incorrect count is obtained if a
+common ancestor that is not an LCA is used to merge 4 and 7. While in
+this example there is a unique LCA for 4 and 7, in general this may
+not be the case. With unrestrained branching and merging, there is no
+bound on the number of LCAs a pair of versions can have.
+For example, in
+Fig.~\ref{fig:criss-cross-lcas}, the merge of 1 with 3 is preceded by
+two ``criss-cross'' merges between their respective branches resulting
+in there being two LCAs (3 and 4) for 1 and 3. Multiple LCAs can
+occur even without criss-cross merges, as demonstrated by
+Fig.~\ref{fig:external-lcas}.
 
@@
 later prove that in the histories generated by the rules in
-Fig.~\ref{fig:opsem}, each version has exactly one least external
-ancestor. We call the unique least external ancestor of a version as
-its \emph{external locus}. Least internal ancestor, by definition, is
-unique, which we sometimes call \emph{internal locus}.
+Fig.~\ref{fig:opsem}, the external ancestors of every version are
+totally ordered, hence there is exactly one least external ancestor
+for each version. We call the unique least external ancestor of a
+version as its \emph{external locus}. Least internal ancestor, by
+definition, is unique, which we sometimes call \emph{internal locus}.
 
@@
 $\under{H}{v_2 \mbleto v_1}$) iff $v_1$'s external locus is an
-ancestor of $v_2$.
+ancestor of $v_2$, or $v_2$'s external locus is an ancestor of $v_1$.
 \end{definition}
@@
 The \rulelabel{E-Pull-Wait} case is where the merge happens. Version
-$v_m$ is merging into version $v$ to produce the next version $v_m$.
-Let $v_2$ be any version. From premise, we know that $v$ and $v_2$
-have a unique LCA (call it $v_0$), and $v'$ and $v_2$ have a unique
-LCA (call it $v_1$). But, we know that $v_2$ has a unique external
-locus, which leads to the following cases:
+$v'$ is merging into version $v$ to produce the next version $v_m$.
+Let $v''$ be any version. From the premise, we know that $v$ and $v''$
+have a unique LCA (call it $v_1$), and $v'$ and $v''$ have a unique
+LCA (call it $v_2$). Also, we know that $v''$ has a unique external
+locus (call it $v_3$). We now consider cases
 \begin{itemize}
-  \item Both $v_0$ and $v_1$ are external ancestors of $v_2$. Hence,
-  $\under{H}{v_0 \preceq v_1} \disj \under{H}{v_1 \preceq v_0}$.
-  Since\under{H}{v_0 \preceq v_m} \disj \under{H}{v_1 \preceq v_m}, 
+  \item $v_1$ and $v_2$ are external to both $v$ and $v'$: Hence $v_1$
+  is an external ancestor of $v$, and $v_2$ is an external ancestor
+  of $v'$. Since $v$ and $v'$ are mergeable, there must exist a $v_4$
+  such that $\under{H}{v_1 \preceq v_4}$ and $\under{H}{v_1 \preceq
+  v_4}$, and either $\under{H}{v_4 \preceq v}$ or $\under{H}{v_4
+  \preceq v'}$.  are external ancestors of
+  $v_2$. Hence,
+  $\under{H}{v_0 \preceq v_3} \conj \under{H}{v_1 \preceq v_3}$.
+  Since $\under{H}{v_0 \preceq v_m} \conj \under{H}{v_1 \preceq v_m}$, 
   $v_2$ and $v_m$ have a unique LCA.

### 2017-02-23 paper/intro-old.tex

\section{Introduction}

Distributed programs are challenging to write because they must juggle
two distinct mechanisms.  The first deals with application logic - how
do we define computation to be robust in the presence of disconnected
non-coherent communication among distributed threads of control?  The
second deals with system concerns - how do define notions of
atomicity, distribution, persistence, and replication?  Because of the
subtleties in dealing with the latter, especially when high
availability (low response latency) is desired, systems typically
separate these mechanisms into separate tools or services (e.g., CRDTs
have a well-defined semantics in an eventually consistent system that
may experience network partitions, while database and related storage
systems provide support for atomic transactions, persistence,
replication.  Unfortunately, because these different mechanisms have
vastly different semantics and goals, program structure becomes
complicated and muddled.  More importantly, it becomes problematic to
relate properties relevant to applications to the guarantees provided
by these lower-level services.  In this paper, we propose a radically
different vision of a distributed application that is predicated on
defining both mechanisms within the same language framework, with
those aspects not directly related to coherence and consistency
encapsulated within a library.  This not only simplifies program
reasoning, but leads to significant expressivity gains since
language-level types and abstractions are directly supported by the
library.  Realizing this vision leads to a radically different
programming model, one which realizes a monadic functional version
control system centered around data, rather than file coherence.  We
have implemented an end-to-end instantiation of this idea in OCaml
called DaLi and demonstrate how efficient distributed applications can
be written and deployed without having to deal with \emph{any}
system-centric features.

Points:
\begin{itemize}
  \item Language is often considered the pivot of the thought process.
  Unsurprisingly, programmers are conditioned to reason in terms of
  the language-level abstractions. However, reality of underlying
  systems often force programmers to think in terms of low-level
  abstractions that are counter-intuitive from the language
  standpoint. Even functional programmers are often exposed to this
  rather unfortunate state of affairs.

  \item Consider, for example, an OCaml programmer who wishes to
  manage an ordered set of elements in a binary tree, while using a
  hashtable to quickly obtain references into the tree. While the
  programmer enjoys high-level abstractions and safety guarantees of
  OCaml insofar as the program only accesses in-memory data, she has
  to let go of this convenience if the application requires data
  persistence. She has to now re-engineer her application around the
  data model enforced by the underlying storage engine (e.g.,
  relational data model, or key-value data model), complicating
  programming, and potentially compormising the language safety
  guarantees in the process. For instance, while OCaml guarantees the
  safety of all references, references between tables in a database
  are unsafe, and have been known to lead to bugs.

  \item Add replication to persistence, and the problems grow
  manifold. The impossibility of obtaining strong consistency
  guarantees without compromising scalability/availability under
  decentralized replication means that applications can now exhibit
  anomalous behaviors that are often incomprehensible to the
  programmer. The standard artifacts of replication made available to
  the programmer, namely mutable key-value pairs, expose the underlying
  system reality in its full ugliness. While the programmer may manage
  to reason about the contents of a key-value cell with much effort,
  the reasoningly unfortunately does not compose; each key-value pair
  can be manipulated independently, which makes it almost impossible
  to obtain guarantees over a collection of data items in the same
  vein as a high-level data structure (e.g., a binary search tree).
  CRDTs are an attempt to provide high-level semantics of replication
  for certain useful data types, however CRDTs do not come with a
  an associated replicated storage system and a programming framework,
  and programmers are again faced with the task of bridging the gap
  between the high-level semantics of the CRDT to the abstractions and
  guarantees offered by the underlying storage system. Moreover,
  standard data structures need significant re-engineering to be
  considered CRDTs (e.g., functions need to return \emph{effects} all
  of which need to commute), and programmers receive little or no tool
  support in this process. Moreover, CRDTs famously do not compose (in
  the sense that the invariants over the composition cannot be
  preserved), requiring programmers to resort to extraneous
  mechanisms with system-specific properties, such as transactions.

  \item In this paper we describe an OCaml library called \name that
  adresses the aforementioned problems. \name lets programmers add
  persistence and fully-decentralized replication to their
  applications without ever having to leave the comforts of high-level
  reasoning in OCaml. Thus, in keeping with the long tradition of
  Functional Programming, programmers are \emph{completely} shielded
  from the complexities of the underlying system model. This elevation
  from the low-level reasoning to the high-level reasoning concretely
  manifests as reasoning in terms of \emph{mergeable} data types
  instead of \emph{replicated} data types; replication being a
  system-level concern. A mergeable data type library is a standard
  OCaml data type library equipped with a \C{merge} function that
  describes how to merge two concurrent versions of the data type that
  originated from a single common ancestor. Thus, any OCaml data type
  for which a meaningful merge function can be written is a mergeable
  type. We give multiple examples of mergeable types useful in
  practice, present a simple theory that guides programmers in writing
  merge functions, and describe the tool support available to them in
  this process. \name also comes with a novel programming model that
  realizes a monadic functional version control system centered around
  data, rather than file coherence. The \name monad lets programmers
  write and compose concurrent computations around multiple versions
  of a mergeable data type. The computation makes progress by forking
  concurrent versions of the existing versions, creating later
  versions of the existing versions along a \emph{branch}, or by
  merging existing versions across the branches. However, as we shall
  see, unconstrained branching may yield a branching structure that
  becomes unmergeable, thus leading to \emph{stuck} states. An
  important property of the \name library is that it suitably
  constrains the branching structure to avoid stuck states without
  introducing additional coordination that compromises the
  decentralized nature of the computation. Futhermore, \name
o  profitably exploits the provenance information available via the
  branching structure to make the application resilient to network
  partitions. As a result, OCaml programmers get the benefits of
  persistence and highly available replication, while they continue to
  enjoy the comforts of high-level reasoning and the familiarity of
  the standard library data types.

  Our contributions are summarized below:
  \begin{itemize}
    \item We formally introduce the concept of a mergeable data type,
    as an alternative to a replicated data type, to admit high-level
    reasoning about replicated data in functional programs.

    \item We describe the \name library that transparently adds
    persistence and replication features to a mergeable type. \name
    comes with a programming model that brings to bear the power and
    flexibility of a version control system on administration of the
    replicated data. \name however hides the complexity of version
    control behind a monad, and exposes its functionality via a simple
    API that lets programmers define and compose concurrent OCaml
    computations around replicated data.

    \item We present a formalization of the \name programming model,
    and prove results that guarantee eventual convergence of
    concurrent versions, and progress (stuck-freeness) guarantees.

    \item We describe an implementation of \name atop the Irmin store,
    and describe multiple examples, and two case studies that
    attest to the practical utility of \name.
  \end{itemize}

\end{itemize}

### 2017-02-23 paper/intro.tex

@@
 Real-world distributed programs are challenging to write and maintain
-because they inevitably juggle and conflate two distinct mechanisms.
-The first concerns the expression of application logic - how do we
-define computations that are robust in the presence of distributed
+because they often conflate two distinct mechanisms.  The first
+concerns the expression of application logic - how do we define
+computations that are robust in the presence of distributed
 communication among concurrently executing threads of control?  The
@@
 networks that connect them unreliable?  To simplify reasoning in such
-complicated environments, models often make strong assumptions on the
-guarantees provided by an implementation (such as
-serializablity~\cite{...} or strong consistency~\cite{...}) that may
-inadvertently mask important albeit unpleasant realities related to
-network partitions~\cite{...}, or which may ignore important system
-architectural features that assume replicated program state~\cite{...}
-to provide high availability and fault tolerance.  On the other hand,
-when these properties are explicitly exposed to the programmer, e.g.,
-by requiring that applications be written in terms of specialized
-distributed data structures~\cite{...} or control
-primitives~\cite{...}, simplicity, composability, and
-ease-of-reasoning suffers.
+complicated environments, programming models often make strong
+assumptions on the guarantees provided by an implementation (such as
+serializability~\cite{Serializability} or strong
+consistency~\cite{CDE+12,DNN+15}) that may inadvertently mask,
+restrict, or ignore important albeit unpleasant realities (e.g.,
+network partitions~\cite{Brewer2000,Gilbert2002}), or which may ignore
+important system architectural features (e.g., replicated underlying
+state~\cite{Walter}).  Given these unwarranted assumptions, program
+behavior is often difficult to predict and verify.  On the other hand,
+when these features are explicitly exposed to the programmer, e.g., by
+requiring that applications be written in terms of specialized
+distributed data structures~\cite{Burckhardt2014,,BFL+12,SPB+11} or
+control primitives~\cite{Calm,Armstrong07}, simplicity, composability,
+and ease-of-reasoning can suffer.
 
@@
 library is written in an idiomatic style in OCaml, with no special
-reasoning principles needed to realize desired functionality.  As long
-as applications use the library on a single machine, this
-implementation suffices well.  If the library is used in the context
-of a more sophisticated application, say one whose computation is
-distributed among a collection of machines, its state may need to be
-replicated.  Unfortunately, replication doesn't come for free.  Global
-synchronous coordination among all nodes is problematic in distributed
-systems.  Thus, applying an \C{Add} operation on one node may not be
-instanteously witnessed on another, which may in turn be
-simultaneously attempting to perform its own \C{Add} or \C{Mult}
-action.  While synchronizing the activities of all nodes to ensure at
-most one such operation is performed at a time is impractical,
-designating a single node to hold the counter state eschewing
-replicating altogether (as in a client-server configuration), is also
-an undesirable solution, given its sensitivity to network partitions
-and server failures.  But, simply replicating counter state among
-nodes, without having any coordination protocol, is an equally
-infeasible approach.
+reasoning principles needed to realize our desired functionality.  As
+long as applications use the library on a single machine, this
+implementation behaves as expected.  However, if the library is used
+in the context of a more sophisticated application, say one whose
+computation is distributed among a collection of machines, its
+behavior becomes significantly harder to understand.  For example, a
+distributed implementation might wish to \emph{replicate} the counter
+state on each node to improve latency (often referred to as
+availability) or fault tolerance.  Unfortunately, replication doesn't
+come for free.  Attempting to update every replicated copy atomically
+is problematic in the absence of sophisticated transaction support,
+which impose sigificant performance and availability penalties.  But,
+without such heavyweight mechanisms, applying an \C{Add} operation on
+one node may not be instanteously witnessed on another, which may in
+turn be simultaneously attempting to perform its own \C{Add} or
+\C{Mult} action.  While synchronizing the activities of all nodes to
+ensure at most one such operation is performed at a time is
+impractical, designating a single node to hold the counter state
+eschewing replication altogether (as in a client-server
+configuration), is also an undesirable solution, given its sensitivity
+to network partitions and server failures.  Removig coordination
+altogether by simply replicating counter state without having any
+supporting consistency protocol is an equally infeasible approach.
 
@@
 when applied at a replica (see \C{apply} in the figure), adds \C{x} to
-the local counter value.  Assuming effects are commutative, they can
-be processed asynchronously using a log local to each replica.  This
-implementation produces an eventually consistent global state in which
-all replicas will eventually contain the same counter value.  While
-effects provide a low-level operational basis for handling replicated
-state, the \emph{ad hoc} nature of the solution confounds desirable
-high-level reasoning principles.  Indeed, the semantic gap between the
-two versions of the counter, one cognizant of replication and the
-other not, breaks backward compatibility.  More significantly, its
-non-trivial construction must be developed in different guises for
+the local counter value.  Note, however, that \C{add} and \C{mult} are
+not commutative operations - assuming two replicas have the same
+initial counter value of 0, applying the effect of adding 3 and then
+multiplying 5 on one replica yields 15, while applying the effecting
+of first multiplying 5 and then adding 3 yields 3 on the other.  Thus,
+implementations must be carefully written to take the lack of
+commutativity into account when defining how effects are applied at a
+replica; in the figure, multiplication is expressed in terms of
+addition to avoid the kind of undesirable behavior described above.
+
+This implementation produces an eventually consistent global state in
+which all replicas will eventually contain the same counter value.
+While effects provide a low-level operational basis for handling
+replicated state, the \emph{ad hoc} nature of the solution confounds
+desirable high-level reasoning principles.  Indeed, the semantic gap
+between the two versions of the counter, one cognizant of replication
+and the other not, breaks backward compatibility.  More significantly,
+its non-trivial construction must be developed in different guises for
 every distributed data structure used by the application.
@@
 \end{figure}
-Lack of composability is another important downside of this approach.
-Consider an application that uses two replicated counters, $c_1$ and
-$c_2$, bound by the invariant $c_2 \ge c_1$, duly enforced by the
-application when updating $c_2$ or $c_1$.  An execution may
+The lack of composability is another important downside of this
+approach.  Consider an application that uses two replicated counters,
+$c_1$ and $c_2$, bound by the invariant $c_2 \ge c_1$, duly enforced
+by the application when updating $c_2$ or $c_1$.  An execution may
 nonetheless witness anamolous states that violate the invariant
-because updates to $c_1$ and $c_2$ are applied independently in any
+because updates to $c_1$ and $c_2$ may be applied independently in any
 order on any replica.  For example, if a replica increments both
@@
 
-On the other hand, a functional programmer is likely to approach the
-problems very differently.  Rather than thinking about a low-level
+On the other hand, a functional programmer is likely to approach this
+problem very differently.  Rather than thinking about a low-level
 operational variant of distributed state given in terms of explicit
@@
 version tree whose root contains the initial counter value, and whose
-branches represent different versions maintained by different nodes,
-with the state produced by a counter's operations on a node recorded
-along the node's local branch.  Now, a globally consistent view of a
-counter can be explained simply in terms of \emph{merge} operations of
-different branch tips.  Treating replication as merging yields a
-counter implementation that bears strong similarity to the original
-sequential one:
+branches represent different (immutable) versions maintained by
+different nodes, with the state produced by a counter's operations on
+a node recorded along the node's local branch.  Now, a globally
+consistent view of a counter can be explained simply in terms of
+\emph{merge} operations of different branch tips.  Treating
+replication as merging yields a counter implementation that bears
+strong similarity to the original sequential one:
   \begin{ocaml}
@@
 contributions of the ancestor.  This interpretation of a replicated
-datatype is thus given in terms of multi-versions of a program state
+datatype is thus given in terms of the evolution of a program state
 impicitly associated with the different nodes that comprise a
-distributed application and a logical notion of a \emph{master}
-representing some consistent, but not necessarily current, view of
-global state.
+distributed application with merge operations serving to communicate
+and reconcile different local states.
 \begin{figure}
@@
 \end{figure}
+The \name programming model realizes a monadic version control system
+in OCaml centered around data, rather than file, coherence.  The \name
+monad lets programmers write and compose concurrent computations
+around multiple (implicit) versions of a mergeable datatype, an
+ordinary datatype additionally equipped with a \emph{merge} function
+responsible for deriving a consistent global state from a collection
+of local versions of that state.  A computation makes progress by
+forking concurrent versions of existing versions, creating later
+versions along a \emph{branch}, or by merging branch tips.
 
-The \name programming model realizes a monadic functional version
-control system centered around data, rather than file coherence.  The
-\name monad lets programmers write and compose concurrent computations
-around multiple (implicit) versions of a mergeable datatype.  A
-computation makes progress by forking concurrent versions of existing
-versions, creating later versions along a \emph{branch}, or by merging
-branch tips.  An important soundness property of the model is that
-instances of a replicated datatype whose merge operation is
-commutative with respect to its local version arguments is guaranteed
-to eventually converge to the same value on all replicas.  As a
-result, OCaml programmers get the benefits of highly available
-(low-latency) distributed computation, while continuing to enjoy the
-comforts of high-level reasoning and familiarity of standard library
-datatypes.  Notably, while \name's programming model makes no explitic
-reference to any specific operational manifestation of a distributed
-system (e.g., programmers do not need to worry about replicas and
-their management), it can be nonetheless efficiently realized on
+While any OCaml datatype can be equipped with a merge operator, and
+thus seamlessly participate in a distributed computation, instances of
+datatypes whose merge operations are commutative with respect to their
+local version state arguments (i.e., \C{v1} and \C{v2} in the counter
+example above) are guaranteed to eventually converge to the same state
+on all replicas.  For such types, OCaml programmers get the benefits
+of realizing highly available (low-latency) distributed computation,
+while continuing to enjoy the comforts of high-level reasoning and
+the familiarity of using standard libraries.  Notably, while \name's
+programming model makes no explicit reference to any specific
+operational manifestation of a distributed system (e.g., programmers
+do not need to worry about replicas and their management), we
+demonstrate that it can be nonetheless efficiently realized on
 existing geo-replicated distributed systems.
@@
 \begin{itemize}
-    \item We formally introduce the concept of a \emph{mergeable} datatype
-      to admit high-level declarative reasoning about distributed
-      computation and replicated state in functional programs.
+    \item We formally introduce the concept of a \emph{mergeable}
+      datatype to admit high-level declarative reasoning about
+      distributed computation and replicated state in ML programs.
 
@@
       behind a monad, and exposes its functionality via a simple API
-      that lets programmers define and compose concurrent OCaml
+      that lets programmers define and compose concurrent ML
       computations around such data.  Issues related to replication
@@
     \item We describe an implementation of \name atop the Irmin
-      persistent store~\cite{...}, a content-addressable storage
+      persistent store~\cite{irmin}, a content-addressable storage
       library for OCaml, along with case studies and experimental
-      results to its practical utility.
+      results that justify its practical utility.
 \end{itemize}

### 2017-02-23 paper/semantics.tex
Edits.

@@
 
-We will not present a series of definitions and results that let us
+We will now present a series of definitions and results that let us
 understand the premises of the \rulelabel{E-Pull-Wait} rule, examine

### 2017-02-23 paper/semantics.tex
LCA Semantics

@@
 For example, in
-Fig.~\ref{fig:criss-cross-lcas}, the merge of 1 with 3 is preceded by
+Fig.~\ref{fig:criss-cross-lcas}, the merge of 0 with 3 is preceded by
 two ``criss-cross'' merges between their respective branches resulting
-in there being two LCAs (3 and 4) for 1 and 3. Multiple LCAs can
-occur even without criss-cross merges, as demonstrated by
-Fig.~\ref{fig:external-lcas}.
+in there being two LCAs (5 and 4) for 0 and 3. Multiple LCAs can occur
+even without criss-cross merges, as demonstrated by
+Fig.~\ref{fig:external-lcas}. The problem with concurrent versions
+with multiple LCAs is that they do not lend themselves to three-way
+merging. This problem also arises in the context of source
+control systems, and often solved via ad-hoc mechanisms.
+GitHub~\cite{github}, for instance, recursively merges LCAs to compute
+a virtual ancestor, which then serves as the LCA for merging the
+concurrent versions. This method is demonstrated for the example in
+Fig.~\ref{fig:criss-cross-lcas}, where LCAs 5 and 4 of 0 and 3 are
+merged (with LCA of 10) to generate -1 as the virtual LCA to merge 5
+and 4. A major downside with this method is that it makes no
+guarantees of the relationship between the virtual ancestor and the
+concurrent versions; the former may not even be a legal ancestor of
+the latter as per the semantics of the data type. For instance, the
+integer type in Fig.~\ref{fig:criss-cross-lcas} may in fact represent
+a bank account, which disallows any activity on the account if balance
+is ever known to be less than zero. Thus, from the perspective of a
+bank account, it doesn't make sense how concurrent versions 3 and 0
+emerged from -1, since the only transition allowed by the semantics
+from -1 is to itself. Clearly, ad hoc mechanisms, which work for the
+text, may not work for more sophisticated data types.
+
+Fortunately, unlike the source control systems where branching
+structure is entirely dictated by the user, \name abstracts away the
+branching structure from the programmer, hence retains the ability to
+constrain it in a way that it deems fit. In particular, \name solves
+the problem of multiple LCAs by suitably constraining the branching
+structure such that the problem never arises. The constraints are
+imposed either implicitly, as a result of how operational semantics
+defines an atomic step, or explicitly, by insisting that certain
+conditions be met before merging a pair of versions
+(\rulelabel{E-Pull-Wait}). Firstly, the operational semantics already
+disables criss-cross merges since it only ever merges versions that
+are latest on their respective branches. For instance let $v_{11}$ and
+$v_{21}$ be the latest versions on branches $b_1$ and $b_2$,
+respectively. Meging $b_2$ into $b_1$ requires entails $v_{21}$ into
+$v_{11}$ to generate version $v_{12}$ on $b_1$. Now, merging $b_1$
+into $b_2$ translates to merging $v_{12}$ into $v_{21}$ (a
+\emph{fast-forward} merge, in Git parlance), but not $v_{11}$ into
+$v_{21}$, thus preventing a criss-cross branching structure. In other
+words, a criss-cross branching structure is prevented due to the
+order among merges between conflicting branches introduced as a result
+of \rulelabel{E-Pull-Wait} being an atomic step.
+
+Secondly, we impose certain pre-conditions on the merging branches to
+preempt the structure shown in Fig.~\ref{fig:external-lcas}. The
+intuition is as follows: consider the branch $b_1$ at the instance of
+merging $b_3$. Since it has already merged $b_2$, a version on $b_2$
+($v_{22}$) could be a common ancestor for a version on $b_1$
+($v_{12}$) and some other version (call it $v$). Now, if $b_1$ merges
+$b_3$, same could be true of $b_3$ and $b_1$: a version on $b_3$
+($v_{32}$) could be a common ancestor for the new version on $b_1$
+($v_{13}$), and the other version $v$. Since $v_{22}$ is also an
+ancestor of $v_{13}$, and both ancestors are not ordered by the
+ancestor relation, $v_{13}$ and $v$ end up with two LCAs. In
+Fig.~\ref{fig:external-lcas}, the role of $v$ is played by the version
+$v_{44}$. We observe that this scenario can be prevented if, when
+merging $b_3$, $b_1$ insists on an ancestor relation between the
+merged version ($v_{22}$) of the $b_2$, the previously merged branch,
+and the latest version of $b_3$, the current merging branch. We call
+the last merged version the \emph{external locus} of the branch. By
+requiring that, for every branch $b$, the external locii of $b$ at
+various points in time (i.e., external locus of every prefix of $b$)
+be totally ordered, we effectively enforce the total order among
+versions that have more than one common ancestor, thereby preventing
+the possibility of multiple LCAs.
+
+%% Multiple versions that merge any pair of versions are totally
+%% ordered. That is, if $v_1$ and $v_2$ are ancestors of $v_3$ and
+%% $v_4$, then $v_3$ and $v_4$ are ordered by the ancestor relation.
 

### 2017-02-23 paper/semantics.tex
Semantics.

@@
 various points in time (i.e., external locus of every prefix of $b$)
-be totally ordered, we effectively enforce the total order among
-versions that have more than one common ancestor, thereby preventing
-the possibility of multiple LCAs.
+be totally ordered, we effectively enforce the invariant that for any
+two common ancestors $v_1$ and $v_2$ between two versions, there
+exists another common ancestor $v_3$ that succeeds $v_1$ and $v_3$ in
+the ancestor relation, thereby preventing the possibility of multiple
+LCAs.
 
+%% <false>
 %% Multiple versions that merge any pair of versions are totally
@@
 %% $v_4$, then $v_3$ and $v_4$ are ordered by the ancestor relation.
+%% </false>
+
+%% If $v_1$ and $v_2$ are ancestors of $v_3$ and $v_4$, then there
+%% exists a $v_5$, a successor of $v_1$ and $v_2$, and an ancestor of
+%% $v_3$ and $v_4$
+%% If I follow one of your locii, or you follow one of my locii, then
+%% we both have same set of external common ancestors.
+
+Let us see how the aforementioned restriction allows us to merge the
+branches in Fig.~\ref{fig:external-lcas} without ever creating
+multiple LCAs. The legal branching history is shown in
+Fig.~\ref{fig:legal-merge}. First, like in
+Fig.~\ref{fig:external-lcas}, $b_2$ is merged into $b_1$ to create
+$v_{12}$, and $b_3$ is merged into $b_4$ to create $v_{42}$. Thus, the
+external locus of $b_1$ is $v_{22}$, and that of $b_4$ is $v_{32}$.
+Observe that now $b_1$ cannot merge $b_3$, neither can $b_4$ merge
+$b_2$ because $v_{22}$ and $v_{32}$ are not related by the ancestor
+relation. Same applies for $b_1$ and $b_4$ because neither one's
+latest version follows from other's external locus.  However, $b_2$
+and $b_3$ can merge among themselves. Let us say they indeed merge to
+to create a version $v_{23}$ on $b_2$. The branch $b_2$ is now
+eligible to be merged into $b_1$ and also $b_4$. These merges do occur
+in Fig.~\ref{fig:legal-merge} to create versions $v_{13}$ and
+$v_{43}$, and making $v_{23}$ the external locus for both $b_1$ and
+$b_4$. Thus, $b_1$ can now merge into $b_4$ to create version $v_{44}$
+on $b_4$.
+
+We now formalize the intuitions described above to precisely define
+the notion of mergeability:
 
@@
 
-\begin{definition} [\bfseries Least Internal and External Ancestors]
-Given a branch $b$ and a version $v\in b$, least internal ancestor is
-the immediate ancestor version ($v_i$) of $v$ on $b$. Least external
-ancestor ($v_o$) is an external ancestor of $v$ that is not an
-ancestor of any other external ancestor of $v$. That is,
-$\under{H}{v_o \preceq v}$, and there does not exist a $v_o' \not\in
-b$ such that $\under{H}{v_o' \preceq v}$ and $\under{H}{v_o \preceq
-v_o'}$. 
-% We lift the definition to the level of branches; $v_o$ is an
-% external locus of a branch $b$ iff it is an external locus of its
-% latest version.
+\begin{definition} [\bfseries External Locus]
+Given a branch $b$ and a version $v\in b$, external locus ($v_o$) of
+$v$ is an external ancestor that is not an ancestor of any other
+external ancestor of $v$. That is, $\under{H}{v_o \preceq v}$, and
+there does not exist a $v_o' \not\in b$ such that $\under{H}{v_o'
+\preceq v}$ and $\under{H}{v_o \preceq v_o'}$. 
 \end{definition}
-
-Note that all branches (except $H(t_{\top})$) begin with a \C{FORK}
-tag, hence every version has at least one least external ancestor. We
-later prove that in the histories generated by the rules in
-Fig.~\ref{fig:opsem}, the external ancestors of every version are
-totally ordered, hence there is exactly one least external ancestor
-for each version. We call the unique least external ancestor of a
-version as its \emph{external locus}. Least internal ancestor, by
-definition, is unique, which we sometimes call \emph{internal locus}.
 
@@
 
-\begin{lemma} [\bfseries Unique External Locus]
-Every version $v$ (except the \C{INIT} version) in the history $H$
-generated by the rules in Fig.~\ref{fig:opsem} has a unique external
-locus.
-\end{lemma}
-\begin{proof}
-By case analysis on the reduction relation. In each case, we assume
-that the unique external locus property holds for the initial history,
-and prove that it holds for the final history.
-
-In \rulelabel{E-Run} case, initial version is the unique external locus of
-the new version on the new branch $H(t)$. In \rulelabel{E-Fork} case,
-which creates the branch $H(t')$ by forking from $H(t)$, the latest
-version on $H(t)$ is the external locus of the first version on
-$H(t)$. In \rulelabel{E-Push} case, the version on $H(t)$ has the same
-external locus as the previous latest version, which, premise says, is
-unique. No new version is created in \rulelabel{E-Pull}. In
-\rulelabel{E-Pull-Wait}, new version is created by merging the latest
-version ($v'$) on $H(t')$ into the latest version $v$ on $H(t)$.
-Premise says that $v$ has a unique external locus (call it $v_o$). The
-new version $v_m$ inherits $v_o$ as a least external ancestor, and
-also has $v'$ as another least external ancestor. However, since
-$\under{H}{v' \mbleto v}$, we have $\under{H}{v_o \preceq v'}$, we
-have $v'$ as the the external locus.
-\qed
-\end{proof}
-
-%% If we are allowed to merge earlier versions on branches, then we
-%% have a possibility of utmost two LCAs.
+Rule \rulelabel{E-Pull-Wait} of the reduction relation
+(Fig.~\ref{fig:opsem}) is enabled only if the latest version ($v'$)
+of thread $t'$'s branch is mergeable into the latest version ($v$) of
+thread $t$'s branch (i.e., $\under{H}{v' \mbleto v}$). Merge computes
+the latest common ancestor of $v$ and $v'$, which is 
+guaranteed to be unique as per the following theorem\footnote{Proof
+included in the appendix}:
 
 \begin{theorem} [\bfseries Unique LCA]
-Every pair of versions $v_1$ and $v_2$ in the history $H$ generated by
-the rules in Fig.~\ref{fig:opsem} have a unique least
-common ancestor. 
+Every pair of versions $v_1$ and $v_2$ in a legal branch history $H$ 
+have a unique least common ancestor. 
 \end{theorem}
-\begin{proof}
-By case analysis on the step relation. In each case, we assume that
-the unique LCA property holds for the initial history, and prove that
-it holds for the final history.
 
-The \rulelabel{E-Run} case is trivial. The \rulelabel{E-Push} case
-creates a new version ($v$) with \C{PUSH} tag at the end of a branch.
-Let $v_1$ be the previous version on the branch, and let $v_2$ be any
-other version. The premise is that $v_1$ and $v_2$ have a unique LCA
-$v_0$. Observe that $v$'s internal locus is $v_1$ and its external
-locus is same as that of $v_1$. Hence, all ancestors of $v_1$
-(including itself) are also ancestors of $v$. Moreover, $v$ has no new
-ancestors except itself, which is the latest version being added.
-Hence, any common ancestor of $v_1$ and $v_2$ is also a common
-ancestor of $v$ and $v_2$, while there cannot be new common ancestors.
-Hence LCA of $v$ and $v_2$ is unique. The reasoning for
-\rulelabel{E-Fork} is similar. The \rulelabel{E-Pull} case is trivial.
-
-The \rulelabel{E-Pull-Wait} case is where the merge happens. Version
-$v'$ is merging into version $v$ to produce the next version $v_m$.
-Let $v''$ be any version. From the premise, we know that $v$ and $v''$
-have a unique LCA (call it $v_1$), and $v'$ and $v''$ have a unique
-LCA (call it $v_2$). Also, we know that $v''$ has a unique external
-locus (call it $v_3$). We now consider cases
-\begin{itemize}
-  \item $v_1$ and $v_2$ are external to both $v$ and $v'$: Hence $v_1$
-  is an external ancestor of $v$, and $v_2$ is an external ancestor
-  of $v'$. Since $v$ and $v'$ are mergeable, there must exist a $v_4$
-  such that $\under{H}{v_1 \preceq v_4}$ and $\under{H}{v_1 \preceq
-  v_4}$, and either $\under{H}{v_4 \preceq v}$ or $\under{H}{v_4
-  \preceq v'}$.  are external ancestors of
-  $v_2$. Hence,
-  $\under{H}{v_0 \preceq v_3} \conj \under{H}{v_1 \preceq v_3}$.
-  Since $\under{H}{v_0 \preceq v_m} \conj \under{H}{v_1 \preceq v_m}$, 
-  $v_2$ and $v_m$ have a unique LCA.
-
-  \item $v_0$ is an internal ancestor of $v_2$, hence an external
-  ancestor of $v$, hence an external ancestor of $v_m$. $v_1$ is
-  an ancestor of $v'$, hence also an external ancestor of $v_m$. 
-\end{itemize}
-
-As proved
-before, the merged version ($v_m$) has a unique external locus, which
-is the merging version ($v'$).
-\end{proof}
+The \rulelabel{E-Pull-Wait} uses the function \C{lca} to compute the
+LCA of (the latest versions on) a pair of branches. The definition of
+the function is standard, hence not discussed.

### 2017-02-23 paper/semantics.tex
Semantics.

@@
 \subcaptionbox[] {\small
-  In this example, 3 and 6 have two LCAs (2 and 1) despite there not
-  being any previous merges between their respective branches.
+  In this example, versions $v_{13}$ and $v_{44}$ have two LCAs
+  ($v_{22}$ and $v_{32}$)  despite there not being any previous merges
+  between their respective branches.
   \label{fig:external-lcas}
@@
 even without criss-cross merges, as demonstrated by
-Fig.~\ref{fig:external-lcas}. The problem with concurrent versions
-with multiple LCAs is that they do not lend themselves to three-way
-merging. This problem also arises in the context of source
-control systems, and often solved via ad-hoc mechanisms.
-GitHub~\cite{github}, for instance, recursively merges LCAs to compute
-a virtual ancestor, which then serves as the LCA for merging the
-concurrent versions. This method is demonstrated for the example in
+Fig.~\ref{fig:external-lcas}. Concurrent versions with multiple LCAs
+do not lend themselves to three-way merging. This problem also arises
+in the context of source control systems, which employ ad hoc
+mechanisms to pave way for three-way merging.  GitHub~\cite{github},
+for instance, recursively merges LCAs to compute a virtual ancestor,
+which then serves as the LCA for merging the concurrent versions. This
+method is demonstrated for the example in
 Fig.~\ref{fig:criss-cross-lcas}, where LCAs 5 and 4 of 0 and 3 are
@@
 
+\begin{figure}
+\centering
+\includegraphics[scale=0.75]{Figures/legal-merge}
+
+\caption{A legal branch history for the example in
+Fig.~\ref{fig:external-lcas}}
+\label{fig:legal-merge}
+\end{figure}
+
 Let us see how the aforementioned restriction allows us to merge the
@@
 in Fig.~\ref{fig:legal-merge} to create versions $v_{13}$ and
-$v_{43}$, and making $v_{23}$ the external locus for both $b_1$ and
-$b_4$. Thus, $b_1$ can now merge into $b_4$ to create version $v_{44}$
-on $b_4$.
+$v_{44}$, and making $v_{23}$ the external locus for both $b_1$ and
+$b_4$. Thus, $b_1$ can now merge into $b_4$ to create version $v_{45}$
+on $b_4$. Note that alternative legal histories are also possible. For
+instance, $b_4$ could have skipped the merge with $v_{23}$ (hence, the
+version $v_{44}$), and directly merged $v_{13}$ in the last step.
+This is legal because $b_4$'s locus before the merge, $v_32$, is an
+ancestor of the merging version $v_{13}$.
 

### 2017-02-23 paper/meta.tex
Meta separated from semantics.

--- 2017-02-23 paper/semantics.tex
@@
-\section{Operational Semantics}
-
-\input{opsem}
-
-We formalize our ideas in the context of a lambda calculus ($\lang$)
-shown in Fig.~\ref{fig:opsem}. Expressions of $\lang$ are variables,
-constants, and \name primitives composed using the lambda combinator.
-For brevity, we use short names for \name primitives: \C{run} for
-\C{with\_init\_version\_do}, and \C{fork} for \C{fork\_version}. To
-simplify the technical development, \name's \C{sync\_next\_version}
-operation is broken down into two primitives - \C{push} and \C{pull},
-which can be lambda-composed get the desired effect:
-\begin{smathpar}
-  \C{sync}\;x \;=\; (\lambda y.\pull)\; (\push\,x)
-\end{smathpar}
-The semantics of\C{get\_current\_version} is subsumed by \C{pull},
-hence elided.  Values ($v$) are constants and lambda abstractions.  A
-program ($p$) is a parallel composition of threads, where each thread
-is an expression ($s$) indexed by the corresponding thread identifier
-($t$).
-
-Fig.~\ref{fig:opsem} also shows the syntax of \emph{branches}, which
-are the artifacts of evaluation and only appear during the run-time. A
-branch is a non-empty sequence of tagged values, where the tag
-captures the abstract run-time operation that led to the creation of
-the value. It is implicitly assumed that each value added to a branch
-is uniquely identifiably, hence no two values on a branch are equal.
-The uniqueness assumption is later extended to a collection of
-branches that constitute a branching structure. A real implementation
-meets this assumption by versioning values across the branches. Thus,
-in reality, branches contain \emph{versions} which denote values. The
-semantics, however, doesn't make this distinction, and uses values and
-versions interchangeably.
-
-Small-step operational semantics of $\lang$ is defined via reduction
-relation ($\stepsto$) that relates \emph{program states}. A program
-state ($p;\,H$) consists of a program $p$ and a \emph{branch history}
-$H$ that maps thread Ids to corresponding branches; each thread is
-associated with a branch during the evaluation. Evaluation contexts
-have been defined separately for expressions ($E$) and programs ($P$),
-with the latter subsuming the former. $E$ is defined to evaluate the
-first argument of a \C{run} expression to a value that constitutes the
-initial version (recall that \C{run} models \name's
-\C{with\_init\_version\_do}). Program evaluation context
-non-deterministically picks one of the threads to evaluate. The admin
-rule that relates transitions of holes to transitions of expressions
-and programs is straightforward, hence elided. Rest of the reduction
-rules are presented in Fig.~\ref{fig:opsem}. For brevity, we write
-$H(t\mapsto (v,f))$ to denote the proposition that $H$ maps $t$ to
-$(v,f)$. The notation $H[t \mapsto (v,f)]$ as usual denotes
-the extension of $H$ with the binding $t \mapsto (v,f)$.
-
-Reduction rules let expression evaluation take a step by rewriting the
-expression and suitably updating the branch history ($H$). The
-\rulelabel{E-Run} rule is applicable only when $H$ is empty, i.e.,
-when no prior branching structure exists. The rule rewrites the
-$\C{run}\;v\;s$ expression to $s$, while creating a new branching
-structure with two branches: a \emph{top} branch that has just the
-initial version (tagged with \C{INIT}), and a branch for the current
-thread ($t$) forked-off from the top branch.  The first version on the
-current branch ($H(t)$) denotes the same value ($v$) as the initial
-version on the top branch, although versions themselves are deemed
-distinct. The new version is tagged with a \C{FORK} tag that keeps the
-record of its orgin, namely the \C{fork} operation and the branch from
-which the current branch is forked. The \rulelabel{E-Fork} rule forks
-a new thread with a fresh id ($t'$) and adds it to the thread pool.
-The corresponding branch ($H(t')$) is forked from the parent thread's
-branch ($H(t)$). The semantics of branch forking is same as described
-above. The \C{fork} expression in the parent thread evaluates to
-\C{()}. The \rulelabel{E-Push} rule creates a new version on the
-current branch ($H(t)$) using the pushed value ($v$).
-\rulelabel{E-App} is the standard beta reduction rule.
-
-The semantics non-deterministically chooses \rulelabel{E-Pull} or
-\rulelabel{E-Pull-Wait} rules to reduce a \C{pull} expression. The
-\rulelabel{E-Pull} rule reduces \C{pull} to \C{()}, and returns the
-latest version on the current branch. The \rulelabel{E-Pull-Wait} rule
-can be thought of as a stutter step; it doesn't reduce \C{pull}, but
-updates the branching structure by merging (the latest version of) a
-concurrent branch ($H(t')$) into (the latest version of) the current
-branch ($H(t)$), and extending the current branch with the merged
-version ($v_m$). The new version is tagged with a \C{MERGE} tag that,
-like a \C{FORK} tag, records its origin. The premises of the
-\rulelabel{E-Pull-Wait} serve the important purpose of contraining the
-branching structure by allowing only the legal merges, and
-consequently preserving certain desirable properties of the system;
-we will go into the details shortly. The \rulelabel{E-Pull-Wait} and
-\rulelabel{E-Pull} rules thus let a thread sync with a subset of
-concurrent threads in multiple steps before returning the result of
-the \C{pull}. Since \C{sync} is a composition of \C{push} and
-\C{pull}, its behavior can be explained thus: \C{sync} pushes the
-given value onto the current (local) branch, merges a (possibly empty)
-subset of concurrent branches into the local branch, and returns the
-result.
-
 We will now present a series of definitions and results that let us
@@
 version $v_{44}$), and directly merged $v_{13}$ in the last step.
-This is legal because $b_4$'s locus before the merge, $v_32$, is an
+This is legal because $b_4$'s locus before the merge, $v_{32}$, is an
 ancestor of the merging version $v_{13}$.

### 2017-02-23 paper/semantics.tex
Meta separated from semantics.

--- 2017-02-23 paper/semantics.tex
@@
 
-We will now present a series of definitions and results that let us
-understand the premises of the \rulelabel{E-Pull-Wait} rule, examine
-the merge operation in closer detail, and appreciate the need for
-constraining the branching structure. First, we formalize the
-intuitive notation of the ancestor relationship between versions of a
-legal branching history (i.e., a branching history generated by the
-rules in Fig.~\ref{fig:opsem}):
-
-\begin{definition} [\bfseries Ancestor]
-Version $v_1$ is a ancestor of version $v_2$ under a history
-$H$ (written $\under{H}{v_1 \preceq v_2}$) if and only if one of the
-following is true:
-\begin{itemize}
-  \item There exists a branch $b$ in $H$ (i.e., $\exists(t\in
-  dom(H)).\,H(t) = b$) in which $v_2$ immediately succeeds
-  $v_1$,
-  \item There exists a branch $b$ in $H$ that contains $(v_2, 
-  \C{FORK}\; (v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
-  \item There exists a branch $b$ in $H$ that contains
-  $(v_2, \C{MERGE}\;(v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
-  \item $v_1 = v_2$, or $v_1$ is transitively a ancestor of
-  $v_2$, i.e., $\exists v.~ \under{H}{v_1 \preceq v} \conj
-  \under{H}{v \preceq v}$ 
-\end{itemize}
-\end{definition}
-
-Ancestor relation is therefore a partial order (reflexive, transitive,
-anti-symmetric) with a greatest lower bound (the initial version).
-Thus, for any two versions in a legal history, there exist at least
-one common ancestor. Ancestor relationships among the common ancestors
-let us define the notion of a least common ancestor (LCA):
-
-\begin{definition} [\bfseries Least Common Ancestor]
-Version $v$ is said to be a common ancestor of versions $v_1$ and
-$v_2$ under a history $H$ if and only if $\under{H}{v \preceq v_1}$
-and $\under{H}{v \preceq v_2}$. It is said to be the least common
-ancestor (LCA) of $v_1$ and $v_2$, iff there does not exist a $v'$
-such that $\under{H}{v' \preceq v_1}$ and $\under{H}{v' \preceq v_2}$
-and $\under{H}{v \preceq v'}$.
-\end{definition}
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.6]{Figures/merge-needs-lca}
-
-\caption{This example of a grow-only counter illustrate why \C{merge}
-needs a least common ancestor, and not just a common ancestor. Both 0
-and 2 are common ancestors of 4 and 7, while 2 is their least common
-ancestor (since $0 \preceq 2$). The result (v) of merging 4 and 7 is
-11 (incorrect) if 0 is used as the common ancestor for merge, and 9
-(correct, because 2+2+3+2 = 9) if 2 is used. }
-\label{fig:merge-needs-lca}
-\end{figure}
-
-\begin{figure}[!t]
-\centering
-\subcaptionbox[] {\small
-  In this example, 1 and 3 have two LCAs (3 and 4) a result of
-  previous merges. The dotted circle denotes a virtual ancestor
-  obtained by merging the two LCAs.
-  \label{fig:criss-cross-lcas}
-} [0.47\columnwidth] {
-  \includegraphics[scale=0.55]{Figures/2-LCAs}
-}
-\hfill
-\subcaptionbox[] {\small
-  In this example, 3 and 6 have two LCAs (2 and 1) despite there not
-  being any previous merges between their respective branches.
-  \label{fig:external-lcas}
-} [0.47\columnwidth] {
-  \includegraphics[scale=0.7]{Figures/2-external-LCAs}
-}
-\caption{Examples where merging versions have more than one LCA}
-\label{fig:many-lcas}
-\end{figure}
-
-When merging two concurrent versions $v_1$ and $v_2$, the common
-ancestor argument for \C{merge} must the LCA of $v_1$ and $v_2$,
-without which \C{merge} may yeild unexpected results. This is
-demonstrated for the grow-only counter in
-Fig.~\ref{fig:merge-needs-lca}, where incorrect count is obtained if a
-common ancestor that is not an LCA is used to merge 4 and 7. While in
-this example there is a unique LCA for 4 and 7, in general this may
-not be the case. With unrestrained branching and merging, there is no
-bound on the number of LCAs a pair of versions can have.
-For example, in
-Fig.~\ref{fig:criss-cross-lcas}, the merge of 0 with 3 is preceded by
-two ``criss-cross'' merges between their respective branches resulting
-in there being two LCAs (5 and 4) for 0 and 3. Multiple LCAs can occur
-even without criss-cross merges, as demonstrated by
-Fig.~\ref{fig:external-lcas}. The problem with concurrent versions
-with multiple LCAs is that they do not lend themselves to three-way
-merging. This problem also arises in the context of source
-control systems, and often solved via ad-hoc mechanisms.
-GitHub~\cite{github}, for instance, recursively merges LCAs to compute
-a virtual ancestor, which then serves as the LCA for merging the
-concurrent versions. This method is demonstrated for the example in
-Fig.~\ref{fig:criss-cross-lcas}, where LCAs 5 and 4 of 0 and 3 are
-merged (with LCA of 10) to generate -1 as the virtual LCA to merge 5
-and 4. A major downside with this method is that it makes no
-guarantees of the relationship between the virtual ancestor and the
-concurrent versions; the former may not even be a legal ancestor of
-the latter as per the semantics of the data type. For instance, the
-integer type in Fig.~\ref{fig:criss-cross-lcas} may in fact represent
-a bank account, which disallows any activity on the account if balance
-is ever known to be less than zero. Thus, from the perspective of a
-bank account, it doesn't make sense how concurrent versions 3 and 0
-emerged from -1, since the only transition allowed by the semantics
-from -1 is to itself. Clearly, ad hoc mechanisms, which work for the
-text, may not work for more sophisticated data types.
-
-Fortunately, unlike the source control systems where branching
-structure is entirely dictated by the user, \name abstracts away the
-branching structure from the programmer, hence retains the ability to
-constrain it in a way that it deems fit. In particular, \name solves
-the problem of multiple LCAs by suitably constraining the branching
-structure such that the problem never arises. The constraints are
-imposed either implicitly, as a result of how operational semantics
-defines an atomic step, or explicitly, by insisting that certain
-conditions be met before merging a pair of versions
-(\rulelabel{E-Pull-Wait}). Firstly, the operational semantics already
-disables criss-cross merges since it only ever merges versions that
-are latest on their respective branches. For instance let $v_{11}$ and
-$v_{21}$ be the latest versions on branches $b_1$ and $b_2$,
-respectively. Meging $b_2$ into $b_1$ requires entails $v_{21}$ into
-$v_{11}$ to generate version $v_{12}$ on $b_1$. Now, merging $b_1$
-into $b_2$ translates to merging $v_{12}$ into $v_{21}$ (a
-\emph{fast-forward} merge, in Git parlance), but not $v_{11}$ into
-$v_{21}$, thus preventing a criss-cross branching structure. In other
-words, a criss-cross branching structure is prevented due to the
-order among merges between conflicting branches introduced as a result
-of \rulelabel{E-Pull-Wait} being an atomic step.
-
-Secondly, we impose certain pre-conditions on the merging branches to
-preempt the structure shown in Fig.~\ref{fig:external-lcas}. The
-intuition is as follows: consider the branch $b_1$ at the instance of
-merging $b_3$. Since it has already merged $b_2$, a version on $b_2$
-($v_{22}$) could be a common ancestor for a version on $b_1$
-($v_{12}$) and some other version (call it $v$). Now, if $b_1$ merges
-$b_3$, same could be true of $b_3$ and $b_1$: a version on $b_3$
-($v_{32}$) could be a common ancestor for the new version on $b_1$
-($v_{13}$), and the other version $v$. Since $v_{22}$ is also an
-ancestor of $v_{13}$, and both ancestors are not ordered by the
-ancestor relation, $v_{13}$ and $v$ end up with two LCAs. In
-Fig.~\ref{fig:external-lcas}, the role of $v$ is played by the version
-$v_{44}$. We observe that this scenario can be prevented if, when
-merging $b_3$, $b_1$ insists on an ancestor relation between the
-merged version ($v_{22}$) of the $b_2$, the previously merged branch,
-and the latest version of $b_3$, the current merging branch. We call
-the last merged version the \emph{external locus} of the branch. By
-requiring that, for every branch $b$, the external locii of $b$ at
-various points in time (i.e., external locus of every prefix of $b$)
-be totally ordered, we effectively enforce the invariant that for any
-two common ancestors $v_1$ and $v_2$ between two versions, there
-exists another common ancestor $v_3$ that succeeds $v_1$ and $v_3$ in
-the ancestor relation, thereby preventing the possibility of multiple
-LCAs.
-
-%% <false>
-%% Multiple versions that merge any pair of versions are totally
-%% ordered. That is, if $v_1$ and $v_2$ are ancestors of $v_3$ and
-%% $v_4$, then $v_3$ and $v_4$ are ordered by the ancestor relation.
-%% </false>
-
-%% If $v_1$ and $v_2$ are ancestors of $v_3$ and $v_4$, then there
-%% exists a $v_5$, a successor of $v_1$ and $v_2$, and an ancestor of
-%% $v_3$ and $v_4$
-%% If I follow one of your locii, or you follow one of my locii, then
-%% we both have same set of external common ancestors.
-
-Let us see how the aforementioned restriction allows us to merge the
-branches in Fig.~\ref{fig:external-lcas} without ever creating
-multiple LCAs. The legal branching history is shown in
-Fig.~\ref{fig:legal-merge}. First, like in
-Fig.~\ref{fig:external-lcas}, $b_2$ is merged into $b_1$ to create
-$v_{12}$, and $b_3$ is merged into $b_4$ to create $v_{42}$. Thus, the
-external locus of $b_1$ is $v_{22}$, and that of $b_4$ is $v_{32}$.
-Observe that now $b_1$ cannot merge $b_3$, neither can $b_4$ merge
-$b_2$ because $v_{22}$ and $v_{32}$ are not related by the ancestor
-relation. Same applies for $b_1$ and $b_4$ because neither one's
-latest version follows from other's external locus.  However, $b_2$
-and $b_3$ can merge among themselves. Let us say they indeed merge to
-to create a version $v_{23}$ on $b_2$. The branch $b_2$ is now
-eligible to be merged into $b_1$ and also $b_4$. These merges do occur
-in Fig.~\ref{fig:legal-merge} to create versions $v_{13}$ and
-$v_{43}$, and making $v_{23}$ the external locus for both $b_1$ and
-$b_4$. Thus, $b_1$ can now merge into $b_4$ to create version $v_{44}$
-on $b_4$.
-
-We now formalize the intuitions described above to precisely define
-the notion of mergeability:
-
-\begin{definition} [\bfseries Internal and External Ancestors]
-Given a branch $b$ and a version $v\in b$, an internal ancestor of $v$
-is an ancestor from the same branch $b$. An external ancestor of $v$
-is an ancestor from a different branch $b'\neq b$. 
-\end{definition}
-
-\begin{definition} [\bfseries External Locus]
-Given a branch $b$ and a version $v\in b$, external locus ($v_o$) of
-$v$ is an external ancestor that is not an ancestor of any other
-external ancestor of $v$. That is, $\under{H}{v_o \preceq v}$, and
-there does not exist a $v_o' \not\in b$ such that $\under{H}{v_o'
-\preceq v}$ and $\under{H}{v_o \preceq v_o'}$. 
-\end{definition}
-
-\begin{definition} [\bfseries Mergeability]
-Given a history $H$, a version $v_1$ and a version $v_2$ that is not
-an ancestor of $v_1$ under $H$, $v_2$ is mergeable into $v_1$ (denoted
-$\under{H}{v_2 \mbleto v_1}$) iff $v_1$'s external locus is an
-ancestor of $v_2$, or $v_2$'s external locus is an ancestor of $v_1$.
-\end{definition}
-
-Rule \rulelabel{E-Pull-Wait} of the reduction relation
-(Fig.~\ref{fig:opsem}) is enabled only if the latest version ($v'$)
-of thread $t'$'s branch is mergeable into the latest version ($v$) of
-thread $t$'s branch (i.e., $\under{H}{v' \mbleto v}$). Merge computes
-the latest common ancestor of $v$ and $v'$, which is 
-guaranteed to be unique as per the following theorem\footnote{Proof
-included in the appendix}:
-
-\begin{theorem} [\bfseries Unique LCA]
-Every pair of versions $v_1$ and $v_2$ in a legal branch history $H$ 
-have a unique least common ancestor. 
-\end{theorem}
-
-The \rulelabel{E-Pull-Wait} uses the function \C{lca} to compute the
-LCA of (the latest versions on) a pair of branches. The definition of
-the function is standard, hence not discussed.
+\input{meta.tex}

### 2017-02-24 paper/meta.tex
Meta

@@
-We will now present a series of definitions and results that let us
-understand the premises of the \rulelabel{E-Pull-Wait} rule, examine
-the merge operation in closer detail, and appreciate the need for
-constraining the branching structure. First, we formalize the
-intuitive notation of the ancestor relationship between versions of a
-legal branching history (i.e., a branching history generated by the
-rules in Fig.~\ref{fig:opsem}):
+To define the semantics of LCA, we formalize the intuitive notation of
+the ancestor relationship between versions of a legal branching
+history (i.e., a branching history generated by the rules in
+Fig.~\ref{fig:opsem}):
 
@@
 branching structure from the programmer, hence retains the ability to
-constrain it in a way that it deems fit. In particular, \name solves
-the problem of multiple LCAs by suitably constraining the branching
-structure such that the problem never arises. The constraints are
-imposed either implicitly, as a result of how operational semantics
-defines an atomic step, or explicitly, by insisting that certain
-conditions be met before merging a pair of versions
-(\rulelabel{E-Pull-Wait}). Firstly, the operational semantics already
-disables criss-cross merges since it only ever merges versions that
-are latest on their respective branches. For instance let $v_{11}$ and
-$v_{21}$ be the latest versions on branches $b_1$ and $b_2$,
-respectively. Meging $b_2$ into $b_1$ requires entails $v_{21}$ into
-$v_{11}$ to generate version $v_{12}$ on $b_1$. Now, merging $b_1$
-into $b_2$ translates to merging $v_{12}$ into $v_{21}$ (a
-\emph{fast-forward} merge, in Git parlance), but not $v_{11}$ into
-$v_{21}$, thus preventing a criss-cross branching structure. In other
-words, a criss-cross branching structure is prevented due to the
-order among merges between conflicting branches introduced as a result
-of \rulelabel{E-Pull-Wait} being an atomic step.
+manifest it in a way that it deems fit. In particular, \name solves
+the problem of multiple LCAs either by constraining the branching
+stucture to preempt multiple LCAs, or by extending it such that
+merging branches have a new unique LCA. The preemption is done for
+criss-cross merges by the operational semantics, which only allows
+merging versions that are latest on their respective branches. For
+instance let $v_{11}$ and $v_{21}$ be the latest versions on branches
+$b_1$ and $b_2$, respectively. Merging $b_2$ into $b_1$ entails
+merging $v_{21}$ into $v_{11}$ to generate version $v_{12}$ on $b_1$.
+Now, merging $b_1$ into $b_2$ involves merging $v_{12}$, the latest
+version on $b_1$, into $v_{21}$, but not $v_{11}$ into $v_{21}$, thus
+preempting a criss-cross branching structure. 
+% In other words, a criss-cross branching structure is prevented due
+% to the order among merges between conflicting branches introduced as
+% a result of \rulelabel{E-Pull-Wait} being an atomic step.
 
-Secondly, we impose certain pre-conditions on the merging branches to
-preempt the structure shown in Fig.~\ref{fig:external-lcas}. The
-intuition is as follows: consider the branch $b_1$ at the instance of
-merging $b_3$. Since it has already merged $b_2$, a version on $b_2$
-($v_{22}$) could be a common ancestor for a version on $b_1$
-($v_{12}$) and some other version (call it $v$). Now, if $b_1$ merges
-$b_3$, same could be true of $b_3$ and $b_1$: a version on $b_3$
-($v_{32}$) could be a common ancestor for the new version on $b_1$
-($v_{13}$), and the other version $v$. Since $v_{22}$ is also an
-ancestor of $v_{13}$, and both ancestors are not ordered by the
-ancestor relation, $v_{13}$ and $v$ end up with two LCAs. In
-Fig.~\ref{fig:external-lcas}, the role of $v$ is played by the version
-$v_{44}$. We observe that this scenario can be prevented if, when
-merging $b_3$, $b_1$ insists on an ancestor relation between the
-merged version ($v_{22}$) of the $b_2$, the previously merged branch,
-and the latest version of $b_3$, the current merging branch. We call
-the last merged version the \emph{external locus} of the branch. By
-requiring that, for every branch $b$, the external locii of $b$ at
-various points in time (i.e., external locus of every prefix of $b$)
-be totally ordered, we effectively enforce the invariant that for any
-two common ancestors $v_1$ and $v_2$ between two versions, there
-exists another common ancestor $v_3$ that succeeds $v_1$ and $v_3$ in
-the ancestor relation, thereby preventing the possibility of multiple
-LCAs.
+Next, for the branching structures where a pair of branches (i.e.,
+their respective latest versions) have multiple externally-located
+LCAs, the problem can be fixed by suitably extending the branching
+structure such that the branches have a new single LCA. For instance,
+consider the example in Fig.~\ref{fig:external-lcas}, where (latest
+versions on) branches $b_1$ and $b_4$ have two externally-located LCAs
+($v_{22}$ and $v_{32}$), hence cannot be merged. Mergeability can
+however be reclaimed by extending the branching structure as shown in
+Fig.~\ref{fig:legal-extension-1}. The extension is done by first merging
+$b_3$ into $b_2$ to create version $v_{23}$ on $b_2$, then merging
+$b_2$ into $b_1$ to create version $v_{14}$ on $b_1$, and finally
+merging $b_2$ into $b_4$ to create version $v_{45}$ on $b_4$. Observe
+that $b_1$ and $b_4$ now have a single LCA ($v_{23}$) hence are now
+mergeable. Mergeability could also have been reclaimed by adopting an
+alternative extension to the branching structure as shown in
+Fig.~\ref{fig:legal-extension-2}.
 
-%% <false>
-%% Multiple versions that merge any pair of versions are totally
-%% ordered. That is, if $v_1$ and $v_2$ are ancestors of $v_3$ and
-%% $v_4$, then $v_3$ and $v_4$ are ordered by the ancestor relation.
-%% </false>
-
-%% If $v_1$ and $v_2$ are ancestors of $v_3$ and $v_4$, then there
-%% exists a $v_5$, a successor of $v_1$ and $v_2$, and an ancestor of
-%% $v_3$ and $v_4$
-%% If I follow one of your locii, or you follow one of my locii, then
-%% we both have same set of external common ancestors.
-
-\begin{figure}
+\begin{figure}[!t]
 \centering
-\includegraphics[scale=0.75]{Figures/legal-merge}
-
-\caption{A legal branch history for the example in
-Fig.~\ref{fig:external-lcas}}
-\label{fig:legal-merge}
+\subcaptionbox[] {\small
+  $b_3$ is merged into $b_2$ to create $v_{23}$ here.
+  \label{fig:legal-extension-1}
+} [0.47\columnwidth] {
+  \includegraphics[scale=0.8]{Figures/legal-extension-1}
+}
+\hfill
+\subcaptionbox[] {\small
+  $b_2$ is merged into $b_3$ to create $v_{33}$ here.
+  \label{fig:legal-extension-2}
+} [0.47\columnwidth] {
+  \includegraphics[scale=0.8]{Figures/legal-extension-2}
+}
+\caption{Legal extensions of the branching structure in
+  Fig.~\ref{fig:external-lcas} that reclaim the mergeability of $b_1$
+  and $b_4$}
+\label{fig:legal-extensions}
 \end{figure}
 
-Let us see how the aforementioned restriction allows us to merge the
-branches in Fig.~\ref{fig:external-lcas} without ever creating
-multiple LCAs. The legal branching history is shown in
-Fig.~\ref{fig:legal-merge}. First, like in
-Fig.~\ref{fig:external-lcas}, $b_2$ is merged into $b_1$ to create
-$v_{12}$, and $b_3$ is merged into $b_4$ to create $v_{42}$. Thus, the
-external locus of $b_1$ is $v_{22}$, and that of $b_4$ is $v_{32}$.
-Observe that now $b_1$ cannot merge $b_3$, neither can $b_4$ merge
-$b_2$ because $v_{22}$ and $v_{32}$ are not related by the ancestor
-relation. Same applies for $b_1$ and $b_4$ because neither one's
-latest version follows from other's external locus.  However, $b_2$
-and $b_3$ can merge among themselves. Let us say they indeed merge to
-to create a version $v_{23}$ on $b_2$. The branch $b_2$ is now
-eligible to be merged into $b_1$ and also $b_4$. These merges do occur
-in Fig.~\ref{fig:legal-merge} to create versions $v_{13}$ and
-$v_{44}$, and making $v_{23}$ the external locus for both $b_1$ and
-$b_4$. Thus, $b_1$ can now merge into $b_4$ to create version $v_{45}$
-on $b_4$. Note that alternative legal histories are also possible. For
-instance, $b_4$ could have skipped the merge with $v_{23}$ (hence, the
-version $v_{44}$), and directly merged $v_{13}$ in the last step.
-This is legal because $b_4$'s locus before the merge, $v_{32}$, is an
-ancestor of the merging version $v_{13}$.
+Note that the extension strategy described above is possible only
+because operational semantics imposes no obligation to merge any two
+concurrent versions. The high-level purpose of merging branches in
+\name is to propagate the data across the branches, which can be done
+by merging any two concurrent versions that contain the data; the
+semantics lets us pick these versions. While mergeability of branches
+with multiple LCAs can be reclaimed by actively extending the
+branching structure as described above, the semantics is
+\emph{self-correcting} in the sense that a fair execution reclaims the
+mergeability of any two branches without any intervention. This is
+possible because the system preempts criss-cross merges, thus letting
+the semantics make progress from any state. The property is formalized
+by the following theorem\footnote{The more precise statement of the
+theorem, and its proof are included in the appendix}:
 
-We now formalize the intuitions described above to precisely define
-the notion of mergeability:
-
-\begin{definition} [\bfseries Internal and External Ancestors]
-Given a branch $b$ and a version $v\in b$, an internal ancestor of $v$
-is an ancestor from the same branch $b$. An external ancestor of $v$
-is an ancestor from a different branch $b'\neq b$. 
-\end{definition}
-
-\begin{definition} [\bfseries External Locus]
-Given a branch $b$ and a version $v\in b$, external locus ($v_o$) of
-$v$ is an external ancestor that is not an ancestor of any other
-external ancestor of $v$. That is, $\under{H}{v_o \preceq v}$, and
-there does not exist a $v_o' \not\in b$ such that $\under{H}{v_o'
-\preceq v}$ and $\under{H}{v_o \preceq v_o'}$. 
-\end{definition}
-
-\begin{definition} [\bfseries Mergeability]
-Given a history $H$, a version $v_1$ and a version $v_2$ that is not
-an ancestor of $v_1$ under $H$, $v_2$ is mergeable into $v_1$ (denoted
-$\under{H}{v_2 \mbleto v_1}$) iff $v_1$'s external locus is an
-ancestor of $v_2$, or $v_2$'s external locus is an ancestor of $v_1$.
-\end{definition}
-
-Rule \rulelabel{E-Pull-Wait} of the reduction relation
-(Fig.~\ref{fig:opsem}) is enabled only if the latest version ($v'$)
-of thread $t'$'s branch is mergeable into the latest version ($v$) of
-thread $t$'s branch (i.e., $\under{H}{v' \mbleto v}$). Merge computes
-the latest common ancestor of $v$ and $v'$, which is 
-guaranteed to be unique as per the following theorem\footnote{Proof
-included in the appendix}:
-
-\begin{theorem} [\bfseries Unique LCA]
-Every pair of versions $v_1$ and $v_2$ in a legal branch history $H$ 
-have a unique least common ancestor. 
+\begin{theorem} [\bfseries Semantics Self-Correcting]
+??
 \end{theorem}

### 2017-02-24 paper/semantics.tex
Meta

@@
 version ($v_m$). The new version is tagged with a \C{MERGE} tag that,
-like a \C{FORK} tag, records its origin. The premises of the
-\rulelabel{E-Pull-Wait} serve the important purpose of contraining the
-branching structure by allowing only the legal merges, and
-consequently preserving certain desirable properties of the system;
-we will go into the details shortly. The \rulelabel{E-Pull-Wait} and
-\rulelabel{E-Pull} rules thus let a thread sync with a subset of
-concurrent threads in multiple steps before returning the result of
-the \C{pull}. Since \C{sync} is a composition of \C{push} and
-\C{pull}, its behavior can be explained thus: \C{sync} pushes the
-given value onto the current (local) branch, merges a (possibly empty)
-subset of concurrent branches into the local branch, and returns the
-result.
+like a \C{FORK} tag, records its origin. The rule assumes a function
+\C{lca} that computes the \emph{least common ancestor} (LCA) for the
+latest versions on the given pair of branches. The semantics of LCA is
+discussed below. The \rulelabel{E-Pull-Wait} and \rulelabel{E-Pull}
+rules thus let a thread sync with a subset of concurrent threads in
+multiple steps before returning the result of the \C{pull}. Since
+\C{sync} is a composition of \C{push} and \C{pull}, its behavior can
+be explained thus: \C{sync} pushes the given value onto the current
+(local) branch, merges a (possibly empty) subset of concurrent
+branches into the local branch, and returns the result.
 

### 2017-02-24 paper/system-model.tex
Instantiations subsection complete.

\subsection{Instantiations}

\begin{figure}[!t]
\centering
\subcaptionbox[] {\small
  Branches $b_1$, $b_2$ and $b_3$ mapped to nodes $n_1$, $n_2$ and
  $n_3$. Each time step captures concurrent operations among the
  nodes.
  \label{fig:instantiation-1}
} [0.47\columnwidth] {
  \includegraphics[scale=0.6]{Figures/instantiation-1}
}
\hfill
\subcaptionbox[] {\small
  Network partitions may leave branches $b_1$ and $b_4$ in one
  partition (red background), and branches $b_2$ and $b_3$ in the
  other (yellow background). Since $b_2$ and $b_3$ cannot merge unless
  $b_1$ and $b_4$ merge (Fig.~\ref{fig:legal-extensions}), they cannot
  communicate among themselves even though they are in the same
  network partition.
} [0.47\columnwidth] {
  \includegraphics[scale=0.8]{Figures/partitions}
}
\caption{Instantiations of the \name concurrency model in a
distributed setting}
\label{fig:partitions}
\end{figure}

The operational semantics of \name describe a concurrency model based
on versions and branches for computations on replicated data, without
referring to any particular instantiation. A useful instantiation of
the model is for a distributed system with multiple nodes, where each
each node hosts a \emph{replica} of the data. Thus, each node
instantiates a branch. The node responds to user requests to perform
updates to the replica, and publishes the results on its local branch
as newer versions. The node also stores (a prefix of) the branching
history - a knowledge about the system that it accumulates over time.
In response to the publication of a new local version, it contacts
other nodes to obtain their latest local versions, in the process
updating its knowledge about the branching history. If a later version
($v_j$) on a remote branch ($b_j$) is mergeable with the latest
version ($v_i$) on the local branch ($b_i$), i.e., if $v_j \not\preceq
v_i$ and LCA of $b_i$ and $b_j$ is unique, the node initiates the
merge process with the remote node, where in it lets the remote node
know that a new version ($v_i'$) later than $v_i$ is being created on
$b_j$ by merging version $v_j$ of $b_j$. This prevents the remote node
from merging versions $v_i$ or earlier of $b_i$, and creating a
criss-cross branching structure. The later merge of $b_i$ into $b_j$
by the remote node will only consider versions $v_i'$ or later. The
merge operation is a function over participating branches, and occurs
in a single time step, extending one of the branches. Concurrent
merges that do not seek to extend a branch participating in the
current merge operation are allowed. The resultant history of merges
will still be linearizable. The example in
Fig.~\ref{fig:instantiation-1} illustrates. Here, in the time step
$t=1$, nodes $n_1$ and $n_2$ communicate to merge branch $b_2$ in
$b_1$. Concurrently, $n_1$ and $n_3$ can also communicate to merge
$b_2$ into $b_3$. Likewise, at $t=1$, merges from $b_1$ into $b_2$ and
$b_1$ into $b_3$ happen concurrently. The resultant history is
equivalent to performing the merges in the order as they were
described above.

In presence of network partitions, the nodes in one partition may not
receive updates from the other partition, hence merges cannot happen
across a partition. This can be a problem if a pair of branches in one
partition (the current partition) have multiple LCAs in the other
partition (the remote partition), hence cannot merge.
Fig.~\ref{fig:partitions} illustrate this problem for the branching
structure of Fig.~\ref{fig:external-lcas}. Fortunately, the access to
full (causal) branching history at every node comes to the rescue.
Relying on the history, the current partition can fork-off (\C{fork})
new branches that start where the remote branches left, and map them
to new (virtual) nodes that emulate the nodes in the remote partition.
For the example in Fig.~\ref{fig:partitions}, the partition of $b_1$
and $b_4$ can fork new branches $b_2'$ and $b_3'$ from $b_2$ and
$b_3$, respectively, and resume the activity as if the partition never
happened. After merging the first versions on $b_2'$ and $b_3'$,
branches $b_1$ and $b_2$ will be in the same state as before (when the
partition happened), except that the branches are now mergeable
because $b_2'$ and $b_3'$ can merge. The ability to track full
provenance information is thus crucial for \name to overcome network
partitions, making it an appropriate programming model for
highly-available replicated data types.

### 2017-02-24 paper/mergeable-types.tex
Sec. 4 beginning.

\section{Mergeable Data Type}

\begin{figure}[!t]
\centering
\subcaptionbox[] {\small
  When a thread \C{pushes} a new version, it is presumably a semantic
  successor of the version it last \C{pulled} into the heap.
  \label{fig:syntactic-ancestor-1}
} [0.47\columnwidth] {
  \includegraphics[scale=0.7]{Figures/semantic-ancestor-1}
}
\hfill
\subcaptionbox[] {\small
  Versions created by the \C{merge} operation are syntactic successors
  of merged versions, but need not necessarily be semantic successors.
  \label{fig:syntactic-ancestor-2}
} [0.47\columnwidth] {
  \includegraphics[scale=0.8]{Figures/semantic-ancestor-2}
}
\caption{New versions are created from existing versions either
through \C{push} or \C{merge}.}
\label{fig:syntactic-ancestors}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=0.4]{Figures/pushouts}

\caption{Commutative diagram representing the merge operation.}
\label{fig:pushouts}
\end{figure}

We now formally define what we mean by a mergeable type. The
definition serves as a guide to write merge functions for functional
data types, and elevate them to the level of replicated data types.

A functional data type library is as a tuple consisting of a type
($t$), and a (possibly infinite) collection of functions ($f$, $g$,
etc.) of type $t \rightarrow t$. We call the functions \emph{primitive
morphisms}\footnote{We borrow terminology from category theory for
convenience}. A morphism ($P$, $Q$, etc.) is either a primitive
morphism, or an associative composition of morphisms ($P \circ Q \circ
R$). An object $B:t$ is called a \emph{semantic successor} of $A:t$
(conversely, $A$ is called a \emph{semantic ancestor} of $B$) if and
only if there exists a morphism $P$ such that $P(A) = B$.

The aim of a three-way merge function over a type $t$ is to merge a
pair of semantic successors, $B:t$ and $C:t$, of an object $A:t$, into
another object $D:t$ such that the relationship between the semantic
successors and $D$ satisfies certain conditions. These conditions can
be understood by observing the \rulelabel{E-Pull-Wait} rule of of
Fig.~\ref{fig:opsem}, which applies the \C{merge} function to a pair
of concurrent versions ($v_1$ and $v_2$) and their least common
ancestor ($v$). Thus, the only relationship that exists between $v$
and $v_1$, and $v$ and $v_2$ is the syntactic ancestor relationship
that follows from the branching structure. The merge function, as
described above, however assumes that the concurrent versions are
semantic successors of their LCA. It is therefore essential to
maintain coherence between the syntactic and semantic ancestor
relations.

The ways in which syntactic ancestor relationships are created among
versions is captured in Fig.~\ref{fig:syntactic-ancestors}. Whenever an
object is pushed onto the branch, an ancestor relationship is created
between the previous version $v_1$, and the newly created version
$v_2$ (Fig.~\ref{fig:syntactic-ancestor-1}). However, since $v_2$ is
pushed by the thread after reading $v_1$ into the heap, it is
reasonable to assume that $v_2$ is a result of applying a morphism $P$
to $v_1$ (i.e., $\exists P.~v_2 = P(v_1)$).  Hence, $v_1$ is a
semantic ancestor of $v_2$.  Fig.~\ref{fig:syntactic-ancestor-2}
captures another way of establishing ancestor relationship, namely by
merging branches. Version $v_{21}$ on branch $b_2$ is merged into
$v_{11}$ on $b_1$ to create $v_{12}$ on $b_1$. Versions $v_{11}$ and
$v_{21}$ are now syntactic ancestors of $v_{12}$, but for them to be
semantic ancestors, merge function has to enforce the relationship
explicitly. In other words, the result of merging of a pair of
semantic successors, $B$ and $C$, of an $A$, has to be an object $D$
that is a semantic successor of both $B$ and $C$. We now formalize
this intuition to define a mergeable type.
\begin{definition} [\bfseries Mergeable Type]
A type $t$ is said to be mergeable, if and only if there exists a
function $M$ of type $t \times t \times t \rightarrow (t \rightarrow
t)\times(t \rightarrow t)$ that satisfies the following property:
$\forall (A, B, C\,:\,t).~ (\exists (P,Q\,:\, t \rightarrow t).~ B =
P(A) \conj C = Q(A)) \Rightarrow (\exists (P',Q': t\rightarrow
t).~M(A,B,C) = (P',Q') \conj  Q'(B) = P'(C))$
\end{definition}
Intuitively, a type $t$ is a mergeable type if, whenever there exist
two morphisms $P$ and $Q$ that map $A:t$ to $B:t$ and $C:t$, there
must also exist morphisms $P'$ and $Q'$, the map both $B$ and $C$ to
$D:t$. This intuition is expressed visusally in
Fig.~\ref{fig:pushouts}.

The three-way merge function \C{merge} can be defined
straightforwardly in terms of the function $M$ described above. Thus,
for a mergeable type $t$, we have the guarantee that a sequence of
values read by a thread from its branch ``makes sense'' as per the
data type semantics:

\begin{theorem} [\bfseries Branch-local consistency]
Let $t$ be a mergeable type, and let $H$ be a legal branching history
over values of type $t$. For every pair of values $v_1:t$ and $v_2:t$
along a branch $b$ in $H$, if $\under{H}{v_1 \preceq v_2}$, then there
exists a morphism $P:t\rightarrow t$ such that $v_2 = P(v_1)$.
\end{theorem}

### 2017-02-25 paper/mergeable-types.tex

@@
 \begin{definition} [\bfseries Mergeable Type]
+\label{def:mergeable-type}
 A type $t$ is said to be mergeable, if and only if there exists a
@@
 
-The three-way merge function \C{merge} can be defined
-straightforwardly in terms of the function $M$ described above. Thus,
-for a mergeable type $t$, we have the guarantee that a sequence of
+For a mergeable type $t$, we have the guarantee that a sequence of
 values read by a thread from its branch ``makes sense'' as per the
@@
 \begin{theorem} [\bfseries Branch-local consistency]
+\label{thm:branch-consistency}
 Let $t$ be a mergeable type, and let $H$ be a legal branching history
@@
 \end{theorem}
+
+Def.~\ref{def:mergeable-type} specifies the merge function ($M$)
+slightly differently; instead of returning a merged value like the
+three-way merge function (\C{merge}), $M$ returns a pair of morphisms.
+This specification is useful to preempt certain non-sensical merge
+functions, and thus enforce the branch-local consistency property
+(Theorem~\ref{thm:branch-consistency}). For instance, consider the
+counter data type from Sec.~\ref{sec:intro}. A three-way merge
+function can be defined to return a constant regardless of its
+arguments:
+\begin{ocaml}
+  let merge lca a b = 0
+\end{ocaml} Such a merge function allows the client (a thread) to
+witness the violation of the branch-local consistency property. For
+example, the client may \C{sync} the counter, whose current local
+value is $10$, and obtain $0$ as a result, thus witnessing a violation
+of counter's monotonicity property. Fortunately,
+Def.~\ref{def:mergeable-type} disallows this semantics. It is
+impossible to define the function $M$ for the counter data type that
+violates the monotonicity invariant, because no counter morphism
+violates the invariant. In general, the specification of $M$
+guarantees that the merge operation for a type $t$ preserves any
+invariants (e.g., balancedness of a tree, sortedness of a list,
+non-negativeness of an integer etc) that are preserved by all of $t$'s
+morphisms. On the hand, the commutativity of the diagram in
+Fig.~\ref{fig:pushouts} may not be enforced statically. It can however
+be verified at run-time by checking that the morphisms returned by $M$
+map the respective concurrent versions to the same value. The
+operational semantics (Fig.~\ref{fig:opsem}) can be extended to this
+effect.
+
+% While the merge function $M$ as specified by
+% Def.~\ref{def:mergeable-type} has more verification value, we continue
+% to use the three-way merge function \C{merge} in our examples for
+% flexibility.
+
+Merge functions for simple data types are easy to write. For instance,
+the merge logic for counter is simple enough that it doesn't require
+explicit efforts to identify $P$ and $Q$ that led to the
+concurrent counter versions, and transform them into $P'$ and $Q'$
+such that the diagram in Fig.~\ref{fig:pushouts} commutes. For more
+sophisticated data types, this process is more involved, and requires
+the programmer to perform non-trivial reasoning. We now present an
+alternative definition of a mergeable type that is stronger than
+Def.~\ref{def:mergeable-type}, but also serves as a recipe to build
+non-trivial merge functions.
+
+\begin{definition} [\bfseries Mergeable Type]
+\label{def:mergeable-type-2}
+A type $t$ is said to be mergeable if and only if the following
+conditions hold:
+\begin{itemize}
+  \item There exists a function $W: t \times t \rightarrow (t
+  \rightarrow t)\; \C{list}$ that accepts an object $A:t$ and its
+  semantic successor $B:t$, and returns a minimal sequence of
+  primitive morphisms $P^*$, whose composition maps $A$ to $B$. That
+  is, $W(A,B) = P^*$, where $\denot{P^*}(A) = B$, and there does not
+  exist an $R^*$ such that $\denot{R^*}(A) = B$ and $\length{R^*} <
+  \length{P^*}$
+
+  \item There exists a function $M: (t \rightarrow t)^*\!\times\!(t
+  \rightarrow t)^* \;\rightarrow\; (t \rightarrow t)\!\times\!(t
+  \rightarrow t)$ that accepts a pair of minimal sequences of
+  primitive morphisms, $P^*$ and $Q^*$, and returns a pair of morphisms,
+  $Q'$ and $P'$, such that for any object $A:t$, $(Q' \circ
+  \denot{P^*})(A) = (P' \circ \denot{Q^*})(A)$.  
+\end{itemize}
+\end{definition}
+
+\paragraph{Strings} Strings are mergeable. Given a string ($s$) and
+its semantic successor ($s'$), an edit script ($P^*$ or $Q^*$) is the
+shortest sequence of \C{insert}, \C{delete}, and \C{substitute}
+operations (string's primitive morphisms), whose composition takes $s$
+to $s'$. The length of the sequence is in fact the Levenshtein
+distance~\cite{levenstein} between $s$ and $s'$, which can be computed
+in polynomial time via Wagner-Fischer algorithm~\cite{wagner-fischer}.
+A slight modification of Wagner-Fischer also lets us compute the edit
+script. Fig.~\ref{fig:mergeable-string-eg} shows the edit script ...
+
+Next, we apply operational transformation to compute $P^*$ given $Q^*$
+($P^* <+ Q^*$), \emph{or} $Q^*$ given $P^*$ ($Q^* <+ P^*$); there is
+no need to compute both unless verification is one of the aims.
+
+\paragraph{}

### 2017-02-25 paper/mergeable-types.tex
Mergeability Def. 2

@@
 
-\caption{Commutative diagram representing the merge operation.}
+\caption{Commutative diagram for Def.~\ref{def:mergeable-type}.}
 \label{fig:pushouts}
@@
 
+\begin{figure}
+\centering
+\includegraphics[scale=0.4]{Figures/pushouts-2}
+
+\caption{Commutative diagram for Def.~\ref{def:mergeable-type-2}.}
+\label{fig:pushouts-2}
+\end{figure}
+
 Def.~\ref{def:mergeable-type} specifies the merge function ($M$)
@@
 
+\paragraph{Notation} As mentioned earlier, we distinguish between the
+named functions defined by a data type library (primitive morphisms),
+and their compositions (morphisms). A sequence (list) of primitive
+morphisms is written $P^*$ or $Q^*$. The composition of a sequence of
+primitive morphisms (written $\denot{P^*}$) is their right-to-left
+composition, i.e., $\C{fold\_left}\;(\lambda f.\lambda g. f \circ
+g)\; id\; P^*$, where $\C{fold\_left}$ has its usual OCaml type. The
+length of a sequence $S$ is written $\length{S}$.
+
 \begin{definition} [\bfseries Mergeable Type]
@@
   \item There exists a function $W: t \times t \rightarrow (t
-  \rightarrow t)\; \C{list}$ that accepts an object $A:t$ and its
+  \rightarrow t)^*$ that accepts an object $A:t$ and its
   semantic successor $B:t$, and returns a minimal sequence of
@@
   primitive morphisms, $P^*$ and $Q^*$, and returns a pair of morphisms,
-  $Q'$ and $P'$, such that for any object $A:t$, $(Q' \circ
+  $P'$ and $Q'$, such that for any object $A:t$, $(Q' \circ
   \denot{P^*})(A) = (P' \circ \denot{Q^*})(A)$.  
 \end{itemize}
+Commutativity diagram of Fig.~\ref{fig:pushouts-2} visualizes this
+definition. 
 \end{definition}
+
+The definition asserts the existence of two functions for a mergeable
+type $t$. First, the function $W$, which computes the \emph{edit
+distance} between an object $A:t$ and its semantic successor $B:t$.
+The sequence of primitive morphisms it returns ($P^*$) is called an
+\emph{edit script}. Replaying the edit script on $A$ produces $B$
+(i.e., $\denot{P^*}(A) = B$). Second, the definition asserts the
+existence of a function $M$, which performs an \emph{operational
+transformation} of the primitive morphisms in an edit script. An
+operational transformation of a primitive morphism $P$ w.r.t a
+morphism $Q$ is another primitive morphism $P'$ such that for any
+object $A:t$, $P'$ has the same \emph{effect} on $Q(A)$ as $P$ has on
+$A$. We will now consider examples of mergeable types that reify the
+notions of edit scripts and operational transformations.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.4]{Figures/string-eg}
+
+\caption{Merging strings with edit scripts and operational
+transformations}
+\label{fig:string-eg}
+\end{figure}
 
@@
 A slight modification of Wagner-Fischer also lets us compute the edit
-script. Fig.~\ref{fig:mergeable-string-eg} shows the edit script ...
+script. Fig.~\ref{fig:string-eg} illustrates.
 

### 2017-02-25 paper/dali.tex
updates to intro
Programming Model.
dali.tex
Canvas concurrent sessions.
edits
Model edits.
opsem.
Sec. 4 beginning.

@@
 %% For double-blind review submission
-\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
+\documentclass[acmlarge,anonymous]{acmart}\settopmatter{printfolios=true}
 %% For single-blind review submission
@@
 %% Some recommended packages.
+\usepackage{wrapfig}
 \usepackage{mathtools}
@@
 \newcommand{\mbleto}{\rightsquigarrow_m}
+\newcommand{\drawsome}{{\sf Canvas}\xspace}
+\newcommand{\lang}{\lambda_b}
+\newcommand{\length}[1]{\left|#1\right|}
 
@@
 \definecolor{Bittersweet}{rgb}{1.0, 0.44, 0.37}
+\definecolor{MidnightBlue}{rgb}{0.0, 0.2, 0.4}
 
@@
 language=ML, % choose the language of the code
-basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
+basicstyle=\footnotesize\ttfamily,       % the size of the fonts that are used for the code
 keywordstyle=\color{Bittersweet},
@@
 %% Title information
-\title[DaLi]{DaLi: Database as a Library}         %% [Short Title] is optional;
+\title[]{Mergeable Types}         %% [Short Title] is optional;
                                         %% when present, will be used in
@@
 \begin{abstract}
-Text of abstract \ldots.
+  Distributed applications often replicate data to improve
+  availability and fault tolerance.  But, the use of replication leads
+  to a significantly more complex programming model.  To ensure a
+  sensible measure of consistency among different replicas of an
+  object, applications are typically structured so that the operations
+  applied to a replicated object on one node are subsequently
+  communicated to other nodes where they can be re-applied locally.
+  As a result, the degree of restructuring and care necessary to
+  transform a sequential or shared-memory concurrent application to a
+  meaningful (replicated) distributed one can be substantial.
+
+  In this paper, we present a new interpretation of replicated types
+  that overcome these concerns.  Our declarative programming model
+  called \name\ treats replication as a form of versioning, in which
+  consistency is enforced through programmer-supplied \emph{merge}
+  functions.  Distributed state is now viewed in terms of a tree of
+  immutable object versions, with merge functions used implicitly by
+  the underlying runtime system to periodically reconcile local
+  versions to yield a consistent global state.  Thus, our model frees
+  the programmer from having to think of low-level operational notions
+  related to distribution and replication, and therefore does not
+  entail any restructuring of application logic.
+
+  We show how any ML datatype can be enriched with mergeable
+  semantics, formalize a rich concurrency model based on such types ,
+  and describe an instantiation of OCaml equipped with these features.
+  Our implementation is integrated with a content-addressable
+  distributed storage layer that allows any OCaml program to be easily
+  transformed into an efficient, highly available, distributed
+  application.
 \end{abstract}
@@
 
-\input{motivation}
+\input{model}
 
 \input{semantics}
+
+\input{mergeable-types}
 
@@
 %% Bibliography
-%\bibliography{bibfile}
+\bibliography{all}
 

### 2017-02-25 paper/intro.tex
Mergeable Type.
edits

@@
 \section{Introduction}
+\label{sec:intro}
 
@@
 restrict, or ignore important albeit unpleasant realities (e.g.,
-network partitions~\cite{Brewer2000,Gilbert2002}), or which may ignore
-important system architectural features (e.g., replicated underlying
-state~\cite{Walter}).  Given these unwarranted assumptions, program
-behavior is often difficult to predict and verify.  On the other hand,
-when these features are explicitly exposed to the programmer, e.g., by
-requiring that applications be written in terms of specialized
-distributed data structures~\cite{Burckhardt2014,,BFL+12,SPB+11} or
-control primitives~\cite{Calm,Armstrong07}, simplicity, composability,
-and ease-of-reasoning can suffer.
+network partitions~\cite{Brewer2000,Gilbert2002}). Given these
+unwarranted assumptions, program behavior is often difficult to
+predict and verify.  On the other hand, when these features are
+explicitly exposed to the programmer, e.g., by requiring that
+applications be written in terms of specialized distributed data
+structures~\cite{Burckhardt2014,BFL+12,SPB+11} or control
+primitives~\cite{Calm}, simplicity, composability, and
+ease-of-reasoning can suffer.
 
@@
 value be added or multiplied to the counter, resp.  Observe that the
-library is written in an idiomatic style in OCaml, with no special
-reasoning principles needed to realize our desired functionality.  As
-long as applications use the library on a single machine, this
+library is written in an idiomatic functional style, with no special
+reasoning principles needed to realize desired functionality.  As long
+as applications use the library on a single machine, this
 implementation behaves as expected.  However, if the library is used
@@
 computation is distributed among a collection of machines, its
-behavior becomes significantly harder to understand.  For example, a
-distributed implementation might wish to \emph{replicate} the counter
-state on each node to improve latency (often referred to as
-availability) or fault tolerance.  Unfortunately, replication doesn't
+behavior can become significantly harder to understand.  In
+particular, a distributed implementation might wish to
+\emph{replicate} the counter state on each node to improve response
+time or fault tolerance.  Unfortunately, adding replication doesn't
 come for free.  Attempting to update every replicated copy atomically
 is problematic in the absence of sophisticated transaction support,
-which impose sigificant performance and availability penalties.  But,
-without such heavyweight mechanisms, applying an \C{Add} operation on
-one node may not be instanteously witnessed on another, which may in
-turn be simultaneously attempting to perform its own \C{Add} or
-\C{Mult} action.  While synchronizing the activities of all nodes to
-ensure at most one such operation is performed at a time is
-impractical, designating a single node to hold the counter state
-eschewing replication altogether (as in a client-server
-configuration), is also an undesirable solution, given its sensitivity
-to network partitions and server failures.  Removig coordination
-altogether by simply replicating counter state without having any
-supporting consistency protocol is an equally infeasible approach.
+which impose significant performance penalties.  But, without such
+heavyweight mechanisms, applying an \C{Add} operation on one node may
+not be instanteously witnessed on another, which may in the process of
+simultaneously attempting to perform its own \C{Add} or \C{Mult}
+action.  While synchronizing the activities of all nodes to ensure at
+most one such operation is performed at a time is impractical,
+designating a single node to hold the counter state eschewing
+replication altogether (as in a client-server
+configuration~\cite{Armstrong}), is also an undesirable solution,
+given its sensitivity to network partitions and server failures, and
+the negative performance impact it incurs in geo-distributed
+environments~\cite{Walter}.  Removig coordination altogether by
+simply replicating counter state without having any supporting
+consistency protocol is an equally infeasible approach.
 
-A method often adopted~\cite{crdts, pldi15, gotsman-popl16} to address
-this tension is to re-define a datatype's operations to return
+A method often adopted~\cite{SPB+11,Burckhardt2014} to address this
+tension is to re-define a datatype's operations to return
 \emph{effects} instead of values.  An \emph{effect} is a tag that
@@
 incorporate the effects of the original
-operaton. Fig.~\ref{fig:counter-rdt} shows the \C{Counter} library
+operation. Fig.~\ref{fig:counter-rdt} shows the \C{Counter} library
 with operations re-defined to returns effects.  Observe that the
@@
 initial counter value of 0, applying the effect of adding 3 and then
-multiplying 5 on one replica yields 15, while applying the effecting
-of first multiplying 5 and then adding 3 yields 3 on the other.  Thus,
+multiplying 5 on one replica yields 15, while applying the effect of
+first multiplying 5 and then adding 3 yields 3 on the other.  Thus,
 implementations must be carefully written to take the lack of
@@
 
-This implementation produces an eventually consistent global state in
-which all replicas will eventually contain the same counter value.
-While effects provide a low-level operational basis for handling
-replicated state, the \emph{ad hoc} nature of the solution confounds
-desirable high-level reasoning principles.  Indeed, the semantic gap
-between the two versions of the counter, one cognizant of replication
-and the other not, breaks backward compatibility.  More significantly,
-its non-trivial construction must be developed in different guises for
-every distributed data structure used by the application.
+In this implementation, all replicas will eventually contain the same
+counter value, assuming computation on the counter quiesces.  While
+transmitting effects provide a low-level operational basis for
+handling replicated state, the \emph{ad hoc} nature of the solution
+confounds desirable high-level reasoning principles.  Indeed, the
+semantic gap between the two versions of the counter, one cognizant of
+replication and the other not, breaks backward compatibility.  Just as
+significantly, its non-trivial construction must be developed in
+different guises for every distributed data structure used by the
+application.
 \begin{figure}
@@
 \end{figure}
-The lack of composability is another important downside of this
+The lack of composability is yet another important downside of this
 approach.  Consider an application that uses two replicated counters,
@@
 counters, a \C{read} operation performed at another replica may
+\begin{wrapfigure}{l}{.5\textwidth}
+  \begin{center}
+    \includegraphics[scale=0.4]{Figures/dali-counter}
+  \end{center}
+  \caption{\small A replicated counter can be expressed as a sequence
+    of versions managed by logically-distributed concurrently
+    executing threads.  Local versions produced by these threads
+    periodically merge their state with one another.  In the figure,
+    when the local version of Node 3, labeled 20, merges with the
+    local version of Node 1, whose counter state is 15 at the time, a
+    new counter value is produced.  This value takes into account the
+    previous merged state (10) from which the two versions were
+    derived to yield a new merged state of 25.  Observe that there is
+    no explicit synchronization or coordination among versions.  In
+    the figure, red circles represent counter state produced through
+    merging, and blue circles represent state produced by applying
+    counter operations on a node-local version.  }
+\vspace*{-.85in}  
+\end{wrapfigure}
 witness the increment to $c_1$, but not $c_2$, thus violating the
@@
 captured in the absence of external mechanisms to compose effects
-(such as effect-based transactions~\cite{pldi15}).
+(such as effect-based transactions~\cite{pldi15}).  Rather than
+viewing a distributed computation in terms of \C{Add} effects produced
+by operations, can we formulate a more declarative interpretation,
+directly in terms of the counter value maintained by each node.  In
+doing so, we observe that each replica essentially operates over its
+own version of a counter.  Local operations on a replicated object can
+be thought of as yielding new versions, thus producing a version tree,
+with one branch for each replica.  Thus, each branch represents
+different (immutable) versions maintained by different nodes, with the
+state produced by a counter's operations on a node recorded along the
+node's local branch.  Now, to generate a globally consistent view of a
+counter, we only need to define a merge operation that explains how to
+combine two local versions to produce a new version that reflects both
+their states.
 
-On the other hand, a functional programmer is likely to approach this
-problem very differently.  Rather than thinking about a low-level
-operational variant of distributed state given in terms of explicit
-effects as described above, a declarative interpretation of a
-decentralized computation could be naturally described in terms of a
-version tree whose root contains the initial counter value, and whose
-branches represent different (immutable) versions maintained by
-different nodes, with the state produced by a counter's operations on
-a node recorded along the node's local branch.  Now, a globally
-consistent view of a counter can be explained simply in terms of
-\emph{merge} operations of different branch tips.  Treating
-replication as merging yields a counter implementation that bears
-strong similarity to the original sequential one:
+Framing replication as merging leads to a counter implementation that
+bears strong similarity to the original sequential one:
   \begin{ocaml}
-    module Counter = struct
+    module Replicated_Counter = struct
       include Counter
-      type mergeable_t = t [@@deriving persistence]
-      let merge lca v1 v2 = lca + (v1 - lca) + (v2 - lca)
+      type mergeable_t = t 
+      let merge lca v1 v2 =
+         lca + (v1 - lca) + (v2 - lca)
     end
   \end{ocaml}
+
 The role of \C{lca} (least common ancestor) here captures salient
@@
 and reconcile different local states.
-\begin{figure}
-  \begin{center}
-    \includegraphics[scale=0.35]{Figures/dali-counter}
-  \end{center}
-  \caption{\small A multi-version counter can be expressed as a
-    sequence of versions managed by logically-distributed concurrently
-    executing threads.  Local versions produced by these threads
-    periodically merge their state with one another.  In the figure,
-    when the local version of Node 3, labeled 20, merges with the
-    local version of Node 1, whose counter state is 15 at the time, a
-    new counter value is produced.  This value takes into account the
-    previous merged state (10) from which the two versions were
-    derived to yield a new merged state of 25.  Observe that there is
-    no explicit synchronization or coordination among versions.  In
-    the figure, red circles represent state produced through merging,
-    and blue circles represent local counter state produced by
-    applying counter operations on a node-local version.  }
-\end{figure}
-The \name programming model realizes a monadic version control system
-in OCaml centered around data, rather than file, coherence.  The \name
-monad lets programmers write and compose concurrent computations
-around multiple (implicit) versions of a mergeable datatype, an
-ordinary datatype additionally equipped with a \emph{merge} function
-responsible for deriving a consistent global state from a collection
-of local versions of that state.  A computation makes progress by
-forking concurrent versions of existing versions, creating later
-versions along a \emph{branch}, or by merging branch tips.
 
-While any OCaml datatype can be equipped with a merge operator, and
-thus seamlessly participate in a distributed computation, instances of
-datatypes whose merge operations are commutative with respect to their
-local version state arguments (i.e., \C{v1} and \C{v2} in the counter
-example above) are guaranteed to eventually converge to the same state
-on all replicas.  For such types, OCaml programmers get the benefits
-of realizing highly available (low-latency) distributed computation,
-while continuing to enjoy the comforts of high-level reasoning and
-the familiarity of using standard libraries.  Notably, while \name's
+In this paper, we develop a programming model that realizes a monadic
+version control system centered around data, rather than file,
+coherence.  The \name monad lets programmers write and compose
+concurrent computations around multiple (implicit) versions of a
+mergeable datatype, an ordinary ML datatype additionally equipped with
+a \emph{merge} function responsible for deriving a consistent global
+state from a collection of local versions of that state.  A
+computation progresses by forking replicating new versions of existing
+versions on geo-distributed nodes, creating new versions along a an
+existing \emph{branch}, or by merging branch tips.
+
+Our model allows ML programmers to get the benefits of realizing
+highly available (low-latency) distributed computation, while
+continuing to enjoy the comforts of high-level reasoning and the
+familiarity of using standard libraries.  Notably, while \name's
 programming model makes no explicit reference to any specific
 operational manifestation of a distributed system (e.g., programmers
-do not need to worry about replicas and their management), we
-demonstrate that it can be nonetheless efficiently realized on
-existing geo-replicated distributed systems.
+do not need to explicitly manage replicas), we demonstrate that it can
+be nonetheless efficiently realized on existing geo-replicated
+distributed systems.
 
@@
       behind a monad, and exposes its functionality via a simple API
-      that lets programmers define and compose concurrent ML
+      that lets programmers define and compose distributed ML
       computations around such data.  Issues related to replication

### 2017-02-25 paper/mergeable-types.tex
edits

--- 2017-02-25 paper/mergeable-types.tex
@@
 
-\begin{figure}
-\centering
-\includegraphics[scale=0.4]{Figures/pushouts}
-
-\caption{Commutative diagram representing the merge operation.}
-\label{fig:pushouts}
-\end{figure}
-
 We now formally define what we mean by a mergeable type. The
 definition serves as a guide to write merge functions for functional
-data types, and elevate them to the level of replicated data types.
+data types, elevating them to the level of replicated data types.
 
-A functional data type library is as a tuple consisting of a type
-($t$), and a (possibly infinite) collection of functions ($f$, $g$,
-etc.) of type $t \rightarrow t$. We call the functions \emph{primitive
-morphisms}\footnote{We borrow terminology from category theory for
-convenience}. A morphism ($P$, $Q$, etc.) is either a primitive
-morphism, or an associative composition of morphisms ($P \circ Q \circ
-R$). An object $B:t$ is called a \emph{semantic successor} of $A:t$
-(conversely, $A$ is called a \emph{semantic ancestor} of $B$) if and
-only if there exists a morphism $P$ such that $P(A) = B$.
+A functional data type library is a tuple consisting of a type ($t$),
+and a collection of functions ($f$, $g$, etc.) of type $t \rightarrow
+t$ that are \emph{primitive morphisms}. A morphism ($P$, $Q$, etc.) is
+either a primitive morphism, or an associative composition of
+morphisms ($P \circ Q \circ R$). An object $B:t$ is called a
+\emph{semantic successor} of $A:t$ (conversely, $A$ is called a
+\emph{semantic ancestor} of $B$) if and only if there exists a
+morphism $P$ such that $P(A) = B$.
 
@@
 successors and $D$ satisfies certain conditions. These conditions can
-be understood by observing the \rulelabel{E-Pull-Wait} rule of of
+be understood by observing the \rulelabel{E-Pull-Wait} rule of
 Fig.~\ref{fig:opsem}, which applies the \C{merge} function to a pair
@@
 and $v_1$, and $v$ and $v_2$ is the syntactic ancestor relationship
-that follows from the branching structure. The merge function, as
-described above, however assumes that the concurrent versions are
-semantic successors of their LCA. It is therefore essential to
+that follows from the computation's branching structure. The merge
+function, as described above, however assumes that concurrent versions
+are semantic successors of their LCA.  It is therefore essential to
 maintain coherence between the syntactic and semantic ancestor
@@
 semantic ancestor of $v_2$.  Fig.~\ref{fig:syntactic-ancestor-2}
-captures another way of establishing ancestor relationship, namely by
+captures another way of establishing an ancestor relationship, namely by
 merging branches. Version $v_{21}$ on branch $b_2$ is merged into
 $v_{11}$ on $b_1$ to create $v_{12}$ on $b_1$. Versions $v_{11}$ and
+\begin{wrapfigure}{l}{.4\textwidth}
+\centering
+\includegraphics[scale=0.35]{Figures/pushouts}
+\caption{Commutative diagram representing the merge operation.}
+\label{fig:pushouts}
+\end{wrapfigure}
 $v_{21}$ are now syntactic ancestors of $v_{12}$, but for them to be
-semantic ancestors, merge function has to enforce the relationship
-explicitly. In other words, the result of merging of a pair of
+semantic ancestors, the merge function has to enforce the relationship
+explicitly. In other words, the result of merging  a pair of
 semantic successors, $B$ and $C$, of an $A$, has to be an object $D$
@@
 two morphisms $P$ and $Q$ that map $A:t$ to $B:t$ and $C:t$, there
-must also exist morphisms $P'$ and $Q'$, the map both $B$ and $C$ to
-$D:t$. This intuition is expressed visusally in
-Fig.~\ref{fig:pushouts}.
-
-For a mergeable type $t$, we have the guarantee that a sequence of
-values read by a thread from its branch ``makes sense'' as per the
-data type semantics:
+also exists morphisms $P'$ and $Q'$, that map both $B$ and $C$ to
+$D:t$. This intuition is expressed visually in
+Fig.~\ref{fig:pushouts}.  For a mergeable type $t$, we have the
+guarantee that a sequence of values read by a thread from its branch
+is sensible, as per the data type semantics:
 
@@
 
-Def.~\ref{def:mergeable-type} specifies the merge function ($M$)
-slightly differently; instead of returning a merged value like the
-three-way merge function (\C{merge}), $M$ returns a pair of morphisms.
-This specification is useful to preempt certain non-sensical merge
-functions, and thus enforce the branch-local consistency property
-(Theorem~\ref{thm:branch-consistency}). For instance, consider the
-counter data type from Sec.~\ref{sec:intro}. A three-way merge
-function can be defined to return a constant regardless of its
-arguments:
-\begin{ocaml}
-  let merge lca a b = 0
-\end{ocaml} Such a merge function allows the client (a thread) to
+Rather than specifying merge ($M$) in terms of an LCA,
+Definition~\ref{def:mergeable-type} specifies $M$ in terms of a pair
+of morphisms.  This specification is useful to avoid the expression of certain
+nonsensical merge functions, and thus enforce a useful branch-local
+consistency property (Theorem~\ref{thm:branch-consistency}). For
+instance, consider the counter data type from Sec.~\ref{sec:intro}. A
+three-way merge function can be defined to return a constant
+regardless of its arguments:
+\begin{ocaml} let merge lca a b = 0 \end{ocaml}
+Such a merge function allows the client to
 witness the violation of the branch-local consistency property. For
@@
 non-negativeness of an integer etc) that are preserved by all of $t$'s
-morphisms. On the hand, the commutativity of the diagram in
-Fig.~\ref{fig:pushouts} may not be enforced statically. It can however
-be verified at run-time by checking that the morphisms returned by $M$
-map the respective concurrent versions to the same value. The
+morphisms.  The commutativity of the diagram in
+Fig.~\ref{fig:pushouts} need not be enforced statically. It can
+equally be verified at run-time by checking that the morphisms returned by $M$
+map respective concurrent versions to the same value. The
 operational semantics (Fig.~\ref{fig:opsem}) can be extended to this
@@
 Merge functions for simple data types are easy to write. For instance,
-the merge logic for counter is simple enough that it doesn't require
-explicit efforts to identify $P$ and $Q$ that led to the
+the merge logic for counter is simple enough that it does not require
+great effort to identify the $P$ and $Q$ that leda to the
 concurrent counter versions, and transform them into $P'$ and $Q'$
 such that the diagram in Fig.~\ref{fig:pushouts} commutes. For more
-sophisticated data types, this process is more involved, and requires
+sophisticated data types, this process is more involved, and may require
 the programmer to perform non-trivial reasoning. We now present an
@@
 shortest sequence of \C{insert}, \C{delete}, and \C{substitute}
-operations (string's primitive morphisms), whose composition takes $s$
+operations (a string's primitive morphisms), whose composition takes $s$
 to $s'$. The length of the sequence is in fact the Levenshtein
 distance~\cite{levenstein} between $s$ and $s'$, which can be computed
-in polynomial time via Wagner-Fischer algorithm~\cite{wagner-fischer}.
+in polynomial time~\cite{wagner-fischer}.
 A slight modification of Wagner-Fischer also lets us compute the edit

### 2017-02-25 paper/meta.tex
edits

@@
 
-Ancestor relation is therefore a partial order (reflexive, transitive,
-anti-symmetric) with a greatest lower bound (the initial version).
-Thus, for any two versions in a legal history, there exist at least
-one common ancestor. Ancestor relationships among the common ancestors
-let us define the notion of a least common ancestor (LCA):
+The ancestor relation is therefore a partial order with a greatest
+lower bound (the initial version).  Thus, for any two versions in a
+legal history, there exists at least one common ancestor.  The ancestor
+relationship among all common ancestors lets us define the notion of a
+least common ancestor (LCA):
 
@@
 
-\begin{figure}
+When merging two concurrent versions $v_1$ and $v_2$, the common
+ancestor argument for \C{merge} must be the LCA of $v_1$ and $v_2$,
+without which \C{merge} may yield unexpected results. This is
+demonstrated for the grow-only counter in
+Fig.~\ref{fig:merge-needs-lca}, where an incorrect count is obtained if a
+common ancestor that is not an LCA is used to merge 4 and 7. While in
+this example there is a unique LCA for 4 and 7, in general this may
+\begin{wrapfigure}{l}{.4\textwidth}
 \centering
 \includegraphics[scale=0.6]{Figures/merge-needs-lca}
-
 \caption{This example of a grow-only counter illustrate why \C{merge}
@@
 \label{fig:merge-needs-lca}
-\end{figure}
+\end{wrapfigure}
+not be the case. With unrestrained branching and merging, there is no
+bound on the number of LCAs a pair of versions can have.  For example,
+in Fig.~\ref{fig:criss-cross-lcas}, the merge of 0 with 3 is preceded
+by two ``criss-cross'' merges between their respective branches
+resulting in there being two LCAs (5 and 4) for 0 and 3. Multiple LCAs
+can occur even without criss-cross merges, as demonstrated by
+Fig.~\ref{fig:external-lcas}. Concurrent versions with multiple LCAs
+do not lend themselves to three-way merging. This problem also arises
+in the context of source control systems, which employ \emph{ad hoc}
+mechanisms to pave the way for three-way merging.
+GitHub~\cite{github}, for instance, recursively merges LCAs to compute
+a virtual ancestor, which then serves as the LCA for merging
+concurrent versions. This method is demonstrated in the branching
+structure for a mergeable, replicated counter as shown in
+Fig.~\ref{fig:criss-cross-lcas}, where LCAs 5 and 4 of 0 and 3 are
+merged (with their LCA being 10) to generate -1 as the virtual LCA to
+merge 5 and 4. A major downside with this approach is that it makes no
+guarantees on the relationship between the virtual ancestor and its
+concurrent versions; the former may not even be a legal ancestor of
+the latter as per the semantics of the data type. For instance,
+suppose the integer type in Fig.~\ref{fig:criss-cross-lcas} represents
+a bank account balance, which is expected to disallows any activity on
+the account if the balance is ever known to be less than zero.  From
+the perspective of the library designer and its clients, there is no
+meaningful scenario in which versions 3 and 0 can emerge from -1,
+since the only transition allowed by the semantics from -1 is to
+itself.  Clearly, \emph{ad hoc} mechanisms like this are error-prone
+and difficult to apply in general.
 
@@
 
-When merging two concurrent versions $v_1$ and $v_2$, the common
-ancestor argument for \C{merge} must the LCA of $v_1$ and $v_2$,
-without which \C{merge} may yeild unexpected results. This is
-demonstrated for the grow-only counter in
-Fig.~\ref{fig:merge-needs-lca}, where incorrect count is obtained if a
-common ancestor that is not an LCA is used to merge 4 and 7. While in
-this example there is a unique LCA for 4 and 7, in general this may
-not be the case. With unrestrained branching and merging, there is no
-bound on the number of LCAs a pair of versions can have.
-For example, in
-Fig.~\ref{fig:criss-cross-lcas}, the merge of 0 with 3 is preceded by
-two ``criss-cross'' merges between their respective branches resulting
-in there being two LCAs (5 and 4) for 0 and 3. Multiple LCAs can occur
-even without criss-cross merges, as demonstrated by
-Fig.~\ref{fig:external-lcas}. Concurrent versions with multiple LCAs
-do not lend themselves to three-way merging. This problem also arises
-in the context of source control systems, which employ ad hoc
-mechanisms to pave way for three-way merging.  GitHub~\cite{github},
-for instance, recursively merges LCAs to compute a virtual ancestor,
-which then serves as the LCA for merging the concurrent versions. This
-method is demonstrated for the example in
-Fig.~\ref{fig:criss-cross-lcas}, where LCAs 5 and 4 of 0 and 3 are
-merged (with LCA of 10) to generate -1 as the virtual LCA to merge 5
-and 4. A major downside with this method is that it makes no
-guarantees of the relationship between the virtual ancestor and the
-concurrent versions; the former may not even be a legal ancestor of
-the latter as per the semantics of the data type. For instance, the
-integer type in Fig.~\ref{fig:criss-cross-lcas} may in fact represent
-a bank account, which disallows any activity on the account if balance
-is ever known to be less than zero. Thus, from the perspective of a
-bank account, it doesn't make sense how concurrent versions 3 and 0
-emerged from -1, since the only transition allowed by the semantics
-from -1 is to itself. Clearly, ad hoc mechanisms, which work for the
-text, may not work for more sophisticated data types.
-
-Fortunately, unlike the source control systems where branching
-structure is entirely dictated by the user, \name abstracts away the
-branching structure from the programmer, hence retains the ability to
+Fortunately, unlike source control systems where branching structure
+is entirely dictated by the user, \name abstracts away branching
+structure from the programmer, and hence retains the ability to
 manifest it in a way that it deems fit. In particular, \name solves
 the problem of multiple LCAs either by constraining the branching
-stucture to preempt multiple LCAs, or by extending it such that
-merging branches have a new unique LCA. The preemption is done for
+stucture to prevent their existence, or by merging branches to ensure
+that they have a new unique LCA.  Preention is done for
 criss-cross merges by the operational semantics, which only allows
-merging versions that are latest on their respective branches. For
-instance let $v_{11}$ and $v_{21}$ be the latest versions on branches
+merging versions that are the latest on their respective branches. For
+instance, suppose $v_{11}$ and $v_{21}$ are the latest versions on branches
 $b_1$ and $b_2$, respectively. Merging $b_2$ into $b_1$ entails
@@
 
-Next, for the branching structures where a pair of branches (i.e.,
-their respective latest versions) have multiple externally-located
-LCAs, the problem can be fixed by suitably extending the branching
-structure such that the branches have a new single LCA. For instance,
-consider the example in Fig.~\ref{fig:external-lcas}, where (latest
-versions on) branches $b_1$ and $b_4$ have two externally-located LCAs
+Next, for branching structures where a pair of branches (i.e., their
+respective latest versions) have multiple externally-located LCAs, we
+suitably extend the branching structure to generate a new single LCA.
+ For instance, consider
+the example in Fig.~\ref{fig:external-lcas}, where (latest versions
+on) branches $b_1$ and $b_4$ have two externally-located LCAs
 ($v_{22}$ and $v_{32}$), hence cannot be merged. Mergeability can
 however be reclaimed by extending the branching structure as shown in
-Fig.~\ref{fig:legal-extension-1}. The extension is done by first merging
-$b_3$ into $b_2$ to create version $v_{23}$ on $b_2$, then merging
-$b_2$ into $b_1$ to create version $v_{14}$ on $b_1$, and finally
-merging $b_2$ into $b_4$ to create version $v_{45}$ on $b_4$. Observe
-that $b_1$ and $b_4$ now have a single LCA ($v_{23}$) hence are now
-mergeable. Mergeability could also have been reclaimed by adopting an
-alternative extension to the branching structure as shown in
-Fig.~\ref{fig:legal-extension-2}.
+Fig.~\ref{fig:legal-extension-1}. The extension is done by first
+merging $b_3$ into $b_2$ to create version $v_{23}$ on $b_2$, then
+merging $b_2$ into $b_1$ to create version $v_{14}$ on $b_1$, and
+finally merging $b_2$ into $b_4$ to create version $v_{45}$ on
+$b_4$. Observe that $b_1$ and $b_4$ now have a single LCA ($v_{23}$)
+hence are now mergeable. Mergeability could also have been reclaimed
+by adopting an alternative extension to the branching structure as
+shown in Fig.~\ref{fig:legal-extension-2}.
 

### 2017-02-25 paper/model.tex
edits

@@
 
-In this section, we describe the \name programming model through the
-example of a collaborative drawing application we call \drawsome.
+In this section, we describe the \name programming model using a
+collaborative drawing application we call \drawsome as a motivating
+example.
 
@@
 containing a single pixel (\C{r},\C{g}, or \C{b}) or a tree, if the
+\begin{wrapfigure}{l}{.4\textwidth}
+\centering
+\includegraphics[scale=0.4]{Figures/canvas-merging}
+\caption{Merging concurrent versions (\C{v1} and \C{v2}) of a drawing
+canvas. \C{lca} is their common ancestor.}
+\label{fig:canvas-merging}
+\end{wrapfigure}
 quadrant contains multiple pixels of different colors. Quadrants are
-expanded into a tree structures as and when pixels are colored.  The
-representation is thus optimized for sparse canvases, such as
+expanded into tree structures lazily as and when pixels are colored.
+The representation is thus optimized for sparse canvases, such as
 whiteboards. The application supports three simple operations:
 creating a new canvas, setting the pixel at a specified coordinate,
-and returning the pixel at a given coordinate. The \C{merge} function
-and \C{deriving versioned} directive are explained below.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.5]{Figures/canvas-merging}
-
-\caption{Merging concurrent versions (\C{v1} and \C{v2} ) of a drawing
-canvas. \C{lca} is their common ancestor.}
-\label{fig:canvas-merging}
-\end{figure}
-
+and returning the pixel at a given coordinate. 
 \drawsome lets multiple users collaborate on a canvas that is
@@
 values on both the canvases. For instance, if the color of a pixel in
-\C{v1} is white, and in \C{v2} it is green, and its color in \C{lca}
-is white, then it means that only \C{v2} modified the color. Hence the
-pixel is colored green in the merged canvas. On the other hand, if the
-pixel is red in \C{v1}, then it means that both \C{v1} and \C{v2} have
-modified the color. In such case, an appopriate color-mixing algorithm
-can be used to determine the color of pixel.  For instance, the pixel
-can be colored yellow - an additive combination of red and green. The
-logic is illustrated in Fig.~\ref{fig:canvas-merging}.
+\C{v1} is white, and the same pixel in \C{v2} is green, and its color
+in \C{lca} is white, then it means that only \C{v2} modified the
+color. Hence the pixel is colored green in the merged canvas. On the
+other hand, if the pixel is red in \C{v1}, then it means that both
+\C{v1} and \C{v2} have modified the color. In such case, an
+appropriate color-mixing algorithm can be used to determine the color
+of pixel.  For instance, the pixel can be colored yellow - an additive
+combination of red and green. The logic is illustrated in
+Fig.~\ref{fig:canvas-merging}.  While the \drawsome\ merge operation
+necessarily reflects domain-specific knowledge about the application,
+it makes no reference to distribution or replication, and does not
+impact the definition of any other operation, whose algorithms can be
+developed for a sequential execution environment and seamlessly
+transplanted to a distributed one.
 \begin{figure}
@@
   else match (lca,v1,v2) with
-    (*
-     * The first three rules isomorphize lca, v1 and v2.
-     *)
     | (_, B _, N px2) -> merge lca v1 @@ b_of_n px2
@@
 \end{figure}
-
-The \name programming model lets programmers define and compose
-concurrent computations around versioned data structures.
 Fig.~\ref{fig:dali-monad} shows the signature of the \name module that
@@
   \end{ocaml}
+  \caption{Signature of the \name monad}
   \label{fig:dali-monad}
-\caption{Signature of the \name monad}
 \end{center}
@@
 next version, which becomes the current version for the rest of the
-computation.  This version is created by merging the proposal with a
-subset of concurrent versions that have become available since the
-last merge or fork. Thus, \C{sync} effectively lets a computation sync
-with a subset of concurrent computations to obtain their latest
-updates.
-
+computation.  This version is created by merging the proposal (the
+local state) with a subset of concurrent versions that have become
+available since the last merge or fork.  Thus, \C{sync} effectively
+lets a computation sync with all computations operating over replicas
+of the same state that are available (i.e., reachable in the
+network)\footnote{The implementation implicitly triggers a sync
+  whenever a detected network partition is repaired.}  to obtain their
+latest updates.
 \begin{figure}
@@
 blank. She then invites Bob for collaboration by forking a new
-concurrent version for Bob. Bob, in turn, invites Cheryl for
+concurrent version for Bob. Bob, in turn, invites Cheryl.
 \begin{figure}
@@
 \end{figure}
-collaboration. All three of them start working on the blank canvas.
+All three start working on the blank canvas on different replicas.
 Alice draws a red horizontal line from $(0,0)$ (top-left) to $(4,0)$
-(top-right) using \C{C.draw\_line}.\footnote{Its definitionis not
+(top-right) using \C{C.draw\_line}.\footnote{Its definition is not
   shown, but can be constructed using \C{set\_px}} Meanwhile, Bob
@@
 decentralized distributed setting requires addressing non-trivial
-theoretical challenges described in the next section. Unlike the
-canvas, the process of writing merge functions for data types with
-non-trivial invariants could be quite involved. The following sections
-describe a principled approach to deriving merge functions for any
-data type given its invariants. Like \C{Canvas.merge}, merge functions
-often need to compare multiple versions of a data type for structural
-equality, which, when performed naively, incurs cost linear in the
-size of versions. The representation of a versioned data type in \name
-is optimized for performing comparisons that allows structural
-equality to be determined in constant time. Such engineering details,
-which contribute to the practicality of our programming model are also
-discussed in subsequent sections.
+theoretical challenges described in
+Section~\ref{sec:semantics}. Unlike the canvas, the process of writing
+merge functions for data types with non-trivial invariants could be
+quite involved. The following section describes a principled approach
+to deriving merge functions for any data type given its
+invariants. Like \C{Canvas.merge}, merge functions often need to
+compare multiple versions of a data type for structural equality,
+which, when performed naively, incurs cost linear in the size of
+versions. The representation of a versioned data type in \name is
+optimized for performing comparisons that allows structural equality
+to be determined in constant time. Such engineering details, which
+contribute to the practicality of our programming model, are discussed
+in Sec.~\ref{sec:impl}.

### 2017-02-25 paper/semantics.tex
System Model
edits

@@
 operation is broken down into two primitives - \C{push} and \C{pull},
-which can be lambda-composed get the desired effect:
+which can be composed to get the desired effect of the original:
 \begin{smathpar}
@@
 Fig.~\ref{fig:opsem} also shows the syntax of \emph{branches}, which
-are the artifacts of evaluation and only appear during the run-time. A
-branch is a non-empty sequence of tagged values, where the tag
-captures the abstract run-time operation that led to the creation of
-the value. It is implicitly assumed that each value added to a branch
-is uniquely identifiably, hence no two values on a branch are equal.
-The uniqueness assumption is later extended to a collection of
-branches that constitute a branching structure. A real implementation
-meets this assumption by versioning values across the branches. Thus,
-in reality, branches contain \emph{versions} which denote values. The
-semantics, however, doesn't make this distinction, and uses values and
-versions interchangeably.
+are artifacts of evaluation and only appear during run-time. A branch
+is a non-empty sequence of tagged values, where the tag captures the
+abstract run-time operation that led to the creation of the value. It
+is implicitly assumed that each value added to a branch is uniquely
+identifiable, and hence no two values on a branch are equal.  The
+uniqueness assumption is later extended to a collection of branches
+that constitute a branching structure. A real implementation meets
+this assumption by versioning values across the branches. Thus, in
+reality, branches contain \emph{versions} which denote values.  To
+simplify the presentation, the semantics, does not make this
+distinction, and uses values and versions interchangeably.
 
-Small-step operational semantics of $\lang$ is defined via reduction
-relation ($\stepsto$) that relates \emph{program states}. A program
-state ($p;\,H$) consists of a program $p$ and a \emph{branch history}
-$H$ that maps thread Ids to corresponding branches; each thread is
-associated with a branch during the evaluation. Evaluation contexts
-have been defined separately for expressions ($E$) and programs ($P$),
-with the latter subsuming the former. $E$ is defined to evaluate the
-first argument of a \C{run} expression to a value that constitutes the
-initial version (recall that \C{run} models \name's
-\C{with\_init\_version\_do}). Program evaluation context
-non-deterministically picks one of the threads to evaluate. The admin
-rule that relates transitions of holes to transitions of expressions
-and programs is straightforward, hence elided. Rest of the reduction
-rules are presented in Fig.~\ref{fig:opsem}. For brevity, we write
-$H(t\mapsto (v,f))$ to denote the proposition that $H$ maps $t$ to
-$(v,f)$. The notation $H[t \mapsto (v,f)]$ as usual denotes
-the extension of $H$ with the binding $t \mapsto (v,f)$.
+The small-step operational semantics of $\lang$ is defined via a
+reduction relation ($\stepsto$) that relates \emph{program states}. A
+program state ($p;\,H$) consists of a program $p$ and a \emph{branch
+  history} $H$ that maps thread identifiers to corresponding branches;
+each thread is associated with a branch during evaluation. Evaluation
+contexts have been defined separately for expressions ($E$) and
+programs ($P$), with the latter subsuming the former.
+
+$E$ is defined to evaluate the first argument of a \C{run} expression
+to a value that constitutes an initial version (recall that \C{run}
+models \name's \C{with\_init\_version\_do}). A program evaluation
+context non-deterministically picks an extant thread to evaluate. The
+administrative rule that relates transitions of holes to transitions
+of expressions and programs is straightforward, and hence elided. The
+remaining core reduction rules are presented in
+Fig.~\ref{fig:opsem}. For brevity, we write $H(t\mapsto (v,f))$ to
+denote the proposition that $H$ maps $t$ to $(v,f)$. The notation $H[t
+  \mapsto (v,f)]$ denotes the extension of $H$ with the binding $t
+\mapsto (v,f)$, as usual.
 
 Reduction rules let expression evaluation take a step by rewriting the
-expression and suitably updating the branch history ($H$). The
+expression and suitably updating the branch history ($H$).
+\rulelabel{E-App} is the standard beta reduction rule.  The
 \rulelabel{E-Run} rule is applicable only when $H$ is empty, i.e.,
-when no prior branching structure exists. The rule rewrites the
+when no prior branching structure exists, and commences a distributed
+computation with a corresponding version tree. The rule rewrites the
 $\C{run}\;v\;s$ expression to $s$, while creating a new branching
@@
 The corresponding branch ($H(t')$) is forked from the parent thread's
-branch ($H(t)$). The semantics of branch forking is same as described
-above. The \C{fork} expression in the parent thread evaluates to
-\C{()}. The \rulelabel{E-Push} rule creates a new version on the
-current branch ($H(t)$) using the pushed value ($v$).
-\rulelabel{E-App} is the standard beta reduction rule.
+branch ($H(t)$). The semantics of branch forking is the same as
+described above. The \C{fork} expression in the parent thread
+evaluates to \C{()}. The \rulelabel{E-Push} rule creates a new version
+on the current branch ($H(t)$) using the pushed value ($v$).  Although
+our semantics does not directly expose heap-allocated values, the
+intention is that $v$ is a replicated object, manipulated on the local
+heap that, after push, now becomes subject to merging and coordination
+with other replicas.
 
@@
 discussed below. The \rulelabel{E-Pull-Wait} and \rulelabel{E-Pull}
-rules thus let a thread sync with a subset of concurrent threads in
-multiple steps before returning the result of the \C{pull}. Since
-\C{sync} is a composition of \C{push} and \C{pull}, its behavior can
-be explained thus: \C{sync} pushes the given value onto the current
-(local) branch, merges a (possibly empty) subset of concurrent
-branches into the local branch, and returns the result.
+rules thus let a thread synchronize with other distributed threads
+manipulating versions of replicas in multiple steps before returning
+the result of the \C{pull}. Since \C{sync} is a composition of
+\C{push} and \C{pull}, its behavior can be explained thus: \C{sync}
+pushes the given value onto the current (local) branch, merges a
+(possibly empty) subset of concurrent branches into the local branch,
+and returns the result.  (The operation is not guaranteed to
+synchronize with all concurrent branches because not all such branches
+may be available.)
 
 \input{meta.tex}
+
+\input{system-model}

### 2017-02-25 paper/mergeable-types.tex
edits

@@
 regardless of its arguments:
-\begin{ocaml} let merge lca a b = 0 \end{ocaml}
+\begin{ocaml}
+  let merge lca a b = 0
+\end{ocaml}
 Such a merge function allows the client to
@@
 such that the diagram in Fig.~\ref{fig:pushouts} commutes. For more
+\begin{wrapfigure}{l}{.4\textwidth}
+\centering
+\includegraphics[scale=0.35]{Figures/pushouts-2}
+\caption{Commutative diagram for Def.~\ref{def:mergeable-type-2}.}
+\label{fig:pushouts-2}
+\end{wrapfigure}
 sophisticated data types, this process is more involved, and may require
@@
 non-trivial merge functions.
+
+\paragraph{Notation} As mentioned earlier, we distinguish between the
+named functions defined by a data type library (primitive morphisms),
+and their compositions (morphisms). A sequence (list) of primitive
+morphisms is written $P^*$ or $Q^*$. The composition of a sequence of
+primitive morphisms (written $\denot{P^*}$) is their right-to-left
+composition, i.e., $\C{fold\_left}\;(\lambda f.\lambda g. f \circ
+g)\; id\; P^*$, where $\C{fold\_left}$ has its usual OCaml type. The
+length of a sequence $S$ is written $\length{S}$.
 
@@
   primitive morphisms, $P^*$ and $Q^*$, and returns a pair of morphisms,
-  $Q'$ and $P'$, such that for any object $A:t$, $(Q' \circ
+  $P'$ and $Q'$, such that for any object $A:t$, $(Q' \circ
   \denot{P^*})(A) = (P' \circ \denot{Q^*})(A)$.  
 \end{itemize}
+Commutativity diagram of Fig.~\ref{fig:pushouts-2} visualizes this
+definition. 
 \end{definition}
 
+The definition asserts the existence of two functions for a mergeable
+type $t$. First, the function $W$, which computes the \emph{edit
+distance} between an object $A:t$ and its semantic successor $B:t$.
+The sequence of primitive morphisms it returns ($P^*$) is called an
+\emph{edit script}. Replaying the edit script on $A$ produces $B$
+(i.e., $\denot{P^*}(A) = B$). Second, the definition asserts the
+existence of a function $M$, which performs an \emph{operational
+transformation} of the primitive morphisms in an edit script. An
+\begin{wrapfigure}{l}{.4\textwidth}
+\centering
+\includegraphics[scale=0.4]{Figures/string-eg}
+\caption{Merging strings with edit scripts and operational
+transformations}
+\label{fig:string-eg}
+\end{wrapfigure}
+operational transformation of a primitive morphism $P$ w.r.t a
+morphism $Q$ is another primitive morphism $P'$ such that for any
+object $A:t$, $P'$ has the same \emph{effect} on $Q(A)$ as $P$ has on
+$A$. We will now consider examples of mergeable types that reify the
+notions of edit scripts and operational transformations.
 \paragraph{Strings} Strings are mergeable. Given a string ($s$) and

### 2017-02-25 paper/mergeable-types.tex
edits

@@
 \end{itemize}
-Commutativity diagram of Fig.~\ref{fig:pushouts-2} visualizes this
+The Commutativity diagram of Fig.~\ref{fig:pushouts-2} visualizes this
 definition. 
@@
 transformation} of the primitive morphisms in an edit script. An
+operational transformation of a primitive morphism $P$ w.r.t a
+morphism $Q$ is another primitive morphism $P'$ such that for any
+object $A:t$, $P'$ has the same \emph{effect} on $Q(A)$ as $P$ has on
+$A$. We will now consider examples of mergeable types that reify the
+notions of edit scripts and operational transformations.
 \begin{wrapfigure}{l}{.4\textwidth}
@@
 \end{wrapfigure}
-operational transformation of a primitive morphism $P$ w.r.t a
-morphism $Q$ is another primitive morphism $P'$ such that for any
-object $A:t$, $P'$ has the same \emph{effect} on $Q(A)$ as $P$ has on
-$A$. We will now consider examples of mergeable types that reify the
-notions of edit scripts and operational transformations.
 \paragraph{Strings} Strings are mergeable. Given a string ($s$) and

### 2017-02-26 paper/dali.tex
List eg.

--- 2017-02-20 paper/dali.tex
@@
 %escapeinside={\%*}{*)},         % if you want to add a comment within your code
-morekeywords={module, @@deriving, not, : , /\\}
+morekeywords={module, when, @@deriving, not, : , /\\}
 }}
@@
 %% Title information
-\title[DaLi]{DaLi: Database as a Library}         %% [Short Title] is optional;
+\title[]{Mergeable Types for Highly-Available Replicated Data}         %% [Short Title] is optional;
                                         %% when present, will be used in

### 2017-02-26 paper/mergeable-types.tex
List eg.

@@
 
+\begin{figure}
+\centering
+\includegraphics[scale=0.4]{Figures/pushouts}
+
+\caption{Commutative diagram for Def.~\ref{def:mergeable-type}.}
+\label{fig:pushouts}
+\end{figure}
+
 We now formally define what we mean by a mergeable type. The
 definition serves as a guide to write merge functions for functional
-data types, elevating them to the level of replicated data types.
+data types, and elevate them to the level of replicated data types.
 
-A functional data type library is a tuple consisting of a type ($t$),
-and a collection of functions ($f$, $g$, etc.) of type $t \rightarrow
-t$ that are \emph{primitive morphisms}. A morphism ($P$, $Q$, etc.) is
-either a primitive morphism, or an associative composition of
-morphisms ($P \circ Q \circ R$). An object $B:t$ is called a
-\emph{semantic successor} of $A:t$ (conversely, $A$ is called a
-\emph{semantic ancestor} of $B$) if and only if there exists a
-morphism $P$ such that $P(A) = B$.
+A functional data type library is as a tuple consisting of a type
+($t$), and a (possibly infinite) collection of functions ($f$, $g$,
+etc.) of type $t \rightarrow t$. We call the functions \emph{primitive
+morphisms}\footnote{We borrow terminology from category theory for
+convenience}. A morphism ($P$, $Q$, etc.) is either a primitive
+morphism, or an associative composition of morphisms ($P \circ Q \circ
+R$). An object $B:t$ is called a \emph{semantic successor} of $A:t$
+(conversely, $A$ is called a \emph{semantic ancestor} of $B$) if and
+only if there exists a morphism $P$ such that $P(A) = B$.
 
@@
 successors and $D$ satisfies certain conditions. These conditions can
-be understood by observing the \rulelabel{E-Pull-Wait} rule of
+be understood by observing the \rulelabel{E-Pull-Wait} rule of of
 Fig.~\ref{fig:opsem}, which applies the \C{merge} function to a pair
@@
 and $v_1$, and $v$ and $v_2$ is the syntactic ancestor relationship
-that follows from the computation's branching structure. The merge
-function, as described above, however assumes that concurrent versions
-are semantic successors of their LCA.  It is therefore essential to
+that follows from the branching structure. The merge function, as
+described above, however assumes that the concurrent versions are
+semantic successors of their LCA. It is therefore essential to
 maintain coherence between the syntactic and semantic ancestor
@@
 semantic ancestor of $v_2$.  Fig.~\ref{fig:syntactic-ancestor-2}
-captures another way of establishing an ancestor relationship, namely by
+captures another way of establishing ancestor relationship, namely by
 merging branches. Version $v_{21}$ on branch $b_2$ is merged into
 $v_{11}$ on $b_1$ to create $v_{12}$ on $b_1$. Versions $v_{11}$ and
-\begin{wrapfigure}{l}{.4\textwidth}
-\centering
-\includegraphics[scale=0.35]{Figures/pushouts}
-\caption{Commutative diagram representing the merge operation.}
-\label{fig:pushouts}
-\end{wrapfigure}
 $v_{21}$ are now syntactic ancestors of $v_{12}$, but for them to be
-semantic ancestors, the merge function has to enforce the relationship
-explicitly. In other words, the result of merging  a pair of
+semantic ancestors, merge function has to enforce the relationship
+explicitly. In other words, the result of merging of a pair of
 semantic successors, $B$ and $C$, of an $A$, has to be an object $D$
@@
 two morphisms $P$ and $Q$ that map $A:t$ to $B:t$ and $C:t$, there
-also exists morphisms $P'$ and $Q'$, that map both $B$ and $C$ to
-$D:t$. This intuition is expressed visually in
-Fig.~\ref{fig:pushouts}.  For a mergeable type $t$, we have the
-guarantee that a sequence of values read by a thread from its branch
-is sensible, as per the data type semantics:
+must also exist morphisms $P'$ and $Q'$, the map both $B$ and $C$ to
+$D:t$. This intuition is expressed visusally in
+Fig.~\ref{fig:pushouts}.
+
+For a mergeable type $t$, we have the guarantee that a sequence of
+values read by a thread from its branch ``makes sense'' as per the
+data type semantics:
 
@@
 
-Rather than specifying merge ($M$) in terms of an LCA,
-Definition~\ref{def:mergeable-type} specifies $M$ in terms of a pair
-of morphisms.  This specification is useful to avoid the expression of certain
-nonsensical merge functions, and thus enforce a useful branch-local
-consistency property (Theorem~\ref{thm:branch-consistency}). For
-instance, consider the counter data type from Sec.~\ref{sec:intro}. A
-three-way merge function can be defined to return a constant
-regardless of its arguments:
+\begin{figure}
+\centering
+\includegraphics[scale=0.4]{Figures/pushouts-2}
+
+\caption{Commutative diagram for Def.~\ref{def:mergeable-type-2}.}
+\label{fig:pushouts-2}
+\end{figure}
+
+Def.~\ref{def:mergeable-type} specifies the merge function ($M$)
+slightly differently; instead of returning a merged value like the
+three-way merge function (\C{merge}), $M$ returns a pair of morphisms.
+This specification is useful to preempt certain non-sensical merge
+functions, and thus enforce the branch-local consistency property
+(Theorem~\ref{thm:branch-consistency}). For instance, consider the
+counter data type from Sec.~\ref{sec:intro}. A three-way merge
+function can be defined to return a constant regardless of its
+arguments:
 \begin{ocaml}
   let merge lca a b = 0
-\end{ocaml}
-Such a merge function allows the client to
+\end{ocaml} Such a merge function allows the client (a thread) to
 witness the violation of the branch-local consistency property. For
@@
 non-negativeness of an integer etc) that are preserved by all of $t$'s
-morphisms.  The commutativity of the diagram in
-Fig.~\ref{fig:pushouts} need not be enforced statically. It can
-equally be verified at run-time by checking that the morphisms returned by $M$
-map respective concurrent versions to the same value. The
+morphisms. On the hand, the commutativity of the diagram in
+Fig.~\ref{fig:pushouts} may not be enforced statically. It can however
+be verified at run-time by checking that the morphisms returned by $M$
+map the respective concurrent versions to the same value. The
 operational semantics (Fig.~\ref{fig:opsem}) can be extended to this
@@
 Merge functions for simple data types are easy to write. For instance,
-the merge logic for counter is simple enough that it does not require
-great effort to identify the $P$ and $Q$ that leda to the
+the merge logic for counter is simple enough that it doesn't require
+explicit efforts to identify $P$ and $Q$ that led to the
 concurrent counter versions, and transform them into $P'$ and $Q'$
 such that the diagram in Fig.~\ref{fig:pushouts} commutes. For more
-\begin{wrapfigure}{l}{.4\textwidth}
-\centering
-\includegraphics[scale=0.35]{Figures/pushouts-2}
-\caption{Commutative diagram for Def.~\ref{def:mergeable-type-2}.}
-\label{fig:pushouts-2}
-\end{wrapfigure}
-sophisticated data types, this process is more involved, and may require
+sophisticated data types, this process is more involved, and requires
 the programmer to perform non-trivial reasoning. We now present an
@@
 \end{itemize}
-The Commutativity diagram of Fig.~\ref{fig:pushouts-2} visualizes this
+Commutativity diagram of Fig.~\ref{fig:pushouts-2} visualizes this
 definition. 
@@
 notions of edit scripts and operational transformations.
-\begin{wrapfigure}{l}{.4\textwidth}
-\centering
-\includegraphics[scale=0.4]{Figures/string-eg}
-\caption{Merging strings with edit scripts and operational
-transformations}
-\label{fig:string-eg}
-\end{wrapfigure}
+
 \paragraph{Strings} Strings are mergeable. Given a string ($s$) and
@@
 shortest sequence of \C{insert}, \C{delete}, and \C{substitute}
-operations (a string's primitive morphisms), whose composition takes $s$
+operations (string's primitive morphisms), whose composition takes $s$
 to $s'$. The length of the sequence is in fact the Levenshtein
 distance~\cite{levenstein} between $s$ and $s'$, which can be computed
-in polynomial time~\cite{wagner-fischer}.
+in polynomial time via Wagner-Fischer algorithm~\cite{wagner-fischer}.
 A slight modification of Wagner-Fischer also lets us compute the edit

### 2017-02-26 paper/model.tex
List eg.

--- 2017-02-23 paper/model.tex
@@
 The \drawsome example demonstrates the utility of mergeable data types
-in building concurrent applications with conceptual sharing of
-state. This example, however, implicitly makes certain assumptions
+and \name programming model in building concurrent applications with
+conceptual sharing of state. The model lets concurrent computations be
+composed around any mergeable type. As exemplified by \drawsome,
+writing a three-way \C{merge} function is the only creative process in
+lifting an OCaml data type to a mergeable type. We now present few
+more examples of mergeable types, and isolate a recurring pattern in
+their merge functions that serves as a guide to write merge functions
+for more sophisticated types.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.4]{Figures/list-eg}
+
+  \caption{Lists of mergeable values are mergeable. }
+  \label{fig:list-eg}
+\end{figure}
+
+\input{list-code}
+
+{\bf Lists}. List, as an abstract data type, supports three
+operations: \C{insert x i l}, that inserts an element \C{x} at
+position \C{i} in the list \C{l}, \C{delete i l}, that deletes the
+element at \C{i}'th position, and \C{subst i x l}, that substitutes
+the element at \C{i}'th position with \C{x}. Lists of mergeable items
+are mergeable. The signature of a mergeable list library is shown in
+Fig.~\ref{fig:mlist-sig}. Note that the type of items in the list
+(\C{A.t}) is also required to be mergeable. The signature
+\C{MERGEABLE} will be discussed shortly.
+
+The merge operation for lists is composed of two separate functions,
+\C{edit\_seq} and \C{op\_transform}.
+\begin{itemize}
+  \item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and
+  computes the shortest sequence of list operations that need to be
+  applied on \C{v} to obtain \C{v'}. Such a sequence is called an
+  \emph{edit sequence}. The length of the sequence corresponds to the
+  standard notion of the \emph{edit distance} between the two lists,
+  which can be computed in polynomial time, for e.g., using the
+  Wagner-Fisher algorithm~\cite{wagner-fischer}. A slight modification
+  of Wagner-Fischer also lets us compute the edit sequence. The
+  implementation in Fig.~\ref{fig:mlist} represents edits using the
+  type \C{edit}. Constructors \C{I}, \C{D}, and \C{S} stand for
+  \C{insert}, \C{delete}, and \C{subst}, respectively; \C{Nop} is
+  there for technical reasons. The subst constructor also carries the
+  \C{A.t} element that was substituted. Our implementation of
+  Wagner-Fischer is more-or-less standard (hence, not shown), except
+  that the edit sequence it returns orders \C{I}'s and \C{D}'s before
+  \C{S}'s.  Fig.~\ref{fig:list-eg} illustrates edit sequences for a
+  sample list. The sequence \C{[I(c,0); S(3,c,s)]} maps the list
+  \C{[a;b;c]} to \C{[c;a;b;s]} because \C{subst 3 s (insert c 0
+  [a;b;c]) = [c;a;b;s]}.
+  \item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and
+  $s_2$, that map a list $v$ to two different lists, $v_1$ and $v_2$
+  (e.g., Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such
+  that $s_1'$ has the same effect on $v_2$ as $s_1$ had on $v$.  For
+  instance, in Fig.~\ref{list-eg}, let $s_1$ be the edit sequence
+  \C{[D(1); S(d,1)]}, which maps the list \C{[a;b;c]} ($v$) to
+  \C{[a;d]} ($v_1$). The \C{D} edit deletes the first element of $v$.
+  However, if $s_1$ is to be applied on the list \C{[c;a;b;s]}
+  ($v_2$), which has a new element at position 0, the \C{D} edit must
+  delete the the second element from $v_2$, if it were to have the
+  same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
+  transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The
+  function \C{op\_transform} of Fig.~\ref{fig:mlist-xform} computes
+  such \emph{operational transformation} for the sequence \C{mine}
+  w.r.t the sequence \C{others}, and returns \C{mine'} as the
+  transformed sequence.  
+\end{itemize}
+
+This example, however, implicitly makes certain assumptions
 about the model and the underlying system, such as the existence of a

### 2017-02-26 paper/mergeable-types.tex

--- 2017-02-25 paper/mergeable-types.tex
@@
 notions of edit scripts and operational transformations.
-\begin{wrapfigure}{l}{.4\textwidth}
-\centering
-\includegraphics[scale=0.4]{Figures/string-eg}
-\caption{Merging strings with edit scripts and operational
-transformations}
-\label{fig:string-eg}
-\end{wrapfigure}
-\paragraph{Strings} Strings are mergeable. Given a string ($s$) and
-its semantic successor ($s'$), an edit script ($P^*$ or $Q^*$) is the
-shortest sequence of \C{insert}, \C{delete}, and \C{substitute}
-operations (a string's primitive morphisms), whose composition takes $s$
-to $s'$. The length of the sequence is in fact the Levenshtein
-distance~\cite{levenstein} between $s$ and $s'$, which can be computed
-in polynomial time~\cite{wagner-fischer}.
-A slight modification of Wagner-Fischer also lets us compute the edit
-script. Fig.~\ref{fig:string-eg} illustrates.
-
-Next, we apply operational transformation to compute $P^*$ given $Q^*$
-($P^* <+ Q^*$), \emph{or} $Q^*$ given $P^*$ ($Q^* <+ P^*$); there is
-no need to compute both unless verification is one of the aims.
-
-\paragraph{}

### 2017-02-26 paper/model.tex

--- 2017-02-26 paper/model.tex
@@
   such \emph{operational transformation} for the sequence \C{mine}
-  w.r.t the sequence \C{others}, and returns \C{mine'} as the
-  transformed sequence.  
+  w.r.t the sequence \C{others}, and returns \C{mine'}, the
+  transformed sequence. To do so, it first pulls the \C{S} edits in
+  \C{mine} to the beginning of the sequence, transforming them w.r.t
+  the \C{I} and \C{D} edits in \C{mine}. Having all the \C{S} edits at
+  the beginning of the \C{mine} sequence makes their transformation
+  easier since the offset of the transformed substitution depends only
+  on the edits in the \C{other} sequence. Furthermore, it facilitates
+  an easy comparison with \C{S} edits at the end of \C{other}. Next,
+  \C{op\_transform} transforms edits in \C{mine} w.r.t \C{other} by
+  comparing their offsets; the rules are straightforward.
 \end{itemize}

### 2017-02-26 paper/list-sig.tex
Edits.

\begin{figure}

\begin{subfigure}[b]{0.7\textwidth}
\begin{ocaml}
module type MList = sig
  module A: sig
    type t
    val merge: t -> t -> t -> t
  end
  type t = A.t list
  ... (* All the standard list functions *)
  val merge: t -> t -> t -> t (* and the merge function *)
  (* The following are exposed only for this presentation *)
  type edit = I of A.t * int
    | D of int
    | S of int * A.t * A.t
  val edit_seq: t -> t -> edit list option
  val op_transform: edit list -> edit list -> edit list
end
\end{ocaml}
\caption{Signature of Mergeable Lists}
\label{fig:mlist-sig}
\end{subfigure}

\caption{Mergeable List Implementation}
\label{fig:mlist}
\end{figure}

### 2017-02-26 paper/model.tex
Edits.

--- 2017-02-26 paper/model.tex
@@
 
-\input{list-code}
+\input{list-sig}
 
 {\bf Lists}. List, as an abstract data type, supports three
-operations: \C{insert x i l}, that inserts an element \C{x} at
-position \C{i} in the list \C{l}, \C{delete i l}, that deletes the
-element at \C{i}'th position, and \C{subst i x l}, that substitutes
-the element at \C{i}'th position with \C{x}. Lists of mergeable items
-are mergeable. The signature of a mergeable list library is shown in
-Fig.~\ref{fig:mlist-sig}. Note that the type of items in the list
-(\C{A.t}) is also required to be mergeable. The signature
-\C{MERGEABLE} will be discussed shortly.
+operations: \C{insert x i l}, to insert an element \C{x} at position
+\C{i} in the list \C{l}, \C{delete i l}, to delete the element at
+\C{i}'th position, and \C{subst i x l}, to substitute the element at
+\C{i}'th position with \C{x}. Lists of mergeable items are mergeable
+while preserving the following properties:
+\begin{itemize}
+  \item Elements present in both the merging lists will be present in
+  the final list. No element is deleted unless it is deleted in at
+  least one list.
+  \item An element deleted in one list will not reappear in the merged
+  list.
+  \item The order of elements is retained. That is, if $x$ occurs
+  before $y$ in one of the lists, and if both elements are present in
+  the merged list, then $x$ occurs before $y$ also in the merged list.
+  \item If an element $y$ substitutes $x$ in one list, then $y$ also
+  substitutes $x$ in the merged list (i.e., $x$ is absent and $y$ is
+  present), unless $x$ is deleted or substituted in the other list. In
+  the former case, deletion wins, and neither $x$ nor $y$ occur in the
+  merged list. In the latter case, if $x$ is substituted by a
+  different element $z$ in the other list, then the merged list
+  substitutes $x$ by a merge of $y$ and $z$, defined as per the merge
+  semantics of the list element type.
+\end{itemize}
+Thus list is an example of a polymorphic mergeable data type that
+demonstrates how mergeability can be composed. The compositionality is
+reified in the the signature of a mergeable list library is shown in
+Fig.~\ref{fig:mlist-sig}; observe that the type of items in the list
+(\C{A.t}) is also required to be mergeable (\C{A.merge}).
 
 The merge operation for lists is composed of two separate functions,
-\C{edit\_seq} and \C{op\_transform}.
+\C{edit\_seq} and \C{op\_transform}. Both these functions have been
+implemented in less than 50 lines of OCaml included in the appendix.
+The functions are described below.
 \begin{itemize}
@@
   which can be computed in polynomial time, for e.g., using the
-  Wagner-Fisher algorithm~\cite{wagner-fischer}. A slight modification
-  of Wagner-Fischer also lets us compute the edit sequence. The
-  implementation in Fig.~\ref{fig:mlist} represents edits using the
-  type \C{edit}. Constructors \C{I}, \C{D}, and \C{S} stand for
-  \C{insert}, \C{delete}, and \C{subst}, respectively; \C{Nop} is
-  there for technical reasons. The subst constructor also carries the
-  \C{A.t} element that was substituted. Our implementation of
-  Wagner-Fischer is more-or-less standard (hence, not shown), except
-  that the edit sequence it returns orders \C{I}'s and \C{D}'s before
-  \C{S}'s.  Fig.~\ref{fig:list-eg} illustrates edit sequences for a
-  sample list. The sequence \C{[I(c,0); S(3,c,s)]} maps the list
-  \C{[a;b;c]} to \C{[c;a;b;s]} because \C{subst 3 s (insert c 0
-  [a;b;c]) = [c;a;b;s]}.
+  Wagner-Fisher algorithm~\cite{wagner-fischer}. \C{edit\_seq} uses a
+  slight modification of Wagner-Fischer, that also lets it compute the
+  edit sequence (an \C{edit list}). The implementation specified in
+  Fig.~\ref{fig:mlist-sig} represents edits using the type \C{edit}.
+  Constructors \C{I}, \C{D}, and \C{S} stand for \C{insert},
+  \C{delete}, and \C{subst}, respectively\footnote{
+    Note that \C{I}, \C{D} and \C{S} are \emph{not} effects; no list
+    function generates them, and they are not exchanged between
+    concurrent threads. They are simply tags used by the list merge
+    operation for convenience.
+  }. The subst constructor also carries the \C{A.t} element that was
+  substituted. The element serves as the \C{lca} argument to
+  \C{A.merge} in case of concurrent conflicting substitutions.
+  Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample list.
+  The sequence \C{[I(c,0); S(3,c,s)]} maps the list \C{[a;b;c]} to
+  \C{[c;a;b;s]} because \C{subst 3 s (insert c 0 [a;b;c]) =
+  [c;a;b;s]}.
   \item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and
@@
   instance, in Fig.~\ref{list-eg}, let $s_1$ be the edit sequence
-  \C{[D(1); S(d,1)]}, which maps the list \C{[a;b;c]} ($v$) to
+  \C{[D(1); S(1,c,d)]}, which maps the list \C{[a;b;c]} ($v$) to
   \C{[a;d]} ($v_1$). The \C{D} edit deletes the first element of $v$.
@@
   transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The
-  function \C{op\_transform} of Fig.~\ref{fig:mlist-xform} computes
-  such \emph{operational transformation} for the sequence \C{mine}
-  w.r.t the sequence \C{others}, and returns \C{mine'} as the
-  transformed sequence.  
+  function \C{op\_transform} (Fig.~\ref{fig:mlist-sig}) computes such
+  \emph{operational transformations} for an edit sequence $s_1$ w.r.t
+  another edit sequence $s_2$. A notable transformation rule is for a
+  substitution w.r.t another substitution at the same position. Since
+  the substituting items could be different, the function relies on
+  the item merge function (\C{A.merge}) to merge them into a single
+  item. For instance, in Fig.~\ref{fig:list-eg}, there are
+  simultaneous substitutions to the item \C{c} in $s_1$ and $s_2$;
+  while $s_1$ substitutes it with \C{d}, $s_2$ does it with \C{s}. The
+  merged list therefore substitutes \C{c} with the merge of \C{d} and
+  \C{s}, defined as per the merge semantics of \C{A.t}.
 \end{itemize}
+The \C{MList.merge} can now be defined as a function that first
+computes edit sequences, $s_1$ and $s_2$, for the two concurrent
+lists, $v_1$ and $v_2$ w.r.t their common ancestor $v$, transforms one
+of them (say, $s_1$) w.r.t $s_2$ to obtain $s_1'$, and finally applies
+(the operations defined by) $s_1'$ to $v_2$ to obtain the merged list.
+
+A signficant advantage of mergeable types is their composability. For
+instance, a mergeble list of counters can be obtained simply by
+instantiating \C{MList} for \C{Counter} (i.e., \C{MList(Counter)}). A
+non-trivial demonstration of the utility of such composability is
+given below.
+
+% The mergeable list data type demonstrates a common pattern of writing
+% \C{merge} functions applicable to a range of functional data
+% structures. For instance, a binary tree (with rotations) can be made
+% mergeable in the same way as list, i.e., by computing edit sequences
+% (\C{edit\_seq}) and then their transformations
+% (\C{op\_transform})\footnote{
+%   Like Wagner-Fischer, established algorithms exist for computing tree
+%   edit distances~\cite{tree-diff}. An alternative naive method to
+%   merge a pair of trees is by merging the lists obtained from their
+%   in-order traversal, followed by reconstructing the tree.
+% }.
+
+{\bf Shopping List}. A shopping list is an
 

### 2017-02-26 paper/related.tex

\section{Related Work}
\label{sec:related}

Our idea of versioning state bears resemblance to Concurrent
Revisions~\cite{BBL+10}, a programming abstraction that provides
deterministic concurrent execution by forking and merging
\emph{versions}, portions of state shared among concurrent threads.
The idea of using revisions as a means to programming eventually
consistent distributed systems was further developed
in~\cite{BFL+12,Burckhardt2012}.  The \name\ programming model,
however, differs substantially from a concurrent revisions model
because it imposes no distinction between \emph{servers}, machines
that hold global state, and \emph{clients}, devices that operate on
local, potentially stale, data - any computation executing in a
distributed environment is free to fork new versions, and synchronize
against other replicated state.  Our model, furthermore, supports
fully decentralized operation and is robust to network partitions and
failures. Just as significantly, \name\ allows applications to
customize join semantics with programmable merge operations.  Indeed,
the integration of a version-based mechanism within OCaml allows a
degree of type safety, composability, and polymorphism not available
in related systems.

\cite{Burckhardt2015} also presents an operational model of a
replicated data store that is based on the abstract system model
presented in ~\cite{Burckhardt2014}; their design is similar to the
model described in~\cite{pldi15}.  In both approaches, coordination
among replicas involves transmitting operations on replicated objects
that are performed locally on each replica.  In contrast,
\name\ allows programmers to use familiar state-based and functional
abstractions when developing distributed applications.  As we
illustrated in Fig.~\ref{fig:counter-rdt}, using effects and
operations to coordinate the activities of replicas may involve
addressing subtleties that do not manifest otherwise.  Our case
studies and experimental results support our contention that using
well-understood state (heap)-based abstractions to build distributed
applications greatly simplifies program reasoning and eases
development, without compromising efficiency.

Modern distributed systems are often equipped with only parsimonious
data models (e.g., key-value model) and poorly understood low-level
consistency guarantees\footnote{Cassandra~\cite{Cassandra}, a popular
  NoSQL data store, comes with various consistency enforcement
  mechanisms, such as anti-entropy protocols, {\sc QUORUM} and {\sc
    LOCAL\_QUORUM} reads and writes, and light-weight transactions,
  each of which can be controlled via configuration knobs or runtime
  parameters.}  that complicate program reasoning, and make it hard to
enforce application integrity. Some authors~\cite{bailis-bolton} have
demonstrated that it is possible to\emph{bolt on} high-level
consistency guarantees (e.g., causal consistency)~\cite{COPS,BEG+17}
as a \emph{shim layer} service over existing stores.

A number of verification techniques, programming abstractions, and
tools have been proposed to reason about program behavior in a
geo-replicated weakly consistent environment.  These techniques treat
replicated storage as a black box with a fixed pre-defined consistency
model~\cite{bailis-vldb, alvaro-calm, gotsman-popl16,redblue-atc,
  redblue-osdi, ecinec}.  On the other hand, compositional proof
techniques and mechanized verification frameworks have been developed
to rigorously reason about various components of a distributed data
store~\cite{verdi,lbc16}. \name seeks to provide a rich high-level
programming model, built on rigorous foundations, that can facilitate
program reasoning and verification.  An important by-product of the
programming model is that it does not require algorithmic
restructuring to transplant a sequential or concurrent program to a
distributed, replicated setting; the only additional burden imposed on
the developer is the need to provide a merge operator, a function that
can be often easily written for many common datatypes.

Several conditions have been proposed to judge whether an operation on
a replicated data object needs coordination or not. ~\cite{Calm}
defines \emph{logical monotonicity} as a sufficient condition for
coordination freedom, and proposes a consistency analysis that marks
code regions performing non-monotonic reasoning (eg: aggregations,
such as \C{COUNT}) as potential coordination
points. ~\cite{IConfluence} and ~\cite{Sieve} define \emph{invariant
  confluence} and \emph{invariant safety}, respectively, as conditions
for safely executing an operation without coordination.
~\cite{indigo} requires programmers to declare application semantics,
and the desired application-specific invariants as formulas in
first-order logic. It performs static analysis on these formulas to
determine $I$-offender sets - sets of operations, which, when
performed concurrently, result in violation of one or more of the
stated invariants. For each offending set of operations, if the
programmer chooses invariant-violation avoidance over violation
repair, the system employs various techniques, such as escrow
reservation, to ensure that the offending set is effectively
serialized.  All of these approaches differ significantly from the
core goals of \name, which is to enable seamless application-driven
techniques for programming geo-replicated systems.  The
\name\ programmer is not concerned with different system-specific
consistency or isolation levels, or invariants that are sensitive to
these levels.  Instead, the only requirements demanded of the
developer is the need to reason about a sensible merge semantics for
data structures.  Such reasoning can be applied without consideration
of system- or architecture-specific details.  This reasoning phase
implicitly expresses salient semantic invariants without having to be
exposed to low-level operational details.

### 2017-02-26 data/merge.py
data

@@ -1,0 +1,40 @@
+import matplotlib
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.ticker import ScalarFormatter, FormatStrFormatter
+import matplotlib.patches as mpatches
+
+nodeKind = ['o-', 's--', 'D-.', '^:', 'X-', 'V--', '>-.', '<:']
+
+matplotlib.rcParams['ps.fonttype'] = 42
+matplotlib.rcParams['pdf.fonttype'] = 42
+
+def main ():
+  infile = open ("merge.dat", "r")
+  merge = []
+  for line in infile.readlines ():
+    if line == "\n" or line.startswith ("#"):
+      continue
+    d = line.split ()
+    merge.append ((int(d[0]), float(d[1])/1000.0, float(d[2])/1000.0, float(d[3])/1000.0))
+
+  font = {'family' : 'normal', 'weight' : 'normal', 'size' : 12}
+  matplotlib.rc('font', **font)
+
+  fig, ax1 = plt.subplots()
+
+  (dist,merge,comp,nw) = zip(*merge)
+
+  fig, ax = plt.subplots()
+  ax.stackplot (dist, [comp, nw], colors=['#55BA87','#7E1137'])
+  ax.set_xlabel("Time (s)")
+  ax.set_ylabel("Distance (ops)")
+  ax.set_xlim([0,256])
+  plt.legend([mpatches.Patch(color='#55BA87'),
+              mpatches.Patch(color='#7E1137')],
+             ['computation','network'], loc='upper left')
+  fig.tight_layout()
+  plt.savefig ("merge.pdf")
+  plt.close ()
+
+main ()

### 2017-02-27 paper/model.tex
Sec. 2

@@
   \item An element deleted in one list will not reappear in the merged
-  list.
+  list (unlike Amazon's replicated shopping cart~\cite{dynamo}).
   \item The order of elements is retained. That is, if $x$ occurs
@@
 
-A signficant advantage of mergeable types is their composability. For
-instance, a mergeble list of counters can be obtained simply by
-instantiating \C{MList} for \C{Counter} (i.e., \C{MList(Counter)}). A
-non-trivial demonstration of the utility of such composability is
-given below.
+\GK{Somewhere here, I want to say:} To better appreciate the virtues
+of mergeable types, we encourage the reader to think about how one
+might implement the mergeable list type with the explicit effect
+representation described in Sec.~\ref{sec:intro}.
+
+\begin{figure}
+  \centering
+  \includegraphics[scale=0.4]{Figures/shopping-list}
+
+  \caption{Merging concurrent shopping lists}
+  \label{fig:shopping-list}
+\end{figure}
+
+A significant advantage of mergeable types is their composability. As
+demonstrated by \C{MList}, the merge semantics of a composed mergeable
+type is a composition of merge semantics of the individual types.  For
+instance, a mergeable list of integers can be obtained by
+instantiating \C{MList} for an \C{Int} type, where \C{Int.merge} is
+same as \C{Counter.merge} (Sec.~\ref{sec:intro}). 
+An alternative instantiation of \C{MList} is for a quantity type
+(\C{Qty}) that, like \C{Int}, supports \C{add} and \C{subtract}
+operations, but defines its \C{merge} semantics a little differently:
+\begin{ocaml}
+let merge lca v1 v2 = if v1 = lca then v2 else (if v2 = lca then v1 else max v1 v2)
+\end{ocaml}
+\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has
+the same semantics under a sequential execution. Their differences
+however manifest under a concurrent execution, where both adopt
+different methods to reconcile concurrent updates.  is given
+below.
 
@@
 
-{\bf Shopping List}. A shopping list is an
+{\bf Shopping List}. A non-trivial demonstration of composability of
+mergeable types is a shopping list application. The application
+allows its users to collaboratively build a grocery shopping
+list (listing items in decreasing ), by adding new items with a
+specified quantity, deleting existing items, and increasing or
+decreasing their quantity. An item (\C{Item.t}) is represented as a
+tuple of its name (a string) and the quantity (A \C{Qty.t}). Merge
+function for \C{Item.t} merges the quantities of items (using
+\C{Qty.merge}) with same names. Items with different names are
+considered distinct. A shopping list is a mergeable list of items
+(i.e., \C{MList(Item)}). \C{MList(Item).merge} automatically lifts the
+merge semantics for items to merge semantics for lists of items.
+Fig.~\ref{fig:shopping-list} illustrates. Here, two users, Alice and
+Bob, concurrently edit a shopping list whose initial contents are
+as milk and eggs. While both simultaneously update the quantity of
+eggs, Bob also removes milk and inserts candy. The merged shopping
+list contains eggs and candy, where the quantity of eggs is obtained
+by merging Alice's and Bob's quantity as per \C{Qty.merge} semantics.
 
-This example, however, implicitly makes certain assumptions
-about the model and the underlying system, such as the existence of a
-single least common ancestor (LCA) for any pair of versions, the
-ability to access a previous version on any branch, and mergeability
-of any two concurrent versions. Enforcing these guarantees in a fully
+The examples described above make certain assumptions about the model
+and the underlying system, such as the existence of a single least
+common ancestor (LCA) for any pair of versions, the ability to access
+a previous version on any branch, and mergeability of any two
+concurrent versions. Enforcing these guarantees in a fully
 decentralized distributed setting requires addressing non-trivial
-theoretical challenges described in the next section. Unlike the
-canvas, the process of writing merge functions for data types with
-non-trivial invariants could be quite involved. The following sections
-describe a principled approach to deriving merge functions for any
-data type given its invariants. Like \C{Canvas.merge}, merge functions
-often need to compare multiple versions of a data type for structural
-equality, which, when performed naively, incurs cost linear in the
-size of versions. The representation of a versioned data type in \name
-is optimized for performing comparisons that allows structural
-equality to be determined in constant time. Such engineering details,
-which contribute to the practicality of our programming model are also
-discussed in subsequent sections.
+theoretical challenges. These, and other details that contribute to
+the practicality of our model, such as containing the complexity of
+merges, are discussed in subsequent sections.

### 2017-02-27 paper/model.tex
Edits.

@@
   standard notion of the \emph{edit distance} between the two lists,
-  which can be computed in polynomial time, for e.g., using the
-  Wagner-Fisher algorithm~\cite{wagner-fischer}. \C{edit\_seq} uses a
-  slight modification of Wagner-Fischer, that also lets it compute the
-  edit sequence (an \C{edit list}). The implementation specified in
+  which can be computed in polynomial time~\cite{wagner-fischer}.
+  \C{edit\_seq} modifies one such algorithm to also compute the edit
+  sequence (an \C{edit list}). The implementation specified in
   Fig.~\ref{fig:mlist-sig} represents edits using the type \C{edit}.

### 2017-02-27 paper/dali.tex
edits to related work
Edits.
Sec. 2
edits

--- 2017-02-25 paper/dali.tex
@@
   communicated to other nodes where they can be re-applied locally.
-  As a result, the degree of restructuring and care necessary to
+  Oftentimes, the degree of restructuring and care necessary to
   transform a sequential or shared-memory concurrent application to a
@@
   functions.  Distributed state is now viewed in terms of a tree of
-  immutable object versions, with merge functions used implicitly by
-  the underlying runtime system to periodically reconcile local
-  versions to yield a consistent global state.  Thus, our model frees
-  the programmer from having to think of low-level operational notions
-  related to distribution and replication, and therefore does not
-  entail any restructuring of application logic.
+  immutable object versions, with merge actions used explicitly by the
+  application or implicitly by the underlying runtime system to
+  periodically reconcile local versions to yield a consistent global
+  state.  Thus, our model frees the programmer from having to think of
+  low-level operational notions related to distribution and
+  replication, and therefore does not entail any restructuring of
+  application logic to enable distributed programming.
 
   We show how any ML datatype can be enriched with mergeable
-  semantics, formalize a rich concurrency model based on such types ,
+  semantics, formalize a rich concurrency model based on such types,
   and describe an instantiation of OCaml equipped with these features.
@@
 <concept>
-<concept_id>10011007.10011006.10011008</concept_id>
-<concept_desc>Software and its engineering~General programming languages</concept_desc>
+<concept_id>10011007.10011006.10011008.10011009.10010177</concept_id>
+<concept_desc>Software and its engineering~Distributed programming languages</concept_desc>
 <concept_significance>500</concept_significance>
@@
 <concept>
-<concept_id>10003456.10003457.10003521.10003525</concept_id>
-<concept_desc>Social and professional topics~History of programming languages</concept_desc>
+<concept_id>10011007.10011006.10011008.10011024.10011028</concept_id>
+<concept_desc>Software and its engineering~Data types and structures</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10011007.10011006.10011039.10011311</concept_id>
+<concept_desc>Software and its engineering~Semantics</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10003752.10010124.10010125.10010130</concept_id>
+<concept_desc>Theory of computation~Type structures</concept_desc>
 <concept_significance>300</concept_significance>
@@
 
-\ccsdesc[500]{Software and its engineering~General programming languages}
-\ccsdesc[300]{Social and professional topics~History of programming languages}
+\ccsdesc[500]{Software and its engineering~Distributed programming languages}
+\ccsdesc[500]{Software and its engineering~Data types and structures}
+\ccsdesc[300]{Software and its engineering~Semantics}
+\ccsdesc[300]{Theory of computation~Type structures}
+
 %% End of generated code
@@
 %% comma separated list
-\keywords{keyword1, keyword2, keyword3}  %% \keywords is optional
+\keywords{Replicated Datatypes, Eventual Consistency, Availability, CRDTs, Versions, ML}  %% \keywords is optional
 
@@
 
+\input{related}
 %% Acknowledgments

### 2017-02-27 paper/intro.tex
edits

@@
 restrict, or ignore important albeit unpleasant realities (e.g.,
-network partitions~\cite{Brewer2000,Gilbert2002}). Given these
-unwarranted assumptions, program behavior is often difficult to
-predict and verify.  On the other hand, when these features are
-explicitly exposed to the programmer, e.g., by requiring that
-applications be written in terms of specialized distributed data
+network partitions~\cite{Brewer2000,Gilbert2002}). When unwarranted
+assumptions are made, program behavior is often difficult to predict
+and verify.  On the other hand, when these features are explicitly
+exposed to the programmer, e.g., by requiring that applications be
+written in terms of specialized distributed data
 structures~\cite{Burckhardt2014,BFL+12,SPB+11} or control
@@
 heavyweight mechanisms, applying an \C{Add} operation on one node may
-not be instanteously witnessed on another, which may in the process of
+not be instanteously witnessed on another, which may be in the process of
 simultaneously attempting to perform its own \C{Add} or \C{Mult}
@@
 configuration~\cite{Armstrong}), is also an undesirable solution,
-given its sensitivity to network partitions and server failures, and
+given the sensitivity of such architectures to network partitions and server failures, and
 the negative performance impact it incurs in geo-distributed
-environments~\cite{Walter}.  Removig coordination altogether by
-simply replicating counter state without having any supporting
-consistency protocol is an equally infeasible approach.
+environments~\cite{Walter}.  Removig coordination altogether by simply
+replicating counter state without having any supporting consistency
+protocol is an equally infeasible approach.
 
-A method often adopted~\cite{SPB+11,Burckhardt2014} to address this
-tension is to re-define a datatype's operations to return
-\emph{effects} instead of values.  An \emph{effect} is a tag that
+A method often adopted to address these concerns is to re-define a
+datatype's operations to return \emph{effects} instead of
+values~\cite{SPB+11,Burckhardt2014}.  An \emph{effect} is a tag that
 identifies the operation to be applied uniformly at all replicas to
@@
 multiplying 5 on one replica yields 15, while applying the effect of
-first multiplying 5 and then adding 3 yields 3 on the other.  Thus,
-implementations must be carefully written to take the lack of
-commutativity into account when defining how effects are applied at a
-replica; in the figure, multiplication is expressed in terms of
-addition to avoid the kind of undesirable behavior described above.
+first multiplying 5 and then adding 3 yields 3 on the other.  Such
+scenarios are possible in a distributed system because there are no
+coherence guarantees on the order in which effects are received by
+different nodes.  Thus, implementations must be carefully written to
+take the lack of commutativity into account when defining how effects
+are applied at a replica; in the figure, multiplication is expressed
+in terms of addition to avoid the kind of undesirable behavior
+described above.
 
 In this implementation, all replicas will eventually contain the same
-counter value, assuming computation on the counter quiesces.  While
-transmitting effects provide a low-level operational basis for
+counter value, assuming updates to the counter eventually quiesces.
+While transmitting effects provide a low-level operational basis for
 handling replicated state, the \emph{ad hoc} nature of the solution
@@
 semantic gap between the two versions of the counter, one cognizant of
-replication and the other not, breaks backward compatibility.  Just as
-significantly, its non-trivial construction must be developed in
-different guises for every distributed data structure used by the
-application.
+replication and the other not, breaks backward compatibility with the
+original state-based implementation.  Just as significantly, its
+non-trivial construction must be developed in different guises for
+every distributed data structure used by the application.
 \begin{figure}
@@
     previous merged state (10) from which the two versions were
-    derived to yield a new merged state of 25.  Observe that there is
-    no explicit synchronization or coordination among versions.  In
-    the figure, red circles represent counter state produced through
-    merging, and blue circles represent state produced by applying
-    counter operations on a node-local version.  }
+    derived to yield a new merged state of 25.  There is no other form
+    of synchronization or coordination among versions except through
+    merging.  In the figure, red circles represent counter state
+    produced through merging, and blue circles represent state
+    produced by applying counter operations on a node-local version.
+  }
 \vspace*{-.85in}  
@@
 be thought of as yielding new versions, thus producing a version tree,
-with one branch for each replica.  Thus, each branch represents
+with one branch for each replica.  Thus, every branch represents
 different (immutable) versions maintained by different nodes, with the
-state produced by a counter's operations on a node recorded along the
-node's local branch.  Now, to generate a globally consistent view of a
-counter, we only need to define a merge operation that explains how to
-combine two local versions to produce a new version that reflects both
-their states.
+state produced by the computation performed over a counter on a node
+recorded along the node's local branch for the counter.  Now, to
+generate a globally consistent view of a counter, we only need to
+define a merge operation that explains how to combine two local
+versions to produce a new version that reflects both their states.
 
@@
 state from a collection of local versions of that state.  A
-computation progresses by forking replicating new versions of existing
-versions on geo-distributed nodes, creating new versions along a an
-existing \emph{branch}, or by merging branch tips.
+computation progresses by forking (i.e., replicating) new versions of
+existing versions on geo-distributed nodes, allowing
+synchronization-free local computation to proceed on those nodes,
+creating new versions along an existing \emph{branch} that represents
+updated local instances, and by merging branches to realize global
+consistency.
 
-Our model allows ML programmers to get the benefits of realizing
+Our model allows ML programmers to get the benefits of achieving
 highly available (low-latency) distributed computation, while
 continuing to enjoy the comforts of high-level reasoning and the
-familiarity of using standard libraries.  Notably, while \name's
-programming model makes no explicit reference to any specific
-operational manifestation of a distributed system (e.g., programmers
-do not need to explicitly manage replicas), we demonstrate that it can
-be nonetheless efficiently realized on existing geo-replicated
-distributed systems.
+familiarity of using standard libraries already provided by the
+language.  Notably, while \name's programming model makes no explicit
+reference to any specific operational manifestation of a distributed
+system (e.g., programmers do not need to explicitly manage replicas),
+we demonstrate that it can be nonetheless efficiently realized on
+existing geo-replicated distributed systems.
 
@@
 
-    \item We describe the \name library that transparently adds
-      persistence and replication features to a mergeable type. \name
-      comes with a programming model that brings to bear the power and
-      flexibility of a version control system to the administration of
-      replicated data. \name hides the complexity of version control
-      behind a monad, and exposes its functionality via a simple API
-      that lets programmers define and compose distributed ML
-      computations around such data.  Issues related to replication
-      only manifest implicitly in the definition of a merge operation
-      that defines coherence among different instances of replicated
-      state.
+    \item We describe \name programming model that brings to bear the
+      power and flexibility of a version control system to the
+      administration of replicated data. \name hides the complexity of
+      version control behind a monad, and exposes its functionality
+      via a simple API that lets programmers define and compose
+      distributed ML computations around such data.  Issues related to
+      replication only manifest implicitly in the definition of a
+      merge operation that defines coherence among different instances
+      of replicated state.
 
@@
 
-    \item We describe an implementation of \name atop the Irmin
-      persistent store~\cite{irmin}, a content-addressable storage
-      library for OCaml, along with case studies and experimental
-      results that justify its practical utility.
+    \item We describe an implementation of the \name library that
+      transparently adds persistence and replication features to a
+      mergeable type, and which sits atop the Irmin persistent
+      store~\cite{irmin}, a content-addressable storage library for
+      OCaml.  We also present case studies and experimental results
+      that justify the practical utility of our approach.
 \end{itemize}

### 2017-02-27 paper/model.tex
edits

--- 2017-02-26 paper/model.tex
@@
 and \C{deriving versioned} directive are explained below.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.5]{Figures/canvas-merging}
-
-\caption{Merging concurrent versions (\C{v1} and \C{v2} ) of a drawing
-canvas. \C{lca} is their common ancestor.}
-\label{fig:canvas-merging}
-\end{figure}
-
 \drawsome lets multiple users collaborate on a canvas that is
@@
 \C{v1} is white, and in \C{v2} it is green, and its color in \C{lca}
+\begin{wrapfigure}{l}{.5\textwidth}
+\centering
+\includegraphics[scale=0.5]{Figures/canvas-merging}
+\caption{Merging concurrent versions (\C{v1} and \C{v2} ) of a drawing
+canvas. \C{lca} is their common ancestor.}
+\label{fig:canvas-merging}
+\end{wrapfigure}
 is white, then it means that only \C{v2} modified the color. Hence the
@@
 
-\input{list-sig}
+\input{list-code}
 
 {\bf Lists}. List, as an abstract data type, supports three
-operations: \C{insert x i l}, to insert an element \C{x} at position
-\C{i} in the list \C{l}, \C{delete i l}, to delete the element at
-\C{i}'th position, and \C{subst i x l}, to substitute the element at
-\C{i}'th position with \C{x}. Lists of mergeable items are mergeable
-while preserving the following properties:
-\begin{itemize}
-  \item Elements present in both the merging lists will be present in
-  the final list. No element is deleted unless it is deleted in at
-  least one list.
-  \item An element deleted in one list will not reappear in the merged
-  list.
-  \item The order of elements is retained. That is, if $x$ occurs
-  before $y$ in one of the lists, and if both elements are present in
-  the merged list, then $x$ occurs before $y$ also in the merged list.
-  \item If an element $y$ substitutes $x$ in one list, then $y$ also
-  substitutes $x$ in the merged list (i.e., $x$ is absent and $y$ is
-  present), unless $x$ is deleted or substituted in the other list. In
-  the former case, deletion wins, and neither $x$ nor $y$ occur in the
-  merged list. In the latter case, if $x$ is substituted by a
-  different element $z$ in the other list, then the merged list
-  substitutes $x$ by a merge of $y$ and $z$, defined as per the merge
-  semantics of the list element type.
-\end{itemize}
-Thus list is an example of a polymorphic mergeable data type that
-demonstrates how mergeability can be composed. The compositionality is
-reified in the the signature of a mergeable list library is shown in
-Fig.~\ref{fig:mlist-sig}; observe that the type of items in the list
-(\C{A.t}) is also required to be mergeable (\C{A.merge}).
+operations: \C{insert x i l}, that inserts an element \C{x} at
+position \C{i} in the list \C{l}, \C{delete i l}, that deletes the
+element at \C{i}'th position, and \C{subst i x l}, that substitutes
+the element at \C{i}'th position with \C{x}. Lists of mergeable items
+are mergeable. The signature of a mergeable list library is shown in
+Fig.~\ref{fig:mlist-sig}. Note that the type of items in the list
+(\C{A.t}) is also required to be mergeable. The signature
+\C{MERGEABLE} will be discussed shortly.
 
 The merge operation for lists is composed of two separate functions,
-\C{edit\_seq} and \C{op\_transform}. Both these functions have been
-implemented in less than 50 lines of OCaml included in the appendix.
-The functions are described below.
+\C{edit\_seq} and \C{op\_transform}.
 \begin{itemize}
@@
   which can be computed in polynomial time, for e.g., using the
-  Wagner-Fisher algorithm~\cite{wagner-fischer}. \C{edit\_seq} uses a
-  slight modification of Wagner-Fischer, that also lets it compute the
-  edit sequence (an \C{edit list}). The implementation specified in
-  Fig.~\ref{fig:mlist-sig} represents edits using the type \C{edit}.
-  Constructors \C{I}, \C{D}, and \C{S} stand for \C{insert},
-  \C{delete}, and \C{subst}, respectively\footnote{
-    Note that \C{I}, \C{D} and \C{S} are \emph{not} effects; no list
-    function generates them, and they are not exchanged between
-    concurrent threads. They are simply tags used by the list merge
-    operation for convenience.
-  }. The subst constructor also carries the \C{A.t} element that was
-  substituted. The element serves as the \C{lca} argument to
-  \C{A.merge} in case of concurrent conflicting substitutions.
-  Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample list.
-  The sequence \C{[I(c,0); S(3,c,s)]} maps the list \C{[a;b;c]} to
-  \C{[c;a;b;s]} because \C{subst 3 s (insert c 0 [a;b;c]) =
-  [c;a;b;s]}.
+  Wagner-Fisher algorithm~\cite{wagner-fischer}. A slight modification
+  of Wagner-Fischer also lets us compute the edit sequence. The
+  implementation in Fig.~\ref{fig:mlist} represents edits using the
+  type \C{edit}. Constructors \C{I}, \C{D}, and \C{S} stand for
+  \C{insert}, \C{delete}, and \C{subst}, respectively; \C{Nop} is
+  there for technical reasons. The subst constructor also carries the
+  \C{A.t} element that was substituted. Our implementation of
+  Wagner-Fischer is more-or-less standard (hence, not shown), except
+  that the edit sequence it returns orders \C{I}'s and \C{D}'s before
+  \C{S}'s.  Fig.~\ref{fig:list-eg} illustrates edit sequences for a
+  sample list. The sequence \C{[I(c,0); S(3,c,s)]} maps the list
+  \C{[a;b;c]} to \C{[c;a;b;s]} because \C{subst 3 s (insert c 0
+  [a;b;c]) = [c;a;b;s]}.
   \item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and
@@
   instance, in Fig.~\ref{list-eg}, let $s_1$ be the edit sequence
-  \C{[D(1); S(1,c,d)]}, which maps the list \C{[a;b;c]} ($v$) to
+  \C{[D(1); S(d,1)]}, which maps the list \C{[a;b;c]} ($v$) to
   \C{[a;d]} ($v_1$). The \C{D} edit deletes the first element of $v$.
@@
   transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The
-  function \C{op\_transform} (Fig.~\ref{fig:mlist-sig}) computes such
-  \emph{operational transformations} for an edit sequence $s_1$ w.r.t
-  another edit sequence $s_2$. A notable transformation rule is for a
-  substitution w.r.t another substitution at the same position. Since
-  the substituting items could be different, the function relies on
-  the item merge function (\C{A.merge}) to merge them into a single
-  item. For instance, in Fig.~\ref{fig:list-eg}, there are
-  simultaneous substitutions to the item \C{c} in $s_1$ and $s_2$;
-  while $s_1$ substitutes it with \C{d}, $s_2$ does it with \C{s}. The
-  merged list therefore substitutes \C{c} with the merge of \C{d} and
-  \C{s}, defined as per the merge semantics of \C{A.t}.
+  function \C{op\_transform} of Fig.~\ref{fig:mlist-xform} computes
+  such \emph{operational transformation} for the sequence \C{mine}
+  w.r.t the sequence \C{others}, and returns \C{mine'}, the
+  transformed sequence. To do so, it first pulls the \C{S} edits in
+  \C{mine} to the beginning of the sequence, transforming them w.r.t
+  the \C{I} and \C{D} edits in \C{mine}. Having all the \C{S} edits at
+  the beginning of the \C{mine} sequence makes their transformation
+  easier since the offset of the transformed substitution depends only
+  on the edits in the \C{other} sequence. Furthermore, it facilitates
+  an easy comparison with \C{S} edits at the end of \C{other}. Next,
+  \C{op\_transform} transforms edits in \C{mine} w.r.t \C{other} by
+  comparing their offsets; the rules are straightforward.
 \end{itemize}
-The \C{MList.merge} can now be defined as a function that first
-computes edit sequences, $s_1$ and $s_2$, for the two concurrent
-lists, $v_1$ and $v_2$ w.r.t their common ancestor $v$, transforms one
-of them (say, $s_1$) w.r.t $s_2$ to obtain $s_1'$, and finally applies
-(the operations defined by) $s_1'$ to $v_2$ to obtain the merged list.
-
-A signficant advantage of mergeable types is their composability. For
-instance, a mergeble list of counters can be obtained simply by
-instantiating \C{MList} for \C{Counter} (i.e., \C{MList(Counter)}). A
-non-trivial demonstration of the utility of such composability is
-given below.
-
-% The mergeable list data type demonstrates a common pattern of writing
-% \C{merge} functions applicable to a range of functional data
-% structures. For instance, a binary tree (with rotations) can be made
-% mergeable in the same way as list, i.e., by computing edit sequences
-% (\C{edit\_seq}) and then their transformations
-% (\C{op\_transform})\footnote{
-%   Like Wagner-Fischer, established algorithms exist for computing tree
-%   edit distances~\cite{tree-diff}. An alternative naive method to
-%   merge a pair of trees is by merging the lists obtained from their
-%   in-order traversal, followed by reconstructing the tree.
-% }.
-
-{\bf Shopping List}. A shopping list is an
 

### 2017-02-27 paper/model.tex
edits

--- 2017-02-27 paper/model.tex
@@
 and returning the pixel at a given coordinate. The \C{merge} function
-and \C{deriving versioned} directive are explained below.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.5]{Figures/canvas-merging}
-
-\caption{Merging concurrent versions (\C{v1} and \C{v2} ) of a drawing
-canvas. \C{lca} is their common ancestor.}
-\label{fig:canvas-merging}
-\end{figure}
-
-\drawsome lets multiple users collaborate on a canvas that is
-conceptually shared among them. Under a shared-memory abstraction,
-there would be a single copy of the canvas that is updated
-concurrently by multiple clients; from the perspective of any single
-client, the canvas could change without any explicit
-intervention. \name ascribes functional semantics to sharing by
-letting each client work on its own version of the state (the tree
-data structure), later merging concurrent versions on-demand.  Thus,
-the primary artifact of the \name programming model is a versioned
-data structure in which different versions are managed by different
-clients.
+and \C{deriving versioned} directive are explained below.  \drawsome
+lets multiple users collaborate on a canvas that is conceptually
+shared among them. Under a shared-memory abstraction, there would be a
+single copy of the canvas that is updated concurrently by multiple
+clients; from the perspective of any single client, the canvas could
+change without any explicit intervention. \name ascribes functional
+semantics to sharing by letting each client work on its own version of
+the state (the tree data structure in this example), later merging
+concurrent versions on-demand.  Thus, the primary artifact of the
+\name programming model is a versioned data structure in which
+different versions are managed by different clients.
 
@@
 The \drawsome example demonstrates the utility of mergeable data types
-and \name programming model in building concurrent applications with
-conceptual sharing of state. The model lets concurrent computations be
-composed around any mergeable type. As exemplified by \drawsome,
-writing a three-way \C{merge} function is the only creative process in
-lifting an OCaml data type to a mergeable type. We now present few
-more examples of mergeable types, and isolate a recurring pattern in
-their merge functions that serves as a guide to write merge functions
-for more sophisticated types.
-
-\begin{figure}
-  \centering
-  \includegraphics[scale=0.4]{Figures/list-eg}
-
-  \caption{Lists of mergeable values are mergeable. }
-  \label{fig:list-eg}
-\end{figure}
+and the \name programming model in building concurrent applications
+with conceptual state sharing. The model lets concurrent computations
+be composed around any mergeable type. As exemplified by \drawsome,
+writing a three-way \C{merge} function is the only creative process
+involved to lift an OCaml data type to a mergeable type. We now
+present some more examples of mergeable types, and isolate a recurring
+pattern in their merge definitions that serves as a guide to write
+merge functions for more sophisticated types.
 
@@
 
-{\bf Lists}. List, as an abstract data type, supports three
+{\bf Lists}.  As an abstract data type, lists support three
 operations: \C{insert x i l}, to insert an element \C{x} at position
@@
 \C{i}'th position, and \C{subst i x l}, to substitute the element at
-\C{i}'th position with \C{x}. Lists of mergeable items are mergeable
-while preserving the following properties:
+the \C{i}'th position with \C{x}.  Lists of mergeable items are
+mergeable if they preserve the following properties:
 \begin{itemize}
@@
   before $y$ in one of the lists, and if both elements are present in
-  the merged list, then $x$ occurs before $y$ also in the merged list.
-  \item If an element $y$ substitutes $x$ in one list, then $y$ also
-  substitutes $x$ in the merged list (i.e., $x$ is absent and $y$ is
-  present), unless $x$ is deleted or substituted in the other list. In
-  the former case, deletion wins, and neither $x$ nor $y$ occur in the
-  merged list. In the latter case, if $x$ is substituted by a
-  different element $z$ in the other list, then the merged list
-  substitutes $x$ by a merge of $y$ and $z$, defined as per the merge
-  semantics of the list element type.
+  the merged list, then $x$ occurs before $y$ in the merged list.
+\item If an element $y$ is substituted for an element $x$ in one list,
+  then $y$ is also substituted for $x$ in the merged list (i.e., $x$
+  is absent and $y$ is present), unless $x$ is deleted or substituted
+  in the other list. In the former case, deletion wins, and neither
+  $x$ nor $y$ occur in the merged list. In the latter case, if $x$ is
+  substituted by a different element $z$ in the other list, then the
+  merged list substitutes $x$ by a merge of $y$ and $z$, defined as
+  per the merge semantics of the list element type.
 \end{itemize}
-Thus list is an example of a polymorphic mergeable data type that
-demonstrates how mergeability can be composed. The compositionality is
-reified in the the signature of a mergeable list library is shown in
+Thus, lists serve as an example of a polymorphic mergeable data type that
+demonstrates how mergeability can be composed.  Compositionality is
+reified in the the signature of a mergeable list library as shown in
 Fig.~\ref{fig:mlist-sig}; observe that the type of items in the list
@@
     applied on \C{v} to obtain \C{v'}. Such a sequence is called an
-  \emph{edit sequence}. The length of the sequence corresponds to the
-  standard notion of the \emph{edit distance} between the two lists,
-  which can be computed in polynomial time~\cite{wagner-fischer}.
-  \C{edit\_seq} modifies one such algorithm to also compute the edit
-  sequence (an \C{edit list}). The implementation specified in
-  Fig.~\ref{fig:mlist-sig} represents edits using the type \C{edit}.
-  Constructors \C{I}, \C{D}, and \C{S} stand for \C{insert},
-  \C{delete}, and \C{subst}, respectively\footnote{
-    Note that \C{I}, \C{D} and \C{S} are \emph{not} effects; no list
-    function generates them, and they are not exchanged between
-    concurrent threads. They are simply tags used by the list merge
-    operation for convenience.
-  }. The subst constructor also carries the \C{A.t} element that was
+    \emph{edit sequence}. The length of the sequence corresponds to
+    the standard notion of the \emph{edit distance} between the two
+    lists, which can be computed in polynomial
+    time~\cite{wagner-fischer}.  \C{edit\_seq} modifies one such
+    algorithm to also compute the edit sequence (an \C{edit
+      list}). The implementation specified in Fig.~\ref{fig:mlist-sig}
+    represents edits using the type \C{edit}.  Constructors \C{I},
+    \C{D}, and \C{S} stand for \C{insert}, \C{delete}, and \C{subst},
+    respectively.\footnote{ Note that \C{I}, \C{D} and \C{S} are
+      \emph{not} effects; no list function generates them, and they
+      are not exchanged between concurrent threads. They are simply
+      tags used by the list merge operation for convenience.  Indeed,
+      to better appreciate the virtues of mergeable types, we
+      encourage the reader to think about how one might implement a
+      replicated list with this functionality using an explicit effect
+      representation as described in Sec.~\ref{sec:intro}.}  The subst
+    constructor also carries the \C{A.t} element that was
     substituted. The element serves as the \C{lca} argument to
     \C{A.merge} in case of concurrent conflicting substitutions.
-  Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample list.
-  The sequence \C{[I(c,0); S(3,c,s)]} maps the list \C{[a;b;c]} to
-  \C{[c;a;b;s]} because \C{subst 3 s (insert c 0 [a;b;c]) =
-  [c;a;b;s]}.
+    Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample
+    list.  The sequence \C{[I(c,0); S(3,c,s)]} maps the list
+    \C{[a;b;c]} to \C{[c;a;b;s]} because \C{subst 3 s (insert c 0
+      [a;b;c]) = [c;a;b;s]}.
   \item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and
@@
 
-\GK{Somewhere here, I want to say:} To better appreciate the virtues
-of mergeable types, we encourage the reader to think about how one
-might implement the mergeable list type with the explicit effect
-representation described in Sec.~\ref{sec:intro}.
-
-\begin{figure}
-  \centering
-  \includegraphics[scale=0.4]{Figures/shopping-list}
-
-  \caption{Merging concurrent shopping lists}
-  \label{fig:shopping-list}
-\end{figure}
-
 A significant advantage of mergeable types is their composability. As
 demonstrated by \C{MList}, the merge semantics of a composed mergeable
-type is a composition of merge semantics of the individual types.  For
+type is a composition of the merge operations of its constituent types.  For
 instance, a mergeable list of integers can be obtained by
 instantiating \C{MList} for an \C{Int} type, where \C{Int.merge} is
-same as \C{Counter.merge} (Sec.~\ref{sec:intro}). 
+the same as \C{Counter.merge} (Sec.~\ref{sec:intro}). 
 An alternative instantiation of \C{MList} is for a quantity type
-(\C{Qty}) that, like \C{Int}, supports \C{add} and \C{subtract}
-operations, but defines its \C{merge} semantics a little differently:
+\begin{wrapfigure}{l}{.45\textwidth}
+  \centering
+  \includegraphics[scale=0.4]{Figures/list-eg}
+  \caption{Lists of mergeable values are mergeable. }
+  \label{fig:list-eg}
+\end{wrapfigure}
+(\C{Qty}) that supports \C{add} and \C{subtract}
+operations, but defines its \C{merge} opeation a little differently:
 \begin{ocaml}
@@
 however manifest under a concurrent execution, where both adopt
-different methods to reconcile concurrent updates.  is given
-below.
+different methods to reconcile concurrent updates.
 
@@
 {\bf Shopping List}. A non-trivial demonstration of composability of
-mergeable types is a shopping list application. The application
-allows its users to collaboratively build a grocery shopping
-list (listing items in decreasing ), by adding new items with a
-specified quantity, deleting existing items, and increasing or
-decreasing their quantity. An item (\C{Item.t}) is represented as a
-tuple of its name (a string) and the quantity (A \C{Qty.t}). Merge
-function for \C{Item.t} merges the quantities of items (using
-\C{Qty.merge}) with same names. Items with different names are
-considered distinct. A shopping list is a mergeable list of items
-(i.e., \C{MList(Item)}). \C{MList(Item).merge} automatically lifts the
-merge semantics for items to merge semantics for lists of items.
-Fig.~\ref{fig:shopping-list} illustrates. Here, two users, Alice and
-Bob, concurrently edit a shopping list whose initial contents are
-as milk and eggs. While both simultaneously update the quantity of
-eggs, Bob also removes milk and inserts candy. The merged shopping
-list contains eggs and candy, where the quantity of eggs is obtained
-by merging Alice's and Bob's quantity as per \C{Qty.merge} semantics.
+mergeable types is a shopping list application. The application allows
+its users to collaboratively build a grocery shopping list (listing
+items by decreasing quantity), by adding new items with a specified
+quantity, deleting existing items, and increasing or decreasing their
+quantity. An item (\C{Item.t}) is represented as a tuple of its name
+(a string) and the quantity (A \C{Qty.t}). The merge function for
+\begin{wrapfigure}{l}{.5\textwidth}
+  \centering
+  \includegraphics[scale=0.3]{Figures/shopping-list}
+  \caption{Merging concurrent shopping lists}
+  \label{fig:shopping-list}
+\end{wrapfigure}
+\C{Item.t} merges the quantities of items (using \C{Qty.merge}) having
+the same names. Items with different names are considered distinct. A
+shopping list is a mergeable list of items (i.e.,
+\C{MList(Item)}). \C{MList(Item).merge} automatically lifts the merge
+semantics for items to merge semantics for lists of items.
+Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
+users, Alice and Bob, concurrently edit a shopping list whose initial
+contents are \C{milk} and \C{eggs}. While both simultaneously update the
+quantity of eggs, Bob also removes milk and inserts candy. The merged
+shopping list contains eggs and candy, where the quantity of eggs is
+obtained by merging Alice's and Bob's quantity as per the definition
+of \C{Qty.merge}.
 

### 2017-02-27 paper/model.tex
Model.

--- 2017-02-27 paper/model.tex
@@
 however manifest under a concurrent execution, where both adopt
-different methods to reconcile concurrent updates.  is given
-below.
-
+different methods to reconcile concurrent updates. 
 % The mergeable list data type demonstrates a common pattern of writing
@@
 {\bf Shopping List}. A non-trivial demonstration of composability of
-mergeable types is a shopping list application. The application
-allows its users to collaboratively build a grocery shopping
-list (listing items in decreasing ), by adding new items with a
+mergeable types is a shopping list application. The application allows
+its users to collaboratively build a grocery shopping list (listing
+items in the order of their priority), by adding new items with a
 specified quantity, deleting existing items, and increasing or
@@
 Fig.~\ref{fig:shopping-list} illustrates. Here, two users, Alice and
-Bob, concurrently edit a shopping list whose initial contents are
-as milk and eggs. While both simultaneously update the quantity of
-eggs, Bob also removes milk and inserts candy. The merged shopping
-list contains eggs and candy, where the quantity of eggs is obtained
-by merging Alice's and Bob's quantity as per \C{Qty.merge} semantics.
+Bob, concurrently edit a shopping list whose initial contents are as
+milk and eggs. While both simultaneously update the quantity of eggs,
+Bob also removes milk and inserts candy. The merged shopping list
+contains eggs and candy, where the quantity of eggs is obtained by
+merging Alice's and Bob's quantity as per \C{Qty.merge} semantics.
 

### 2017-02-27 paper/mergeable-types.tex

@@
-\section{Mergeable Data Type}
+\subsection{Mergeable Data Type}
 
@@
 
-We now formally define what we mean by a mergeable type. The
-definition serves as a guide to write merge functions for functional
-data types, elevating them to the level of replicated data types.
-
-A functional data type library is a tuple consisting of a type ($t$),
-and a collection of functions ($f$, $g$, etc.) of type $t \rightarrow
-t$ that are \emph{primitive morphisms}. A morphism ($P$, $Q$, etc.) is
-either a primitive morphism, or an associative composition of
-morphisms ($P \circ Q \circ R$). An object $B:t$ is called a
-\emph{semantic successor} of $A:t$ (conversely, $A$ is called a
-\emph{semantic ancestor} of $B$) if and only if there exists a
-morphism $P$ such that $P(A) = B$.
+We now formally define a mergeable type.  A functional data type
+library is a tuple consisting of a type ($t$), and a collection of
+functions ($f$, $g$, etc.) of type $t \rightarrow t$ that are
+\emph{primitive morphisms}. A morphism ($P$, $Q$, etc.) is either a
+primitive morphism, or an associative composition of morphisms ($P
+\circ Q \circ R$). An object $B:t$ is called a \emph{semantic
+successor} of $A:t$ (conversely, $A$ is called a \emph{semantic
+ancestor} of $B$) if and only if there exists a morphism $P$ such that
+$P(A) = B$.
 
@@
 map respective concurrent versions to the same value. The
-operational semantics (Fig.~\ref{fig:opsem}) can be extended to this
-effect.
+\rulelabel{E-Pull-Wait} rule of the operational semantics
+(Fig.~\ref{fig:opsem}) can be extended to this effect:
+\begin{smathpar}
+\begin{array}{c}
+\RULE
+{
+  t\neq t' \spc
+% \under{H}{v' \mbleto v} \spc
+% \C{world}(H,t') \semsucceq \C{world}(H,t)\spc 
+  v' \not\preceq v \spc
+  (P',Q') = M(\C{lca}(H(t),H(t')), v, v') \spc
+  Q'(v) = P'(v')
+}
+{
+  (\pull)_t;H(t \mapsto (v,f)::m)(t' \mapsto (v',\_)::\_) ~\stepsto~
+  (\pull)_t;H[t \mapsto (v_m,\C{MERGE}\; H(t'))::(v,f)::m]
+}
+\end{array}
+\end{smathpar}
 
@@
 
-Merge functions for simple data types are easy to write. For instance,
-the merge logic for counter is simple enough that it does not require
-great effort to identify the $P$ and $Q$ that leda to the
-concurrent counter versions, and transform them into $P'$ and $Q'$
-such that the diagram in Fig.~\ref{fig:pushouts} commutes. For more
+Def.~\ref{def:mergeable-type} is immediately applicable to simple data
+types like counter to explain why they are mergeable. The merge logic
+for counter is simple enough that it does not require great effort to
+identify the $P$ and $Q$ that lead to the concurrent counter versions,
+and transform them into $P'$ and $Q'$ such that the diagram in
+Fig.~\ref{fig:pushouts} commutes. For more
 \begin{wrapfigure}{l}{.4\textwidth}
@@
 \end{wrapfigure}
-sophisticated data types, this process is more involved, and may require
-the programmer to perform non-trivial reasoning. We now present an
-alternative definition of a mergeable type that is stronger than
-Def.~\ref{def:mergeable-type}, but also serves as a recipe to build
-non-trivial merge functions.
+sophisticated data types, such as list, this process is more involved,
+and may require the programmer to perform non-trivial reasoning. We
+now present an alternative definition of a mergeable type that is
+stronger than Def.~\ref{def:mergeable-type}, but also serves as a
+recipe to build non-trivial merge functions, like that of a list.
 
@@
 named functions defined by a data type library (primitive morphisms),
-and their compositions (morphisms). A sequence (list) of primitive
-morphisms is written $P^*$ or $Q^*$. The composition of a sequence of
-primitive morphisms (written $\denot{P^*}$) is their right-to-left
-composition, i.e., $\C{fold\_left}\;(\lambda f.\lambda g. f \circ
-g)\; id\; P^*$, where $\C{fold\_left}$ has its usual OCaml type. The
-length of a sequence $S$ is written $\length{S}$.
+and their compositions (morphisms). A sequence of primitive morphisms
+is written $P^*$ or $Q^*$. The composition of a sequence of primitive
+morphisms (written $\denot{P^*}$) is their right-to-left composition,
+i.e., $\C{fold\_left}\;(\lambda f.\lambda g. f \circ g)\; id\; P^*$,
+where $\C{fold\_left}$ has its usual OCaml type. The length of a
+sequence $S$ is written $\length{S}$.
 
@@
 The definition asserts the existence of two functions for a mergeable
-type $t$. First, the function $W$, which computes the \emph{edit
-distance} between an object $A:t$ and its semantic successor $B:t$.
-The sequence of primitive morphisms it returns ($P^*$) is called an
-\emph{edit script}. Replaying the edit script on $A$ produces $B$
-(i.e., $\denot{P^*}(A) = B$). Second, the definition asserts the
-existence of a function $M$, which performs an \emph{operational
-transformation} of the primitive morphisms in an edit script. An
+type $t$. First, the function $W$, which (implicitly) computes the
+edit distance between an object $A:t$ and its semantic successor
+$B:t$.  The sequence of primitive morphisms it returns ($P^*$) is
+called an edit sequence. Applying the edit sequence on $A$ produces
+$B$ (i.e., $\denot{P^*}(A) = B$). Second, the definition asserts the
+existence of a function $M$, which performs an operational
+transformation of the primitive morphisms in an edit script. An
 operational transformation of a primitive morphism $P$ w.r.t a
@@
 object $A:t$, $P'$ has the same \emph{effect} on $Q(A)$ as $P$ has on
-$A$. We will now consider examples of mergeable types that reify the
-notions of edit scripts and operational transformations.
+$A$.
+
+Def.~\ref{def:mergeable-type-2} is immediately applicable to the list
+type, and lets us explain why it is a mergeable type.  \C{MList.edit\_seq}
+and \C{MList.op\_transform}, respectively, are the evidences for the
+existence of $W$ and $M$ functions.

### 2017-02-27 paper/meta.tex

@@
-To define the semantics of LCA, we formalize the intuitive notation of
-the ancestor relationship between versions of a legal branching
-history (i.e., a branching history generated by the rules in
-Fig.~\ref{fig:opsem}):
-
-\begin{definition} [\bfseries Ancestor]
-Version $v_1$ is a ancestor of version $v_2$ under a history
-$H$ (written $\under{H}{v_1 \preceq v_2}$) if and only if one of the
-following is true:
-\begin{itemize}
-  \item There exists a branch $b$ in $H$ (i.e., $\exists(t\in
-  dom(H)).\,H(t) = b$) in which $v_2$ immediately succeeds
-  $v_1$,
-  \item There exists a branch $b$ in $H$ that contains $(v_2, 
-  \C{FORK}\; (v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
-  \item There exists a branch $b$ in $H$ that contains
-  $(v_2, \C{MERGE}\;(v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
-  \item $v_1 = v_2$, or $v_1$ is transitively a ancestor of
-  $v_2$, i.e., $\exists v.~ \under{H}{v_1 \preceq v} \conj
-  \under{H}{v \preceq v}$ 
-\end{itemize}
-\end{definition}
-
-The ancestor relation is therefore a partial order with a greatest
-lower bound (the initial version).  Thus, for any two versions in a
-legal history, there exists at least one common ancestor.  The ancestor
-relationship among all common ancestors lets us define the notion of a
-least common ancestor (LCA):
-
-\begin{definition} [\bfseries Least Common Ancestor]
-Version $v$ is said to be a common ancestor of versions $v_1$ and
-$v_2$ under a history $H$ if and only if $\under{H}{v \preceq v_1}$
-and $\under{H}{v \preceq v_2}$. It is said to be the least common
-ancestor (LCA) of $v_1$ and $v_2$, iff there does not exist a $v'$
-such that $\under{H}{v' \preceq v_1}$ and $\under{H}{v' \preceq v_2}$
-and $\under{H}{v \preceq v'}$.
-\end{definition}
+\subsection{Properties of the LCA}
 
@@
 by two ``criss-cross'' merges between their respective branches
-resulting in there being two LCAs (5 and 4) for 0 and 3. Multiple LCAs
-can occur even without criss-cross merges, as demonstrated by
-Fig.~\ref{fig:external-lcas}. Concurrent versions with multiple LCAs
-do not lend themselves to three-way merging. This problem also arises
-in the context of source control systems, which employ \emph{ad hoc}
-mechanisms to pave the way for three-way merging.
-GitHub~\cite{github}, for instance, recursively merges LCAs to compute
-a virtual ancestor, which then serves as the LCA for merging
-concurrent versions. This method is demonstrated in the branching
-structure for a mergeable, replicated counter as shown in
-Fig.~\ref{fig:criss-cross-lcas}, where LCAs 5 and 4 of 0 and 3 are
-merged (with their LCA being 10) to generate -1 as the virtual LCA to
-merge 5 and 4. A major downside with this approach is that it makes no
-guarantees on the relationship between the virtual ancestor and its
-concurrent versions; the former may not even be a legal ancestor of
-the latter as per the semantics of the data type. For instance,
-suppose the integer type in Fig.~\ref{fig:criss-cross-lcas} represents
-a bank account balance, which is expected to disallows any activity on
-the account if the balance is ever known to be less than zero.  From
-the perspective of the library designer and its clients, there is no
-meaningful scenario in which versions 3 and 0 can emerge from -1,
-since the only transition allowed by the semantics from -1 is to
-itself.  Clearly, \emph{ad hoc} mechanisms like this are error-prone
-and difficult to apply in general.
+resulting in there being two LCAs (5 and 4) for 0 and 3. 
+% Multiple LCAs can occur even without criss-cross merges, as
+% demonstrated by Fig.~\ref{fig:external-lcas}. 
+Concurrent versions with multiple LCAs do not lend themselves to
+three-way merging. If such versions are latest on their respective
+branches, they render the branches unmergeable (since  \C{lca} is no
+longer a function). This problem also arises in the context of source
+control systems, which employ \emph{ad hoc} mechanisms to pave the way
+for three-way merging.  GitHub~\cite{github}, for instance,
+recursively merges LCAs to compute a virtual ancestor, which then
+serves as the LCA for merging concurrent versions. This method is
+demonstrated in the branching structure for a mergeable, replicated
+counter as shown in Fig.~\ref{fig:criss-cross-lcas}, where LCAs 5 and
+4 of 0 and 3 are merged (with their LCA being 10) to generate -1 as
+the virtual LCA to merge 5 and 4. A major downside with this approach
+is that it makes no guarantees on the relationship between the virtual
+ancestor and its concurrent versions; the former may not even be a
+legal ancestor of the latter as per the semantics of the data type.
+For instance, suppose the integer type in
+Fig.~\ref{fig:criss-cross-lcas} represents a bank account balance,
+which is expected to disallows any activity on the account if the
+balance is ever known to be less than zero.  From the perspective of
+the library designer and its clients, there is no meaningful scenario
+in which versions 3 and 0 can emerge from -1, since the only
+transition allowed by the semantics from -1 is to itself.  Clearly,
+\emph{ad hoc} mechanisms like this are error-prone and difficult to
+apply in general.
 
@@
 }
-\hfill
-\subcaptionbox[] {\small
-  In this example, versions $v_{13}$ and $v_{44}$ have two LCAs
-  ($v_{22}$ and $v_{32}$)  despite there not being any previous merges
-  between their respective branches.
-  \label{fig:external-lcas}
-} [0.47\columnwidth] {
-  \includegraphics[scale=0.7]{Figures/2-external-LCAs}
-}
+% \hfill
+% \subcaptionbox[] {\small
+%   In this example, versions $v_{13}$ and $v_{44}$ have two LCAs
+%   ($v_{22}$ and $v_{32}$)  despite there not being any previous merges
+%   between their respective branches.
+%   \label{fig:external-lcas}
+% } [0.47\columnwidth] {
+%   \includegraphics[scale=0.7]{Figures/2-external-LCAs}
+% }
 \caption{Examples where merging versions have more than one LCA}
@@
 manifest it in a way that it deems fit. In particular, \name solves
-the problem of multiple LCAs either by constraining the branching
-stucture to prevent their existence, or by merging branches to ensure
-that they have a new unique LCA.  Preention is done for
-criss-cross merges by the operational semantics, which only allows
-merging versions that are the latest on their respective branches. For
-instance, suppose $v_{11}$ and $v_{21}$ are the latest versions on branches
-$b_1$ and $b_2$, respectively. Merging $b_2$ into $b_1$ entails
-merging $v_{21}$ into $v_{11}$ to generate version $v_{12}$ on $b_1$.
-Now, merging $b_1$ into $b_2$ involves merging $v_{12}$, the latest
-version on $b_1$, into $v_{21}$, but not $v_{11}$ into $v_{21}$, thus
-preempting a criss-cross branching structure.
+the problem of criss-cross merges by constraining the branching
+stucture to prevent their existence. As captured in the operational
+semantics, this is done by restricting the merge operation only to the
+tips (latest versions) of the branches. For instance, suppose $v_{11}$
+and $v_{21}$ are the latest versions on branches $b_1$ and $b_2$,
+respectively. Merging $b_2$ into $b_1$ entails merging $v_{21}$ into
+$v_{11}$ to generate version $v_{12}$ on $b_1$.  Now, merging $b_1$
+into $b_2$ involves merging $v_{12}$, the latest version on $b_1$,
+into $v_{21}$, but not $v_{11}$ into $v_{21}$, thus preempting a
+criss-cross branching structure.
 % In other words, a criss-cross branching structure is prevented due
@@
 
-Next, for branching structures where a pair of branches (i.e., their
-respective latest versions) have multiple externally-located LCAs, we
-suitably extend the branching structure to generate a new single LCA.
- For instance, consider
-the example in Fig.~\ref{fig:external-lcas}, where (latest versions
-on) branches $b_1$ and $b_4$ have two externally-located LCAs
-($v_{22}$ and $v_{32}$), hence cannot be merged. Mergeability can
-however be reclaimed by extending the branching structure as shown in
-Fig.~\ref{fig:legal-extension-1}. The extension is done by first
-merging $b_3$ into $b_2$ to create version $v_{23}$ on $b_2$, then
-merging $b_2$ into $b_1$ to create version $v_{14}$ on $b_1$, and
-finally merging $b_2$ into $b_4$ to create version $v_{45}$ on
-$b_4$. Observe that $b_1$ and $b_4$ now have a single LCA ($v_{23}$)
-hence are now mergeable. Mergeability could also have been reclaimed
-by adopting an alternative extension to the branching structure as
-shown in Fig.~\ref{fig:legal-extension-2}.
+% Next, for branching structures where a pair of branches (i.e., their
+% respective latest versions) have multiple externally-located LCAs, we
+% suitably extend the branching structure to generate a new single LCA.
+%  For instance, consider
+% the example in Fig.~\ref{fig:external-lcas}, where (latest versions
+% on) branches $b_1$ and $b_4$ have two externally-located LCAs
+% ($v_{22}$ and $v_{32}$), hence cannot be merged. Mergeability can
+% however be reclaimed by extending the branching structure as shown in
+% Fig.~\ref{fig:legal-extension-1}. The extension is done by first
+% merging $b_3$ into $b_2$ to create version $v_{23}$ on $b_2$, then
+% merging $b_2$ into $b_1$ to create version $v_{14}$ on $b_1$, and
+% finally merging $b_2$ into $b_4$ to create version $v_{45}$ on
+% $b_4$. Observe that $b_1$ and $b_4$ now have a single LCA ($v_{23}$)
+% hence are now mergeable. Mergeability could also have been reclaimed
+% by adopting an alternative extension to the branching structure as
+% shown in Fig.~\ref{fig:legal-extension-2}.
 
@@
 \subcaptionbox[] {\small
-  $b_3$ is merged into $b_2$ to create $v_{23}$ here.
+  In this example, versions $v_{13}$ and $v_{44}$ have two LCAs
+  ($v_{22}$ and $v_{32}$) rendering their respective branches
+  unmergeable.
+  \label{fig:external-lcas}
+} [0.47\columnwidth] {
+  \includegraphics[scale=0.7]{Figures/2-external-LCAs}
+}
+\hfill
+\subcaptionbox[] {\small
+  However, execution can still progress to the depicted state, where
+  $b_1$ and $b_4$ are again mergeable.
   \label{fig:legal-extension-1}
@@
 }
-\hfill
-\subcaptionbox[] {\small
-  $b_2$ is merged into $b_3$ to create $v_{33}$ here.
-  \label{fig:legal-extension-2}
-} [0.47\columnwidth] {
-  \includegraphics[scale=0.8]{Figures/legal-extension-2}
-}
-\caption{Legal extensions of the branching structure in
-  Fig.~\ref{fig:external-lcas} that reclaim the mergeability of $b_1$
-  and $b_4$}
+% \hfill
+% \subcaptionbox[] {\small
+%   $b_2$ is merged into $b_3$ to create $v_{33}$ here.
+%   \label{fig:legal-extension-2}
+% } [0.47\columnwidth] {
+%   \includegraphics[scale=0.8]{Figures/legal-extension-2}
+% }
+\caption{Multiple LCAs without criss-cross merges do not threaten
+the progress guarantee.}
 \label{fig:legal-extensions}
@@
 
-Note that the extension strategy described above is possible only
-because operational semantics imposes no obligation to merge any two
-concurrent versions. The high-level purpose of merging branches in
-\name is to propagate the data across the branches, which can be done
-by merging any two concurrent versions that contain the data; the
-semantics lets us pick these versions. While mergeability of branches
-with multiple LCAs can be reclaimed by actively extending the
-branching structure as described above, the semantics is
-\emph{self-correcting} in the sense that a fair execution reclaims the
-mergeability of any two branches without any intervention. This is
-possible because the system preempts criss-cross merges, thus letting
-the semantics make progress from any state. The property is formalized
-by the following theorem\footnote{The more precise statement of the
-theorem, and its proof are included in the appendix}:
+There can be other scenarios that lead to multiple LCAs, such as the
+one shown in Fig.~\ref{fig:external-lcas}. However, preemption of
+criss-cross merges is sufficient to make progress despite multiple
+LCAs. For instance, the branching history in
+Fig.~\ref{fig:external-lcas}, where $b_1$ and $b_4$ cannot merge due
+to multiple LCAs, can still make progress to the history shown in
+~\ref{fig:legal-extension-1}, where $b_1$ and $b_4$ are now mergeable.
+We formalize the progress guarantee in the following theorem, whose
+proof is included in the appendix.
 
-\begin{theorem} [\bfseries Semantics Self-Correcting]
-??
+\begin{theorem} [\bfseries Progress]
+In a legal branching history $H$ produced by the operational
+semantics, there exist at least two branches that have a unique LCA,
+hence mergeable.
 \end{theorem}
-
-The \rulelabel{E-Pull-Wait} uses the function \C{lca} to compute the
-LCA of (the latest versions on) a pair of branches. The definition of
-the function is standard, hence not discussed.

### 2017-02-27 paper/model.tex

--- 2017-02-27 paper/model.tex
@@
 \section{Programming Model}
+\label{sec:model}
 
@@
 \begin{ocaml}
-let merge lca v1 v2 = if v1 = lca then v2 else (if v2 = lca then v1 else max v1 v2)
+let merge lca v1 v2 = if v1 = lca then v2 
+    else (if v2 = lca then v1 else max v1 v2)
 \end{ocaml}
@@
 its users to collaboratively build a grocery shopping list (listing
-items by decreasing quantity), by adding new items with a specified
-quantity, deleting existing items, and increasing or decreasing their
-quantity. An item (\C{Item.t}) is represented as a tuple of its name
-(a string) and the quantity (A \C{Qty.t}). The merge function for
+items in the decreasing order of their priority), by adding new items
+with a specified quantity, deleting existing items, and increasing or
+decreasing their quantity. An item (\C{Item.t}) is represented as a
+tuple of its name (a string) and the quantity (A \C{Qty.t}). The merge
+function for
 \begin{wrapfigure}{l}{.5\textwidth}

### 2017-02-27 paper/semantics.tex

@@
-\section{Operational Semantics}
+\section{Formalization}
+
+\subsection{Operational Semantics}
 
@@
 \C{lca} that computes the \emph{least common ancestor} (LCA) for the
-latest versions on the given pair of branches. The semantics of LCA is
-discussed below. The \rulelabel{E-Pull-Wait} and \rulelabel{E-Pull}
-rules thus let a thread synchronize with other distributed threads
-manipulating versions of replicas in multiple steps before returning
-the result of the \C{pull}. Since \C{sync} is a composition of
-\C{push} and \C{pull}, its behavior can be explained thus: \C{sync}
-pushes the given value onto the current (local) branch, merges a
-(possibly empty) subset of concurrent branches into the local branch,
-and returns the result.  (The operation is not guaranteed to
+latest versions on the given pair of branches. The formal definition
+of LCA is given below. The \rulelabel{E-Pull-Wait} and
+\rulelabel{E-Pull} rules thus let a thread synchronize with other
+distributed threads manipulating versions of replicas in multiple
+steps before returning the result of the \C{pull}. Since \C{sync} is a
+composition of \C{push} and \C{pull}, its behavior can be explained
+thus: \C{sync} pushes the given value onto the current (local) branch,
+merges a (possibly empty) subset of concurrent branches into the local
+branch, and returns the result.  (The operation is not guaranteed to
 synchronize with all concurrent branches because not all such branches
@@
 
+To define LCA, we formalize the intuitive notation of the ancestor
+relationship between versions of a legal branching history (i.e., a
+branching history generated by the rules in Fig.~\ref{fig:opsem}):
+
+\begin{definition} [\bfseries Ancestor]
+Version $v_1$ is a ancestor of version $v_2$ under a history
+$H$ (written $\under{H}{v_1 \preceq v_2}$) if and only if one of the
+following is true:
+\begin{itemize}
+  \item There exists a branch $b$ in $H$ (i.e., $\exists(t\in
+  dom(H)).\,H(t) = b$) in which $v_2$ immediately succeeds
+  $v_1$,
+  \item There exists a branch $b$ in $H$ that contains $(v_2, 
+  \C{FORK}\; (v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
+  \item There exists a branch $b$ in $H$ that contains
+  $(v_2, \C{MERGE}\;(v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
+  \item $v_1 = v_2$, or $v_1$ is transitively a ancestor of
+  $v_2$, i.e., $\exists v.~ \under{H}{v_1 \preceq v} \conj
+  \under{H}{v \preceq v}$ 
+\end{itemize}
+\end{definition}
+
+The ancestor relation is therefore a partial order with a greatest
+lower bound (the initial version).  Thus, for any two versions in a
+legal history, there exists at least one common ancestor.  The ancestor
+relationship among all common ancestors lets us define the notion of a
+least common ancestor (LCA):
+
+\begin{definition} [\bfseries Least Common Ancestor]
+Version $v$ is said to be a common ancestor of versions $v_1$ and
+$v_2$ under a history $H$ if and only if $\under{H}{v \preceq v_1}$
+and $\under{H}{v \preceq v_2}$. It is said to be the least common
+ancestor (LCA) of $v_1$ and $v_2$, iff there does not exist a $v'$
+such that $\under{H}{v' \preceq v_1}$ and $\under{H}{v' \preceq v_2}$
+and $\under{H}{v \preceq v'}$.
+\end{definition}
+
+The function \C{lca} used in the premise of \rulelabel{E-Pull-Wait}
+computes the LCA of two branches. The function is in fact a partial
+function because the definition of LCA doesn't guarantee its
+uniqueness for a pair of branches. Such nuances of LCA, and the
+guarantees provided by the system notwithstanding these nuances, are
+discussed in the subsection~\ref{sec:meta}.
+
+\input{mergeable-types}
+
 \input{meta.tex}
-
-\input{system-model}

### 2017-02-27 paper/evaluation.tex
edits

\section{Evaluation}
\label{sec:evaluation}

In this section, we present an evaluation of \name programming model with the
aim of assesing its practical utility for describing geo-distributed eventually
consistent applications. In particular, we

To this end, we have implemented a collaborative editing application that
simulates concurrent editing of the same document by several authors. This
application is not unlike Google Docs, but also differs from it by the fact
that we do not need a central server that coordinates the edits. Instead, edits
from each author is asynchronously sent to other authors. The shared document
itself is represented by a mergeable lists of strings
(Section~\ref{sec:model}), and hence, the remote edits can be correctly
reconciled into the local document. We resolve conflicts on concurrent
substitutions at the same position (such as two authors concurrently editing
\C{abc} to \C{xbc} and \C{ybc}, respectively) by choosing the character with
the smaller ASCII code.

### 2017-02-27 paper/model.tex
edits

@@
 
-{\bf Lists}.  As an abstract data type, lists support three
+{\bf Lists}.  \KC{Why don't we call these ropes?} As an abstract data type, lists support three
 operations: \C{insert x i l}, to insert an element \C{x} at position
@@
 \begin{itemize}
-  \item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and
-    computes the shortest sequence of list operations that need to be
-    applied on \C{v} to obtain \C{v'}. Such a sequence is called an
-    \emph{edit sequence}. The length of the sequence corresponds to
-    the standard notion of the \emph{edit distance} between the two
-    lists, which can be computed in polynomial
-    time~\cite{wagner-fischer}.  \C{edit\_seq} modifies one such
-    algorithm to also compute the edit sequence (an \C{edit
-      list}). The implementation specified in Fig.~\ref{fig:mlist-sig}
-    represents edits using the type \C{edit}.  Constructors \C{I},
-    \C{D}, and \C{S} stand for \C{insert}, \C{delete}, and \C{subst},
-    respectively.\footnote{ Note that \C{I}, \C{D} and \C{S} are
-      \emph{not} effects; no list function generates them, and they
-      are not exchanged between concurrent threads. They are simply
-      tags used by the list merge operation for convenience.  Indeed,
-      to better appreciate the virtues of mergeable types, we
-      encourage the reader to think about how one might implement a
-      replicated list with this functionality using an explicit effect
-      representation as described in Sec.~\ref{sec:intro}.}  The subst
-    constructor also carries the \C{A.t} element that was
-    substituted. The element serves as the \C{lca} argument to
+	\item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and computes
+		the shortest sequence of list operations that need to be applied on \C{v}
+		to obtain \C{v'}. Such a sequence is called an \emph{edit sequence}. The
+		length of the sequence corresponds to the standard notion of the \emph{edit
+		distance} between the two lists, which can be computed in polynomial
+		time~\cite{wagner-fischer}.  \C{edit\_seq} modifies one such algorithm to
+		also compute the edit sequence (an \C{edit list}). The implementation
+		specified in Fig.~\ref{fig:mlist-sig} represents edits using the type
+		\C{edit}.  Constructors \C{I}, \C{D}, and \C{S} stand for \C{insert},
+		\C{delete}, and \C{subst}, respectively.\footnote{ Note that \C{I}, \C{D}
+		and \C{S} are \emph{not} effects; no list function generates them, and
+		they are not exchanged between concurrent threads. They are simply tags
+		used by the list merge operation for convenience.  Indeed, to better
+		appreciate the virtues of mergeable types, we encourage the reader to
+		think about how one might implement a replicated list with this
+		functionality using an explicit effect representation as described in
+		Sec.~\ref{sec:intro}.}  The subst constructor also carries the \C{A.t}
+		element that was substituted. The element serves as the \C{lca} argument to
 		\C{A.merge} in case of concurrent conflicting substitutions.
-    Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample
-    list.  The sequence \C{[I(c,0); S(3,c,s)]} maps the list
-    \C{[a;b;c]} to \C{[c;a;b;s]} because \C{subst 3 s (insert c 0
-      [a;b;c]) = [c;a;b;s]}.
-  \item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and
-  $s_2$, that map a list $v$ to two different lists, $v_1$ and $v_2$
-  (e.g., Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such
-  that $s_1'$ has the same effect on $v_2$ as $s_1$ had on $v$.  For
-  instance, in Fig.~\ref{list-eg}, let $s_1$ be the edit sequence
-  \C{[D(1); S(1,c,d)]}, which maps the list \C{[a;b;c]} ($v$) to
-  \C{[a;d]} ($v_1$). The \C{D} edit deletes the first element of $v$.
-  However, if $s_1$ is to be applied on the list \C{[c;a;b;s]}
-  ($v_2$), which has a new element at position 0, the \C{D} edit must
-  delete the the second element from $v_2$, if it were to have the
-  same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
-  transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The
-  function \C{op\_transform} (Fig.~\ref{fig:mlist-sig}) computes such
-  \emph{operational transformations} for an edit sequence $s_1$ w.r.t
-  another edit sequence $s_2$. A notable transformation rule is for a
-  substitution w.r.t another substitution at the same position. Since
-  the substituting items could be different, the function relies on
-  the item merge function (\C{A.merge}) to merge them into a single
-  item. For instance, in Fig.~\ref{fig:list-eg}, there are
-  simultaneous substitutions to the item \C{c} in $s_1$ and $s_2$;
-  while $s_1$ substitutes it with \C{d}, $s_2$ does it with \C{s}. The
-  merged list therefore substitutes \C{c} with the merge of \C{d} and
-  \C{s}, defined as per the merge semantics of \C{A.t}.
+		Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample list.  The
+		sequence \C{[I(c,0); S(3,c,s)]} maps the list \C{[a;b;c]} to \C{[c;a;b;s]}
+		because \C{subst 3 s (insert c 0 [a;b;c]) = [c;a;b;s]}.
+	\item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and $s_2$,
+		that map a list $v$ to two different lists, $v_1$ and $v_2$ (e.g.,
+		Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such that $s_1'$
+		has the same effect on $v_2$ as $s_1$ had on $v$.  For instance, in
+		Fig.~\ref{list-eg}, let $s_1$ be the edit sequence \C{[D(1); S(1,c,d)]},
+		which maps the list \C{[a;b;c]} ($v$) to \C{[a;d]} ($v_1$). The \C{D} edit
+		deletes the first element of $v$. However, if $s_1$ is to be applied on the
+		list \C{[c;a;b;s]} ($v_2$), which has a new element at position 0, the
+		\C{D} edit must delete the the second element from $v_2$, if it were to
+		have the same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
+		transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The function
+		\C{op\_transform} (Fig.~\ref{fig:mlist-sig}) computes such
+		\emph{operational transformations} for an edit sequence $s_1$ w.r.t another
+		edit sequence $s_2$. A notable transformation rule is for a substitution
+		w.r.t another substitution at the same position. Since the substituting
+		items could be different, the function relies on the item merge function
+		(\C{A.merge}) to merge them into a single item. For instance, in
+		Fig.~\ref{fig:list-eg}, there are simultaneous substitutions to the item
+		\C{c} in $s_1$ and $s_2$; while $s_1$ substitutes it with \C{d}, $s_2$ does
+		it with \C{s}. The merged list therefore substitutes \C{c} with the merge
+		of \C{d} and \C{s}, defined as per the merge semantics of \C{A.t}.
 \end{itemize}

### 2017-02-27 paper/model.tex
Added merge tree code
Sec. 2 -> Sec. 2 + Sec. 3

--- 2017-02-27 paper/model.tex
@@
 versions.
-
-The \drawsome example demonstrates the utility of mergeable data types
-and the \name programming model in building concurrent applications
-with conceptual state sharing. The model lets concurrent computations
-be composed around any mergeable type. As exemplified by \drawsome,
-writing a three-way \C{merge} function is the only creative process
-involved to lift an OCaml data type to a mergeable type. We now
-present some more examples of mergeable types, and isolate a recurring
-pattern in their merge definitions that serves as a guide to write
-merge functions for more sophisticated types.
-
-\input{list-sig}
-
-{\bf Lists}.  As an abstract data type, lists support three
-operations: \C{insert x i l}, to insert an element \C{x} at position
-\C{i} in the list \C{l}, \C{delete i l}, to delete the element at
-\C{i}'th position, and \C{subst i x l}, to substitute the element at
-the \C{i}'th position with \C{x}.  Lists of mergeable items are
-mergeable if they preserve the following properties:
-\begin{itemize}
-  \item Elements present in both the merging lists will be present in
-  the final list. No element is deleted unless it is deleted in at
-  least one list.
-  \item An element deleted in one list will not reappear in the merged
-  list (unlike Amazon's replicated shopping cart~\cite{dynamo}).
-  \item The order of elements is retained. That is, if $x$ occurs
-  before $y$ in one of the lists, and if both elements are present in
-  the merged list, then $x$ occurs before $y$ in the merged list.
-\item If an element $y$ is substituted for an element $x$ in one list,
-  then $y$ is also substituted for $x$ in the merged list (i.e., $x$
-  is absent and $y$ is present), unless $x$ is deleted or substituted
-  in the other list. In the former case, deletion wins, and neither
-  $x$ nor $y$ occur in the merged list. In the latter case, if $x$ is
-  substituted by a different element $z$ in the other list, then the
-  merged list substitutes $x$ by a merge of $y$ and $z$, defined as
-  per the merge semantics of the list element type.
-\end{itemize}
-Thus, lists serve as an example of a polymorphic mergeable data type that
-demonstrates how mergeability can be composed.  Compositionality is
-reified in the the signature of a mergeable list library as shown in
-Fig.~\ref{fig:mlist-sig}; observe that the type of items in the list
-(\C{A.t}) is also required to be mergeable (\C{A.merge}).
-
-The merge operation for lists is composed of two separate functions,
-\C{edit\_seq} and \C{op\_transform}. Both these functions have been
-implemented in less than 50 lines of OCaml included in the appendix.
-The functions are described below.
-\begin{itemize}
-  \item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and
-    computes the shortest sequence of list operations that need to be
-    applied on \C{v} to obtain \C{v'}. Such a sequence is called an
-    \emph{edit sequence}. The length of the sequence corresponds to
-    the standard notion of the \emph{edit distance} between the two
-    lists, which can be computed in polynomial
-    time~\cite{wagner-fischer}.  \C{edit\_seq} modifies one such
-    algorithm to also compute the edit sequence (an \C{edit
-      list}). The implementation specified in Fig.~\ref{fig:mlist-sig}
-    represents edits using the type \C{edit}.  Constructors \C{I},
-    \C{D}, and \C{S} stand for \C{insert}, \C{delete}, and \C{subst},
-    respectively.\footnote{ Note that \C{I}, \C{D} and \C{S} are
-      \emph{not} effects; no list function generates them, and they
-      are not exchanged between concurrent threads. They are simply
-      tags used by the list merge operation for convenience.  Indeed,
-      to better appreciate the virtues of mergeable types, we
-      encourage the reader to think about how one might implement a
-      replicated list with this functionality using an explicit effect
-      representation as described in Sec.~\ref{sec:intro}.}  The subst
-    constructor also carries the \C{A.t} element that was
-    substituted. The element serves as the \C{lca} argument to
-    \C{A.merge} in case of concurrent conflicting substitutions.
-    Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample
-    list.  The sequence \C{[I(c,0); S(3,c,s)]} maps the list
-    \C{[a;b;c]} to \C{[c;a;b;s]} because \C{subst 3 s (insert c 0
-      [a;b;c]) = [c;a;b;s]}.
-  \item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and
-  $s_2$, that map a list $v$ to two different lists, $v_1$ and $v_2$
-  (e.g., Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such
-  that $s_1'$ has the same effect on $v_2$ as $s_1$ had on $v$.  For
-  instance, in Fig.~\ref{list-eg}, let $s_1$ be the edit sequence
-  \C{[D(1); S(1,c,d)]}, which maps the list \C{[a;b;c]} ($v$) to
-  \C{[a;d]} ($v_1$). The \C{D} edit deletes the first element of $v$.
-  However, if $s_1$ is to be applied on the list \C{[c;a;b;s]}
-  ($v_2$), which has a new element at position 0, the \C{D} edit must
-  delete the the second element from $v_2$, if it were to have the
-  same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
-  transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The
-  function \C{op\_transform} (Fig.~\ref{fig:mlist-sig}) computes such
-  \emph{operational transformations} for an edit sequence $s_1$ w.r.t
-  another edit sequence $s_2$. A notable transformation rule is for a
-  substitution w.r.t another substitution at the same position. Since
-  the substituting items could be different, the function relies on
-  the item merge function (\C{A.merge}) to merge them into a single
-  item. For instance, in Fig.~\ref{fig:list-eg}, there are
-  simultaneous substitutions to the item \C{c} in $s_1$ and $s_2$;
-  while $s_1$ substitutes it with \C{d}, $s_2$ does it with \C{s}. The
-  merged list therefore substitutes \C{c} with the merge of \C{d} and
-  \C{s}, defined as per the merge semantics of \C{A.t}.
-\end{itemize}
-The \C{MList.merge} can now be defined as a function that first
-computes edit sequences, $s_1$ and $s_2$, for the two concurrent
-lists, $v_1$ and $v_2$ w.r.t their common ancestor $v$, transforms one
-of them (say, $s_1$) w.r.t $s_2$ to obtain $s_1'$, and finally applies
-(the operations defined by) $s_1'$ to $v_2$ to obtain the merged list.
-
-A significant advantage of mergeable types is their composability. As
-demonstrated by \C{MList}, the merge semantics of a composed mergeable
-type is a composition of the merge operations of its constituent types.  For
-instance, a mergeable list of integers can be obtained by
-instantiating \C{MList} for an \C{Int} type, where \C{Int.merge} is
-the same as \C{Counter.merge} (Sec.~\ref{sec:intro}). 
-An alternative instantiation of \C{MList} is for a quantity type
-\begin{wrapfigure}{l}{.45\textwidth}
-  \centering
-  \includegraphics[scale=0.4]{Figures/list-eg}
-  \caption{Lists of mergeable values are mergeable. }
-  \label{fig:list-eg}
-\end{wrapfigure}
-(\C{Qty}) that supports \C{add} and \C{subtract}
-operations, but defines its \C{merge} opeation a little differently:
-\begin{ocaml}
-let merge lca v1 v2 = if v1 = lca then v2 
-    else (if v2 = lca then v1 else max v1 v2)
-\end{ocaml}
-\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has
-the same semantics under a sequential execution. Their differences
-however manifest under a concurrent execution, where both adopt
-different methods to reconcile concurrent updates.
-
-% The mergeable list data type demonstrates a common pattern of writing
-% \C{merge} functions applicable to a range of functional data
-% structures. For instance, a binary tree (with rotations) can be made
-% mergeable in the same way as list, i.e., by computing edit sequences
-% (\C{edit\_seq}) and then their transformations
-% (\C{op\_transform})\footnote{
-%   Like Wagner-Fischer, established algorithms exist for computing tree
-%   edit distances~\cite{tree-diff}. An alternative naive method to
-%   merge a pair of trees is by merging the lists obtained from their
-%   in-order traversal, followed by reconstructing the tree.
-% }.
-
-{\bf Shopping List}. A non-trivial demonstration of composability of
-mergeable types is a shopping list application. The application allows
-its users to collaboratively build a grocery shopping list (listing
-items in the decreasing order of their priority), by adding new items
-with a specified quantity, deleting existing items, and increasing or
-decreasing their quantity. An item (\C{Item.t}) is represented as a
-tuple of its name (a string) and the quantity (A \C{Qty.t}). The merge
-function for
-\begin{wrapfigure}{l}{.5\textwidth}
-  \centering
-  \includegraphics[scale=0.3]{Figures/shopping-list}
-  \caption{Merging concurrent shopping lists}
-  \label{fig:shopping-list}
-\end{wrapfigure}
-\C{Item.t} merges the quantities of items (using \C{Qty.merge}) having
-the same names. Items with different names are considered distinct. A
-shopping list is a mergeable list of items (i.e.,
-\C{MList(Item)}). \C{MList(Item).merge} automatically lifts the merge
-semantics for items to merge semantics for lists of items.
-Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
-users, Alice and Bob, concurrently edit a shopping list whose initial
-contents are \C{milk} and \C{eggs}. While both simultaneously update the
-quantity of eggs, Bob also removes milk and inserts candy. The merged
-shopping list contains eggs and candy, where the quantity of eggs is
-obtained by merging Alice's and Bob's quantity as per the definition
-of \C{Qty.merge}.
-
-The examples described above make certain assumptions about the model
-and the underlying system, such as the existence of a single least
-common ancestor (LCA) for any pair of versions, the ability to access
-a previous version on any branch, and mergeability of any two
-concurrent versions. Enforcing these guarantees in a fully
-decentralized distributed setting requires addressing non-trivial
-theoretical challenges. These, and other details that contribute to
-the practicality of our model, such as containing the complexity of
-merges, are discussed in subsequent sections.

### 2017-02-27 paper/examples.tex
Examples file.

\section{Mergeable Types}

The \drawsome example demonstrates the utility of the \name
programming model in building concurrent applications with conceptual
state sharing. The model lets concurrent computations be composed
around any mergeable type. We now present a series of examples that
demonstrate how mergeable types are built: either by defining merge
functions for OCaml data types, or by composing existing mergeable
types.

{\bf Ropes}. A rope is a data structure that faciliates efficient
implementations of operations such as concatenation, splitting, lookup
and insertion at a random position etc., for a list of values. The
values are usually characters; their lists being strings. The data
structure is a binary tree, whose leaf nodes store substrings of a
larger string such that their left-to-right (in-order) concatenation
yeilds the larger string. The idea is to facilitate efficient lookups
and insertions at random positions in the string by storing
position indexes in the internal nodes of the tree.

Fig.~\ref{fig:rope-sig} shows a rope interface in OCaml. The type
of the list of values is left abstract (\C{A.t}); it can be
instantiated with any module that supports \C{concat} and \C{split}
operations. The interface shown in Fig.~\ref{fig:rope-sig} is that of
a mergeable rope (\C{MRope});

\input{list-sig}

{\bf Lists}. As an abstract data type, lists support three
operations: \C{insert x i l}, to insert an element \C{x} at position
\C{i} in the list \C{l}, \C{delete i l}, to delete the element at
\C{i}'th position, and \C{subst i x l}, to substitute the element at
the \C{i}'th position with \C{x}.  Lists of mergeable items are
mergeable if they preserve the following properties:
\begin{itemize}
  \item Elements present in both the merging lists will be present in
  the final list. No element is deleted unless it is deleted in at
  least one list.
  \item An element deleted in one list will not reappear in the merged
  list (unlike Amazon's replicated shopping cart~\cite{dynamo}).
  \item The order of elements is retained. That is, if $x$ occurs
  before $y$ in one of the lists, and if both elements are present in
  the merged list, then $x$ occurs before $y$ in the merged list.
\item If an element $y$ is substituted for an element $x$ in one list,
  then $y$ is also substituted for $x$ in the merged list (i.e., $x$
  is absent and $y$ is present), unless $x$ is deleted or substituted
  in the other list. In the former case, deletion wins, and neither
  $x$ nor $y$ occur in the merged list. In the latter case, if $x$ is
  substituted by a different element $z$ in the other list, then the
  merged list substitutes $x$ by a merge of $y$ and $z$, defined as
  per the merge semantics of the list element type.
\end{itemize}
Thus, lists serve as an example of a polymorphic mergeable data type that
demonstrates how mergeability can be composed.  Compositionality is
reified in the the signature of a mergeable list library as shown in
Fig.~\ref{fig:mlist-sig}; observe that the type of items in the list
(\C{A.t}) is also required to be mergeable (\C{A.merge}).

The merge operation for lists is composed of two separate functions,
\C{edit\_seq} and \C{op\_transform}. Both these functions have been
implemented in less than 50 lines of OCaml included in the appendix.
The functions are described below.
\begin{itemize}
	\item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and computes
		the shortest sequence of list operations that need to be applied on \C{v}
		to obtain \C{v'}. Such a sequence is called an \emph{edit sequence}. The
		length of the sequence corresponds to the standard notion of the \emph{edit
		distance} between the two lists, which can be computed in polynomial
		time~\cite{wagner-fischer}.  \C{edit\_seq} modifies one such algorithm to
		also compute the edit sequence (an \C{edit list}). The implementation
		specified in Fig.~\ref{fig:mlist-sig} represents edits using the type
		\C{edit}.  Constructors \C{I}, \C{D}, and \C{S} stand for \C{insert},
		\C{delete}, and \C{subst}, respectively.\footnote{ Note that \C{I}, \C{D}
		and \C{S} are \emph{not} effects; no list function generates them, and
		they are not exchanged between concurrent threads. They are simply tags
		used by the list merge operation for convenience.  Indeed, to better
		appreciate the virtues of mergeable types, we encourage the reader to
		think about how one might implement a replicated list with this
		functionality using an explicit effect representation as described in
		Sec.~\ref{sec:intro}.}  The subst constructor also carries the \C{A.t}
		element that was substituted. The element serves as the \C{lca} argument to
		\C{A.merge} in case of concurrent conflicting substitutions.
		Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample list.  The
		sequence \C{[I(c,0); S(3,c,s)]} maps the list \C{[a;b;c]} to \C{[c;a;b;s]}
		because \C{subst 3 s (insert c 0 [a;b;c]) = [c;a;b;s]}.
	\item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and $s_2$,
		that map a list $v$ to two different lists, $v_1$ and $v_2$ (e.g.,
		Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such that $s_1'$
		has the same effect on $v_2$ as $s_1$ had on $v$.  For instance, in
		Fig.~\ref{list-eg}, let $s_1$ be the edit sequence \C{[D(1); S(1,c,d)]},
		which maps the list \C{[a;b;c]} ($v$) to \C{[a;d]} ($v_1$). The \C{D} edit
		deletes the first element of $v$. However, if $s_1$ is to be applied on the
		list \C{[c;a;b;s]} ($v_2$), which has a new element at position 0, the
		\C{D} edit must delete the the second element from $v_2$, if it were to
		have the same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
		transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The function
		\C{op\_transform} (Fig.~\ref{fig:mlist-sig}) computes such
		\emph{operational transformations} for an edit sequence $s_1$ w.r.t another
		edit sequence $s_2$. A notable transformation rule is for a substitution
		w.r.t another substitution at the same position. Since the substituting
		items could be different, the function relies on the item merge function
		(\C{A.merge}) to merge them into a single item. For instance, in
		Fig.~\ref{fig:list-eg}, there are simultaneous substitutions to the item
		\C{c} in $s_1$ and $s_2$; while $s_1$ substitutes it with \C{d}, $s_2$ does
		it with \C{s}. The merged list therefore substitutes \C{c} with the merge
		of \C{d} and \C{s}, defined as per the merge semantics of \C{A.t}.
\end{itemize}
The \C{MList.merge} can now be defined as a function that first
computes edit sequences, $s_1$ and $s_2$, for the two concurrent
lists, $v_1$ and $v_2$ w.r.t their common ancestor $v$, transforms one
of them (say, $s_1$) w.r.t $s_2$ to obtain $s_1'$, and finally applies
(the operations defined by) $s_1'$ to $v_2$ to obtain the merged list.

A significant advantage of mergeable types is their composability. As
demonstrated by \C{MList}, the merge semantics of a composed mergeable
type is a composition of the merge operations of its constituent types.  For
instance, a mergeable list of integers can be obtained by
instantiating \C{MList} for an \C{Int} type, where \C{Int.merge} is
the same as \C{Counter.merge} (Sec.~\ref{sec:intro}).
An alternative instantiation of \C{MList} is for a quantity type
\begin{wrapfigure}{l}{.45\textwidth}
  \centering
  \includegraphics[scale=0.4]{Figures/list-eg}
  \caption{Lists of mergeable values are mergeable. }
  \label{fig:list-eg}
\end{wrapfigure}
(\C{Qty}) that supports \C{add} and \C{subtract}
operations, but defines its \C{merge} opeation a little differently:
\begin{ocaml}
let merge lca v1 v2 = if v1 = lca then v2
    else (if v2 = lca then v1 else max v1 v2)
\end{ocaml}
\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has
the same semantics under a sequential execution. Their differences
however manifest under a concurrent execution, where both adopt
different methods to reconcile concurrent updates.

% The mergeable list data type demonstrates a common pattern of writing
% \C{merge} functions applicable to a range of functional data
% structures. For instance, a binary tree (with rotations) can be made
% mergeable in the same way as list, i.e., by computing edit sequences
% (\C{edit\_seq}) and then their transformations
% (\C{op\_transform})\footnote{
%   Like Wagner-Fischer, established algorithms exist for computing tree
%   edit distances~\cite{tree-diff}. An alternative naive method to
%   merge a pair of trees is by merging the lists obtained from their
%   in-order traversal, followed by reconstructing the tree.
% }.

{\bf Shopping List}. A non-trivial demonstration of composability of
mergeable types is a shopping list application. The application allows
its users to collaboratively build a grocery shopping list (listing
items in the decreasing order of their priority), by adding new items
with a specified quantity, deleting existing items, and increasing or
decreasing their quantity. An item (\C{Item.t}) is represented as a
tuple of its name (a string) and the quantity (A \C{Qty.t}). The merge
function for
\begin{wrapfigure}{l}{.5\textwidth}
  \centering
  \includegraphics[scale=0.3]{Figures/shopping-list}
  \caption{Merging concurrent shopping lists}
  \label{fig:shopping-list}
\end{wrapfigure}
\C{Item.t} merges the quantities of items (using \C{Qty.merge}) having
the same names. Items with different names are considered distinct. A
shopping list is a mergeable list of items (i.e.,
\C{MList(Item)}). \C{MList(Item).merge} automatically lifts the merge
semantics for items to merge semantics for lists of items.
Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
users, Alice and Bob, concurrently edit a shopping list whose initial
contents are \C{milk} and \C{eggs}. While both simultaneously update the
quantity of eggs, Bob also removes milk and inserts candy. The merged
shopping list contains eggs and candy, where the quantity of eggs is
obtained by merging Alice's and Bob's quantity as per the definition
of \C{Qty.merge}.

The examples described above make certain assumptions about the model
and the underlying system, such as the existence of a single least
common ancestor (LCA) for any pair of versions, the ability to access
a previous version on any branch, and mergeability of any two
concurrent versions. Enforcing these guarantees in a fully
decentralized distributed setting requires addressing non-trivial
theoretical challenges. These, and other details that contribute to
the practicality of our model, such as containing the complexity of
merges, are discussed in subsequent sections.

### 2017-02-27 paper/examples.tex
Rope Edits.

@@
 \section{Mergeable Types}
+
+\input{rope-sig}
 
@@
 
-Fig.~\ref{fig:rope-sig} shows a rope interface in OCaml. The type
-of the list of values is left abstract (\C{A.t}); it can be
-instantiated with any module that supports \C{concat} and \C{split}
-operations. The interface shown in Fig.~\ref{fig:rope-sig} is that of
-a mergeable rope (\C{MRope});
+Fig.~\ref{fig:mrope-sig} shows a rope interface in OCaml. The type of
+the list of values is left abstract (\C{A.t}); it can be instantiated
+with any module that supports \C{concat} and \C{split} operations. The
+interface shown in Fig.~\ref{fig:mrope-sig} is that of a mergeable
+rope (\C{MRope}), whose \C{merge} implementation is shown
+in Fig.~\ref{fig:mrope-merge}. The function returns one of the
+versions (\C{l} or \C{r}), if the other one is same as the common
+ancestor. Otherwise, it calls \C{merge\_rec}, which recursively
+compares the structures of merging ropes, and their common ancestor.
+The function recurses until one of the ropes differs from the other
+two, or all of them are leaf nodes, at which point it relies on the
+(abstract) list merge function (\C{A.merge}) to merge the lists
+represented by the ropes. \C{merge\_rec} relies on auxiliary
+functions, \C{mk\_rope} and \C{create\_index}, whose definitions are
+not shown. The former constructs a rope given a list (\C{A.t}), and
+the latter computes the index at a node, given its left sub-tree.
+
+One feature of the mergeable rope implementation in
+Fig.~\ref{fig:rope-sig}, that is characterisic of mergeable types in
+general, is compositionality; rope is a polymorphic mergeable data
+type that composes mergeability. The compositionality is concretized
+by the \C{MRope} signature , which requires the type of items in the
+list (\C{A.t}) to also be mergeable (\C{A.merge}).
 
@@
 \end{itemize}
-Thus, lists serve as an example of a polymorphic mergeable data type that
-demonstrates how mergeability can be composed.  Compositionality is
-reified in the the signature of a mergeable list library as shown in
-Fig.~\ref{fig:mlist-sig}; observe that the type of items in the list
-(\C{A.t}) is also required to be mergeable (\C{A.merge}).
 
@@
 		list \C{[c;a;b;s]} ($v_2$), which has a new element at position 0, the
-		\C{D} edit must delete the the second element from $v_2$, if it were to
+		\C{D} edit must delete the second element from $v_2$, if it were to
 		have the same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be

### 2017-02-27 paper/related.tex
edits

@@
 exposed to low-level operational details.
+
+There have been numerous proposals over the years for realizing
+distributed functionality in a functional programming context.
+Poly/ML~\cite{Mat97} and Facile~\cite{TLK+97} both extend Standard ML
+with distributed programming abstractions.  Kali-Scheme~\cite{CJK95}
+defines a distributed extension for Scheme-48, and Cloud
+Haskell~\cite{EPPJ11} describes a domain-specific language that
+enables development of distributed Haskell programs.
+Obliq~\cite{Car95} explores issues related to mobility, lexical
+scoping, and network references in an object-oriented framework that
+enables safe (i.e., lexical) transmission of closures.  However,
+Unlike \name, none of these systems consider issues of object
+replication or consistency as central to their programming models.
+The realities of modern-day scalable distributed systems, especially
+those that are constructed from geographically distributed
+participants, dictate that we must necessarily grapple with
+consistency issues to achieve any kind of reasonable performance; it
+is noteworthy that replication is inherent in many widely-used
+distributed system implementations today (e.g.,
+Cassandra~\cite{Cassandra} or Dynamo~\cite{Dynamo}), which provide
+only weak consistency guarantees.  In this context, we view our
+primary contribution as one that provides a principled high-level
+approach to framing consistency issues leveraging functional
+programming abstractions and techniques that minimizes disruption to
+the way programmers structure and reason about their applications,
+without compromising efficiency or performance.  Indeed, our semantics
+of mergeable types and state versioning has utility in any concurrent
+context with non-trivial coordination and synchronization costs, even
+without taking distribution or replication into account.

### 2017-02-27 paper/related.tex
related work

--- 2017-02-26 paper/related.tex
@@
 exposed to low-level operational details.
+
+There have been numerous proposals over the years for realizing
+distributed functionality in a functional programming context.
+Poly/ML~\cite{Mat97} and Facile~\cite{TLK+93} both extend Standard ML
+with distributed programming abstractions.  Kali-Scheme~\cite{CJK95}
+defines a distributed extension for Scheme-48, and Cloud
+Haskell~\cite{EBPJ11} describes a domain-specific language that
+enables development of distributed Haskell programs.
+Obliq~\cite{Car95} explores issues related to mobility, lexical
+scoping, and network references in an object-oriented framework that
+enables safe (i.e., lexical) transmission of closures.  However,
+unlike \name, none of these systems consider issues of object
+replication or consistency as central to their programming models.
+The realities of modern-day scalable distributed systems, especially
+those that are constructed from geographically distributed
+participants, dictate that we must necessarily grapple with
+consistency issues to achieve any kind of reasonable performance; it
+is noteworthy that replication is inherent in many widely-used
+distributed system implementations today (e.g.,
+Cassandra~\cite{Cassandra} or Dynamo~\cite{Dynamo}), which provide
+only weak consistency guarantees.  Rather than providing, as other
+approaches have, a low-level operational treatment of replication, our
+primary contribution is the development of a principled high-level
+approach to framing issues of consistency and replication.  Our goal
+is to leverage functional programming principles to minimize
+disruption to the way programmers structure and reason about their
+applications, without compromising efficiency or performance.  Indeed,
+the semantics of mergeable types and versioning has utility in any
+concurrent setting that must deal with non-trivial coordination and
+synchronization costs, even without taking distribution or replication
+issues into account.

### 2017-02-27 paper/examples.tex
EDits.

@@
 
-{\bf Ropes}. A rope is a data structure that faciliates efficient
+\subsection{Ropes}
+
+A rope is a data structure that faciliates efficient
 implementations of operations such as concatenation, splitting, lookup
@@
 
+\subsection{Lists}
+
 \input{list-sig}
 
-{\bf Lists}. As an abstract data type, lists support three
+As an abstract data type, lists support three
 operations: \C{insert x i l}, to insert an element \C{x} at position
@@
 
-A significant advantage of mergeable types is their composability. As
-demonstrated by \C{MList}, the merge semantics of a composed mergeable
-type is a composition of the merge operations of its constituent types.  For
-instance, a mergeable list of integers can be obtained by
-instantiating \C{MList} for an \C{Int} type, where \C{Int.merge} is
-the same as \C{Counter.merge} (Sec.~\ref{sec:intro}).
-An alternative instantiation of \C{MList} is for a quantity type
+\subsection{Lists of Integers and Quantities}.
+
+A mergeable list of integers can be composed using mergeable lists and
+mergeable integers.  Concretely, it is a composition of \C{MList} and
+\C{Int} types (i.e., \C{MList(Int)}), where \C{Int.merge} is the same
+as \C{Counter.merge} (Sec.~\ref{sec:intro}).  An alternative
+instantiation of \C{MList} is for a quantity type
 \begin{wrapfigure}{l}{.45\textwidth}
@@
 \end{wrapfigure}
-(\C{Qty}) that supports \C{add} and \C{subtract}
-operations, but defines its \C{merge} opeation a little differently:
+(\C{Qty}) that supports \C{add} and \C{subtract} operations, but
+defines its \C{merge} opeation a little differently:
 \begin{ocaml}
@@
 different methods to reconcile concurrent updates.
+
+If efficient insertions and lookups (by position) are desired over a
+large list of integers, while admitting concurrent updates, we require
+a mergeable rope of integers, which can be obtained by simply
+composing \C{MList(Int)} with \C{MRope} (i.e., \C{MRope(MList(Int))}).
 
@@
 
-{\bf Shopping List}. A non-trivial demonstration of composability of
-mergeable types is a shopping list application. The application allows
-its users to collaboratively build a grocery shopping list (listing
-items in the decreasing order of their priority), by adding new items
-with a specified quantity, deleting existing items, and increasing or
+\subsection{Shopping List}
+
+\begin{figure}
+\centering
+\begin{tabular}{l||l}
+\begin{ocaml}
+let alice_f : L.t unit t =
+  fork bob_f >>= fun () -> (* Invite Bob *)
+  get () >>= fun c0 ->
+  let qty = L.assoc "eggs" in
+  let c0' = L.subst 1 ("eggs",qty+1) c0 in
+  sync () ~v:c0' >>=
+  fun c1 -> return ()
+\end{ocaml}
+&
+\begin{ocaml}
+let bob_f : L.t unit t =
+  get () >>= fun c0 ->
+  let c0' = L.insert ("candy",Qty.of_int 1) 1 @@ 
+            L.delete @@ L.lookup "eggs" in
+  sync () ~v:c0' >>=
+  fun c1 -> return ()
+\end{ocaml}
+\\
+\end{tabular}
+\caption{Bob and Alice collaboratively build a grocery shopping list}
+\label{fig:shopping-list-code}
+\end{figure}
+
+A non-trivial demonstration of composability of mergeable types is a
+shopping list application. The application allows its users to
+collaboratively build a grocery shopping list (listing items in the
+decreasing order of their priority), by adding new items with a
+specified quantity, deleting existing items, and increasing or
 decreasing their quantity. An item (\C{Item.t}) is represented as a
@@
 the same names. Items with different names are considered distinct. A
-shopping list is a mergeable list of items (i.e.,
-\C{MList(Item)}). \C{MList(Item).merge} automatically lifts the merge
-semantics for items to merge semantics for lists of items.
+shopping list is a mergeable list of items, i.e., \C{MList(Item)} (a
+grocery shopping list may not be large enough to justify using ropes,
+but \C{MRope(MList(Item))} is definitely a possibility).
+\C{MList(Item).merge} automatically lifts the merge semantics for
+items to merge semantics for lists of items.
 Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
 users, Alice and Bob, concurrently edit a shopping list whose initial
-contents are \C{milk} and \C{eggs}. While both simultaneously update the
-quantity of eggs, Bob also removes milk and inserts candy. The merged
-shopping list contains eggs and candy, where the quantity of eggs is
-obtained by merging Alice's and Bob's quantity as per the definition
-of \C{Qty.merge}.
+contents are \C{milk} and \C{eggs}. While both simultaneously update
+the quantity of eggs, Bob also removes milk and inserts candy
+(Fig.~\ref{fig:shopping-list-code} shows the \name code for the
+concurrent sessions). The merged shopping list contains eggs and
+candy, where the quantity of eggs is obtained by merging Alice's and
+Bob's quantity as per the definition of \C{Qty.merge}.
 

### 2017-02-27 paper/evaluation.tex
Added merge tree code
edits

@@
 In this section, we present an evaluation of \name programming model with the
-aim of assesing its practical utility for describing geo-distributed eventually
-consistent applications. In particular, we
+aim of assessing its practical utility for describing geo-distributed
+eventually consistent applications. In particular, we are interested in the
+horizontal scalability of applications built using \name. Typical distributed
+eventually consistent data stores use custom synchronization and dissemination
+protocols for transferring updates between replicas. Unlike this, \name is
+realized over Irmin which uses Git transfer protocol~\cite{..}, a protocol not designed
+with distributed data stores in mind. Hence, we also evaluate the performance
+of synchronization and merging.
 
-To this end, we have implemented a collaborative editing application that
-simulates concurrent editing of the same document by several authors. This
-application is not unlike Google Docs, but also differs from it by the fact
-that we do not need a central server that coordinates the edits. Instead, edits
-from each author is asynchronously sent to other authors. The shared document
-itself is represented by a mergeable lists of strings
+\subsection{Benchmark: Collaborative editing}
+
+In order to evaluate this, we have implemented a collaborative editing
+application that simulates concurrent editing of the same document by several
+authors. This application is not unlike Google Docs, but also differs from it
+by the fact that we do not need a central server that coordinates the edits.
+Instead, edits from each author is asynchronously sent to other authors. The
+shared document itself is represented by a mergeable rope of characters
 (Section~\ref{sec:model}), and hence, the remote edits can be correctly
@@
 substitutions at the same position (such as two authors concurrently editing
-\C{abc} to \C{xbc} and \C{ybc}, respectively) by choosing the character with
-the smaller ASCII code.
+\C{"abc"} to \C{"xbc"} and \C{"ybc"}, respectively) by choosing the character
+with the smaller ASCII code. For the benchmarks, the initial document that we
+use has 1576 characters. The workload consists of 4000 edit operations at
+random indices with 85\% insertions and 15\% deletions.
+
+\subsection{Experimental setup}
+
+\begin{figure}[t]
+  \centering
+	\begin{subfigure}{0.45\textwidth}
+		\includegraphics[width=0.9\textwidth]{Figures/cluster.pdf}
+		\caption{8-node ring cluster on Google Cloud Platform used for
+			benchmarking. US-C, EU-W, ASIA-NE and ASIA-E are availability zones
+			us-central1-b, europe-west1-b, asia-northeast1-b and asia-east1-b
+			respectively. Edge labels are inter-node latencies in milliseconds.}
+		\label{fig:cluster}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}{0.45\textwidth}
+		\includegraphics[width=\textwidth]{Graphs/scalability.pdf}
+		\caption{Scalability: Overall throughput of the cluster and latency of each
+			operation.}
+		\label{grf:scalability}
+	\end{subfigure}
+	\par\bigskip
+	\begin{subfigure}{0.45\textwidth}
+		\includegraphics[width=\textwidth]{Graphs/synchronization.pdf}
+		\caption{Synchronization: Overall throughput of the cluster while actively
+			synchronizing with the successor node.}
+		\label{grf:synchronization}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}{0.45\textwidth}
+		\includegraphics[width=\textwidth]{Graphs/merge.pdf}
+		\caption{Merge performance: Cost of merging concurrent operations across
+		nodes.}
+		\label{grf:merge}
+	\end{subfigure}
+	\caption{\name performance evaluation.}
+  \label{grf:evaluation}
+\end{figure}
+
+For our experiments, we instantiate an 8-node geo-distributed cluster in Google
+Cloud Platform~\cite{gcp}. Each node is an \C{n1-standard-4} instance with 2
+virtual CPUs and 7.5 GB of memory. The nodes are arranged in a ring as shown in
+Figure~\ref{fig:cluster} and synchronize (push and pull updates) when necessary
+with its successor. While \name does not impose any restrictions on
+synchronizing with other nodes, we choose to only synchronize with the
+successor since we are guaranteed eventual convergence with a ring cluster. If
+the successor becomes unavailable, the node synchronizes with other available
+nodes. Operations are generated and applied locally on each node before they
+are propagated eventually to other nodes with the help of a background thread
+that synchronizes with its successor every second.
+
+\subsection{Results}
+
+We evaluate the scalability of concurrent editing application by increasing the
+cluster size from 1 to 8 (the 4 node ring cluster consists of nodes numbered 0
+to 3), with each node performing concurrent edits to the same document. In each
+case, we measure the overall cluster throughput and latency of each operation.
+The results are presented in Figure~\ref{grf:scalability}. The results show
+that the cluster throughput increases linearly with the number of concurrent
+editors, while the latency for each operation remains the same. This is because
+each operation is performed locally and does not require synchronization with
+other nodes. The nodes remain available to accept requests even if the node
+gets disconnected. Since the document type is mergeable, eventually when the
+node comes back online, the updates are synchronized with the cluster.
+
+While updates are propagated asynchronously, we also measured the impact of
+propagating updates synchronously by forcing synchronization every few
+operations. This corresponds to a partially synchronous system, which is also
+unavailable; if the cluster is unreachable, local operations will not be
+accepted. While this is not a realistic deployment, the results help us
+understand the overhead of synchronization. The results are shown in
+Figure~\ref{grf:synchronization}. The result show that frequent blocking
+synchronization is prohibitively expensive with the throughput dropping to 3
+ops/s for synchronizing after every operation. \name uses Git transfer protocol
+for synchronization which involves multiple roundtrips between the nodes for
+synchronization. But as the synchronizations get less frequent, we can see that
+the throughput asymptotically approaches the asynchronous case
+(Figure~\ref{grf:scalability}).
+
+Finally, we evaluate the performance of merge functions. For this experiment,
+we consider the nodes 1 (\C{us-central1-b}) and 2 (\C{europe-west1-b}). Both
+nodes start with the initial document and perform \C{n} concurrent edits
+without synchronization. After this, the nodes synchronize pushing the local
+edits and pulling the remote edits, and merge to produce the final document. In
+this case, we indicate the distance of synchronization as \C{2n}.
+Figure~\ref{grf:merge} presents the time share between network and computation
+for performing 3-way merge. We see that the merge time is dominated by the
+network overhead. This is because of the transfer protocol involving multiple
+roundtrips and the need to transfer objects not only for application objects
+but also Git internal objects for history, directory structure and branches. We
+observe that we can improve the network performance by implementing a streaming
+protocol that eagerly transfers objects.
+
+More importantly, the cost of computation grows sub-linearly. This is because
+of efficient merge function on ropes enabled by Irmin's content-addressed
+storage backend, and more expensive operational transformation of merging
+concurrent edits to leaf nodes is invoked rarely on small strings.

### 2017-02-27 data/merge.py

@@ -16,23 +16,24 @@ def main ():
     if line == "\n" or line.startswith ("#"):
       continue
     d = line.split ()
-    merge.append ((int(d[0]), float(d[1])/1000.0, float(d[2])/1000.0, float(d[3])/1000.0))
+    merge.append ((int(d[0]), float(d[2]) * 100.0 / float(d[1]), float(d[3]) * 100.0/float(d[1])))
 
   font = {'family' : 'normal', 'weight' : 'normal', 'size' : 12}
   matplotlib.rc('font', **font)
 
   fig, ax1 = plt.subplots()
 
-  (dist,merge,comp,nw) = zip(*merge)
+  (dist,comp,nw) = zip(*merge)
 
-  fig, ax = plt.subplots()
-  ax.stackplot (dist, [comp, nw], colors=['#55BA87','#7E1137'])
-  ax.set_xlabel("Time (s)")
-  ax.set_ylabel("Distance (ops)")
-  ax.set_xlim([0,256])
-  plt.legend([mpatches.Patch(color='#55BA87'),
-              mpatches.Patch(color='#7E1137')],
-             ['computation','network'], loc='upper left')
+  ind = np.arange(len(dist))
+  width = 0.35
+  p1 = plt.bar (ind, comp, width, color='#FFCE00')
+  p2 = plt.bar (ind, nw, width, color='#0375B4', bottom=comp)
+
+  plt.xlabel("Distance (ops)")
+  plt.ylabel("Time share (%)")
+  plt.xticks(ind, dist)
+  plt.legend((p1[0],p2[0]),('computation','network'))
   fig.tight_layout()
   plt.savefig ("merge.pdf")
   plt.close ()

### 2017-02-27 paper/rope-sig.tex

\begin{figure}

\begin{subfigure}{0.35\textwidth}
\begin{ocaml}
module type MRope = sig
  module A: sig
    type t
    val concat: t -> t -> t
    val split: t -> t*t
    val merge: t -> t -> t -> t
  end
  type t =
  | Leaf of A.t
  | Node of t * int * t
  ... 
  (* Standard rope functions *)
  val merge: t -> t -> t -> t 
end
\end{ocaml}
\caption{Signature of Mergeable Ropes}
\label{fig:mrope-sig}
\end{subfigure}
\begin{subfigure}{0.6\textwidth}
\begin{ocaml}
let rec flatten = function
  | Leaf x -> x
  | Node (l,_,r) -> A.concat (flatten l) (flatten r)

let rec merge old l r =
  if l = r then l else if old = l then r
  else if old = r then l else merge_rec old l r

and merge_rec old l r = match (old,l,r) with
  | Leaf _, _, _ | _, Leaf _, _ 
  | _, _, Leaf _ -> 
      rope_of @@ A.merge (flatten old) 
                    (flatten l) (flatten r)
  | Node (oldl,_,oldr), Node (ll,_,lr), 
    Node (rl,_,rr) ->
      let newl = merge oldl ll rl in
      let newr = merge oldr lr rr in
        Node (newl, create_index newl, newr)
\end{ocaml}
\caption{Signature of Mergeable Lists}
\label{fig:mrope-merge}
\end{subfigure}

\caption{Mergeable List Implementation}
\label{fig:rope}
\end{figure}

### 2017-02-27 paper/examples.tex
edits
Evaluation

@@
 \section{Mergeable Types}
+\label{sec:merge-examples}
 
@@
 programming model in building concurrent applications with conceptual
-state sharing. The model lets concurrent computations be composed
-around any mergeable type. We now present a series of examples that
-demonstrate how mergeable types are built: either by defining merge
-functions for OCaml data types, or by composing existing mergeable
-types.
+state sharing that make them suitable for transparent deployment in a
+distributed setting. The model lets concurrent computations to be
+composed around any mergeable type. We now present a series of
+examples that demonstrate how mergeable types are built: either by
+defining merge functions for OCaml data types, or by composing
+existing mergeable types.
 
@@
   \item An element deleted in one list will not reappear in the merged
-  list (unlike Amazon's replicated shopping cart~\cite{dynamo}).
+  list (unlike Amazon's replicated shopping cart~\cite{Dynamo}).
   \item The order of elements is retained. That is, if $x$ occurs

### 2017-02-27 paper/intro.tex
edits

@@
       include Counter
-      type mergeable_t = t 
       let merge lca v1 v2 =
@@
   \end{ocaml}
-
 The role of \C{lca} (least common ancestor) here captures salient
@@
 
+The remainder of the paper is organized as follows.  We present the
+\name\ programming model informally in the next section.
+Sec.~\ref{sec:merge-examples} develops a series of examples that
+illustrate how merge operations can be defined and used.
+Sec.~\ref{sec:formalization} defines an operational semantics, and
+formalizes our intution of a mergeable type in terms of morphisms over
+datatype operations.  We also present soundness (consistency) and
+progress guarantees enjoyed by the
+semantics. Sec.~\ref{sec:system-model} considers an instantiation of
+the model suitable for distributed environments with network
+partitions and failures.  Experimental results
+
+, along with
+a series of examples illustrating how merge operations can be defined
+and used.  Sec.~\ref{sec:
+
 %% Persistence adds yet another dimension to the problem. Functional data

### 2017-02-27 paper/evaluation.tex

@@
 ops/s for synchronizing after every operation. \name uses Git transfer protocol
-for synchronization which involves multiple roundtrips between the nodes for
+for synchronization which involves multiple round trips between the nodes for
 synchronization. But as the synchronizations get less frequent, we can see that
@@
 Figure~\ref{grf:merge} presents the time share between network and computation
-for performing 3-way merge. We see that the merge time is dominated by the
-network overhead. This is because of the transfer protocol involving multiple
-roundtrips and the need to transfer objects not only for application objects
+for performing 3-way merge as we increase the operation distance between the
+remote branches. We see that the merge time is dominated by the network
+overhead. This is because of the transfer protocol involving multiple
+round trips and the need to transfer objects not only for application objects
 but also Git internal objects for history, directory structure and branches. We
@@
 More importantly, the cost of computation grows sub-linearly. This is because
-of efficient merge function on ropes enabled by Irmin's content-addressed
-storage backend, and more expensive operational transformation of merging
-concurrent edits to leaf nodes is invoked rarely on small strings.
+of efficient merge function on ropes~\ref{ropes} enabled by Irmin's
+content-addressed storage backend, and more expensive operational
+transformation of merging concurrent edits to leaf nodes~\ref{lists} is invoked
+rarely on small. On average, we measured that at a distance of 256, for merging
+the two documents with 1743 characters, 50 list merges were performed on
+strings of average length 13. A purely list based approach would require the
+polynomial time edit sequence algorithm to be applied on the entire document
+twice. Compare to this, the rope implementation has much better algorithmic
+complexity. We measured that by representing the entire document using
+mergeable list type, in addition to slowing down each edit operation, the merge
+at a distance of 256 was 2.3 times slower than the rope implementation.

### 2017-02-27 paper/intro.tex

@@
 \name\ programming model informally in the next section.
-Sec.~\ref{sec:merge-examples} develops a series of examples that
+Sec.~\ref{sec:mergeable_types} develops a series of examples that
 illustrate how merge operations can be defined and used.
@@
 the model suitable for distributed environments with network
-partitions and failures.  Experimental results
+partitions and failures.  Implementation details describing how we
+incorporate mergeable types into OCaml are presented in
+Sec.~\ref{sec:impl}.  Experimental results and benchmarks are given
+in Sec.~\ref{sec:evaluation}.
 

### 2017-02-27 paper/evaluation.tex
edits

@@
 More importantly, the cost of computation grows sub-linearly. This is because
-of efficient merge function on ropes~\ref{ropes} enabled by Irmin's
+of efficient merge function on ropes~\ref{sec:ropes} enabled by Irmin's
 content-addressed storage backend, and more expensive operational
-transformation of merging concurrent edits to leaf nodes~\ref{lists} is invoked
+transformation of merging concurrent edits to leaf nodes~\ref{sec:lists} is invoked
 rarely on small. On average, we measured that at a distance of 256, for merging

### 2017-02-27 paper/dali.tex
Minor edit
edits
Sec. 2 -> Sec. 2 + Sec. 3

@@
 \newcommand{\bind}{>\!\!>\!\!=}
+\ifx\coloneqq\undefined
   \newcommand{\coloneqq}{::=}
+\else
+  \renewcommand{\coloneqq}{::=}
+\fi
 \newcommand{\stepsto}{\longrightarrow}
@@
 \newcommand{\drawsome}{{\sf Canvas}\xspace}
+\newcommand{\buylist}{{\sf ToBuy}\xspace}
 \newcommand{\lang}{\lambda_b}
 \newcommand{\length}[1]{\left|#1\right|}
+\newcommand{\GK}[1]{\textcolor{red}{GK: #1}}
 
@@
 %escapeinside={\%*}{*)},         % if you want to add a comment within your code
-morekeywords={module, when, @@deriving, not, : , /\\}
+morekeywords={module, match, when, @@deriving, not, : , /\\}
 }}
@@
 \citestyle{acmauthoryear}   %% For author/year citations
+
+\newcommand{\KC}[1]{{\textcolor{red}{{KC: #1}}}}
 
@@
   availability and fault tolerance.  But, the use of replication leads
-  to a significantly more complex programming model.  To ensure a
-  sensible measure of consistency among different replicas of an
-  object, applications are typically structured so that the operations
-  applied to a replicated object on one node are subsequently
-  communicated to other nodes where they can be re-applied locally.
-  Oftentimes, the degree of restructuring and care necessary to
-  transform a sequential or shared-memory concurrent application to a
-  meaningful (replicated) distributed one can be substantial.
+  to a significantly more complex and onerous programming model.  To
+  ensure a sensible measure of consistency among different replicas of
+  an object, applications are typically structured so that the
+  operations applied to a replicated object on one node are
+  subsequently communicated to other nodes where they can be
+  re-applied locally.  Oftentimes, the degree of restructuring and
+  care necessary to transform a sequential or shared-memory concurrent
+  application to a meaningful (replicated) distributed one can be
+  substantial.
 
@@
   periodically reconcile local versions to yield a consistent global
-  state.  Thus, our model frees the programmer from having to think of
+  state.  Our model frees the programmer from having to think of
   low-level operational notions related to distribution and
@@
 
-  We show how any ML datatype can be enriched with mergeable
-  semantics, formalize a rich concurrency model based on such types,
-  and describe an instantiation of OCaml equipped with these features.
-  Our implementation is integrated with a content-addressable
-  distributed storage layer that allows any OCaml program to be easily
-  transformed into an efficient, highly available, distributed
-  application.
+  We show how any ML datatype can be enriched to support mergeability,
+  formalize a concurrency semantics based on such types, and describe
+  an instantiation of OCaml equipped with these features.  Our
+  implementation is integrated with a content-addressable distributed
+  storage layer that allows any OCaml program to be easily transformed
+  into an efficient, highly available, distributed application.
 \end{abstract}
@@
 
+\input{examples}
+
 \input{semantics}
 
-\input{mergeable-types}
+\input{system-model}
+
+\input{evaluation}
 
@@
 \appendix
-\section{Appendix}
-
-Text of appendix \ldots
+\input{appendix}
 

### 2017-02-27 paper/examples.tex
edits

@@
 \section{Mergeable Types}
-\label{sec:merge-examples}
+\label{sec:mergeable_types}
 
@@
 state sharing that make them suitable for transparent deployment in a
-distributed setting. The model lets concurrent computations to be
+distributed setting. The model allows concurrent computations to be
 composed around any mergeable type. We now present a series of
 examples that demonstrate how mergeable types are built: either by
-defining merge functions for OCaml data types, or by composing
-existing mergeable types.
+defining merge functions for ML data types, or by composing existing
+mergeable types.
 
 \subsection{Ropes}
+\ref{sec:ropes}
 
-A rope is a data structure that faciliates efficient
-implementations of operations such as concatenation, splitting, lookup
-and insertion at a random position etc., for a list of values. The
-values are usually characters; their lists being strings. The data
-structure is a binary tree, whose leaf nodes store substrings of a
-larger string such that their left-to-right (in-order) concatenation
-yeilds the larger string. The idea is to facilitate efficient lookups
-and insertions at random positions in the string by storing
-position indexes in the internal nodes of the tree.
+A rope is a data structure that faciliates efficient implementations
+of operations such as concatenation, splitting, lookup and insertion
+at a random position etc., for a list of values in a distributed
+setting. The values are usually characters, with their lists being
+strings. The data structure is a binary tree, whose leaf nodes store
+substrings of a larger string such that their left-to-right (in-order)
+concatenation yeilds the larger string. The idea is to facilitate
+efficient lookups and insertions at random positions in the string by
+storing position indexes in the internal nodes of the tree.
 
@@
 interface shown in Fig.~\ref{fig:mrope-sig} is that of a mergeable
-rope (\C{MRope}), whose \C{merge} implementation is shown
-in Fig.~\ref{fig:mrope-merge}. The function returns one of the
-versions (\C{l} or \C{r}), if the other one is same as the common
+rope (\C{MRope}), whose \C{merge} implementation is shown in
+Fig.~\ref{fig:mrope-merge}. The function returns one of its versions
+(\C{l} or \C{r}), if the other one is the same as the common
 ancestor. Otherwise, it calls \C{merge\_rec}, which recursively
@@
 (abstract) list merge function (\C{A.merge}) to merge the lists
-represented by the ropes. \C{merge\_rec} relies on auxiliary
-functions, \C{mk\_rope} and \C{create\_index}, whose definitions are
-not shown. The former constructs a rope given a list (\C{A.t}), and
-the latter computes the index at a node, given its left sub-tree.
+represented by the ropes. \C{merge\_rec} uses auxiliary functions,
+\C{mk\_rope} and \C{create\_index}, whose definitions are not shown to
+construct a rope given a list (\C{A.t}), and to compute the index at a
+node, given its left sub-tree, resp.
 
-One feature of the mergeable rope implementation in
-Fig.~\ref{fig:rope-sig}, that is characterisic of mergeable types in
-general, is compositionality; rope is a polymorphic mergeable data
-type that composes mergeability. The compositionality is concretized
-by the \C{MRope} signature , which requires the type of items in the
-list (\C{A.t}) to also be mergeable (\C{A.merge}).
+One feature of the mergeable rope implementation that is characterisic
+of mergeable types in general is compositionality; a rope is a
+polymorphic mergeable data type that composes mergeability over its
+polymorphic constituents.   Compositionality is concretized by the
+\C{MRope} signature , which requires the type of items in the list
+(\C{A.t}) to also be mergeable (\C{A.merge}).
 
 \subsection{Lists}
+\label{sec:lists}
 
@@
 
-As an abstract data type, lists support three
-operations: \C{insert x i l}, to insert an element \C{x} at position
-\C{i} in the list \C{l}, \C{delete i l}, to delete the element at
-\C{i}'th position, and \C{subst i x l}, to substitute the element at
-the \C{i}'th position with \C{x}.  Lists of mergeable items are
-mergeable if they preserve the following properties:
+As an abstract data type, lists support three operations: \C{insert x
+  i l}, to insert an element \C{x} at position \C{i} in the list
+\C{l}, \C{delete i l}, to delete the element at \C{i}'th position, and
+\C{subst i x l}, to substitute the element at the \C{i}'th position
+with \C{x}.  Lists of mergeable items are mergeable if they preserve
+the following properties:
 \begin{itemize}
@@
   \item An element deleted in one list will not reappear in the merged
-  list (unlike Amazon's replicated shopping cart~\cite{Dynamo}).
+  list (unlike, for example, Amazon's replicated shopping cart~\cite{Dynamo}).
   \item The order of elements is retained. That is, if $x$ occurs
@@
 
-The merge operation for lists is composed of two separate functions,
+\noindent The merge operation for lists is composed of two separate functions,
 \C{edit\_seq} and \C{op\_transform}. Both these functions have been
@@
 instantiation of \C{MList} is for a quantity type
+(\C{Qty}) that supports \C{add} and \C{subtract} operations, but
 \begin{wrapfigure}{l}{.45\textwidth}
@@
   \label{fig:list-eg}
+  \vspace*{-.2in}
 \end{wrapfigure}
-(\C{Qty}) that supports \C{add} and \C{subtract} operations, but
 defines its \C{merge} opeation a little differently:
@@
 
-A non-trivial demonstration of composability of mergeable types is a
-shopping list application. The application allows its users to
-collaboratively build a grocery shopping list (listing items in the
-decreasing order of their priority), by adding new items with a
+A non-trivial demonstration illustrating the compositional power of
+mergeable types is shopping list application, that allows its users to
+collaboratively build a shopping list (listing items in the decreasing
+order of their priority), by adding new items with a
 specified quantity, deleting existing items, and increasing or
-decreasing their quantity. An item (\C{Item.t}) is represented as a
-tuple of its name (a string) and the quantity (A \C{Qty.t}). The merge
-function for
 \begin{wrapfigure}{l}{.5\textwidth}
@@
   \label{fig:shopping-list}
+  \vspace*{-.2in}
 \end{wrapfigure}
-\C{Item.t} merges the quantities of items (using \C{Qty.merge}) having
-the same names. Items with different names are considered distinct. A
-shopping list is a mergeable list of items, i.e., \C{MList(Item)} (a
-grocery shopping list may not be large enough to justify using ropes,
-but \C{MRope(MList(Item))} is definitely a possibility).
-\C{MList(Item).merge} automatically lifts the merge semantics for
-items to merge semantics for lists of items.
+decreasing their quantity. An item (\C{Item.t}) is represented as a
+tuple of its name (a string) and the quantity (A \C{Qty.t}). The merge
+function for \C{Item.t} merges the quantities of items (using
+\C{Qty.merge}) having the same names. Items with different names are
+considered distinct. A shopping list is a mergeable list of items,
+i.e., \C{MList(Item)} (a grocery shopping list may not be large enough
+to justify using ropes, but \C{MRope(MList(Item))} is definitely a
+possibility).  \C{MList(Item).merge} automatically lifts the merge
+semantics for items to merge semantics for lists of items.
 Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
@@
 contents are \C{milk} and \C{eggs}. While both simultaneously update
-the quantity of eggs, Bob also removes milk and inserts candy
-(Fig.~\ref{fig:shopping-list-code} shows the \name code for the
-concurrent sessions). The merged shopping list contains eggs and
-candy, where the quantity of eggs is obtained by merging Alice's and
-Bob's quantity as per the definition of \C{Qty.merge}.
+the quantity of eggs, Bob also removes \C{milk} and inserts \C{candy}.
+Fig.~\ref{fig:shopping-list-code} shows the \name code for two
+concurrent sessions. The merged shopping list contains \C{eggs} and
+\C{candy}, where the quantity of eggs is obtained by merging Alice's
+and Bob's quantity as per the definition of \C{Qty.merge}.
 

### 2017-02-27 paper/semantics.tex
Semantics.

@@
 The function \C{lca} used in the premise of \rulelabel{E-Pull-Wait}
-computes the LCA of two branches. The function is in fact a partial
-function because the definition of LCA doesn't guarantee its
-uniqueness for a pair of branches. Such nuances of LCA, and the
-guarantees provided by the system notwithstanding these nuances, are
-discussed in the subsection~\ref{sec:meta}.
+computes the LCA of two branches, and assumes it to be unique.  But,
+observe that the definition of LCA doesn't guarantee its uniqueness;
+it has to be enforced explicitly.  The semantics enforces the
+uniqueness of LCA by constraining the branching structure via the
+\rulelabel{E-Pull-Wait} rule. The rule merges version $v'$ into
+version $v$ only if they are \emph{safely mergeable} (denoted
+$\under{H}{v' \mbleto t}$), i.e., they have a single LCA, and merging
+both versions does not lead to a case where a branch has multiple LCAs
+with rest of the branches, hence gets \emph{stuck}. Such nuances of
+LCA, and the guarantees provided by the system notwithstanding these
+nuances, are discussed in the subsection~\ref{sec:meta}.
 

### 2017-02-27 paper/model.tex

@@
 example of a collaborative drawing application we call \drawsome.
-
 Fig.~\ref{fig:canvas-sig} shows the signature of the \drawsome
@@
 and returning the pixel at a given coordinate. The \C{merge} function
-and \C{deriving versioned} directive are explained below.  \drawsome
-lets multiple users collaborate on a canvas that is conceptually
-shared among them. Under a shared-memory abstraction, there would be a
-single copy of the canvas that is updated concurrently by multiple
-clients; from the perspective of any single client, the canvas could
-change without any explicit intervention. \name ascribes functional
-semantics to sharing by letting each client work on its own version of
-the state (the tree data structure in this example), later merging
-concurrent versions on-demand.  Thus, the primary artifact of the
-\name programming model is a versioned data structure in which
-different versions are managed by different clients.
+is explained below.  \drawsome lets multiple users collaborate on a
+canvas that is conceptually shared among them. Under a shared-memory
+abstraction, there would be a single copy of the canvas that is
+updated concurrently by multiple clients; from the perspective of any
+single client, the canvas could change without any explicit
+intervention. \name ascribes functional semantics to sharing by
+letting each client work on its own version of the state (the tree
+data structure in this example), later merging concurrent versions
+on-demand.  Thus, the primary artifact of the \name programming model
+is a versioned data structure in which different versions are managed
+by different clients.
 
@@
 
-\name requires a three-way \C{merge} function to merge the concurrent
+\name requires a three-way \C{merge} function to merge concurrent
 versions of a drawing canvas (see Fig.~\ref{fig:merge-canvas}). The
@@
 \C{v1} is white, and in \C{v2} it is green, and its color in \C{lca}
-\begin{wrapfigure}{l}{.5\textwidth}
+\begin{wrapfigure}{l}{.4\textwidth}
 \centering
-\includegraphics[scale=0.5]{Figures/canvas-merging}
-\caption{Merging concurrent versions (\C{v1} and \C{v2} ) of a drawing
+\includegraphics[scale=0.43]{Figures/canvas-merging}
+\caption{Merging concurrent versions (\C{v1} and \C{v2}) of a drawing
 canvas. \C{lca} is their common ancestor.}

### 2017-02-27 paper/dali.tex
Semantics.
edits

@@
   that overcome these concerns.  Our declarative programming model
-  called \name\ treats replication as a form of versioning, in which
-  consistency is enforced through programmer-supplied \emph{merge}
-  functions.  Distributed state is now viewed in terms of a tree of
-  immutable object versions, with merge actions used explicitly by the
-  application or implicitly by the underlying runtime system to
-  periodically reconcile local versions to yield a consistent global
-  state.  Our model frees the programmer from having to think of
-  low-level operational notions related to distribution and
-  replication, and therefore does not entail any restructuring of
-  application logic to enable distributed programming.
+  called \name\ treats replication as a compositional form of
+  versioning, in which consistency is enforced through
+  programmer-supplied \emph{merge} functions.  Distributed state is
+  now viewed in terms of a tree of immutable object versions, with
+  merge actions used explicitly by the application or implicitly by
+  the underlying runtime system to periodically reconcile local
+  versions to yield a consistent global state.  Our model frees the
+  programmer from having to think of low-level operational notions
+  related to distribution and replication, and therefore does not
+  entail any restructuring of application logic to enable distributed
+  programming.
 

### 2017-02-27 paper/related.tex
edits

@@
 the integration of a version-based mechanism within OCaml allows a
-degree of type safety, composability, and polymorphism not available
-in related systems.
+degree of type safety, composability, and profitable use of
+polymorphism not available in related systems.
 

### 2017-02-27 paper/model.tex

@@
 \name requires a three-way \C{merge} function to merge concurrent
-versions of a drawing canvas (see Fig.~\ref{fig:merge-canvas}). The
-three arguments include two concurrent versions (\C{v1} and \C{v2}),
+versions of a drawing canvas (see Fig.~\ref{fig:merge-canvas}) that
+includes includes two concurrent versions (\C{v1} and \C{v2}),
 and their least common ancestor (\C{lca}) - the version from which the
 two concurrent versions evolved independently. The merge function can
-make use of the pixel values of the common ancestor to merge the pixel
-values on both the canvases. For instance, if the color of a pixel in
-\C{v1} is white, and in \C{v2} it is green, and its color in \C{lca}
 \begin{wrapfigure}{l}{.4\textwidth}
@@
 \end{wrapfigure}
+make use of the pixel values of the common ancestor to merge the pixel
+values on both the canvases. For instance, if the color of a pixel in
+\C{v1} is white, and in \C{v2} it is green, and its color in \C{lca}
 is white, then it means that only \C{v2} modified the color. Hence the
@@
     val fork_version : ('a, 'b) t -> 'a unit t
-    val sync_next_version: unit -> ?v:'a -> ('a, 'a) t
+    val sync_next_version: unit -> 'a -> ('a, 'a) t
   end
@@
 monad's \C{get}; it returns the versioned state encapsulated by the
-monad. \C{with\_init\_version\_do} runs a monadic computation against
-a given initial version and returns the result. \C{fork\_version}
-returns a computation that forks a new concurrent version from the
-current version, and runs the given monadic computation asynchronously
-against the forked version.  \C{sync\_next\_version} (simply called
-\C{sync}) accepts a \emph{proposal} for the next version of the state;
-this proposal is the current local version of the state that reflects
-local modifications not yet witnessed by any other concurrently
-executing computation.  The operation returns (via a monad) the actual
-next version, which becomes the current version for the rest of the
+monad. To initiate a computation, we use \C{with\_init\_version\_do}
+which runs a monadic computation against a given initial version and
+returns the result. The \C{fork\_version} operation returns a
+computation that forks a new concurrent version from the current
+version, and runs the given monadic computation asynchronously against
+the forked version.  \C{sync\_next\_version} (simply called \C{sync})
+accepts a \emph{proposal} for the next version of the state; this
+proposal is the current local version of the state that reflects local
+modifications not yet witnessed by any other concurrently executing
+computation.  The operation returns (via a monad) the actual next
+version, which becomes the current version for the rest of the
 computation.  This version is created by merging the proposal with a
 subset of concurrent versions that have become available since the
-last merge or fork. Thus, \C{sync} effectively lets a computation sync
-with a subset of concurrent computations to obtain their latest
-updates.
+last merge or fork. Thus, \C{sync} effectively lets a computation
+synchronize with a subset of concurrent computations to obtain their
+latest updates.
 
@@
 Alice draws a red horizontal line from $(0,0)$ (top-left) to $(4,0)$
-(top-right) using \C{C.draw\_line}.\footnote{Its definitionis not
-  shown, but can be constructed using \C{set\_px}} Meanwhile, Bob
+(top-right) using \C{C.draw\_line}.\footnote{Its definition is not
+  shown, but can be constructed using \C{set\_px}.} Meanwhile, Bob
 draws a green vertical line from $(0,0)$ to $(0,4)$, and Cheryl draws

### 2017-02-27 paper/dali.tex
edits

@@
   storage layer that allows any OCaml program to be easily transformed
-  into an efficient, highly available, distributed application.
+  into a composable, efficient, and highly-available distributed
+  application.
 \end{abstract}
@@
 
-\input{evaluation}
+%\input{evaluation}
 

### 2017-02-27 paper/examples.tex
edits

@@
 \subsection{Ropes}
-\ref{sec:ropes}
+\label{sec:ropes}
 
@@
   let c0' = L.insert ("candy",Qty.of_int 1) 1 @@
-            L.delete @@ L.lookup "eggs" in
+            L.delete @@ L.lookup "milk" in
   sync () ~v:c0' >>=
@@
 possibility).  \C{MList(Item).merge} automatically lifts the merge
-semantics for items to merge semantics for lists of items.
+semantics for items to a merge semantics for lists of items.
 Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
@@
 
-The examples described above make certain assumptions about the model
-and the underlying system, such as the existence of a single least
-common ancestor (LCA) for any pair of versions, the ability to access
-a previous version on any branch, and mergeability of any two
-concurrent versions. Enforcing these guarantees in a fully
+The examples described above make certain assumptions about the
+concurrency model and the underlying system, such as the existence of
+a single least common ancestor (LCA) for any pair of versions, the
+ability to access a previous version on any branch, and mergeability
+of any two concurrent versions. Enforcing these guarantees in a fully
 decentralized distributed setting requires addressing non-trivial

### 2017-02-27 paper/mergeable-types.tex
Semantics.
edits

@@
 
-Rather than specifying merge ($M$) in terms of an LCA,
-Definition~\ref{def:mergeable-type} specifies $M$ in terms of a pair
-of morphisms.  This specification is useful to avoid the expression of certain
-nonsensical merge functions, and thus enforce a useful branch-local
-consistency property (Theorem~\ref{thm:branch-consistency}). For
-instance, consider the counter data type from Sec.~\ref{sec:intro}. A
-three-way merge function can be defined to return a constant
-regardless of its arguments:
+\noindent Rather than specifying merge ($M$) in terms of an LCA as we
+did in describing our programming model,
+Def.~\ref{def:mergeable-type} specifies $M$ in terms of a pair
+of morphisms.  This specification is useful to avoid the expression of
+certain nonsensical merge functions, and thus enforce a useful
+branch-local consistency property
+(Theorem~\ref{thm:branch-consistency}). For instance, consider the
+counter data type from Sec.~\ref{sec:intro}. A three-way merge
+function can be defined to return a constant regardless of its
+arguments:
 \begin{ocaml}
@@
 \end{ocaml}  
-Such a merge function allows the client to
-witness the violation of the branch-local consistency property. For
-example, the client may \C{sync} the counter, whose current local
-value is $10$, and obtain $0$ as a result, thus witnessing a violation
-of counter's monotonicity property. Fortunately,
-Def.~\ref{def:mergeable-type} disallows this semantics. It is
-impossible to define the function $M$ for the counter data type that
-violates the monotonicity invariant, because no counter morphism
-violates the invariant. In general, the specification of $M$
+Such a merge function allows the client to witness a violation of
+branch-local consistency. For example, the client may
+\C{sync} the counter, whose current local value is $10$, and obtain
+$0$ as a result, thus witnessing a violation of a counter's monotonicity
+property. Fortunately, Def.~\ref{def:mergeable-type} disallows this
+semantics. It is impossible to define the function $M$ for the counter
+data type that violates the monotonicity invariant, because no counter
+morphism violates the invariant. In general, the specification of $M$
 guarantees that the merge operation for a type $t$ preserves any
@@
 Fig.~\ref{fig:pushouts} need not be enforced statically. It can
-equally be verified at run-time by checking that the morphisms returned by $M$
-map respective concurrent versions to the same value. The
-\rulelabel{E-Pull-Wait} rule of the operational semantics
+equally be verified at run-time by checking that the morphisms
+returned by $M$ map respective concurrent versions to the same
+value. The \rulelabel{E-Pull-Wait} rule of the operational semantics
 (Fig.~\ref{fig:opsem}) can be extended to this effect:
@@
 Fig.~\ref{fig:pushouts} commutes. For more
-\begin{wrapfigure}{l}{.4\textwidth}
-\centering
-\includegraphics[scale=0.35]{Figures/pushouts-2}
-\caption{Commutative diagram for Def.~\ref{def:mergeable-type-2}.}
-\label{fig:pushouts-2}
-\end{wrapfigure}
 sophisticated data types, such as list, this process is more involved,
@@
 \end{itemize}
-The Commutativity diagram of Fig.~\ref{fig:pushouts-2} visualizes this
+The commutativity diagram of Fig.~\ref{fig:pushouts-2} visualizes this
 definition. 
@@
 called an edit sequence. Applying the edit sequence on $A$ produces
+\begin{wrapfigure}{l}{.4\textwidth}
+\centering
+\includegraphics[scale=0.35]{Figures/pushouts-2}
+\caption{Commutative diagram for Def.~\ref{def:mergeable-type-2}.}
+\label{fig:pushouts-2}
+\end{wrapfigure}
 $B$ (i.e., $\denot{P^*}(A) = B$). Second, the definition asserts the

### 2017-02-27 paper/meta.tex
Semantics.
edits

@@
 \subsection{Properties of the LCA}
+\label{sec:meta}
 
@@
 this example there is a unique LCA for 4 and 7, in general this may
+not be the case. With unrestrained branching and merging, there is no
+bound on the number of LCAs a pair of versions can have.  For example,
+in Fig.~\ref{fig:criss-cross-lcas}, the merge of 0 with 3 is preceded
+by two ``criss-cross'' merges between their respective branches
+resulting in there being two LCAs (5 and 4) for 0 and 3. 
+% Multiple LCAs can occur even without criss-cross merges, as
+% demonstrated by Fig.~\ref{fig:external-lcas}. 
+Concurrent versions with multiple LCAs do not lend themselves to
+three-way merging. If such versions represent the tips of their respective
+branches, they render the branches unmergeable (since  \C{lca} is no
+longer a function). This problem also arises in the context of source
+control systems, which employ \emph{ad hoc} mechanisms to pave the way
+for three-way merging.  GitHub~\cite{github}, for instance,
+recursively merges LCAs to compute a virtual ancestor, which then
 \begin{wrapfigure}{l}{.4\textwidth}
 \centering
-\includegraphics[scale=0.6]{Figures/merge-needs-lca}
+\includegraphics[scale=0.45]{Figures/merge-needs-lca}
 \caption{This example of a grow-only counter illustrate why \C{merge}
@@
 \end{wrapfigure}
-not be the case. With unrestrained branching and merging, there is no
-bound on the number of LCAs a pair of versions can have.  For example,
-in Fig.~\ref{fig:criss-cross-lcas}, the merge of 0 with 3 is preceded
-by two ``criss-cross'' merges between their respective branches
-resulting in there being two LCAs (5 and 4) for 0 and 3. 
-% Multiple LCAs can occur even without criss-cross merges, as
-% demonstrated by Fig.~\ref{fig:external-lcas}. 
-Concurrent versions with multiple LCAs do not lend themselves to
-three-way merging. If such versions are latest on their respective
-branches, they render the branches unmergeable (since  \C{lca} is no
-longer a function). This problem also arises in the context of source
-control systems, which employ \emph{ad hoc} mechanisms to pave the way
-for three-way merging.  GitHub~\cite{github}, for instance,
-recursively merges LCAs to compute a virtual ancestor, which then
 serves as the LCA for merging concurrent versions. This method is

### 2017-02-27 paper/semantics.tex
edits

@@
 \section{Formalization}
+\label{sec:formalization}
 
@@
 branch ($H(t)$), and extending the current branch with the merged
-version ($v_m$). The new version is tagged with a \C{MERGE} tag that,
+version ($v_m$). The versions are merged only if they are \emph{safely
+mergeable} (denoted $\under{H}{v' \mbleto v}$) - a notion that we will
+explain shorty. The new version is tagged with a \C{MERGE} tag that,
 like a \C{FORK} tag, records its origin. The rule assumes a function
@@
 \rulelabel{E-Pull-Wait} rule. The rule merges version $v'$ into
-version $v$ only if they are \emph{safely mergeable} (denoted
-$\under{H}{v' \mbleto t}$), i.e., they have a single LCA, and merging
-both versions does not lead to a case where a branch has multiple LCAs
-with rest of the branches, hence gets \emph{stuck}. Such nuances of
-LCA, and the guarantees provided by the system notwithstanding these
-nuances, are discussed in the subsection~\ref{sec:meta}.
+version $v$ only if they are safely mergeable ($\under{H}{v' \mbleto
+v}$), i.e., they have a single LCA, and merging both versions does not
+lead to a case where a branch has multiple LCAs with rest of the
+branches, hence gets \emph{stuck}. Such nuances of LCA, and the
+guarantees provided by the system notwithstanding these nuances, are
+discussed in the Sec.~\ref{sec:meta}.
 

### 2017-02-27 paper/model.tex
fix wrapfig
minor edits: missing parens and spelling fix

@@
 two concurrent versions evolved independently. The merge function can
-\begin{wrapfigure}{l}{.4\textwidth}
+\begin{wrapfigure}{L}{.4\textwidth}
 \centering
@@
     val with_init_version_do: 'a -> ('a, 'b) t -> 'b
-    val fork_version : ('a, 'b) t -> 'a unit t
+    val fork_version : ('a, 'b) t -> ('a, unit) t
     val sync_next_version: unit -> 'a -> ('a, 'a) t

### 2017-02-27 paper/meta.tex
fix wrapfig
Meta updated.

--- 2017-02-27 paper/meta.tex
@@
 in Fig.~\ref{fig:criss-cross-lcas}, the merge of 0 with 3 is preceded
-by two ``criss-cross'' merges between their respective branches
+by two ``criss-cross'' merges between their respective
+branches\footnote{
+  When discussing merges and LCAs, we often attribute the properties
+  of latest versions on branches to the branches themselves.  For
+  instance, when we say two branches merge, in fact their latest
+  versions merge. Likewise, LCA of two branches means the LCA of their
+  latest versions.
+}
 resulting in there being two LCAs (5 and 4) for 0 and 3. 
-% Multiple LCAs can occur even without criss-cross merges, as
-% demonstrated by Fig.~\ref{fig:external-lcas}. 
+Multiple LCAs can occur even without criss-cross merges, as
+demonstrated by Fig.~\ref{fig:external-lcas}. 
 Concurrent versions with multiple LCAs do not lend themselves to
@@
 branches, they render the branches unmergeable (since  \C{lca} is no
-longer a function). This problem also arises in the context of source
+longer a function) as demonstrated by examples in
+Fig.~\ref{fig:many-lcas}. Note that for both the examples in
+Fig.~\ref{fig:many-lcas}, no extension of the branching structure
+can make the branches merge again. Thus the system is effectively
+\emph{partitioned} permanently. This is clearly a problem.
+
+\begin{figure}[!t]
+\centering
+\subcaptionbox[] {\small
+  In this example, 1 and 3 have two LCAs (3 and 4) a result of
+  previous merges. The dotted circle denotes a virtual ancestor
+  obtained by merging the two LCAs.
+  \label{fig:criss-cross-lcas}
+} [0.47\columnwidth] {
+  \includegraphics[scale=0.55]{Figures/2-LCAs}
+}
+\hfill
+\subcaptionbox[] {\small
+  In this example, versions $v_{13}$ and $v_{44}$ have two LCAs
+  ($v_{22}$ and $v_{32}$)  despite there not being any previous merges
+  between their respective branches.
+  \label{fig:external-lcas}
+} [0.47\columnwidth] {
+  \includegraphics[scale=0.7]{Figures/2-external-LCAs}
+}
+\caption{Examples where two branches have more than one LCA, hence
+cannot merge. }
+\label{fig:many-lcas}
+\end{figure}
+
+The problem of multiple LCAs also arises in the context of source
 control systems, which employ \emph{ad hoc} mechanisms to pave the way
@@
 
-\begin{figure}[!t]
-\centering
-\subcaptionbox[] {\small
-  In this example, 1 and 3 have two LCAs (3 and 4) a result of
-  previous merges. The dotted circle denotes a virtual ancestor
-  obtained by merging the two LCAs.
-  \label{fig:criss-cross-lcas}
-} [0.47\columnwidth] {
-  \includegraphics[scale=0.55]{Figures/2-LCAs}
-}
-% \hfill
-% \subcaptionbox[] {\small
-%   In this example, versions $v_{13}$ and $v_{44}$ have two LCAs
-%   ($v_{22}$ and $v_{32}$)  despite there not being any previous merges
-%   between their respective branches.
-%   \label{fig:external-lcas}
-% } [0.47\columnwidth] {
-%   \includegraphics[scale=0.7]{Figures/2-external-LCAs}
-% }
-\caption{Examples where merging versions have more than one LCA}
-\label{fig:many-lcas}
-\end{figure}
-
 Fortunately, unlike source control systems where branching structure
@@
 manifest it in a way that it deems fit. In particular, \name solves
-the problem of criss-cross merges by constraining the branching
-stucture to prevent their existence. As captured in the operational
-semantics, this is done by restricting the merge operation only to the
-tips (latest versions) of the branches. For instance, suppose $v_{11}$
-and $v_{21}$ are the latest versions on branches $b_1$ and $b_2$,
-respectively. Merging $b_2$ into $b_1$ entails merging $v_{21}$ into
-$v_{11}$ to generate version $v_{12}$ on $b_1$.  Now, merging $b_1$
-into $b_2$ involves merging $v_{12}$, the latest version on $b_1$,
-into $v_{21}$, but not $v_{11}$ into $v_{21}$, thus preempting a
-criss-cross branching structure.
-% In other words, a criss-cross branching structure is prevented due
-% to the order among merges between conflicting branches introduced as
-% a result of \rulelabel{E-Pull-Wait} being an atomic step.
+the problem of multiple LCAs by suitably constraining the branching
+structure such that the problem never arises. The constraints are
+imposed either implicitly, as a result of how operational semantics
+defines an atomic step, or explicitly, by insisting that certain
+conditions be met before merging a pair of versions
+(\rulelabel{E-Pull-Wait}). Firstly, the operational semantics already
+disables criss-cross merges since it only ever merges versions that
+are latest on their respective branches. For instance, suppose
+$v_{11}$ and $v_{21}$ are the latest versions on branches $b_1$ and
+$b_2$, respectively. Merging $b_2$ into $b_1$ entails merging $v_{21}$
+into $v_{11}$ to generate version $v_{12}$ on $b_1$.  Now, merging
+$b_1$ into $b_2$ involves merging $v_{12}$, the latest version on
+$b_1$, into $v_{21}$, but not $v_{11}$ into $v_{21}$, thus preempting
+a criss-cross branching structure.
 
-% Next, for branching structures where a pair of branches (i.e., their
-% respective latest versions) have multiple externally-located LCAs, we
-% suitably extend the branching structure to generate a new single LCA.
-%  For instance, consider
-% the example in Fig.~\ref{fig:external-lcas}, where (latest versions
-% on) branches $b_1$ and $b_4$ have two externally-located LCAs
-% ($v_{22}$ and $v_{32}$), hence cannot be merged. Mergeability can
-% however be reclaimed by extending the branching structure as shown in
-% Fig.~\ref{fig:legal-extension-1}. The extension is done by first
-% merging $b_3$ into $b_2$ to create version $v_{23}$ on $b_2$, then
-% merging $b_2$ into $b_1$ to create version $v_{14}$ on $b_1$, and
-% finally merging $b_2$ into $b_4$ to create version $v_{45}$ on
-% $b_4$. Observe that $b_1$ and $b_4$ now have a single LCA ($v_{23}$)
-% hence are now mergeable. Mergeability could also have been reclaimed
-% by adopting an alternative extension to the branching structure as
-% shown in Fig.~\ref{fig:legal-extension-2}.
+Secondly, we impose certain pre-conditions on the merging branches to
+preempt the structure shown in Fig.~\ref{fig:external-lcas}. The
+intuition is as follows: consider the branch $b_3$ at the instance of
+merging $b_1$. Since it has already merged $b_2$, a version on $b_2$
+($v_{21}$) could be a common ancestor for $b_3$ and some other branch
+(call it $b$). Now, if $b_3$ merges $b_1$, same could be true of $b_3$
+and $b_1$: a version on $b_1$ ($v_{11}$) could be a common ancestor
+for $b_3$ and the other branch $v$. Since $v_{21}$ is also a common
+ancestor for both the branches, and the ancestors are not ordered by
+the ancestor relation, $b_3$ and $b$ end up with two LCAs. In
+Fig.~\ref{fig:external-lcas}, the role of $b$ is played both by $b_1$
+and $b_2$. We observe that this scenario can be prevented if, when
+merging $b_1$, $b_3$ insists on an ancestor relation between the last
+merged version ($v_{22}$ of the $b_2$), and the latest version of
+$b_1$, the currently merging branch. We call the last merged version
+the \emph{external locus} of the branch. By requiring that, for every
+branch $b$, the external locii of $b$ at various points in time (i.e.,
+external locus of every prefix of $b$) be totally ordered, and given
+that the semantics already preempts criss-cross merges, we effectively
+avoid the case of there ever being more than 2 LCAs for a pair of
+branches. Thus, uniqueness of LCAs is thus guaranteed.
 
-\begin{figure}[!t]
-\centering
-\subcaptionbox[] {\small
-  In this example, versions $v_{13}$ and $v_{44}$ have two LCAs
-  ($v_{22}$ and $v_{32}$) rendering their respective branches
-  unmergeable.
-  \label{fig:external-lcas}
-} [0.47\columnwidth] {
-  \includegraphics[scale=0.7]{Figures/2-external-LCAs}
-}
-\hfill
-\subcaptionbox[] {\small
-  However, execution can still progress to the depicted state, where
-  $b_1$ and $b_4$ are again mergeable.
-  \label{fig:legal-extension-1}
-} [0.47\columnwidth] {
-  \includegraphics[scale=0.8]{Figures/legal-extension-1}
-}
-% \hfill
-% \subcaptionbox[] {\small
-%   $b_2$ is merged into $b_3$ to create $v_{33}$ here.
-%   \label{fig:legal-extension-2}
-% } [0.47\columnwidth] {
-%   \includegraphics[scale=0.8]{Figures/legal-extension-2}
-% }
-\caption{Multiple LCAs without criss-cross merges do not threaten
-the progress guarantee.}
-\label{fig:legal-extensions}
-\end{figure}
+We now formalize the aforementioned intuitions via a series of
+definitions, and theorems.
 
-There can be other scenarios that lead to multiple LCAs, such as the
-one shown in Fig.~\ref{fig:external-lcas}. However, preemption of
-criss-cross merges is sufficient to make progress despite multiple
-LCAs. For instance, the branching history in
-Fig.~\ref{fig:external-lcas}, where $b_1$ and $b_4$ cannot merge due
-to multiple LCAs, can still make progress to the history shown in
-~\ref{fig:legal-extension-1}, where $b_1$ and $b_4$ are now mergeable.
-We formalize the progress guarantee in the following theorem, whose
-proof is included in the appendix.
+\begin{definition} [\bfseries Internal and External Ancestors]
+Given a branch $b$ and a version $v\in b$, an internal ancestor of $v$
+is an ancestor from the same branch $b$. An external ancestor of $v$
+is an ancestor from a different branch $b'\neq b$. 
+\end{definition}
 
-\begin{theorem} [\bfseries Progress]
+\begin{definition} [\bfseries External Locus]
+Given a branch $b$ and a version $v\in b$, external locus ($v_o$) of
+$v$ is an external ancestor that is not an ancestor of any other
+external ancestor of $v$. That is, $\under{H}{v_o \preceq v}$, and
+there does not exist a $v_o' \not\in b$ such that $\under{H}{v_o'
+\preceq v}$ and $\under{H}{v_o \preceq v_o'}$. 
+\end{definition}
+
+Note that internal ancestor relation ($\preceq_i$) is, by definition,
+a total order. We later prove that the external ancestor relation
+($\preceq_o$) is also a total order. Hence, for any given version $v$,
+least internal and external ancestors, if exist, must be unique.  We
+call the unique least internal ancestor as its \emph{internal locus},
+and the unique least external ancestor version as its \emph{external
+locus}.
+
+\begin{definition} [\bfseries Mergeability]
+\label{def:mergeability}
+Given a history $H$, a version $v_1$ and a version $v_2$ that is not
+an ancestor of $v_1$ under $H$, $v_2$ is mergeable into $v_1$ (denoted
+$\under{H}{v_2 \mbleto v_1}$) iff $v_1$'s external locus is an
+ancestor of $v_2$.
+\end{definition}
+
+\begin{lemma} [\bfseries Unique External Locus]
+Every version $v$ (except the \C{INIT} version) in the history $H$
+generated by the rules in Fig.~\ref{fig:opsem} has a unique external
+locus.
+\end{lemma}
+
+\begin{lemma} [\bfseries Unique LCA]
+Every pair of branches in the history $H$ generated by the rules in
+Fig.~\ref{fig:opsem} have a unique least common ancestor. 
+\end{lemma}
+
+\begin{lemma} [\bfseries Non-Monotonicity of Mergeability]
 In a legal branching history $H$ produced by the operational
-semantics, there exist at least two branches that have a unique LCA,
-hence mergeable.
+semantics, if two branches, $b_i$ and $b_j$ are not mergeable (as per
+Def.~\ref{def:mergeability}), then there exist a series of merges
+involving only the mergeable branches that can be performed on $H$ to
+yeild a new history $H'$, where $b_i$ and $b_j$ are now mergeable.
+\end{lemma}
+
+\begin{theorem} [\bfseries Progress and Eventual Convergence] Given a
+legal branching history $H$, and a program $p$, where every thread
+expression in $p$ is a \C{pull}, there exist a series of reduction steps
+that reduce $p; H$ to $p'; H'$, where every thread expression in $p$
+is the same value $v$.
 \end{theorem}

### 2017-02-27 paper/semantics.tex
Meta updated.

@@
 v}$), i.e., they have a single LCA, and merging both versions does not
-lead to a case where a branch has multiple LCAs with rest of the
-branches, hence gets \emph{stuck}. Such nuances of LCA, and the
-guarantees provided by the system notwithstanding these nuances, are
-discussed in the Sec.~\ref{sec:meta}.
+lead to a case where a branch (i.e., its latest version) has multiple
+LCAs with rest of the branches, hence gets \emph{stuck}. Such nuances
+of LCA, and the guarantees provided by the system notwithstanding
+these nuances, are discussed in the Sec.~\ref{sec:meta}.
 

### 2017-02-27 paper/intro.tex
edits
fix wrapfig

@@
 counters, a \C{read} operation performed at another replica may
-\begin{wrapfigure}{l}{.5\textwidth}
+\begin{wrapfigure}{L}{.5\textwidth}
   \begin{center}
@@
     \item We present a formalization of the \name programming model,
-      and establish the comditions under which eventual convergence of
+      and establish the conditions under which eventual convergence of
       concurrent version state can be guaranteed.
@@
 incorporate mergeable types into OCaml are presented in
-Sec.~\ref{sec:impl}.  Experimental results and benchmarks are given
-in Sec.~\ref{sec:evaluation}.
-
-, along with
-a series of examples illustrating how merge operations can be defined
-and used.  Sec.~\ref{sec:
+Sec.~\ref{sec:impl}.  Experimental results and benchmarks are given in
+Sec.~\ref{sec:evaluation}.  Secs.~\ref{sec:related}
+and~\ref{sec:conclusions} describes related work and conclusions.
 

### 2017-02-28 paper/examples.tex
fix wrapfig
minor edits: spelling fixes

@@
 
-A rope is a data structure that faciliates efficient implementations
+A rope is a data structure that facilitates efficient implementations
 of operations such as concatenation, splitting, lookup and insertion
@@
 
-One feature of the mergeable rope implementation that is characterisic
+One feature of the mergeable rope implementation that is characteristic
 of mergeable types in general is compositionality; a rope is a
@@
 (\C{Qty}) that supports \C{add} and \C{subtract} operations, but
-\begin{wrapfigure}{l}{.45\textwidth}
+\begin{wrapfigure}{L}{.45\textwidth}
   \centering
@@
 \end{wrapfigure}
-defines its \C{merge} opeation a little differently:
+defines its \C{merge} operation a little differently:
 \begin{ocaml}
@@
 specified quantity, deleting existing items, and increasing or
-\begin{wrapfigure}{l}{.5\textwidth}
+\begin{wrapfigure}{L}{.5\textwidth}
   \centering

### 2017-02-28 paper/appendix.tex

\section{Appendix}
\input{list-code}
\begin{lemma} [\bfseries Unique External Locus]
Every version $v$ (except the \C{INIT} version) in the history $H$
generated by the rules in Fig.~\ref{fig:opsem} has a unique external
locus.
\end{lemma}
\begin{proof}
By case analysis on the reduction relation. In each case, we assume
that the unique external locus property holds for the initial history,
and prove that it holds for the final history.

In \rulelabel{E-Run} case, initial version is the unique external locus of
the new version on the new branch $H(t)$. In \rulelabel{E-Fork} case,
which creates the branch $H(t')$ by forking from $H(t)$, the latest
version on $H(t)$ is the external locus of the first version on
$H(t)$. In \rulelabel{E-Push} case, the version on $H(t)$ has the same
external locus as the previous latest version, which, premise says, is
unique. No new version is created in \rulelabel{E-Pull}. In
\rulelabel{E-Pull-Wait}, new version is created by merging the latest
version ($v'$) on $H(t')$ into the latest version $v$ on $H(t)$.
Premise says that $v$ has a unique external locus (call it $v_o$). The
new version $v_m$ inherits $v_o$ as a least external ancestor, and
also has $v'$ as another least external ancestor. However, since
$\under{H}{v' \mbleto v}$, we have $\under{H}{v_o \preceq v'}$, we
have $v'$ as the the external locus.
\qed
\end{proof}

\begin{lemma} [\bfseries Unique LCA]
Every pair of versions $v_1$ and $v_2$ in the history $H$ generated by
the rules in Fig.~\ref{fig:opsem} have a unique least
common ancestor. 
\end{lemma}
\begin{proof}
By case analysis on the step relation. In each case, we assume that
the unique LCA property holds for the initial history, and prove that
it holds for the final history.

The \rulelabel{E-Run} case is trivial. The \rulelabel{E-Push} case
creates a new version ($v$) with \C{PUSH} tag at the end of a branch.
Let $v_1$ be the previous version on the branch, and let $v_2$ be any
other version. The premise is that $v_1$ and $v_2$ have a unique LCA
$v_0$. Observe that $v$'s internal locus is $v_1$ and its external
locus is same as that of $v_1$. Hence, all ancestors of $v_1$
(including itself) are also ancestors of $v$. Moreover, $v$ has no new
ancestors except itself, which is the latest version being added.
Hence, any common ancestor of $v_1$ and $v_2$ is also a common
ancestor of $v$ and $v_2$, while there cannot be new common ancestors.
Hence LCA of $v$ and $v_2$ is unique. The reasoning for
\rulelabel{E-Fork} is similar. The \rulelabel{E-Pull} case is trivial.

The \rulelabel{E-Pull-Wait} case is where the merge happens. Version
$v'$ is merging into version $v$ to produce the next version $v_m$.
Let $v_2$ be any version. From the premise, we know that $v$ and $v_2$
have a unique LCA (call it $v_0$), and $v'$ and $v_2$ have a unique
LCA (call it $v_1$). But, we know that $v_2$ has a unique external
locus (call it $v_3$), hence we know that 
\begin{itemize}
  \item Both $v_0$ and $v_1$ are external ancestors of $v_2$. Hence,
  $\under{H}{v_0 \preceq v_3} \conj \under{H}{v_1 \preceq v_3}$.
  Since $\under{H}{v_0 \preceq v_m} \conj \under{H}{v_1 \preceq v_m}$, 
  $v_2$ and $v_m$ have a unique LCA.

  \item $v_0$ is an internal ancestor of $v_2$, hence an external
  ancestor of $v$, hence an external ancestor of $v_m$. $v_1$ is
  an ancestor of $v'$, hence also an external ancestor of $v_m$. 
\end{itemize}
\end{proof}

### 2017-02-28 paper/intro.tex

@@
 heavyweight mechanisms, applying an \C{Add} operation on one node may
-not be instanteously witnessed on another, which may be in the process of
+not be instantaneously witnessed on another, which may be in the process of
 simultaneously attempting to perform its own \C{Add} or \C{Mult}
@@
 the negative performance impact it incurs in geo-distributed
-environments~\cite{Walter}.  Removig coordination altogether by simply
+environments~\cite{Walter}.  Removing coordination altogether by simply
 replicating counter state without having any supporting consistency
@@
 by the application when updating $c_2$ or $c_1$.  An execution may
-nonetheless witness anamolous states that violate the invariant
+nonetheless witness anomalous states that violate the invariant
 because updates to $c_1$ and $c_2$ may be applied independently in any
@@
 datatype is thus given in terms of the evolution of a program state
-impicitly associated with the different nodes that comprise a
+implicitly associated with the different nodes that comprise a
 distributed application with merge operations serving to communicate
@@
 Sec.~\ref{sec:formalization} defines an operational semantics, and
-formalizes our intution of a mergeable type in terms of morphisms over
+formalizes our intuition of a mergeable type in terms of morphisms over
 datatype operations.  We also present soundness (consistency) and

### 2017-02-28 paper/meta.tex

@@
 involving only the mergeable branches that can be performed on $H$ to
-yeild a new history $H'$, where $b_i$ and $b_j$ are now mergeable.
+yield a new history $H'$, where $b_i$ and $b_j$ are now mergeable.
 \end{lemma}

### 2017-02-28 paper/motivation.tex
edits

@@
 
-In this section, we motivate mergeable data types and \name
+In this section, we motivate mergeable data types and the \name
 programming model via a series of examples.
@@
 identifies the operation to be applied uniformly at all
-replicas to incorporate the effects of the original operaton. For
+replicas to incorporate the effects of the original operation. For
 instance, \C{Counter.add x v} operation returns \C{Add x} effect,
@@
 enforces when updating $c_2$ and $c_1$. Nonetheless, it may still
-witness anamolous states that violate the invariant because updates to
+witness anomalous states that violate the invariant because updates to
 $c_1$ and $c_2$ are applied independently in any order. For example,
@@
 and the possibility of new bugs. Thus, adding persistence to
-applications while preserving high-level safety gurantees offered by
+applications while preserving high-level safety guarantees offered by
 the language abstraction is a non-trivial problem.

### 2017-02-28 paper/semantics.tex

@@
 distinct. The new version is tagged with a \C{FORK} tag that keeps the
-record of its orgin, namely the \C{fork} operation and the branch from
+record of its origin, namely the \C{fork} operation and the branch from
 which the current branch is forked. The \rulelabel{E-Fork} rule forks

### 2017-02-28 paper/evaluation.tex
Semantics.
edits

--- 2017-02-27 paper/evaluation.tex
@@
 
-In this section, we present an evaluation of \name programming model with the
-aim of assessing its practical utility for describing geo-distributed
-eventually consistent applications. In particular, we are interested in the
-horizontal scalability of applications built using \name. Typical distributed
-eventually consistent data stores use custom synchronization and dissemination
-protocols for transferring updates between replicas. Unlike this, \name is
-realized over Irmin which uses Git transfer protocol~\cite{..}, a protocol not designed
-with distributed data stores in mind. Hence, we also evaluate the performance
-of synchronization and merging.
+In this section, we present an evaluation of \name programming model
+with the aim of assessing its practical utility for describing
+geo-distributed eventually consistent applications. In particular, we
+are interested in the horizontal scalability of applications built
+using \name. Typical distributed eventually consistent data stores use
+custom synchronization and dissemination protocols for transferring
+updates between replicas. In contrast, \name is realized over Irmin~\cite{Irmin}
+which uses Git transfer protocol~\cite{..}, a protocol not designed
+with distributed data stores in mind. Hence, we also evaluate the
+performance of synchronization and merging.
 
@@
 
-In order to evaluate this, we have implemented a collaborative editing
-application that simulates concurrent editing of the same document by several
-authors. This application is not unlike Google Docs, but also differs from it
-by the fact that we do not need a central server that coordinates the edits.
-Instead, edits from each author is asynchronously sent to other authors. The
-shared document itself is represented by a mergeable rope of characters
-(Section~\ref{sec:model}), and hence, the remote edits can be correctly
-reconciled into the local document. We resolve conflicts on concurrent
-substitutions at the same position (such as two authors concurrently editing
-\C{"abc"} to \C{"xbc"} and \C{"ybc"}, respectively) by choosing the character
-with the smaller ASCII code. For the benchmarks, the initial document that we
-use has 1576 characters. The workload consists of 4000 edit operations at
-random indices with 85\% insertions and 15\% deletions.
+To evaluate the effectiveness of our model, we have implemented a
+collaborative editing application that simulates concurrent editing of
+the same document by several authors. This application is not unlike
+Google Docs, but differs from it by the fact that we do not need a
+central server that coordinates the edits.  Instead, edits from each
+author is asynchronously sent to other authors. The shared document
+itself is represented by a mergeable rope of characters
+(Section~\ref{sec:model}), and hence, the remote edits can be
+correctly reconciled into each local document. We resolve conflicts on
+concurrent substitutions at the same position (such as two authors
+concurrently editing \C{"abc"} to \C{"xbc"} and \C{"ybc"},
+respectively) by choosing the character with the smaller ASCII
+code. For the benchmarks, the initial document that we use has 1576
+characters. The workload consists of 4000 edit operations at random
+indices with 85\% insertions and 15\% deletions.
 
@@
 		\includegraphics[width=0.9\textwidth]{Figures/cluster.pdf}
-		\caption{8-node ring cluster on Google Cloud Platform used for
-			benchmarking. US-C, EU-W, ASIA-NE and ASIA-E are availability zones
-			us-central1-b, europe-west1-b, asia-northeast1-b and asia-east1-b
-			respectively. Edge labels are inter-node latencies in milliseconds.}
+		\caption{Our experimental configuration consists of an
+                  8-node ring cluster on executing on a Google Cloud
+                  Platform. US-C, EU-W, ASIA-NE and ASIA-E are
+                  availability zones us-central1-b, europe-west1-b,
+                  asia-northeast1-b and asia-east1-b
+                  respectively. Edge labels are inter-node latencies
+                  in milliseconds.}
 		\label{fig:cluster}
@@
 
-For our experiments, we instantiate an 8-node geo-distributed cluster in Google
-Cloud Platform~\cite{gcp}. Each node is an \C{n1-standard-4} instance with 2
-virtual CPUs and 7.5 GB of memory. The nodes are arranged in a ring as shown in
-Figure~\ref{fig:cluster} and synchronize (push and pull updates) when necessary
-with its successor. While \name does not impose any restrictions on
-synchronizing with other nodes, we choose to only synchronize with the
-successor since we are guaranteed eventual convergence with a ring cluster. If
-the successor becomes unavailable, the node synchronizes with other available
-nodes. Operations are generated and applied locally on each node before they
-are propagated eventually to other nodes with the help of a background thread
-that synchronizes with its successor every second.
+For our experiments, we instantiate an 8-node geo-distributed cluster
+using the Google Cloud Platform~\cite{gcp}. Each node is an
+\C{n1-standard-4} instance with 2 virtual CPUs and 7.5 GB of
+memory. The nodes are arranged in a ring as shown in
+Figure~\ref{fig:cluster} and synchronize (push and pull updates) when
+necessary with its successor. While \name does not impose any
+restrictions on synchronizing with other nodes, we choose to only
+synchronize with the successor since we are guaranteed eventual
+convergence with a ring cluster. If the successor becomes unavailable,
+the node synchronizes with other available nodes. Operations are
+generated and applied locally on each node before they are propagated
+eventually to other nodes with the help of a background thread that
+synchronizes with its successor every second.
 
@@
 we consider the nodes 1 (\C{us-central1-b}) and 2 (\C{europe-west1-b}). Both
-nodes start with the initial document and perform \C{n} concurrent edits
+nodes start with the initial document and perform $n$ concurrent edits
 without synchronization. After this, the nodes synchronize pushing the local
 edits and pulling the remote edits, and merge to produce the final document. In
-this case, we indicate the distance of synchronization as \C{2n}.
+this case, we indicate the distance of synchronization as $2n$.
 Figure~\ref{grf:merge} presents the time share between network and computation
@@
 
-More importantly, the cost of computation grows sub-linearly. This is because
-of efficient merge function on ropes~\ref{ropes} enabled by Irmin's
-content-addressed storage backend, and more expensive operational
-transformation of merging concurrent edits to leaf nodes~\ref{lists} is invoked
-rarely on small. On average, we measured that at a distance of 256, for merging
-the two documents with 1743 characters, 50 list merges were performed on
-strings of average length 13. A purely list based approach would require the
-polynomial time edit sequence algorithm to be applied on the entire document
-twice. Compare to this, the rope implementation has much better algorithmic
-complexity. We measured that by representing the entire document using
-mergeable list type, in addition to slowing down each edit operation, the merge
-at a distance of 256 was 2.3 times slower than the rope implementation.
+More importantly, the cost of computation grows sub-linearly. This is
+because of efficient merge function on ropes (Sec.~\ref{sec:ropes})
+enabled by Irmin's content-addressed storage backend, and more
+expensive operational transformation of merging concurrent edits to
+leaf nodes (Sec.~\ref{sec:lists}) is invoked rarely on small. On
+average, we measured that with a distance of 256, for merging the two
+documents with 1743 characters, 50 list merges were performed on
+strings of average length 13. A purely list based approach would
+require a polynomial time edit sequence algorithm to be applied on the
+entire document twice. Compared to this, the rope implementation has
+much better algorithmic complexity.  While not depicted here, we
+measured that by representing the entire document using a mergeable
+list type, in addition to slowing down each edit operation, the merge
+at a distance of 256 was 2.3 times slower than the rope
+implementation.  We conclude that versioning and merging as found in
+\name\ can yield substantial throughput and latency gains with low
+computational cost on modern eventually-consistent cloud systems.

### 2017-02-28 paper/examples.tex
edits

@@
 		transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The function
-		\C{op\_transform} (Fig.~\ref{fig:mlist-sig}) computes such
+		\C{op\_transform} (shown in the Appendix) computes such
 		\emph{operational transformations} for an edit sequence $s_1$ w.r.t another
@@
 \C{Int} types (i.e., \C{MList(Int)}), where \C{Int.merge} is the same
-as \C{Counter.merge} (Sec.~\ref{sec:intro}).  An alternative
+as \C{Counter.merge} described in the introduction.  An alternative
 instantiation of \C{MList} is for a quantity type
-(\C{Qty}) that supports \C{add} and \C{subtract} operations, but
-\begin{wrapfigure}{L}{.45\textwidth}
+(\C{Qty}) that supports \C{add} and \C{subtract} operations, and
+defines its \C{merge} operation a little differently:
+\begin{wrapfigure}{l}{.45\textwidth}
   \centering
@@
   \label{fig:list-eg}
-  \vspace*{-.2in}
+  \vspace*{-.45in}
 \end{wrapfigure}
-defines its \C{merge} operation a little differently:
 \begin{ocaml}
@@
 
+A non-trivial demonstration illustrating the compositional power of
+mergeable types is a shopping list application that allows its users
+to collaboratively build a shopping list (listing items in the
+decreasing order of their priority), by adding new items with a
+specified quantity, deleting existing items, and increasing or
+decreasing their quantity. An item (\C{Item.t}) is represented as a
+tuple of its name (a string) and the quantity (A \C{Qty.t}).  The
+merge function for \C{Item.t} merges the quantities of items (using
+\C{Qty.merge}) having the same names. Items with different names are
+considered distinct. A shopping list is a mergeable list of items,
+\begin{wrapfigure}{l}{.5\textwidth}
+  \centering
+  \includegraphics[scale=0.3]{Figures/shopping-list}
+  \caption{Merging concurrent shopping lists}
+  \label{fig:shopping-list}
+  \vspace*{-.35in}
+\end{wrapfigure}
+i.e., \C{MList(Item)} (a grocery shopping list may not be large enough
+to justify using ropes, but \C{MRope(MList(Item))} is certainly
+allowed).  \C{MList(Item).merge} automatically lifts the merge
+semantics for items to a merge semantics for lists of items.
+Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
+users, Alice and Bob, concurrently edit a shopping list whose initial
+contents are \C{milk} and \C{eggs}. While both simultaneously update
+the quantity of eggs, Bob also removes \C{milk} and inserts \C{candy}.
+Fig.~\ref{fig:shopping-list-code} shows the \name code for two
+concurrent sessions. The merged shopping list contains \C{eggs} and
+\C{candy}, where the quantity of eggs is obtained by merging Alice's
+and Bob's quantity as per the definition of \C{Qty.merge}.
+The examples described above make certain assumptions about the
+concurrency model and the underlying system, such as the existence of
+a single least common ancestor (LCA) for any pair of versions, the
+ability to access a previous version on any branch, and mergeability
+of any two concurrent versions. Enforcing these guarantees in a fully
+decentralized distributed setting requires addressing non-trivial
+theoretical challenges. These, and other details that contribute to
+the practicality of our model, such as containing the complexity of
+merges, are discussed in subsequent sections.
+
 \begin{figure}
@@
 \end{figure}
-
-A non-trivial demonstration illustrating the compositional power of
-mergeable types is shopping list application, that allows its users to
-collaboratively build a shopping list (listing items in the decreasing
-order of their priority), by adding new items with a
-specified quantity, deleting existing items, and increasing or
-\begin{wrapfigure}{L}{.5\textwidth}
-  \centering
-  \includegraphics[scale=0.3]{Figures/shopping-list}
-  \caption{Merging concurrent shopping lists}
-  \label{fig:shopping-list}
-  \vspace*{-.2in}
-\end{wrapfigure}
-decreasing their quantity. An item (\C{Item.t}) is represented as a
-tuple of its name (a string) and the quantity (A \C{Qty.t}). The merge
-function for \C{Item.t} merges the quantities of items (using
-\C{Qty.merge}) having the same names. Items with different names are
-considered distinct. A shopping list is a mergeable list of items,
-i.e., \C{MList(Item)} (a grocery shopping list may not be large enough
-to justify using ropes, but \C{MRope(MList(Item))} is definitely a
-possibility).  \C{MList(Item).merge} automatically lifts the merge
-semantics for items to a merge semantics for lists of items.
-Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
-users, Alice and Bob, concurrently edit a shopping list whose initial
-contents are \C{milk} and \C{eggs}. While both simultaneously update
-the quantity of eggs, Bob also removes \C{milk} and inserts \C{candy}.
-Fig.~\ref{fig:shopping-list-code} shows the \name code for two
-concurrent sessions. The merged shopping list contains \C{eggs} and
-\C{candy}, where the quantity of eggs is obtained by merging Alice's
-and Bob's quantity as per the definition of \C{Qty.merge}.
-
-The examples described above make certain assumptions about the
-concurrency model and the underlying system, such as the existence of
-a single least common ancestor (LCA) for any pair of versions, the
-ability to access a previous version on any branch, and mergeability
-of any two concurrent versions. Enforcing these guarantees in a fully
-decentralized distributed setting requires addressing non-trivial
-theoretical challenges. These, and other details that contribute to
-the practicality of our model, such as containing the complexity of
-merges, are discussed in subsequent sections.

### 2017-02-28 paper/meta.tex
meta

--- 2017-02-27 paper/meta.tex
@@
 
-We now formalize the aforementioned intuitions via a series of
-definitions, and theorems.
+We now formalize the notion of mergeability, and the guarantees
+provided by the operational semantics.
 
 \begin{definition} [\bfseries Internal and External Ancestors]
-Given a branch $b$ and a version $v\in b$, an internal ancestor of $v$
-is an ancestor from the same branch $b$. An external ancestor of $v$
-is an ancestor from a different branch $b'\neq b$. 
+Given a branch $b$ and a version $v\in b$, an internal ancestor
+($\preceq_i$) of $v$ is an ancestor from the same branch $b$. An
+external ancestor ($\preceq_o$) of $v$ is an ancestor from a different
+branch $b'\neq b$. 
 \end{definition}
 
-\begin{definition} [\bfseries External Locus]
-Given a branch $b$ and a version $v\in b$, external locus ($v_o$) of
-$v$ is an external ancestor that is not an ancestor of any other
-external ancestor of $v$. That is, $\under{H}{v_o \preceq v}$, and
-there does not exist a $v_o' \not\in b$ such that $\under{H}{v_o'
-\preceq v}$ and $\under{H}{v_o \preceq v_o'}$. 
+\begin{definition} [\bfseries Locus]
+Given a branch $b$ and a version $v\in b$, locus ($v_o$) of $v$ is an
+ancestor that is not an ancestor of any other external ancestor of
+$v$. That is, $\under{H}{v_o \preceq v}$, and there does not exist a
+$v_o' \not\in b$ such that $\under{H}{v_o' \preceq v}$ and
+$\under{H}{v_o \preceq v_o'}$. We write $L(v)$ to denote the locus
+of version $v$.
 \end{definition}
 
-Note that internal ancestor relation ($\preceq_i$) is, by definition,
-a total order. We later prove that the external ancestor relation
-($\preceq_o$) is also a total order. Hence, for any given version $v$,
-least internal and external ancestors, if exist, must be unique.  We
-call the unique least internal ancestor as its \emph{internal locus},
-and the unique least external ancestor version as its \emph{external
-locus}.
+In a legal branching history, every version has a unique
+locus\footnote{lemma and proof in the appendix}. Thus, locus of a
+version is the maxima of its causal history. As usual, the locus of
+the latest version on a branch is called the locus of the branch.
 
@@
 an ancestor of $v_1$ under $H$, $v_2$ is mergeable into $v_1$ (denoted
-$\under{H}{v_2 \mbleto v_1}$) iff $v_1$'s external locus is an
-ancestor of $v_2$.
+$\under{H}{v_2 \mbleto v_1}$) if and only if $L(v_1)$ is an ancestor
+of $v_2$, and no internal ancestor of $v_1$ that is not also an
+ancestor of $v_o$ is an ancestor of $v_2$. 
+If $v_1$ and $v_2$ are the latest versions on their respective
+branches $b_1$ and $b_2$, we say $b_2$ is mergeable into $b_1$, or
+more generally, $b_1$ and $b_2$ are mergeable.  
 \end{definition}
 
-\begin{lemma} [\bfseries Unique External Locus]
-Every version $v$ (except the \C{INIT} version) in the history $H$
-generated by the rules in Fig.~\ref{fig:opsem} has a unique external
-locus.
-\end{lemma}
+The following theorem asserts that every pair of branches deemed
+mergeable by the above definition are infact mergeable, i.e., they are
+guaranteed to have a unique LCA.
 
 \begin{lemma} [\bfseries Unique LCA]
-Every pair of branches in the history $H$ generated by the rules in
-Fig.~\ref{fig:opsem} have a unique least common ancestor. 
+Every pair of branches in a legal history $H$, if mergeable as per
+Def.~\ref{fig:mergeability}, have a unique least common ancestor.
 \end{lemma}
@@
 
+Thus, a pair of branches that are not currently mergeable (as per
+Def.~\ref{fig:mergeability}), will become mergeable as the branching
+history is extended.
+
 \begin{theorem} [\bfseries Progress and Eventual Convergence] Given a

### 2017-02-28 paper/evaluation.tex

@@
 
-In this section, we present an evaluation of \name programming model
-with the aim of assessing its practical utility for describing
-geo-distributed eventually consistent applications. In particular, we
-are interested in the horizontal scalability of applications built
-using \name. Typical distributed eventually consistent data stores use
-custom synchronization and dissemination protocols for transferring
-updates between replicas. In contrast, \name is realized over Irmin~\cite{Irmin}
-which uses Git transfer protocol~\cite{..}, a protocol not designed
-with distributed data stores in mind. Hence, we also evaluate the
-performance of synchronization and merging.
+In this section, we present an evaluation of \name programming model with the
+aim of assessing its practical utility for describing geo-distributed
+eventually consistent applications. In particular, we are interested in the
+horizontal scalability of applications built using \name. Typical distributed
+eventually consistent data stores use custom synchronization and dissemination
+protocols for transferring updates between replicas. In contrast, \name is
+realized over Irmin~\cite{irmin} which uses Git transfer protocol~\cite{..}, a
+protocol not designed with distributed data stores in mind. Hence, we also
+evaluate the performance of synchronization and merging.
 
@@
 
-To evaluate the effectiveness of our model, we have implemented a
-collaborative editing application that simulates concurrent editing of
-the same document by several authors. This application is not unlike
-Google Docs, but differs from it by the fact that we do not need a
-central server that coordinates the edits.  Instead, edits from each
-author is asynchronously sent to other authors. The shared document
-itself is represented by a mergeable rope of characters
-(Section~\ref{sec:model}), and hence, the remote edits can be
-correctly reconciled into each local document. We resolve conflicts on
-concurrent substitutions at the same position (such as two authors
-concurrently editing \C{"abc"} to \C{"xbc"} and \C{"ybc"},
-respectively) by choosing the character with the smaller ASCII
-code. For the benchmarks, the initial document that we use has 1576
-characters. The workload consists of 4000 edit operations at random
-indices with 85\% insertions and 15\% deletions.
+To evaluate the effectiveness of our model, we have implemented a collaborative
+editing application that simulates concurrent editing of the same document by
+several authors. This application is not unlike Google Docs, but differs from
+it by the fact that we do not need a central server that coordinates the edits.
+Instead, edits from each author is asynchronously sent to other authors. The
+shared document itself is represented by a mergeable rope of characters
+(Section~\ref{sec:model}), and hence, the remote edits can be correctly
+reconciled into each local document. We resolve conflicts on concurrent
+substitutions at the same position (such as two authors concurrently editing
+\C{"abc"} to \C{"xbc"} and \C{"ybc"}, respectively) by choosing the character
+with the smaller ASCII code. For the benchmarks, the initial document that we
+use has 1576 characters. The workload consists of 4000 edit operations at
+random indices with 85\% insertions and 15\% deletions.
 
@@
 		\includegraphics[width=0.9\textwidth]{Figures/cluster.pdf}
-		\caption{Our experimental configuration consists of an
-                  8-node ring cluster on executing on a Google Cloud
-                  Platform. US-C, EU-W, ASIA-NE and ASIA-E are
-                  availability zones us-central1-b, europe-west1-b,
-                  asia-northeast1-b and asia-east1-b
-                  respectively. Edge labels are inter-node latencies
-                  in milliseconds.}
+		\caption{Our experimental configuration consists of an 8-node ring cluster
+			on executing on a Google Cloud Platform. US-C, EU-W, ASIA-NE and ASIA-E
+			are availability zones us-central1-b, europe-west1-b, asia-northeast1-b
+			and asia-east1-b respectively. Edge labels are inter-node latencies in
+		milliseconds.}
 		\label{fig:cluster}
@@
 
-For our experiments, we instantiate an 8-node geo-distributed cluster
-using the Google Cloud Platform~\cite{gcp}. Each node is an
-\C{n1-standard-4} instance with 2 virtual CPUs and 7.5 GB of
-memory. The nodes are arranged in a ring as shown in
-Figure~\ref{fig:cluster} and synchronize (push and pull updates) when
-necessary with its successor. While \name does not impose any
-restrictions on synchronizing with other nodes, we choose to only
-synchronize with the successor since we are guaranteed eventual
-convergence with a ring cluster. If the successor becomes unavailable,
-the node synchronizes with other available nodes. Operations are
-generated and applied locally on each node before they are propagated
-eventually to other nodes with the help of a background thread that
-synchronizes with its successor every second.
+For our experiments, we instantiate an 8-node geo-distributed cluster using the
+Google Cloud Platform~\cite{gcp}. Each node is an \C{n1-standard-4} instance
+with 2 virtual CPUs and 7.5 GB of memory. The nodes are arranged in a ring as
+shown in Figure~\ref{fig:cluster} and synchronize (push and pull updates) when
+necessary with its successor. While \name does not impose any restrictions on
+synchronizing with other nodes, we choose to only synchronize with the
+successor since we are guaranteed eventual convergence with a ring cluster. If
+the successor becomes unavailable, the node synchronizes with other available
+nodes. Operations are generated and applied locally on each node before they
+are propagated eventually to other nodes with the help of a background thread
+that synchronizes with its successor every second.
 
@@
 we consider the nodes 1 (\C{us-central1-b}) and 2 (\C{europe-west1-b}). Both
-nodes start with the initial document and perform $n$ concurrent edits
-without synchronization. After this, the nodes synchronize pushing the local
-edits and pulling the remote edits, and merge to produce the final document. In
-this case, we indicate the distance of synchronization as $2n$.
+nodes start with the initial document and perform $n$ concurrent edits without
+synchronization. After this, the nodes synchronize pushing the local edits and
+pulling the remote edits, and merge to produce the final document. In this
+case, we indicate the distance of synchronization as $2n$.
 Figure~\ref{grf:merge} presents the time share between network and computation
@@
 remote branches. We see that the merge time is dominated by the network
-overhead. This is because of the transfer protocol involving multiple
-round trips and the need to transfer objects not only for application objects
-but also Git internal objects for history, directory structure and branches. We
+overhead. This is because of the transfer protocol involving multiple round
+trips and the need to transfer objects not only for application objects but
+also Git internal objects for history, directory structure and branches. We
 observe that we can improve the network performance by implementing a streaming
@@
 
-More importantly, the cost of computation grows sub-linearly. This is
-because of efficient merge function on ropes (Sec.~\ref{sec:ropes})
-enabled by Irmin's content-addressed storage backend, and more
-expensive operational transformation of merging concurrent edits to
-leaf nodes (Sec.~\ref{sec:lists}) is invoked rarely on small. On
-average, we measured that with a distance of 256, for merging the two
-documents with 1743 characters, 50 list merges were performed on
-strings of average length 13. A purely list based approach would
-require a polynomial time edit sequence algorithm to be applied on the
-entire document twice. Compared to this, the rope implementation has
-much better algorithmic complexity.  While not depicted here, we
-measured that by representing the entire document using a mergeable
-list type, in addition to slowing down each edit operation, the merge
-at a distance of 256 was 2.3 times slower than the rope
-implementation.  We conclude that versioning and merging as found in
-\name\ can yield substantial throughput and latency gains with low
+More importantly, the cost of computation grows sub-linearly. This is because
+of efficient merge function on ropes (Sec.~\ref{sec:ropes}) enabled by Irmin's
+content-addressed storage backend, and more expensive operational
+transformation of merging concurrent edits to leaf nodes (Sec.~\ref{sec:lists})
+is invoked rarely on small. On average, we measured that with a distance of
+256, for merging the two documents with 1743 characters, 50 list merges were
+performed on strings of average length 13. A purely list based approach would
+require a polynomial time edit sequence algorithm to be applied on the entire
+document twice. Compared to this, the rope implementation has much better
+algorithmic complexity.  While not depicted here, we measured that by
+representing the entire document using a mergeable list type, in addition to
+slowing down each edit operation, the merge at a distance of 256 was 2.3 times
+slower than the rope implementation.  We conclude that versioning and merging
+as found in \name\ can yield substantial throughput and latency gains with low
 computational cost on modern eventually-consistent cloud systems.

### 2017-02-28 paper/intro.tex

@@
 
-To illustrate how these kinds of problems manifest in practice,
-consider how we might write a simple counter library (see
+In this paper, we consider how declarative abstractions can be used to
+overcome this tension to provide the best of both worlds: functional
+programs whose logical structure can be seamlessly transplanted to a
+distributed setting, while nonetheless being resilient to system
+realities such as network failure and latency.  Our key insight is the
+development of a principled methodology that allows us transform
+low-level operational details such as replication and consistency
+into high-level compositional reasoning on language-level datatypes.
+
+To illustrate how the kinds of problems described above manifest in
+practice, consider how we might write a simple counter library (see
 Fig.~\ref{fig:counter-adt}).  A \C{Counter} supports two (update)
@@
 heavyweight mechanisms, applying an \C{Add} operation on one node may
-not be instantaneously witnessed on another, which may be in the process of
-simultaneously attempting to perform its own \C{Add} or \C{Mult}
-action.  While synchronizing the activities of all nodes to ensure at
-most one such operation is performed at a time is impractical,
-designating a single node to hold the counter state eschewing
-replication altogether (as in a client-server
+not be instantaneously witnessed on another, which may be in the
+process of simultaneously attempting to perform its own \C{Add} or
+\C{Mult} action.  While synchronizing the activities of all nodes to
+ensure at most one such operation is performed at a time is
+impractical, designating a single node to hold the counter state
+eschewing replication altogether (as in a client-server
 configuration~\cite{Armstrong}), is also an undesirable solution,
-given the sensitivity of such architectures to network partitions and server failures, and
-the negative performance impact it incurs in geo-distributed
-environments~\cite{Walter}.  Removing coordination altogether by simply
-replicating counter state without having any supporting consistency
-protocol is an equally infeasible approach.
+given the sensitivity of such architectures to network partitions and
+server failures, and the negative performance impact it incurs in
+geo-distributed environments~\cite{Walter}.  Removing coordination
+altogether by simply replicating counter state without having any
+supporting consistency protocol is an equally infeasible approach.
 
@@
   }
-\vspace*{-.85in}
+\vspace*{-1.2in}
 \end{wrapfigure}
@@
 by operations, can we formulate a more declarative interpretation,
-directly in terms of the counter value maintained by each node.  In
-doing so, we observe that each replica essentially operates over its
-own version of a counter.  Local operations on a replicated object can
-be thought of as yielding new versions, thus producing a version tree,
-with one branch for each replica.  Thus, every branch represents
-different (immutable) versions maintained by different nodes, with the
-state produced by the computation performed over a counter on a node
-recorded along the node's local branch for the counter.  Now, to
-generate a globally consistent view of a counter, we only need to
-define a merge operation that explains how to combine two local
-versions to produce a new version that reflects both their states.
+directly in terms of the counter value maintained by each node?  To do
+so, we first observe that each replica essentially operates over its
+own version of a counter.  Thus, local operations on a replicated
+object can be thought of as yielding new versions, collectively
+producing a version tree, with one branch for each replica.  Every
+branch represents different (immutable) versions maintained by
+different nodes, with the state produced by the computation performed
+over a counter on a node recorded along the node's local branch for
+the counter.  Now, to generate a globally consistent view of a
+counter, we only need to define a merge operation that explains how to
+combine two local versions to produce a new version that reflects both
+their states.  This operation is defined not in terms of replicas or
+other system-specific artifacts, but in terms of the semantics of the
+datatype itself.
 
@@
 
-In this paper, we develop a programming model that realizes a monadic
-version control system centered around data, rather than file,
-coherence.  The \name monad lets programmers write and compose
-concurrent computations around multiple (implicit) versions of a
-mergeable datatype, an ordinary ML datatype additionally equipped with
-a \emph{merge} function responsible for deriving a consistent global
-state from a collection of local versions of that state.  A
-computation progresses by forking (i.e., replicating) new versions of
-existing versions on geo-distributed nodes, allowing
-synchronization-free local computation to proceed on those nodes,
-creating new versions along an existing \emph{branch} that represents
-updated local instances, and by merging branches to realize global
-consistency.
+Our main innovation is thus the development of a programming model
+that realizes a monadic version control system centered around data,
+rather than file, coherence.  The \name monad lets programmers write
+and compose concurrent computations around multiple (implicit)
+versions of a mergeable datatype, an ordinary ML datatype additionally
+equipped with a \emph{merge} function responsible for deriving a
+consistent global state from a collection of local versions of that
+state.  A computation progresses by forking (i.e., replicating) new
+versions of existing versions, allowing synchronization-free local
+computation to proceed on those nodes, creating new versions along an
+existing \emph{branch} that represents updated local instances, and by
+merging branches to realize global consistency.
 
@@
 we demonstrate that it can be nonetheless efficiently realized on
-existing geo-replicated distributed systems.
+existing real-world geo-replicated distributed systems.
 
@@
 
-The remainder of the paper is organized as follows.  We present the
+\noindent The remainder of the paper is organized as follows.  We present the
 \name\ programming model informally in the next section.

### 2017-02-28 paper/model.tex

@@
 quadrant contains multiple pixels of different colors. Quadrants are
+\begin{wrapfigure}{l}{.4\textwidth}
+\centering
+\includegraphics[scale=0.43]{Figures/canvas-merging}
+\caption{Merging concurrent versions (\C{v1} and \C{v2}) of a drawing
+canvas. \C{lca} is their common ancestor.}
+\label{fig:canvas-merging}
+\end{wrapfigure}
 expanded into a tree structures as and when pixels are colored.  The
@@
 two concurrent versions evolved independently. The merge function can
-\begin{wrapfigure}{L}{.4\textwidth}
-\centering
-\includegraphics[scale=0.43]{Figures/canvas-merging}
-\caption{Merging concurrent versions (\C{v1} and \C{v2}) of a drawing
-canvas. \C{lca} is their common ancestor.}
-\label{fig:canvas-merging}
-\end{wrapfigure}
 make use of the pixel values of the common ancestor to merge the pixel
@@
 pixel is red in \C{v1}, then it means that both \C{v1} and \C{v2} have
-modified the color. In such case, an appopriate color-mixing algorithm
+modified the color. In such case, an appropriate color-mixing algorithm
 can be used to determine the color of pixel.  For instance, the pixel

### 2017-02-28 paper/related.tex

@@
 Revisions~\cite{BBL+10}, a programming abstraction that provides
-deterministic concurrent execution by forking and merging
-\emph{versions}, portions of state shared among concurrent threads.
-The idea of using revisions as a means to programming eventually
-consistent distributed systems was further developed
-in~\cite{BFL+12,Burckhardt2012}.  The \name\ programming model,
-however, differs substantially from a concurrent revisions model
-because it imposes no distinction between \emph{servers}, machines
-that hold global state, and \emph{clients}, devices that operate on
-local, potentially stale, data - any computation executing in a
-distributed environment is free to fork new versions, and synchronize
-against other replicated state.  Our model, furthermore, supports
-fully decentralized operation and is robust to network partitions and
-failures. Just as significantly, \name\ allows applications to
-customize join semantics with programmable merge operations.  Indeed,
-the integration of a version-based mechanism within OCaml allows a
-degree of type safety, composability, and profitable use of
-polymorphism not available in related systems.
+deterministic concurrent execution.  The idea of using revisions as a
+means to programming eventually consistent distributed systems was
+further developed in~\cite{BFL+12,BLF+12}.  The
+\name\ programming model, however, differs substantially from a
+concurrent revisions model because it imposes no distinction between
+\emph{servers}, machines that hold global state, and \emph{clients},
+devices that operate on local, potentially stale, data - any
+computation executing in a distributed environment is free to fork new
+versions, and synchronize against other replicated state.  Our model,
+furthermore, supports fully decentralized operation and is robust to
+network partitions and failures. Just as significantly, \name\ allows
+applications to customize join semantics with programmable merge
+operations.  Indeed, the integration of a version-based mechanism
+within OCaml allows a degree of type safety, composability, and
+profitable use of polymorphism not available in related systems.
 
@@
 operations to coordinate the activities of replicas may involve
-addressing subtleties that do not manifest otherwise.  Our case
-studies and experimental results support our contention that using
-well-understood state (heap)-based abstractions to build distributed
-applications greatly simplifies program reasoning and eases
-development, without compromising efficiency.
+addressing subtleties that would not manifest in their absence.  Our
+case studies and experimental results support our contention that
+using well-understood state (heap)-based abstractions to build
+distributed applications greatly simplifies program reasoning and
+eases development, without compromising efficiency.
 
@@
 distributed functionality in a functional programming context.
-Poly/ML~\cite{Mat97} and Facile~\cite{TLK+93} both extend Standard ML
-with distributed programming abstractions.  Kali-Scheme~\cite{CJK95}
+Poly/ML~\cite{Mat97} and Facile~\cite{TLK+93} extend Standard ML,
+while Acute~\cite{SLW+05} and HashCaml~\cite{BSS+06} extend OCaml, with
+distributed programming abstractions and support, with particular
+focus on type-safe marshalling and mobility.  Kali-Scheme~\cite{CJK95}
 defines a distributed extension for Scheme-48, and Cloud
@@
 is noteworthy that replication is inherent in many widely-used
-distributed system implementations today (e.g.,
+cloud-based distributed system implementations today (e.g.,
 Cassandra~\cite{Cassandra} or Dynamo~\cite{Dynamo}), which provide
-only weak consistency guarantees.  Rather than providing, as other
-approaches have, a low-level operational treatment of replication, our
-primary contribution is the development of a principled high-level
-approach to framing issues of consistency and replication.  Our goal
-is to leverage functional programming principles to minimize
+only weak consistency guarantees.  Rather than exposing a low-level
+operational treatment of replication, foisting the burden of reasoning
+about replicated state onto the programmer, our primary contribution
+is the development of a principled high-level approach to framing
+these issues that abstracts these kinds of system-level details.  Our
+goal is to leverage functional programming principles to minimize
 disruption to the way programmers structure and reason about their

### 2017-02-28 paper/appendix.tex
edits

@@
 Premise says that $v$ has a unique external locus (call it $v_o$). The
-new version $v_m$ inherits $v_o$ as a least external ancestor, and
-also has $v'$ as another least external ancestor. However, since
+new version $v_m$ inherits $v_o$ as a lowest external ancestor, and
+also has $v'$ as another lowest external ancestor. However, since
 $\under{H}{v' \mbleto v}$, we have $\under{H}{v_o \preceq v'}$, we
@@
 Every pair of versions $v_1$ and $v_2$ in the history $H$ generated by
-the rules in Fig.~\ref{fig:opsem} have a unique least
+the rules in Fig.~\ref{fig:opsem} have a unique lowest
 common ancestor.

### 2017-02-28 paper/meta.tex
edits

--- 2017-02-28 paper/meta.tex
@@
 \caption{This example of a grow-only counter illustrate why \C{merge}
-needs a least common ancestor, and not just a common ancestor. Both 0
-and 2 are common ancestors of 4 and 7, while 2 is their least common
+needs a lowest common ancestor, and not just a common ancestor. Both 0
+and 2 are common ancestors of 4 and 7, while 2 is their lowest common
 ancestor (since $0 \preceq 2$). The result (v) of merging 4 and 7 is
@@
 ($\preceq_o$) is also a total order. Hence, for any given version $v$,
-least internal and external ancestors, if exist, must be unique.  We
-call the unique least internal ancestor as its \emph{internal locus},
-and the unique least external ancestor version as its \emph{external
+lowest internal and external ancestors, if exist, must be unique.  We
+call the unique lowest internal ancestor as its \emph{internal locus},
+and the unique lowest external ancestor version as its \emph{external
 locus}.
@@
 Every pair of branches in the history $H$ generated by the rules in
-Fig.~\ref{fig:opsem} have a unique least common ancestor. 
+Fig.~\ref{fig:opsem} have a unique lowest common ancestor.
 \end{lemma}

### 2017-02-28 paper/model.tex
edits

@@
 includes includes two concurrent versions (\C{v1} and \C{v2}),
-and their least common ancestor (\C{lca}) - the version from which the
+and their lowest common ancestor (\C{lca}) - the version from which the
 two concurrent versions evolved independently. The merge function can
@@
 version, and merges it with its proposal to produce the next version
-$C_1$, which is then installed on Bob's branch.  The least common
+$C_1$, which is then installed on Bob's branch.  The lowest common
 ancestor (LCA) for this merge is the initial version ($C_0$), and the

### 2017-02-28 paper/semantics.tex
edits

@@
 \subsection{Operational Semantics}
-
-\input{opsem}
 
@@
 ($t$).
+
+\input{opsem}
 
@@
 like a \C{FORK} tag, records its origin. The rule assumes a function
-\C{lca} that computes the \emph{least common ancestor} (LCA) for the
+\C{lca} that computes the \emph{lowest common ancestor} (LCA) for the
 latest versions on the given pair of branches. The formal definition
@@
 relationship among all common ancestors lets us define the notion of a
-least common ancestor (LCA):
+lowest common ancestor (LCA):
 
-\begin{definition} [\bfseries Least Common Ancestor]
+\begin{definition} [\bfseries Lowest Common Ancestor]
 Version $v$ is said to be a common ancestor of versions $v_1$ and
 $v_2$ under a history $H$ if and only if $\under{H}{v \preceq v_1}$
-and $\under{H}{v \preceq v_2}$. It is said to be the least common
+and $\under{H}{v \preceq v_2}$. It is said to be the lowest common
 ancestor (LCA) of $v_1$ and $v_2$, iff there does not exist a $v'$

### 2017-02-28 paper/implementation.tex
Implementation

\section{Implementation}
\label{sec:implementation}

We describe the realization of \name programming model on top of a persistent
distributed store and the language extensions to enable and enforce the
constraints of the \name.

\name programming model is realized on top of Irmin~\cite{irmin}, an OCaml
library database implementation that is part of the MirageOS
project~\cite{mirage}. Irmin provides a persistent multi-versioned store with
with a content-addressable heap abstraction. Simply put, content-addressability
means that the address of a data block is determined by its content. If the
content changes, then so does the address. Old content continues to be
available the old address. Content-addressability also results in constant time
structural equality checks, which we exploit in our mergeable rope
implementation (Section~\ref{sec:ropes)}.

Irmin provides support for distribution, fault-tolerance and concurrency
control by incorporating Git distributed version control~\cite{git} protocol
over its object model. Indeed, Irmin is fully compatible with the Git command
line tools. Distributed replicas in \name are created by cloning a \name
repository. Due to \name's mergeable type, each replica can operate completely
independently, accepting client requests, even when disconnected from other
replicas, resulting in a fully available distributed database system.

Concurrent operations are tracked within the notion of branches, but
importantly, Irmin supports user-defined three-way merge functions for
reconciling concurrent operations. While Irmin's merge functions are defined
over objects on Irmin's content-addressable heap, \name's merge functions are
defined over OCaml types. We address this representational mis-match with the
help of OCaml's PPX metaprogramming support to derive bi-directional
transformations between objects on OCaml and Irmin heaps. We also derive the
various serialization functions required Irmin. This address the representation
mismatch between the application and storage layers. Synchronization between
replicas is performed using Git's notion of pushing and pulling updates from
remotes over the Git transfer protocol~\cite{git-tp}. As a result, we reap the
benefits of efficient delta-transfer (only missing objects are transferred
between replicas), compression and end-to-end encrypted communication between
the replicas.

### 2017-02-28 paper/appendix.tex
Committing meta.

--- 2017-02-28 paper/appendix.tex
@@
 \begin{proof}
-By case analysis on the reduction relation. In each case, we assume
-that the unique external locus property holds for the initial history,
-and prove that it holds for the final history.
-
-In \rulelabel{E-Run} case, initial version is the unique external locus of
-the new version on the new branch $H(t)$. In \rulelabel{E-Fork} case,
-which creates the branch $H(t')$ by forking from $H(t)$, the latest
-version on $H(t)$ is the external locus of the first version on
-$H(t)$. In \rulelabel{E-Push} case, the version on $H(t)$ has the same
-external locus as the previous latest version, which, premise says, is
-unique. No new version is created in \rulelabel{E-Pull}. In
-\rulelabel{E-Pull-Wait}, new version is created by merging the latest
-version ($v'$) on $H(t')$ into the latest version $v$ on $H(t)$.
-Premise says that $v$ has a unique external locus (call it $v_o$). The
-new version $v_m$ inherits $v_o$ as a least external ancestor, and
-also has $v'$ as another least external ancestor. However, since
-$\under{H}{v' \mbleto v}$, we have $\under{H}{v_o \preceq v'}$, we
-have $v'$ as the the external locus.
-\qed
+Follows from the definitions~\ref{def:external-locus},
+\ref{mergeability}, and the \rulelabel{E-Pull-Wait} rule.
 \end{proof}
@@
 \begin{lemma} [\bfseries Unique LCA]
-Every pair of versions $v_1$ and $v_2$ in the history $H$ generated by
-the rules in Fig.~\ref{fig:opsem} have a unique least
-common ancestor. 
+Every pair of branches in a legal history $H$, if mergeable as per
+Def.~\ref{fig:mergeability}, have a unique least common ancestor. 
 \end{lemma}
 \begin{proof}
-By case analysis on the step relation. In each case, we assume that
-the unique LCA property holds for the initial history, and prove that
-it holds for the final history.
+By construction. If $b_j$ is mergeable into $b_i$, the locus of
+$b_i$ is the least common ancestor of $b_i$ and $b_j$
+\end{proof}
 
-The \rulelabel{E-Run} case is trivial. The \rulelabel{E-Push} case
-creates a new version ($v$) with \C{PUSH} tag at the end of a branch.
-Let $v_1$ be the previous version on the branch, and let $v_2$ be any
-other version. The premise is that $v_1$ and $v_2$ have a unique LCA
-$v_0$. Observe that $v$'s internal locus is $v_1$ and its external
-locus is same as that of $v_1$. Hence, all ancestors of $v_1$
-(including itself) are also ancestors of $v$. Moreover, $v$ has no new
-ancestors except itself, which is the latest version being added.
-Hence, any common ancestor of $v_1$ and $v_2$ is also a common
-ancestor of $v$ and $v_2$, while there cannot be new common ancestors.
-Hence LCA of $v$ and $v_2$ is unique. The reasoning for
-\rulelabel{E-Fork} is similar. The \rulelabel{E-Pull} case is trivial.
-
-The \rulelabel{E-Pull-Wait} case is where the merge happens. Version
-$v'$ is merging into version $v$ to produce the next version $v_m$.
-Let $v_2$ be any version. From the premise, we know that $v$ and $v_2$
-have a unique LCA (call it $v_0$), and $v'$ and $v_2$ have a unique
-LCA (call it $v_1$). But, we know that $v_2$ has a unique external
-locus (call it $v_3$), hence we know that 
-\begin{itemize}
-  \item Both $v_0$ and $v_1$ are external ancestors of $v_2$. Hence,
-  $\under{H}{v_0 \preceq v_3} \conj \under{H}{v_1 \preceq v_3}$.
-  Since $\under{H}{v_0 \preceq v_m} \conj \under{H}{v_1 \preceq v_m}$, 
-  $v_2$ and $v_m$ have a unique LCA.
-
-  \item $v_0$ is an internal ancestor of $v_2$, hence an external
-  ancestor of $v$, hence an external ancestor of $v_m$. $v_1$ is
-  an ancestor of $v'$, hence also an external ancestor of $v_m$. 
-\end{itemize}
-\end{proof}
+\begin{lemma}
+Every pair of 
+\end{lemma}
+% If $b_j$ is not mergeble into $b_i$, it could be because:
+% 1. $b_j$ included some internal nodes of $b_i$ that are not in
+% $L(b_i)$, then simply merge $b(L(b_i))$ into $b_i$, and then that back
+% into $b(L(b_i))$. Now the problem reduces to merging $b(L(b_i))$
+% 2. $If

### 2017-02-28 paper/meta.tex
Committing meta.

--- 2017-02-28 paper/meta.tex
@@
 merging $b_1$, $b_3$ insists on an ancestor relation between the last
-merged version ($v_{22}$ of the $b_2$), and the latest version of
-$b_1$, the currently merging branch. We call the last merged version
-the \emph{external locus} of the branch. By requiring that, for every
-branch $b$, the external locii of $b$ at various points in time (i.e.,
-external locus of every prefix of $b$) be totally ordered, and given
-that the semantics already preempts criss-cross merges, we effectively
-avoid the case of there ever being more than 2 LCAs for a pair of
-branches. Thus, uniqueness of LCAs is thus guaranteed.
+merged version ($v_{22}$) and the currently merging version.  We call
+the last merged version of a branch its \emph{locus}. By maintaining a
+total order of among locii of a branch $b$ as new versions are
+created, we effectively guarantee the presence of a version (the
+locus) that is lower than all the ancestors of the latest version on
+$b$. Next, by requiring the locus to be an ancestor of the merging
+version, we ensure that all the common ancestors have a single lowest
+element, thus enforcing the uniqueness of the LCA.
 
-We now formalize the notion of mergeability, and the guarantees
-provided by the operational semantics.
+We now formalize the intuitions described above via a series of
+definitions that help us state the guarantees offered by the system.
 
@@
 $v_o' \not\in b$ such that $\under{H}{v_o' \preceq v}$ and
-$\under{H}{v_o \preceq v_o'}$. We write $L(v)$ to denote the locus
-of version $v$.
+$\under{H}{v_o \preceq v_o'}$. We lift the notion of locus to the
+level of branches by defining the locus of a branch as the locus of
+its latest version.
 \end{definition}
 
-In a legal branching history, every version has a unique
-locus\footnote{lemma and proof in the appendix}. Thus, locus of a
-version is the maxima of its causal history. As usual, the locus of
-the latest version on a branch is called the locus of the branch.
+In a legal branching history, every version has a unique locus
+(property follows from Def.~\ref{def:mergeability} below). Let
+\emph{causal history} of a version be defined as the set of all
+ancestors of that version. A locus ($v_o$) of a version ($v$) is
+therefore the maximum element (lowest element in the ancestor
+relation) of the causal history of $v$'s internal ancestor that was
+the result of the last merge. We make use of this observation while
+defining mergeability.
 
@@
 an ancestor of $v_1$ under $H$, $v_2$ is mergeable into $v_1$ (denoted
-$\under{H}{v_2 \mbleto v_1}$) if and only if $L(v_1)$ is an ancestor
-of $v_2$, and no internal ancestor of $v_1$ that is not also an
-ancestor of $v_o$ is an ancestor of $v_2$. 
-If $v_1$ and $v_2$ are the latest versions on their respective
-branches $b_1$ and $b_2$, we say $b_2$ is mergeable into $b_1$, or
-more generally, $b_1$ and $b_2$ are mergeable.  
+$\under{H}{v_2 \mbleto v_1}$) if and only if the locus ($v_o$) of
+$v_1$ is an ancestor of $v_2$, and no internal ancestor of $v_1$ that
+is not also an ancestor of $v_o$ is an ancestor of $v_2$.    
 \end{definition}
 
-The following theorem asserts that every pair of branches deemed
-mergeable by the above definition are infact mergeable, i.e., they are
-guaranteed to have a unique LCA.
+If $v_1$ and $v_2$ referred in the above definition are the latest
+versions on their respective branches $b_1$ and $b_2$, we say $b_2$ is
+mergeable into $b_1$, or more generally, $b_1$ and $b_2$ are
+mergeable. The definition essentially requires all of $v_1$'s causal
+history that preceeds its locus $v_o$ to be included in $v_2$'s causal
+history, and none of $v_1$'s history that succeeds $v_o$ to be
+included in $v_2$'s history. This allows us to view versions $v_1$ and
+$v_2$ as having been independently evolved from a common causal
+history whose maximum (w.r.t the ancestor relation) is the version
+$v_o$. Consquently, $v_o$ becomes the LCA for $v_2$'s merge into
+$v_1$. Generalizing this observation for the latest versions on
+any two branches, we state the following theorem:
 
-\begin{lemma} [\bfseries Unique LCA]
+\begin{theorem} [\bfseries Unique LCA]
 Every pair of branches in a legal history $H$, if mergeable as per
 Def.~\ref{fig:mergeability}, have a unique least common ancestor.
-\end{lemma}
+\end{theorem}
 
-\begin{lemma} [\bfseries Non-Monotonicity of Mergeability]
+While the definition of mergeability is sufficient to enforce
+uniqueness of LCAs, it may hinder progress in the sense that there may
+not exist a pair of branches in a legal history satisfying the
+definition, thus making the progress impossible. The following theorem
+asserts that this is not the case:
+
+\begin{theorem} [\bfseries Progress]
 In a legal branching history $H$ produced by the operational
 semantics, if two branches, $b_i$ and $b_j$ are not mergeable (as per
-Def.~\ref{def:mergeability}), then there exist a series of merges
-involving only the mergeable branches that can be performed on $H$ to
-yield a new history $H'$, where $b_i$ and $b_j$ are now mergeable.
-\end{lemma}
+Def.~\ref{def:mergeability}), then there exist a sequence of fork and
+merge operations (between mergeable versions) that can be performed on
+$H$ to yeild a new history $H'$, where $b_i$ and $b_j$ are mergeable.  
+\end{theorem}
 
-Thus, a pair of branches that are not currently mergeable (as per
-Def.~\ref{fig:mergeability}), will become mergeable as the branching
-history is extended.
+Proof is by the induction on the size of branching history. The
+observation that the smallest history that is the union of causal
+histories of two unmergeable versions (call them $v_i$ and $v_j$) is
+smaller than the history that includes the unmergeable versions. Since
+causal histories have maximum versions (respective locii), which, by
+inductive hypothesis, can be made mergeable, we merge them to yeild a
+new version $v$ that is the maximum of the causal histories of both
+$v_i$ and $v_j$. If $v_i$ and $v_j$ are the latest versions on their
+respective branches, the branches are now mergeable.
 
-\begin{theorem} [\bfseries Progress and Eventual Convergence] Given a
-legal branching history $H$, and a program $p$, where every thread
-expression in $p$ is a \C{pull}, there exist a series of reduction steps
-that reduce $p; H$ to $p'; H'$, where every thread expression in $p$
-is the same value $v$.
-\end{theorem}
+% Observe that $v$ and $v_i$ are mergeable (with
+% $v_i$'s locus as the LCA), and so do $v$ and $v_j$. If $v_i$ and $v_j$
+% are the latest versions on their respective branches, then merging $v$
+% into $v_i$ and $v$ into $v_j$ makes $v$ the locus of both the
+% branches, thus making them mergeable again.
+
+% \begin{theorem} [\bfseries Progress and Eventual Convergence] Given a
+% legal branching history $H$, and a program $p$, where every thread
+% expression in $p$ is a \C{pull}, there exist a series of reduction steps
+% that reduce $p; H$ to $p'; H'$, where every thread expression in $p$
+% is the same value $v$.
+% \end{theorem}

### 2017-02-28 paper/examples.tex

@@
 A mergeable list of integers can be composed using mergeable lists and
-mergeable integers.  Concretely, it is a composition of \C{MList} and
-\C{Int} types (i.e., \C{MList(Int)}), where \C{Int.merge} is the same
-as \C{Counter.merge} described in the introduction.  An alternative
-instantiation of \C{MList} is for a quantity type
-(\C{Qty}) that supports \C{add} and \C{subtract} operations, and
-defines its \C{merge} operation a little differently:
+mergeable integers.  Concretely, it is a composition of \C{MList} and \C{Int}
+types (i.e., \C{MList(Int)}), where \C{Int.merge} is the same as
+\C{Counter.merge} described in the introduction.  An alternative instantiation
+of \C{MList} is for a quantity type (\C{Qty}) that supports \C{add} and
+\C{subtract} operations, and defines its \C{merge} operation a little
+differently:
+\begin{ocaml}
+let merge lca v1 v2 = if v1 = lca then v2
+    else (if v2 = lca then v1 else max v1 v2)
+\end{ocaml}
 \begin{wrapfigure}{l}{.45\textwidth}
@@
 \end{wrapfigure}
-\begin{ocaml}
-let merge lca v1 v2 = if v1 = lca then v2
-    else (if v2 = lca then v1 else max v1 v2)
-\end{ocaml}
-\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has
-the same semantics under a sequential execution. Their differences
-however manifest under a concurrent execution, where both adopt
-different methods to reconcile concurrent updates.
+\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has the same
+semantics under a sequential execution. Their differences however manifest
+under a concurrent execution, where both adopt different methods to reconcile
+concurrent updates.
 
@@
 considered distinct. A shopping list is a mergeable list of items,
-\begin{wrapfigure}{l}{.5\textwidth}
+i.e., \C{MList(Item)} (a grocery shopping list may not be large enough
+to justify using ropes, but \C{MRope(MList(Item))} is certainly
+allowed).  \C{MList(Item).merge} automatically lifts the merge
+semantics for items to a merge semantics for lists of items.
+\begin{wrapfigure}{r}{.35\textwidth}
   \centering
-  \includegraphics[scale=0.3]{Figures/shopping-list}
+  \includegraphics[width=0.25\textwidth]{Figures/shopping-list}
   \caption{Merging concurrent shopping lists}
@@
 \end{wrapfigure}
-i.e., \C{MList(Item)} (a grocery shopping list may not be large enough
-to justify using ropes, but \C{MRope(MList(Item))} is certainly
-allowed).  \C{MList(Item).merge} automatically lifts the merge
-semantics for items to a merge semantics for lists of items.
 Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
@@
 concurrency model and the underlying system, such as the existence of
-a single least common ancestor (LCA) for any pair of versions, the
+a single lowest common ancestor (LCA) for any pair of versions, the
 ability to access a previous version on any branch, and mergeability

### 2017-02-28 paper/implementation.tex

@@
 
-Concurrent operations are tracked within the notion of branches, but
-importantly, Irmin supports user-defined three-way merge functions for
+Concurrent operations in Irmin are tracked within the notion of branches,
+allowing the programmer to explicitly merge the branches on demand. \name
+realizes the concurrent programming model on top of the notion of branches;
+each replica operates on its own branch, and thus is isolated from actions on
+other replicas. \name monad conceals the branching structure, but also
+transparently performs the necessary merges to obtain a history graph with a
+unique lowest common ancestor.
+
+Importantly, Irmin supports user-defined three-way merge functions for
 reconciling concurrent operations. While Irmin's merge functions are defined

### 2017-02-28 paper/intro-sj.tex

@@
   \end{ocaml}
-The role of \C{lca} (least common ancestor) here captures salient
+The role of \C{lca} (lowest common ancestor) here captures salient
 history - the state resulting from the merge of two versions derived

### 2017-02-28 paper/mergeable-types.tex
fix wrapfig

@@
 Fig.~\ref{fig:opsem}, which applies the \C{merge} function to a pair
-of concurrent versions ($v_1$ and $v_2$) and their least common
+of concurrent versions ($v_1$ and $v_2$) and their lowest common
 ancestor ($v$). Thus, the only relationship that exists between $v$
@@
 $v_{11}$ on $b_1$ to create $v_{12}$ on $b_1$. Versions $v_{11}$ and
-\begin{wrapfigure}{l}{.4\textwidth}
+\begin{wrapfigure}{L}{.4\textwidth}
 \centering
@@
   t\neq t' \spc
-% \under{H}{v' \mbleto v} \spc
+  \under{H}{v' \mbleto v} \spc
 % \C{world}(H,t') \semsucceq \C{world}(H,t)\spc
@@
 called an edit sequence. Applying the edit sequence on $A$ produces
-\begin{wrapfigure}{l}{.4\textwidth}
+\begin{wrapfigure}{L}{.4\textwidth}
 \centering

### 2017-02-28 paper/implementation.tex
edits

@@
-\section{Implementation}
+\section{Distributed Instantiation}
 \label{sec:implementation}
@@
 structural equality checks, which we exploit in our mergeable rope
-implementation (Section~\ref{sec:ropes)}.
+implementation (Section~\ref{sec:ropes}).
 
@@
 
+\begin{wrapfigure}{L}{.5\textwidth}
+	\begin{center}
+	\includegraphics[scale=0.8]{Figures/partitions}
+	\end{center}
+	\caption{Network partitions may leave branches $b_1$ and $b_4$ in one
+		partition (red background), and branches $b_2$ and $b_3$ in the other (yellow
+		background). Since $b_2$ and $b_3$ cannot merge unless $b_1$ and $b_4$ merge
+		(Fig.~\ref{fig:legal-extensions}), they cannot communicate among themselves
+		even though they are in the same network partition.}
+	\label{fig:partitions}
+\end{wrapfigure}
+
 Concurrent operations in Irmin are tracked within the notion of branches,
@@
 each replica operates on its own branch, and thus is isolated from actions on
-other replicas. \name monad conceals the branching structure, but also
+other replicas. In presence of network partitions, the nodes in one partition
+may not receive updates from the other partition, hence merges cannot happen
+across a partition. This can be a problem if a pair of branches in one
+partition (the current partition) have multiple LCAs in the other partition
+(the remote partition), hence cannot merge. Fig.~\ref{fig:partitions}
+illustrate this problem for the branching structure of
+Fig.~\ref{fig:external-lcas}.
+
+Fortunately, the access to full (causal) branching history assured by Irmin at
+every node comes to the rescue. Relying on the history, the current partition
+can fork-off (\C{fork}) new branches that start where the remote branches left,
+and map them to new (virtual) nodes that emulate the nodes in the remote
+partition. For the example in Fig.~\ref{fig:partitions}, the partition of $b_1$
+and $b_4$ can fork new branches $b_2'$ and $b_3'$ from $b_2$ and $b_3$,
+respectively, and resume the activity as if the partition never happened. After
+merging the first versions on $b_2'$ and $b_3'$, branches $b_1$ and $b_2$ will
+be in the same state as before (when the partition happened), except that the
+branches are now mergeable because $b_2'$ and $b_3'$ can merge. The ability to
+track full provenance information is thus crucial for \name to overcome network
+partitions, making it an appropriate programming model for highly-available
+replicated data types. \name monad conceals the branching structure, but also
 transparently performs the necessary merges to obtain a history graph with a
@@
 defined over OCaml types. We address this representational mis-match with the
-help of OCaml's PPX metaprogramming support to derive bi-directional
+help of OCaml's PPX metaprogramming support~\cite{ppx} to derive bi-directional
 transformations between objects on OCaml and Irmin heaps. We also derive the

### 2017-02-28 paper/system-model.tex
edits

@@
-\subsection{Instantiations}
-
-\begin{figure}[!t]
-\centering
-\subcaptionbox[] {\small
-  Branches $b_1$, $b_2$ and $b_3$ mapped to nodes $n_1$, $n_2$ and
-  $n_3$. Each time step captures concurrent operations among the
-  nodes.
-  \label{fig:instantiation-1}
-} [0.47\columnwidth] {
-  \includegraphics[scale=0.6]{Figures/instantiation-1}
-}
-\hfill
-\subcaptionbox[] {\small
-  Network partitions may leave branches $b_1$ and $b_4$ in one
-  partition (red background), and branches $b_2$ and $b_3$ in the
-  other (yellow background). Since $b_2$ and $b_3$ cannot merge unless
-  $b_1$ and $b_4$ merge (Fig.~\ref{fig:legal-extensions}), they cannot
-  communicate among themselves even though they are in the same
-  network partition.
-} [0.47\columnwidth] {
-  \includegraphics[scale=0.8]{Figures/partitions}
-}
-\caption{Instantiations of the \name concurrency model in a
-distributed setting}
-\label{fig:partitions}
-\end{figure}
+\section{Distributed Instantiation}
+\label{sec:system-model}
 
@@
 described above.
-
-In presence of network partitions, the nodes in one partition may not
-receive updates from the other partition, hence merges cannot happen
-across a partition. This can be a problem if a pair of branches in one
-partition (the current partition) have multiple LCAs in the other
-partition (the remote partition), hence cannot merge.
-Fig.~\ref{fig:partitions} illustrate this problem for the branching
-structure of Fig.~\ref{fig:external-lcas}. Fortunately, the access to
-full (causal) branching history at every node comes to the rescue.
-Relying on the history, the current partition can fork-off (\C{fork})
-new branches that start where the remote branches left, and map them
-to new (virtual) nodes that emulate the nodes in the remote partition.
-For the example in Fig.~\ref{fig:partitions}, the partition of $b_1$
-and $b_4$ can fork new branches $b_2'$ and $b_3'$ from $b_2$ and
-$b_3$, respectively, and resume the activity as if the partition never
-happened. After merging the first versions on $b_2'$ and $b_3'$,
-branches $b_1$ and $b_2$ will be in the same state as before (when the
-partition happened), except that the branches are now mergeable
-because $b_2'$ and $b_3'$ can merge. The ability to track full
-provenance information is thus crucial for \name to overcome network
-partitions, making it an appropriate programming model for
-highly-available replicated data types.

### 2017-02-28 paper/intro.tex
edits

@@
 \name\ programming model informally in the next section.
-Sec.~\ref{sec:mergeable_types} develops a series of examples that
-illustrate how merge operations can be defined and used.
-Sec.~\ref{sec:formalization} defines an operational semantics, and
-formalizes our intuition of a mergeable type in terms of morphisms over
-datatype operations.  We also present soundness (consistency) and
-progress guarantees enjoyed by the
-semantics. Sec.~\ref{sec:system-model} considers an instantiation of
-the model suitable for distributed environments with network
-partitions and failures.  Implementation details describing how we
-incorporate mergeable types into OCaml are presented in
-Sec.~\ref{sec:impl}.  Experimental results and benchmarks are given in
-Sec.~\ref{sec:evaluation}.  Secs.~\ref{sec:related}
-and~\ref{sec:conclusions} describes related work and conclusions.
+Sec.~\ref{sec:mergeable_types} develops a series of examples that illustrate
+how merge operations can be defined and used. Sec.~\ref{sec:formalization}
+defines an operational semantics, and formalizes our intuition of a mergeable
+type in terms of morphisms over datatype operations.  We also present soundness
+(consistency) and progress guarantees enjoyed by the semantics.
+Sec.~\ref{sec:implementation} presents an instantiation of the model suitable
+for highly-available distributed environments with network partitions and
+failures. Experimental results and benchmarks are given in
+Sec.~\ref{sec:evaluation}.  Secs.~\ref{sec:related} and~\ref{sec:conclusions}
+describes related work and conclusions.
 

### 2017-02-28 paper/intro.tex
edits

--- 2017-02-28 paper/intro.tex
@@
 realities such as network failure and latency.  Our key insight is the
-development of a principled methodology that allows us transform
-low-level operational details such as replication and consistency
-into high-level compositional reasoning on language-level datatypes.
+development of a principled methodology that allows us to transform
+low-level operational details such as replication and consistency into
+high-level compositional abstractions over language-level datatypes.
 
@@
 that realizes a monadic version control system centered around data,
-rather than file, coherence.  The \name monad lets programmers write
-and compose concurrent computations around multiple (implicit)
-versions of a mergeable datatype, an ordinary ML datatype additionally
-equipped with a \emph{merge} function responsible for deriving a
-consistent global state from a collection of local versions of that
-state.  A computation progresses by forking (i.e., replicating) new
-versions of existing versions, allowing synchronization-free local
-computation to proceed on those nodes, creating new versions along an
-existing \emph{branch} that represents updated local instances, and by
-merging branches to realize global consistency.
+rather than file, coherence.  The \name (Versioned ML) monad lets
+programmers write and compose concurrent computations around multiple
+(implicit) versions of a mergeable datatype, an ordinary ML datatype
+additionally equipped with a \emph{merge} function responsible for
+deriving a consistent global state from a collection of local versions
+of that state.  A computation progresses by forking (i.e.,
+replicating) new versions of existing versions, allowing
+synchronization-free local computation to proceed on those nodes,
+creating new versions along an existing \emph{branch} that represents
+updated local instances, and by merging branches to realize global
+consistency.
 

### 2017-02-28 paper/implementation.tex

@@
 	\caption{Network partitions may leave branches $b_1$ and $b_4$ in one
-		partition (red background), and branches $b_2$ and $b_3$ in the other (yellow
-		background). Since $b_2$ and $b_3$ cannot merge unless $b_1$ and $b_4$ merge
-		(Fig.~\ref{fig:legal-extensions}), they cannot communicate among themselves
-		even though they are in the same network partition.}
+		partition (red background), and branches $b_2$ and $b_3$ in the other
+		(yellow background). Since $b_2$ and $b_3$ cannot merge unless $b_1$ and
+		$b_4$ merge, they cannot communicate among themselves even though they are in
+		the same network partition.}
 	\label{fig:partitions}

### 2017-02-28 paper/meta.tex
Edits.

@@
 \subcaptionbox[] {\small
-  In this example, versions $v_{13}$ and $v_{44}$ have two LCAs
-  ($v_{22}$ and $v_{32}$)  despite there not being any previous merges
-  between their respective branches.
+  In this example, latest versions on $b_1$ and $b_3$, and
+  $b_2$ and $b_3$ have two LCAs each, hence are unmergeable. 
   \label{fig:external-lcas}
@@
 disables criss-cross merges since it only ever merges versions that
-are latest on their respective branches. For instance, suppose
-$v_{11}$ and $v_{21}$ are the latest versions on branches $b_1$ and
-$b_2$, respectively. Merging $b_2$ into $b_1$ entails merging $v_{21}$
-into $v_{11}$ to generate version $v_{12}$ on $b_1$.  Now, merging
-$b_1$ into $b_2$ involves merging $v_{12}$, the latest version on
-$b_1$, into $v_{21}$, but not $v_{11}$ into $v_{21}$, thus preempting
-a criss-cross branching structure.
-
-Secondly, we impose certain pre-conditions on the merging branches to
-preempt the structure shown in Fig.~\ref{fig:external-lcas}. The
-intuition is as follows: consider the branch $b_3$ at the instance of
-merging $b_1$. Since it has already merged $b_2$, a version on $b_2$
-($v_{21}$) could be a common ancestor for $b_3$ and some other branch
-(call it $b$). Now, if $b_3$ merges $b_1$, same could be true of $b_3$
-and $b_1$: a version on $b_1$ ($v_{11}$) could be a common ancestor
-for $b_3$ and the other branch $v$. Since $v_{21}$ is also a common
-ancestor for both the branches, and the ancestors are not ordered by
-the ancestor relation, $b_3$ and $b$ end up with two LCAs. In
-Fig.~\ref{fig:external-lcas}, the role of $b$ is played both by $b_1$
-and $b_2$. We observe that this scenario can be prevented if, when
+are latest on their respective branches. Secondly, we impose certain
+pre-conditions on the merging branches to preempt the structure shown
+in Fig.~\ref{fig:external-lcas}. The intuition is as follows: consider
+the branch $b_3$ at the instance of merging $v_{12}$. Since it has
+already merged $v_{22}$, $v_{22}$ could be a common ancestor for $b_3$
+and some other branch (call it $b$). Now, if $b_3$ merges $v_{12}$,
+same could be true of $v_{12}$. Since $v_{12}$ and $v_{22}$ are not
+orderd by the ancestor relation, both become LCAs of
+$b_3$ and $b$. We observe that this scenario can be prevented if, when
 merging $b_1$, $b_3$ insists on an ancestor relation between the last

### 2017-02-28 paper/model.tex

@@
   \begin{ocaml}
-  module type DALI = sig
+  module type VST = sig
     type ('a, 'b) t

### 2017-02-28 paper/meta.tex
minor edit

--- 2017-02-28 paper/meta.tex
@@
 control systems, which employ \emph{ad hoc} mechanisms to pave the way
-for three-way merging.  GitHub~\cite{github}, for instance,
-recursively merges LCAs to compute a virtual ancestor, which then
+for three-way merging.  Git~\cite{git}, for instance,
+recursively merges LCAs by default to compute a virtual ancestor, which then
 serves as the LCA for merging concurrent versions. This method is
@@
 Every pair of branches in a legal history $H$, if mergeable as per
-Def.~\ref{fig:mergeability}, have a unique least common ancestor.
+Def.~\ref{def:mergeability}, have a unique least common ancestor.
 \end{theorem}

### 2017-02-28 paper/model.tex
Fix bib

@@
   \begin{ocaml}
-  module type VST = sig
+  module type VML = sig
     type ('a, 'b) t

### 2017-02-28 paper/examples.tex
edits

@@
 
-The \drawsome example demonstrates the utility of the \name
-programming model in building concurrent applications with conceptual
-state sharing that make them suitable for transparent deployment in a
-distributed setting. The model allows concurrent computations to be
-composed around any mergeable type. We now present a series of
-examples that demonstrate how mergeable types are built: either by
-defining merge functions for ML data types, or by composing existing
-mergeable types.
+The \drawsome example demonstrates the \name's utility in building
+concurrent applications with conceptual state sharing that make them
+suitable for transparent deployment in a distributed setting. The
+model allows concurrent computations to be composed around any
+mergeable type. We now present a series of examples that demonstrate
+how mergeable types are built - either by defining merge functions for
+ML data types, or by composing existing mergeable types.
 
@@
 substrings of a larger string such that their left-to-right (in-order)
-concatenation yeilds the larger string. The idea is to facilitate
+concatenation yields the larger string. The idea is to facilitate
 efficient lookups and insertions at random positions in the string by
@@
 considered distinct. A shopping list is a mergeable list of items,
-i.e., \C{MList(Item)} (a grocery shopping list may not be large enough
-to justify using ropes, but \C{MRope(MList(Item))} is certainly
-allowed).  \C{MList(Item).merge} automatically lifts the merge
-semantics for items to a merge semantics for lists of items.
-\begin{wrapfigure}{r}{.35\textwidth}
-  \centering
-  \includegraphics[width=0.25\textwidth]{Figures/shopping-list}
-  \caption{Merging concurrent shopping lists}
-  \label{fig:shopping-list}
-  \vspace*{-.35in}
-\end{wrapfigure}
+i.e., \C{MList(Item)} (although a grocery shopping list may not be
+large enough to justify using ropes, \C{MRope(MList(Item))} is
+certainly allowed).  \C{MList(Item).merge} automatically lifts the
+merge semantics for items to a merge semantics for lists of items.
 Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
@@
 \C{candy}, where the quantity of eggs is obtained by merging Alice's
+\begin{wrapfigure}{r}{.35\textwidth}
+  \centering
+  \includegraphics[width=0.25\textwidth]{Figures/shopping-list}
+  \caption{Merging concurrent shopping lists}
+  \label{fig:shopping-list}
+  \vspace*{.25in}
+\end{wrapfigure}
 and Bob's quantity as per the definition of \C{Qty.merge}.

### 2017-02-28 paper/implementation.tex
edits

@@
 
-We describe the realization of \name programming model on top of a persistent
-distributed store and the language extensions to enable and enforce the
-constraints of the \name.
+We describe the realization of \name on top of a persistent
+distributed store and the language extensions to enable and enforce
+the constraints of the \name.
 

### 2017-02-28 paper/intro.tex
edits

@@
 impractical, designating a single node to hold the counter state
-eschewing replication altogether (as in a client-server
-configuration~\cite{Armstrong}), is also an undesirable solution,
-given the sensitivity of such architectures to network partitions and
-server failures, and the negative performance impact it incurs in
-geo-distributed environments~\cite{Walter}.  Removing coordination
-altogether by simply replicating counter state without having any
-supporting consistency protocol is an equally infeasible approach.
+eschewing replication altogether (as in a typical client-server
+configuration), is also an undesirable solution, given the sensitivity
+of such architectures to network partitions and server failures, and
+the negative performance impact it incurs in geo-distributed
+environments~\cite{Walter}.  Removing coordination altogether by
+simply replicating counter state without having any supporting
+consistency protocol is an equally infeasible approach.
 
@@
   }
-	\vspace*{-1.2in}
+	\vspace*{-1in}
 \end{wrapfigure}
@@
 that realizes a monadic version control system centered around data,
-rather than file, coherence.  The \name (Versioned ML) monad lets
-programmers write and compose concurrent computations around multiple
-(implicit) versions of a mergeable datatype, an ordinary ML datatype
-additionally equipped with a \emph{merge} function responsible for
-deriving a consistent global state from a collection of local versions
-of that state.  A computation progresses by forking (i.e.,
-replicating) new versions of existing versions, allowing
-synchronization-free local computation to proceed on those nodes,
-creating new versions along an existing \emph{branch} that represents
-updated local instances, and by merging branches to realize global
-consistency.
+rather than file, coherence.  The \name (Versioned ML) programming
+model comprises the {\sc vst} monad which lets programmers write and
+compose concurrent computations around multiple (implicit) versions of
+a mergeable datatype, an ordinary ML datatype additionally equipped
+with a \emph{merge} function responsible for deriving a consistent
+global state from a collection of local versions of that state.  A
+computation progresses by forking (i.e., replicating) new versions of
+existing versions, allowing synchronization-free local computation to
+proceed on those nodes, creating new versions along an existing
+\emph{branch} that represents updated local instances, and by merging
+branches to realize global consistency.
 
@@
 familiarity of using standard libraries already provided by the
-language.  Notably, while \name's programming model makes no explicit
-reference to any specific operational manifestation of a distributed
-system (e.g., programmers do not need to explicitly manage replicas),
-we demonstrate that it can be nonetheless efficiently realized on
-existing real-world geo-replicated distributed systems.
+language.  Notably, while \name\ makes no explicit reference to any
+specific operational manifestation of a distributed system (e.g.,
+programmers do not need to explicitly manage replicas), we demonstrate
+that it can be nonetheless efficiently realized on existing real-world
+geo-replicated distributed systems.
 
@@
 
-    \item We describe \name programming model that brings to bear the
-      power and flexibility of a version control system to the
-      administration of replicated data. \name hides the complexity of
-      version control behind a monad, and exposes its functionality
-      via a simple API that lets programmers define and compose
-      distributed ML computations around such data.  Issues related to
-      replication only manifest implicitly in the definition of a
-      merge operation that defines coherence among different instances
-      of replicated state.
+    \item We describe \name, a programming model that extends ML to
+      brings to bear the power and flexibility of a version control
+      system to the administration of replicated data. \name hides the
+      complexity of version control behind a monad, and exposes its
+      functionality via a simple API that lets programmers define and
+      compose distributed ML computations around such data.  Issues
+      related to replication only manifest implicitly in the
+      definition of a merge operation that defines coherence among
+      different instances of (presumably) replicated state.
 

### 2017-02-28 paper/meta.tex
edits

@@
 Fig.~\ref{fig:criss-cross-lcas} represents a bank account balance,
-which is expected to disallows any activity on the account if the
+which is expected to disallow any activity on the account if the
 balance is ever known to be less than zero.  From the perspective of
@@
 structure such that the problem never arises. The constraints are
-imposed either implicitly, as a result of how operational semantics
+imposed either implicitly, as a result of how the operational semantics
 defines an atomic step, or explicitly, by insisting that certain
 conditions be met before merging a pair of versions
-(\rulelabel{E-Pull-Wait}). Firstly, the operational semantics already
+(\rulelabel{E-Pull-Wait}). First, the operational semantics already
 disables criss-cross merges since it only ever merges versions that
-are latest on their respective branches. Secondly, we impose certain
+are latest on their respective branches. Second, we impose certain
 pre-conditions on the merging branches to preempt the structure shown
@@
 and some other branch (call it $b$). Now, if $b_3$ merges $v_{12}$,
-same could be true of $v_{12}$. Since $v_{12}$ and $v_{22}$ are not
+the same could be true of $v_{12}$. Since $v_{12}$ and $v_{22}$ are not
 orderd by the ancestor relation, both become LCAs of
@@
 \begin{definition} [\bfseries Locus]
-Given a branch $b$ and a version $v\in b$, locus ($v_o$) of $v$ is an
+Given a branch $b$ and a version $v\in b$, the locus ($v_o$) of $v$ is an
 ancestor that is not an ancestor of any other external ancestor of
@@
 In a legal branching history, every version has a unique locus
-(property follows from Def.~\ref{def:mergeability} below). Let
-\emph{causal history} of a version be defined as the set of all
+(property follows from Def.~\ref{def:mergeability} below). We
+define the \emph{causal history} of a version as the set of all
 ancestors of that version. A locus ($v_o$) of a version ($v$) is
@@
 semantics, if two branches, $b_i$ and $b_j$ are not mergeable (as per
-Def.~\ref{def:mergeability}), then there exist a sequence of fork and
+Def.~\ref{def:mergeability}), then there exists a sequence of fork and
 merge operations (between mergeable versions) that can be performed on
@@
 
-Proof is by the induction on the size of branching history. The
-observation that the smallest history that is the union of causal
-histories of two unmergeable versions (call them $v_i$ and $v_j$) is
-smaller than the history that includes the unmergeable versions. Since
-causal histories have maximum versions (respective locii), which, by
-inductive hypothesis, can be made mergeable, we merge them to yeild a
-new version $v$ that is the maximum of the causal histories of both
-$v_i$ and $v_j$. If $v_i$ and $v_j$ are the latest versions on their
-respective branches, the branches are now mergeable.
+The proof is by the induction on the size of branching history. The
+key observation is that the smallest history that is the union of
+causal histories of two unmergeable versions (call them $v_i$ and
+$v_j$) is smaller than the history that includes the unmergeable
+versions. Since causal histories have maximum versions (respective
+locii), which, by the inductive hypothesis, can be made mergeable, we
+merge them to yield a new version $v$ that is the maximum of the
+causal histories of both $v_i$ and $v_j$. If $v_i$ and $v_j$ are the
+latest versions on their respective branches, the branches are now
+mergeable.
 

### 2017-02-28 paper/model.tex
edits

--- 2017-02-28 paper/model.tex
@@
 
-The \name programming model lets programmers define and compose
-concurrent computations around versioned data structures.
-Fig.~\ref{fig:dali-monad} shows the signature of the \name module that
-implements the programming model along the lines of the well-known
-\C{State} monad~\cite{wadler-monad}. The monad encapsulates a
-versioned functional state (\C{'a}) and the type \C{('a, 'b) t}
-represents a monadic computation that returns a \C{'b} result.
-Functions \C{return} and \C{bind} have their usual monadic
+\name\ lets programmers define and compose concurrent computations
+around versioned data structures.  Fig.~\ref{fig:dali-monad} shows the
+signature of the \nameMonad module that implements the programming
+model along the lines of the well-known \C{State} monad. The monad
+encapsulates a versioned functional state (\C{'a}) and the type
+\C{('a, 'b) t} represents a monadic computation that returns a \C{'b}
+result.  Functions \C{return} and \C{bind} have their usual monadic
 \begin{figure}

### 2017-02-28 paper/implementation.tex
Edits.

--- 2017-02-28 paper/implementation.tex
@@
 	\end{center}
-	\caption{Network partitions may leave branches $b_1$ and $b_4$ in one
-		partition (red background), and branches $b_2$ and $b_3$ in the other
-		(yellow background). Since $b_2$ and $b_3$ cannot merge unless $b_1$ and
-		$b_4$ merge, they cannot communicate among themselves even though they are in
-		the same network partition.}
+  \caption{Network partitions may leave branches $b_1$ and $b_4$ in
+  one partition (red background), and branches $b_2$ and $b_3$ in the
+  other (yellow background). Access to full history on both sides of
+  the partition lets both sides make progress.}
 	\label{fig:partitions}
@@
 other replicas. In presence of network partitions, the nodes in one partition
-may not receive updates from the other partition, hence merges cannot happen
-across a partition. This can be a problem if a pair of branches in one
-partition (the current partition) have multiple LCAs in the other partition
-(the remote partition), hence cannot merge. Fig.~\ref{fig:partitions}
-illustrate this problem for the branching structure of
+may not receive updates from the other partition, hence merges cannot
+happen across a partition. This can be a problem if a pair of branches
+in one partition (the current partition) have their loci
+(Sec.~\ref{sec:meta}) in the other partition (the remote partition),
+hence cannot merge. Fig.~\ref{fig:partitions} illustrate this problem
+for the branching structure of
 Fig.~\ref{fig:external-lcas}.
 
-Fortunately, the access to full (causal) branching history assured by Irmin at
-every node comes to the rescue. Relying on the history, the current partition
-can fork-off (\C{fork}) new branches that start where the remote branches left,
-and map them to new (virtual) nodes that emulate the nodes in the remote
-partition. For the example in Fig.~\ref{fig:partitions}, the partition of $b_1$
-and $b_4$ can fork new branches $b_2'$ and $b_3'$ from $b_2$ and $b_3$,
-respectively, and resume the activity as if the partition never happened. After
-merging the first versions on $b_2'$ and $b_3'$, branches $b_1$ and $b_2$ will
-be in the same state as before (when the partition happened), except that the
-branches are now mergeable because $b_2'$ and $b_3'$ can merge. The ability to
-track full provenance information is thus crucial for \name to overcome network
-partitions, making it an appropriate programming model for highly-available
-replicated data types. \name monad conceals the branching structure, but also
-transparently performs the necessary merges to obtain a history graph with a
-unique lowest common ancestor.
+Fortunately, the access to full (causal) branching history assured by
+Irmin at every node comes to the rescue. Relying on the history, the
+current partition can fork-off (\C{fork}) new branches that start
+where the remote branches left, and map them to new (virtual) nodes
+that emulate the nodes in the remote partition. For the example in
+Fig.~\ref{fig:partitions}, the partition of $b_1$ and $b_4$ can fork
+new branches $b_2'$ and $b_3'$ from $b_2$ and $b_3$, respectively, and
+resume the activity as if the partition never happened. The ability to
+track full provenance information is thus crucial for \name to
+overcome network partitions, making it an appropriate programming
+model for highly-available replicated data types. \name monad conceals
+the branching structure, but also transparently performs the necessary
+merges to obtain a history graph with a unique lowest common ancestor.
 

### 2017-02-28 paper/evaluation.tex

@@
 protocols for transferring updates between replicas. In contrast, \name is
-realized over Irmin~\cite{irmin} which uses Git transfer protocol~\cite{..}, a
-protocol not designed with distributed data stores in mind. Hence, we also
-evaluate the performance of synchronization and merging.
+realized over Irmin~\cite{irmin} which uses Git transfer
+protocol~\cite{git-tp}, a protocol not designed with distributed data stores in
+mind. Hence, we also evaluate the performance of synchronization and merging.
 

### 2017-02-28 paper/semantics.tex
edits

@@
 \subsection{Operational Semantics}
+\label{sec:opsem}
 
@@
 which can be composed to get the desired effect of the original:
-\begin{smathpar}
-  \C{sync}\;x \;=\; (\lambda y.\pull)\; (\push\,x)
-\end{smathpar}
-The semantics of\C{get\_current\_version} is subsumed by \C{pull},
-hence elided.  Values ($v$) are constants and lambda abstractions.  A
-program ($p$) is a parallel composition of threads, where each thread
-is an expression ($s$) indexed by the corresponding thread identifier
-($t$).
+$\C{sync}\;x \;=\; (\lambda y.\pull)\; (\push\,x)$.  The semantics
+of\C{get\_current\_version} is subsumed by \C{pull}, hence elided.
+Values ($v$) are constants and lambda abstractions.  A program ($p$)
+is a parallel composition of threads, where each thread is an
+expression ($s$) indexed by the corresponding thread identifier ($t$).
 
@@
 to a value that constitutes an initial version (recall that \C{run}
-models \name's \C{with\_init\_version\_do}). A program evaluation
+models \nameMonad's \C{with\_init\_version\_do}). A program evaluation
 context non-deterministically picks an extant thread to evaluate. The
@@
 synchronize with all concurrent branches because not all such branches
-may be available.)
+may be available, because of e.g., network partitions.)
 
@@
   $(v_2, \C{MERGE}\;(v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
-  \item $v_1 = v_2$, or $v_1$ is transitively a ancestor of
+  \item $v_1 = v_2$, or $v_1$ is transitively an ancestor of
   $v_2$, i.e., $\exists v.~ \under{H}{v_1 \preceq v} \conj
@@
 version $v$ only if they are safely mergeable ($\under{H}{v' \mbleto
-v}$), i.e., they have a single LCA, and merging both versions does not
-lead to a case where a branch (i.e., its latest version) has multiple
-LCAs with rest of the branches, hence gets \emph{stuck}. Such nuances
-of LCA, and the guarantees provided by the system notwithstanding
-these nuances, are discussed in the Sec.~\ref{sec:meta}.
+  v}$), i.e., they have a single LCA, and merging both versions does
+not lead to a case where a branch (i.e., its latest version) has
+multiple LCAs with remaining branches, andd hence gets
+\emph{stuck}. Such nuances of LCA, and the guarantees provided by the
+system notwithstanding these nuances, are discussed in 
+Sec.~\ref{sec:meta}.
 

### 2017-02-28 paper/implementation.tex
edits

--- 2017-02-28 paper/implementation.tex
@@
 
-\name programming model is realized on top of Irmin~\cite{irmin}, an OCaml
-library database implementation that is part of the MirageOS
-project~\cite{mirage}. Irmin provides a persistent multi-versioned store with
-with a content-addressable heap abstraction. Simply put, content-addressability
-means that the address of a data block is determined by its content. If the
-content changes, then so does the address. Old content continues to be
-available the old address. Content-addressability also results in constant time
+The \name programming model is realized on top of Irmin~\cite{irmin},
+an OCaml library database implementation that is part of the MirageOS
+project~\cite{mirage}. Irmin provides a persistent multi-versioned
+store with a content-addressable heap abstraction. Simply put,
+content-addressability means that the address of a data block is
+determined by its content. If the content changes, then so does the
+address. Old content continues to be available from the old
+address. Content-addressability also results in constant time
 structural equality checks, which we exploit in our mergeable rope
-implementation (Section~\ref{sec:ropes}).
+implementation (Section~\ref{sec:ropes}), among others.
 
-Irmin provides support for distribution, fault-tolerance and concurrency
-control by incorporating Git distributed version control~\cite{git} protocol
-over its object model. Indeed, Irmin is fully compatible with the Git command
-line tools. Distributed replicas in \name are created by cloning a \name
-repository. Due to \name's mergeable type, each replica can operate completely
-independently, accepting client requests, even when disconnected from other
-replicas, resulting in a fully available distributed database system.
+Irmin provides support for distribution, fault-tolerance and
+concurrency control by incorporating the Git distributed version
+control~\cite{git} protocol over its object model. Indeed, Irmin is
+fully compatible with Git command line tools. Distributed replicas in
+\name are created by cloning a \name repository. Due to \name's
+support for mergeable types, each replica can operate completely
+independently, accepting client requests, even when disconnected from
+other replicas, resulting in a highly available distributed system.
 
@@
 
-Concurrent operations in Irmin are tracked within the notion of branches,
-allowing the programmer to explicitly merge the branches on demand. \name
-realizes the concurrent programming model on top of the notion of branches;
-each replica operates on its own branch, and thus is isolated from actions on
-other replicas. In presence of network partitions, the nodes in one partition
-may not receive updates from the other partition, hence merges cannot happen
-across a partition. This can be a problem if a pair of branches in one
-partition (the current partition) have multiple LCAs in the other partition
-(the remote partition), hence cannot merge. Fig.~\ref{fig:partitions}
-illustrate this problem for the branching structure of
-Fig.~\ref{fig:external-lcas}.
+Concurrent operations in Irmin are tracked using the notion of
+branches, allowing the programmer to explicitly merge branches on
+demand. \name's concurrency and distribution model is realized in
+terms of branches; each replica operates on its own branch, and thus
+is isolated from the actions on other replicas. In the presence of
+network partitions, the nodes in one partition may not receive updates
+from the other partition, preventing merges from happening across a
+partition. This can be a problem if a pair of branches in one
+partition (the current partition) have multiple LCAs in the other
+partition (the remote partition), and hence cannot
+merge. Fig.~\ref{fig:partitions} illustrate this problem for the
+branching structure of Fig.~\ref{fig:external-lcas}.
 
-Fortunately, the access to full (causal) branching history assured by Irmin at
-every node comes to the rescue. Relying on the history, the current partition
-can fork-off (\C{fork}) new branches that start where the remote branches left,
-and map them to new (virtual) nodes that emulate the nodes in the remote
-partition. For the example in Fig.~\ref{fig:partitions}, the partition of $b_1$
-and $b_4$ can fork new branches $b_2'$ and $b_3'$ from $b_2$ and $b_3$,
-respectively, and resume the activity as if the partition never happened. After
-merging the first versions on $b_2'$ and $b_3'$, branches $b_1$ and $b_2$ will
-be in the same state as before (when the partition happened), except that the
-branches are now mergeable because $b_2'$ and $b_3'$ can merge. The ability to
-track full provenance information is thus crucial for \name to overcome network
-partitions, making it an appropriate programming model for highly-available
-replicated data types. \name monad conceals the branching structure, but also
-transparently performs the necessary merges to obtain a history graph with a
-unique lowest common ancestor.
+Fortunately, access to full (causal) branching history assured by
+Irmin at every node comes to the rescue. Relying on the history, the
+current partition can fork-off (\C{fork}) new branches that start
+where the remote branches left, and map them to new (virtual) nodes
+that emulate the nodes in the remote partition. For the example in
+Fig.~\ref{fig:partitions}, the partition of $b_1$ and $b_4$ can fork
+new branches $b_2'$ and $b_3'$ from $b_2$ and $b_3$, respectively, and
+resume the activity as if the partition never happened. After merging
+the first versions on $b_2'$ and $b_3'$, branches $b_1$ and $b_2$ will
+be in the same state as before (when the partition happened), except
+that the branches are now mergeable because $b_2'$ and $b_3'$ can
+merge. The ability to track full provenance information is thus
+crucial for \name to overcome network partitions, making it an
+appropriate programming model for highly-available replicated data
+types. The \nameMonad conceals branching structure, but also
+transparently performs the necessary merges to obtain a history graph
+with a unique lowest common ancestor.
 
@@
 transformations between objects on OCaml and Irmin heaps. We also derive the
-various serialization functions required Irmin. This address the representation
+various serialization functions required by Irmin. This address the representation
 mismatch between the application and storage layers. Synchronization between

### 2017-02-28 paper/implementation.tex

@@
 branches, allowing the programmer to explicitly merge branches on
-demand. \name's concurrency and distribution model is realized in
+demand. \name's concurrency and distribution model is also realized in
 terms of branches; each replica operates on its own branch, and thus
@@
 
-Importantly, Irmin supports user-defined three-way merge functions for
-reconciling concurrent operations. While Irmin's merge functions are defined
+Importantly, Irmin supports user-defined three-way merge functions to
+reconcile concurrent operations.  While Irmin's merge functions are defined
 over objects on Irmin's content-addressable heap, \name's merge functions are

### 2017-02-28 paper/implementation.tex
Edits.

--- 2017-02-28 paper/implementation.tex
@@
 Importantly, Irmin supports user-defined three-way merge functions for
-reconciling concurrent operations. While Irmin's merge functions are defined
-over objects on Irmin's content-addressable heap, \name's merge functions are
-defined over OCaml types. We address this representational mis-match with the
-help of OCaml's PPX metaprogramming support~\cite{ppx} to derive bi-directional
-transformations between objects on OCaml and Irmin heaps. We also derive the
-various serialization functions required Irmin. This address the representation
-mismatch between the application and storage layers. Synchronization between
-replicas is performed using Git's notion of pushing and pulling updates from
-remotes over the Git transfer protocol~\cite{git-tp}. As a result, we reap the
-benefits of efficient delta-transfer (only missing objects are transferred
-between replicas), compression and end-to-end encrypted communication between
+reconciling concurrent operations. While Irmin's merge functions are
+defined over objects on Irmin's content-addressable heap, \name's
+merge functions are defined over OCaml types. We address this
+representational mismatch with the help of OCaml's PPX metaprogramming
+support~\cite{ppx} to derive bi-directional transformations between
+objects on OCaml and Irmin heaps. We also derive the various
+serialization functions required by Irmin\footnote{
+  The Since \nameMonad monad relies on Irmin for realizing the branch
+  structure, it is in fact parameterized over a mergeable type that
+  implements all these functions. Thus the monad is functorized over
+  the mergeable type, but not polymorphic. The convenience of type
+  parameter inference that polymorphism offers can however be obtained
+  for module parameters using OCaml extensioms, such as Modular
+  Implicits~\cite{implicits}.
+}.
+Synchronization between replicas is performed using Git's notion of
+pushing and pulling updates from remotes over the Git transfer
+protocol~\cite{git-tp}. As a result, we reap the benefits of efficient
+delta-transfer (only missing objects are transferred between
+replicas), compression and end-to-end encrypted communication between
 the replicas.

### 2017-02-28 paper/implementation.tex

--- 2017-02-28 paper/implementation.tex
@@
 
-Importantly, Irmin supports user-defined three-way merge functions to
-reconcile concurrent operations.  While Irmin's merge functions are defined
-over objects on Irmin's content-addressable heap, \name's merge functions are
-defined over OCaml types. We address this representational mis-match with the
-help of OCaml's PPX metaprogramming support~\cite{ppx} to derive bi-directional
-transformations between objects on OCaml and Irmin heaps. We also derive the
-various serialization functions required by Irmin. This address the representation
-mismatch between the application and storage layers. Synchronization between
-replicas is performed using Git's notion of pushing and pulling updates from
-remotes over the Git transfer protocol~\cite{git-tp}. As a result, we reap the
-benefits of efficient delta-transfer (only missing objects are transferred
-between replicas), compression and end-to-end encrypted communication between
-the replicas.
+Importantly, Irmin supports user-defined three-way merge functions for
+reconciling concurrent operations. While Irmin's merge functions are
+defined over objects on Irmin's content-addressable heap, \name's
+merge functions are defined over OCaml types. We address this
+representational mismatch with the help of OCaml's PPX metaprogramming
+support~\cite{ppx} to derive bi-directional transformations between
+objects on OCaml and Irmin heaps. We also derive the various
+serialization functions required by Irmin\footnote{
+  The Since \nameMonad monad relies on Irmin for realizing the branch
+  structure, it is in fact parameterized over a mergeable type that
+  implements all these functions. Thus the monad is functorized over
+  the mergeable type, but not polymorphic. The convenience of type
+  parameter inference that polymorphism offers can however be obtained
+  for module parameters using OCaml extensioms, such as Modular
+  Implicits~\cite{implicits}.
+}.
+Synchronization between replicas is performed using Git's notion of
+pushing and pulling updates from remotes over the Git transfer
+protocol~\cite{git-tp}. As a result, we reap the benefits of efficient
+delta-transfer (only missing objects are transferred between
+replicas), compression and end-to-end encrypted communication between
+he replicas.

### 2017-02-28 paper/conc.tex

\section{Conclusions}
\label{sec:conclusions}

Replication is a critical feature in geo-distributed systems used to
improve latency and availability.  Existing approaches to dealing with
replicated data often involve complex and subtle code restructuring,
and the use of specialized datatypes that interact poorly with other
language structures.  \name\ is a programming model that extends ML
with implicit support for replication that overcomes these concerns.
Its notion of concurrency and synchronization is encapsulated within a
monad that considers distrbuted computation in terms of typed versions
of replicated state.  Notably, any ML data structure can participate
in a distributed computation using this monad by providing a merge
function that dictates how different versions of that structure can be
reconciled.  Our versioning semantics enjoys strong consistency and
progress guarantees that make it useful in a fully decentralized (and
unreliable) cloud environment.  Furthermore, our experimental results
demonstrate that using such high-level abstractions need not come at
the expense of efficient implementations.

### 2017-02-28 paper/mergeable-types.tex
edits
Added a note.

@@
 
+The proof follows directly from the fact that the values on branches
+are computed from previous values either by applying morphisms
+locally, or by merging from the remote branches. In the latter case,
+Def.~\ref{def:mergeable-type} guarantees that the existence of a
+morphism from the merging value to the merged value.
+
 \noindent Rather than specifying merge ($M$) in terms of an LCA as we
@@
 
-\begin{definition} [\bfseries Mergeable Type]
+\begin{definition} [\bfseries Mergeable Type (Strengthened)]
 \label{def:mergeable-type-2}
@@
 type, and lets us explain why it is a mergeable type.  \C{MList.edit\_seq}
-and \C{MList.op\_transform}, respectively, are the evidences for the
-existence of $W$ and $M$ functions.
+and \C{MList.op\_transform}, respectively, provide evidences for the
+existence of $W$ and $M$.

### 2017-02-28 paper/related.tex

@@
 means to programming eventually consistent distributed systems was
-further developed in~\cite{BFL+12,BLF+12}.  The
+further developed in~\cite{BFL+12}.  The
 \name\ programming model, however, differs substantially from a
@@
 
-Modern distributed systems are often equipped with only parsimonious
-data models (e.g., key-value model) and poorly understood low-level
-consistency guarantees\footnote{Cassandra~\cite{Cassandra}, a popular
-  NoSQL data store, comes with various consistency enforcement
-  mechanisms, such as anti-entropy protocols, {\sc QUORUM} and {\sc
-    LOCAL\_QUORUM} reads and writes, and light-weight transactions,
-  each of which can be controlled via configuration knobs or runtime
-  parameters.}  that complicate program reasoning, and make it hard to
-enforce application integrity. Some authors~\cite{bailis-bolton} have
-demonstrated that it is possible to\emph{bolt on} high-level
-consistency guarantees (e.g., causal consistency)~\cite{COPS,BEG+17}
-as a \emph{shim layer} service over existing stores.
+Modern distributed systems are often equipped with only parsimonious data
+models (e.g., key-value model) and poorly understood low-level consistency
+guarantees\footnote{Cassandra~\cite{Cassandra}, a popular NoSQL data store,
+comes with various consistency enforcement mechanisms, such as anti-entropy
+protocols, {\sc QUORUM} and {\sc LOCAL\_QUORUM} reads and writes, and
+light-weight transactions, each of which can be controlled via configuration
+knobs or runtime parameters.}  that complicate program reasoning, and make it
+hard to enforce application integrity. Some authors~\cite{bailis-bolton} have
+demonstrated that it is possible to\emph{bolt on} high-level consistency
+guarantees (e.g., causal consistency)~\cite{COPS,BEG+17} as a \emph{shim layer}
+service over existing stores.
 
@@
 
-There have been numerous proposals over the years for realizing
-distributed functionality in a functional programming context.
-Poly/ML~\cite{Mat97} and Facile~\cite{TLK+93} extend Standard ML,
-while Acute~\cite{SLW+05} and HashCaml~\cite{BSS+06} extend OCaml, with
-distributed programming abstractions and support, with particular
-focus on type-safe marshalling and mobility.  Kali-Scheme~\cite{CJK95}
-defines a distributed extension for Scheme-48, and Cloud
-Haskell~\cite{EBPJ11} describes a domain-specific language that
-enables development of distributed Haskell programs.
-Obliq~\cite{Car95} explores issues related to mobility, lexical
-scoping, and network references in an object-oriented framework that
-enables safe (i.e., lexical) transmission of closures.  However,
-unlike \name, none of these systems consider issues of object
-replication or consistency as central to their programming models.
-The realities of modern-day scalable distributed systems, especially
-those that are constructed from geographically distributed
-participants, dictate that we must necessarily grapple with
-consistency issues to achieve any kind of reasonable performance; it
-is noteworthy that replication is inherent in many widely-used
+There have been numerous proposals over the years for realizing distributed
+functionality in a functional programming context. Poly/ML~\cite{Mat97} and
+Facile~\cite{TLK+93} extend Standard ML, while Acute~\cite{SLW+05} and
+HashCaml~\cite{BSS+06} extend OCaml, with distributed programming abstractions
+and support, with particular focus on type-safe marshalling and mobility.
+Kali-Scheme~\cite{CJK95} defines a distributed extension for Scheme-48, and
+Cloud Haskell~\cite{EBPJ11} describes a domain-specific language that enables
+development of distributed Haskell programs. Obliq~\cite{Car95} explores issues
+related to mobility, lexical scoping, and network references in an
+object-oriented framework that enables safe (i.e., lexical) transmission of
+closures. Orleans~\cite{Orleans} provides persistent actor-based programming
+model for elastic cloud application, but the focus is on strong consistency,
+mirroring a single-threaded execution.
+
+However, unlike \name, none of these systems consider issues of object
+replication or consistency as central to their programming models. The
+realities of modern-day scalable distributed systems, especially those that are
+constructed from geographically distributed participants, dictate that we must
+necessarily grapple with consistency issues to achieve any kind of reasonable
+performance; it is noteworthy that replication is inherent in many widely-used
 cloud-based distributed system implementations today (e.g.,
-Cassandra~\cite{Cassandra} or Dynamo~\cite{Dynamo}), which provide
-only weak consistency guarantees.  Rather than exposing a low-level
-operational treatment of replication, foisting the burden of reasoning
-about replicated state onto the programmer, our primary contribution
-is the development of a principled high-level approach to framing
-these issues that abstracts these kinds of system-level details.  Our
-goal is to leverage functional programming principles to minimize
-disruption to the way programmers structure and reason about their
-applications, without compromising efficiency or performance.  Indeed,
-the semantics of mergeable types and versioning has utility in any
+Cassandra~\cite{Cassandra} or Dynamo~\cite{Dynamo}), which provide only weak
+consistency guarantees.  Rather than exposing a low-level operational treatment
+of replication, foisting the burden of reasoning about replicated state onto
+the programmer, our primary contribution is the development of a principled
+high-level approach to framing these issues that abstracts these kinds of
+system-level details.  Our goal is to leverage functional programming
+principles to minimize disruption to the way programmers structure and reason
+about their applications, without compromising efficiency or performance.
+Indeed, the semantics of mergeable types and versioning has utility in any
 concurrent setting that must deal with non-trivial coordination and
-synchronization costs, even without taking distribution or replication
-issues into account.
+synchronization costs, even without taking distribution or replication issues
+into account.
+
+\name shares resemblance to conflict-free replicated data types
+(CRDT)~\cite{crdt}. CRDTs define abstract data types such as counters, sets,
+etc., with commutative operations such that the state of the data type always
+converges. Unlike CRDTs, the operations on mergeable types in \name need not
+commute and the reconciliation protocol is defined by user-defined merge
+functions. Moreover, CRDTs are not composable and each CRDT tends to be built
+from scratch over the network protocol. Compare to this, \name types are
+composable; mergeable ropes are polymorphic over its mergeable content. \name
+uses 3-way merges using the lowest common ancestor, which is critical for all
+of our user-defined merges. However, CRDTs do not have the benefit of lowest
+common ancestor for merges and are only presented with the two concurrent
+versions. If a 3-way merge is desired, then the causal history has to be
+explicitly encoded in the data type. As a result, constructing even simple data
+types like counters are more complicated using CRDTs~\cite{crdt} compared to
+their implementation in \name.

### 2017-02-28 paper/related.tex
edits

@@
 from scratch over the network protocol. Compare to this, \name types are
-composable; mergeable ropes are polymorphic over its mergeable content. \name
-uses 3-way merges using the lowest common ancestor, which is critical for all
-of our user-defined merges. However, CRDTs do not have the benefit of lowest
-common ancestor for merges and are only presented with the two concurrent
-versions. If a 3-way merge is desired, then the causal history has to be
-explicitly encoded in the data type. As a result, constructing even simple data
-types like counters are more complicated using CRDTs~\cite{crdt} compared to
-their implementation in \name.
+composable; mergeable ropes are polymorphic over its mergeable content. A
+programmer working working with CRDTs is provided with a fixed set of mergeable
+data types, with no recourse to constructing their own from the available
+types.
+
+\name uses 3-way merges using the lowest common ancestor, which is critical for
+all of our user-defined merges. However, CRDTs do not have the benefit of
+lowest common ancestor for merges and are only presented with the two
+concurrent versions. If a 3-way merge is desired, then the causal history has
+to be explicitly encoded in the data type. As a result, constructing even
+simple data types like counters are more complicated using CRDTs~\cite{crdt}
+compared to their implementation in \name. CRDTs also tend to be implemented
+directly over the network protocols. Hence, low-level concerns such as
+duplicate delivery, lost messages, message reordering are explicitly handled in
+the data type definition. In \name, the programmer does not have to deal with
+underlying network issues since \name is realized on Irmin, a high-level
+branch-consistent distributed store.

### 2017-02-28 paper/model.tex

--- 2017-02-28 paper/model.tex
@@
 
-The \name programming model lets programmers define and compose
-concurrent computations around versioned data structures.
-Fig.~\ref{fig:dali-monad} shows the signature of the \name module that
-implements the programming model along the lines of the well-known
-\C{State} monad~\cite{wadler-monad}. The monad encapsulates a
-versioned functional state (\C{'a}) and the type \C{('a, 'b) t}
-represents a monadic computation that returns a \C{'b} result.
-Functions \C{return} and \C{bind} have their usual monadic
+\name\ lets programmers define and compose concurrent computations
+around versioned data structures.  Fig.~\ref{fig:dali-monad} shows the
+signature of the \nameMonad module that implements the programming
+model along the lines of the well-known \C{State}
+monad~\cite{wadler-monad}. The monad encapsulates a versioned
+functional state (\C{'a}) and the type \C{('a, 'b) t} represents a
+monadic computation that returns a \C{'b} result.  Functions
+\C{return} and \C{bind} have their usual monadic
 \begin{figure}

### 2017-02-28 paper/examples.tex
edits

@@
 \C{edit\_seq} and \C{op\_transform}. Both these functions have been
-implemented in less than 50 lines of OCaml included in the appendix.
-The functions are described below.
+implemented in less than 50 lines of OCaml.  The functions are described below.
 \begin{itemize}
@@
 		time~\cite{wagner-fischer}.  \C{edit\_seq} modifies one such algorithm to
-		also compute the edit sequence (an \C{edit list}). The implementation
-		specified in Fig.~\ref{fig:mlist-sig} represents edits using the type
+		also compute the edit sequence (an \C{edit list}). The signature
+		specified in Fig.~\ref{fig:mlist} represents edits using the type
 		\C{edit}.  Constructors \C{I}, \C{D}, and \C{S} stand for \C{insert},

### 2017-02-28 paper/related.tex
edits

--- 2017-02-28 paper/related.tex
@@
 
-There have been numerous proposals over the years for realizing distributed
-functionality in a functional programming context. Poly/ML~\cite{Mat97} and
-Facile~\cite{TLK+93} extend Standard ML, while Acute~\cite{SLW+05} and
-HashCaml~\cite{BSS+06} extend OCaml, with distributed programming abstractions
-and support, with particular focus on type-safe marshalling and mobility.
-Kali-Scheme~\cite{CJK95} defines a distributed extension for Scheme-48, and
-Cloud Haskell~\cite{EBPJ11} describes a domain-specific language that enables
-development of distributed Haskell programs. Obliq~\cite{Car95} explores issues
-related to mobility, lexical scoping, and network references in an
-object-oriented framework that enables safe (i.e., lexical) transmission of
-closures. Orleans~\cite{Orleans} provides persistent actor-based programming
-model for elastic cloud application, but the focus is on strong consistency,
-mirroring a single-threaded execution.
+There have been numerous proposals over the years for realizing
+distributed programming functionality in a functional programming
+context.  Erlang~\cite{Armstrong}, which is the most well-known, is
+based on a ``shared-nothing'' message-passing programming methodology.
+Poly/ML~\cite{Mat97} and Facile~\cite{TLK+93} extend Standard ML,
+while Acute~\cite{SLW+05} and HashCaml~\cite{BSS+06} extend OCaml,
+with distributed programming abstractions and support, with particular
+focus on type-safe marshalling and mobility.  Kali-Scheme~\cite{CJK95}
+defines a distributed extension for Scheme-48, and Cloud
+Haskell~\cite{EBPJ11} describes a domain-specific language that
+enables development of distributed Haskell
+programs. Obliq~\cite{Car95} explores issues related to mobility,
+lexical scoping, and network references in an object-oriented
+framework that enables safe (i.e., lexical) transmission of
+closures. Orleans~\cite{Orleans} provides persistent actor-based
+programming model for elastic cloud application, but the focus is on
+strong consistency, mirroring a single-threaded execution.
 
@@
 replication or consistency as central to their programming models. The
-realities of modern-day scalable distributed systems, especially those that are
-constructed from geographically distributed participants, dictate that we must
-necessarily grapple with consistency issues to achieve any kind of reasonable
-performance; it is noteworthy that replication is inherent in many widely-used
+realities of modern-day scalable cloud-based distributed systems,
+especially those that are constructed from geographically distributed
+participants, dictate that we must necessarily grapple with
+consistency issues to achieve any kind of reasonable performance; it
+is noteworthy that replication is inherent in many widely-used
 cloud-based distributed system implementations today (e.g.,
-Cassandra~\cite{Cassandra} or Dynamo~\cite{Dynamo}), which provide only weak
-consistency guarantees.  Rather than exposing a low-level operational treatment
-of replication, foisting the burden of reasoning about replicated state onto
-the programmer, our primary contribution is the development of a principled
-high-level approach to framing these issues that abstracts these kinds of
-system-level details.  Our goal is to leverage functional programming
-principles to minimize disruption to the way programmers structure and reason
-about their applications, without compromising efficiency or performance.
-Indeed, the semantics of mergeable types and versioning has utility in any
+Cassandra~\cite{Cassandra} or Dynamo~\cite{Dynamo}), which provide
+only weak consistency guarantees.  Rather than exposing a low-level
+operational treatment of replication, foisting the burden of reasoning
+about replicated state onto the programmer, our primary contribution
+is the development of a principled high-level approach to framing
+these issues that abstracts these kinds of system-level details.  Our
+goal is to leverage functional programming principles to minimize
+disruption to the way programmers structure and reason about their
+applications, without compromising efficiency or performance.  Indeed,
+the semantics of mergeable types and versioning has utility in any
 concurrent setting that must deal with non-trivial coordination and
-synchronization costs, even without taking distribution or replication issues
-into account.
+synchronization costs, even without taking distribution or replication
+issues into account.
 
-\name shares resemblance to conflict-free replicated data types
-(CRDT)~\cite{crdt}. CRDTs define abstract data types such as counters, sets,
-etc., with commutative operations such that the state of the data type always
-converges. Unlike CRDTs, the operations on mergeable types in \name need not
-commute and the reconciliation protocol is defined by user-defined merge
-functions. Moreover, CRDTs are not composable and each CRDT tends to be built
-from scratch over the network protocol. Compare to this, \name types are
-composable; mergeable ropes are polymorphic over its mergeable content. \name
-uses 3-way merges using the lowest common ancestor, which is critical for all
-of our user-defined merges. However, CRDTs do not have the benefit of lowest
-common ancestor for merges and are only presented with the two concurrent
-versions. If a 3-way merge is desired, then the causal history has to be
-explicitly encoded in the data type. As a result, constructing even simple data
-types like counters are more complicated using CRDTs~\cite{crdt} compared to
-their implementation in \name.
+Finally, \name shares some resemblance to conflict-free replicated
+data types (CRDT)~\cite{crdt}. CRDTs define abstract data types such
+as counters, sets, etc., with commutative operations such that the
+state of the data type always converges. Unlike CRDTs, the operations
+on mergeable types in \name need not commute and the reconciliation
+protocol is defined by user-defined merge functions. Moreover, CRDTs
+are not composable and each CRDT tends to be built from scratch over
+the network protocol. Compared to this, \name types are composable;
+mergeable ropes are polymorphic over its mergeable content. \name uses
+3-way merges using the lowest common ancestor, which is critical for
+all of our user-defined merges. However, CRDTs do not have the benefit
+of lowest common ancestor for merges and are only presented with the
+two concurrent versions. If a 3-way merge is desired, then the causal
+history has to be explicitly encoded in the data type. As a result,
+constructing even simple data types like counters are more complicated
+using CRDTs~\cite{crdt} compared to their implementation in \name.

### 2017-02-28 paper/meta.tex
Fig. added causal hist.

@@
 
+\begin{wrapfigure}{l}{.4\textwidth}
+\centering
+\includegraphics[scale=0.6]{Figures/causal-history}
+\caption{The cloud represents the set of all ancestors of versions $v_i$ and
+$v_{i+1}$, i.e., their causal history. The locus $v_k$ of both is the
+version that succeeds every other ancestor in the causal history.}
+\label{fig:causal-history}
+\end{wrapfigure}
 In a legal branching history, every version has a unique locus
-(property follows from Def.~\ref{def:mergeability} below). We
-define the \emph{causal history} of a version as the set of all
-ancestors of that version. A locus ($v_o$) of a version ($v$) is
-therefore the maximum element (lowest element in the ancestor
-relation) of the causal history of $v$'s internal ancestor that was
-the result of the last merge. We make use of this observation while
-defining mergeability.
+(property follows from Def.~\ref{def:mergeability} below). We define
+the \emph{causal history} of a version as the set of all ancestors of
+that version. A locus ($v_o$) of a version ($v$) is therefore the
+maximum element (lowest element in the ancestor relation) of the
+causal history of $v$'s internal ancestor that was the result of the
+last merge. This is illustrated in Fig.~\ref{fig:causal-history}. We
+make use of this observation while defining mergeability.
 

### 2017-02-28 paper/implementation.tex

@@
 Synchronization between replicas is performed using Git's notion of
-pushing and pulling updates from remotes over the Git transfer
-protocol~\cite{git-tp}. As a result, we reap the benefits of efficient
-delta-transfer (only missing objects are transferred between
+pushing and pulling updates from remotes~\cite{git-tp}. As a result, we reap the
+benefits of efficient delta-transfer (only missing objects are transferred between
 replicas), compression and end-to-end encrypted communication between
-he replicas.
+the replicas.

### 2017-02-28 paper/related.tex

@@
 using CRDTs~\cite{crdt} compared to their implementation in \name.
+
+\name uses 3-way merges using the lowest common ancestor, which is critical for
+all of our user-defined merges. However, CRDTs do not have the benefit of
+lowest common ancestor for merges and are only presented with the two
+concurrent versions. If a 3-way merge is desired, then the causal history has
+to be explicitly encoded in the data type. As a result, constructing even
+simple data types like counters are more complicated using CRDTs~\cite{crdt}
+compared to their implementation in \name. CRDTs also tend to be implemented
+directly over the network protocols. Hence, low-level concerns such as
+duplicate delivery, lost messages, message reordering are explicitly handled in
+the data type definition. In \name, the programmer does not have to deal with
+underlying network issues since \name is realized on Irmin, a high-level
+branch-consistent distributed store.

### 2017-02-28 paper/meta.tex
Minor edits in meta.tex

@@
 the same could be true of $v_{12}$. Since $v_{12}$ and $v_{22}$ are not
-orderd by the ancestor relation, both become LCAs of
+ordered by the ancestor relation, both become LCAs of
 $b_3$ and $b$. We observe that this scenario can be prevented if, when
@@
 mergeable. The definition essentially requires all of $v_1$'s causal
-history that preceeds its locus $v_o$ to be included in $v_2$'s causal
+history that precedes its locus $v_o$ to be included in $v_2$'s causal
 history, and none of $v_1$'s history that succeeds $v_o$ to be
@@
 history whose maximum (w.r.t the ancestor relation) is the version
-$v_o$. Consquently, $v_o$ becomes the LCA for $v_2$'s merge into
+$v_o$. Consequently, $v_o$ becomes the LCA for $v_2$'s merge into
 $v_1$. Generalizing this observation for the latest versions on
@@
 Every pair of branches in a legal history $H$, if mergeable as per
-Def.~\ref{def:mergeability}, have a unique least common ancestor.
+Def.~\ref{def:mergeability}, have a unique lowest common ancestor.
 \end{theorem}
@@
 merge operations (between mergeable versions) that can be performed on
-$H$ to yeild a new history $H'$, where $b_i$ and $b_j$ are mergeable.
+$H$ to yield a new history $H'$, where $b_i$ and $b_j$ are mergeable.
 \end{theorem}

### 2017-02-28 paper/implementation.tex
edit

@@
 serialization functions required by Irmin\footnote{
-  The Since \nameMonad monad relies on Irmin for realizing the branch
+  Since the \nameMonad monad relies on Irmin for realizing the branch
   structure, it is in fact parameterized over a mergeable type that

### 2017-02-28 paper/intro.tex

@@
 low-level operational details such as replication and consistency into
-high-level compositional abstractions over language-level datatypes.
+high-level compositional abstractions over language-level data types.
 

### 2017-02-28 paper/model.tex

@@
 versions of a drawing canvas (see Fig.~\ref{fig:merge-canvas}) that
-includes includes two concurrent versions (\C{v1} and \C{v2}),
+includes two concurrent versions (\C{v1} and \C{v2}),
 and their lowest common ancestor (\C{lca}) - the version from which the

### 2017-02-28 paper/mergeable-types.tex
edit

@@
 
-\noindent Rather than specifying merge ($M$) in terms of an LCA as we
+Rather than specifying merge ($M$) in terms of an LCA as we
 did in describing our programming model,

### 2017-02-28 paper/implementation.tex
URL

--- 2017-02-28 paper/implementation.tex
@@
 
-We describe the realization of \name on top of a persistent
+We describe the
+realization\footnote{\url{https://github.com/icfp2017/vml}} of \name on top of a persistent
 distributed store and the language extensions to enable and enforce

### 2017-02-28 paper/implementation.tex
Typos.

@@
   parameter inference that polymorphism offers can however be obtained
-  for module parameters using OCaml extensioms, such as Modular
+  for module parameters using OCaml extensions, such as Modular
   Implicits~\cite{implicits}.

### 2017-02-28 paper/meta.tex
Typos.

@@
 the last merged version of a branch its \emph{locus}. By maintaining a
-total order of among locii of a branch $b$ as new versions are
+total order of among loci of a branch $b$ as new versions are
 created, we effectively guarantee the presence of a version (the
@@
 versions. Since causal histories have maximum versions (respective
-locii), which, by the inductive hypothesis, can be made mergeable, we
+loci), which, by the inductive hypothesis, can be made mergeable, we
 merge them to yield a new version $v$ that is the maximum of the

### 2017-02-28 paper/examples.tex
missing ref

@@
 		has the same effect on $v_2$ as $s_1$ had on $v$.  For instance, in
-		Fig.~\ref{list-eg}, let $s_1$ be the edit sequence \C{[D(1); S(1,c,d)]},
+		Fig.~\ref{fig:list-eg}, let $s_1$ be the edit sequence \C{[D(1); S(1,c,d)]},
 		which maps the list \C{[a;b;c]} ($v$) to \C{[a;d]} ($v_1$). The \C{D} edit

### 2017-02-28 paper/meta.tex
Edits.

@@
 $v_j$) is smaller than the history that includes the unmergeable
-versions. Since causal histories have maximum versions (respective
-loci), which, by the inductive hypothesis, can be made mergeable, we
-merge them to yield a new version $v$ that is the maximum of the
-causal histories of both $v_i$ and $v_j$. If $v_i$ and $v_j$ are the
-latest versions on their respective branches, the branches are now
-mergeable.
+versions. As shown in Fig.~\ref{fig:causal-history}, causal
+histories have maximum versions (respective loci of $v_i$ and $v_j$),
+which, by the inductive hypothesis, can be made mergeable. We
+therefore can merge them to yield a new version $v$. Since $v$
+immediately follows the loci of both $v_i$ and $v_j$, and since both
+the versions are the latest versions on their respective branches, $v$
+can be merged into $v_i$ and $v_j$ to create new versions $v_i'$ and
+$v_j'$ on their respective branches, which have a single LCA - $v$.
+Thus the branches are now mergeable.
 

### 2017-03-15 paper/examples.tex
Fixed many errors.

@@
 
-The \drawsome example demonstrates the \name's utility in building
-concurrent applications with conceptual state sharing that make them
-suitable for transparent deployment in a distributed setting. The
+The \drawsome example demonstrates \name's utility in building
+concurrent applications with conceptual state sharing, which makes
+them suitable for transparent deployment in a distributed setting. The
 model allows concurrent computations to be composed around any
-mergeable type. We now present a series of examples that demonstrate
-how mergeable types are built - either by defining merge functions for
-ML data types, or by composing existing mergeable types.
+mergeable type. We now present a series of examples of mergeable types
+built either by defining merge functions for ML data types, or by
+composing existing mergeable types.
 
@@
 of operations such as concatenation, splitting, lookup and insertion
-at a random position etc., for a list of values in a distributed
-setting. The values are usually characters, with their lists being
-strings. The data structure is a binary tree, whose leaf nodes store
-substrings of a larger string such that their left-to-right (in-order)
-concatenation yields the larger string. The idea is to facilitate
-efficient lookups and insertions at random positions in the string by
-storing position indexes in the internal nodes of the tree.
+at a random position etc., for a list of values. The values are
+usually characters, their lists being strings. A rope is essentially a
+binary tree, whose leaf nodes store substrings of a larger string such
+that their left-to-right (in-order) concatenation yields the larger
+string. The idea is to facilitate efficient lookups and insertions at
+random positions in the string by storing position indexes in the
+internal nodes of the tree.
 
@@
 Fig.~\ref{fig:mrope-merge}. The function returns one of its versions
-(\C{l} or \C{r}), if the other one is the same as the common
-ancestor. Otherwise, it calls \C{merge\_rec}, which recursively
-compares the structures of merging ropes, and their common ancestor.
-The function recurses until one of the ropes differs from the other
-two, or all of them are leaf nodes, at which point it relies on the
-(abstract) list merge function (\C{A.merge}) to merge the lists
-represented by the ropes. \C{merge\_rec} uses auxiliary functions,
-\C{mk\_rope} and \C{create\_index}, whose definitions are not shown to
-construct a rope given a list (\C{A.t}), and to compute the index at a
-node, given its left sub-tree, resp.
+(\C{l} or \C{r}), if the other one is the same as the common ancestor.
+Otherwise, it calls \C{merge\_rec}, which recursively compares the
+structures of merging ropes, and their common ancestor.  The function
+recurses until one of the ropes differs from the other two, or all of
+them are leaf nodes, at which point it relies on the (abstract) list
+merge function (\C{A.merge}) to merge the lists represented by the
+ropes. \C{merge\_rec} uses auxiliary functions, \C{mk\_rope} and
+\C{create\_index} (definitions not shown), respectively, to construct
+a rope given a list (\C{A.t}), and to compute the index at a node
+given its left sub-tree.
 
-One feature of the mergeable rope implementation that is characteristic
-of mergeable types in general is compositionality; a rope is a
-polymorphic mergeable data type that composes mergeability over its
-polymorphic constituents.   Compositionality is concretized by the
-\C{MRope} signature , which requires the type of items in the list
+A notable feature of the mergeable rope implementation, which is also
+a characteristic of mergeable types in general, is compositionality. A
+rope is a polymorphic mergeable data type that composes mergeability
+over its polymorphic constituents.  Compositionality is concretized by
+the \C{MRope} signature, which requires the type of items in the list
 (\C{A.t}) to also be mergeable (\C{A.merge}).
@@
 
-As an abstract data type, lists support three operations: \C{insert x
-  i l}, to insert an element \C{x} at position \C{i} in the list
-\C{l}, \C{delete i l}, to delete the element at \C{i}'th position, and
-\C{subst i x l}, to substitute the element at the \C{i}'th position
-with \C{x}.  Lists of mergeable items are mergeable if they preserve
-the following properties:
+As an abstract data type, \C{list} supports three operations:
+\C{insert x i l}, to insert an element \C{x} at position \C{i} in the
+list \C{l}, \C{delete i l}, to delete the element at \C{i}'th position
+in \C{l}, and \C{subst i x l}, to substitute the element at the
+\C{i}'th position with \C{x} in \C{l}.  Lists of mergeable items are
+mergeable if they preserve the following properties:
 \begin{itemize}
@@
 		\C{D} edit must delete the second element from $v_2$, if it were to
-		have the same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
-		transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The function
-		\C{op\_transform} (shown in the Appendix) computes such
-		\emph{operational transformations} for an edit sequence $s_1$ w.r.t another
-		edit sequence $s_2$. A notable transformation rule is for a substitution
-		w.r.t another substitution at the same position. Since the substituting
-		items could be different, the function relies on the item merge function
-		(\C{A.merge}) to merge them into a single item. For instance, in
-		Fig.~\ref{fig:list-eg}, there are simultaneous substitutions to the item
-		\C{c} in $s_1$ and $s_2$; while $s_1$ substitutes it with \C{d}, $s_2$ does
-		it with \C{s}. The merged list therefore substitutes \C{c} with the merge
-		of \C{d} and \C{s}, defined as per the merge semantics of \C{A.t}.
+    have the same effect as it did on $v$. We say \C{D(1)} in $s_1$
+    has to be transformed w.r.t the sequence $s_2 =
+    \C{[I(c,0);S(3,c,s)]}$.  The function \C{op\_transform} computes
+    such \emph{operational transformations} for an edit sequence $s_1$
+    w.r.t another edit sequence $s_2$. A notable transformation rule
+    is for a substitution w.r.t another substitution at the same
+    position. Since the substituting items could be different, the
+    function relies on the item merge function (\C{A.merge}) to merge
+    them into a single item. For instance, in Fig.~\ref{fig:list-eg},
+    there are simultaneous substitutions to the item \C{c} in $s_1$
+    and $s_2$; while $s_1$ substitutes it with \C{d}, $s_2$ does it
+    with \C{s}. The merged list therefore substitutes \C{c} with the
+    merge of \C{d} and \C{s}, defined as per the merge semantics of
+    \C{A.t}.
 \end{itemize}
@@
 
+\begin{figure}
+\centering
+\begin{tabular}{l||l}
+\begin{ocaml}
+let alice_f : L.t unit t =
+  fork bob_f >>= fun () -> (* Invite Bob *)
+  get () >>= fun c0 ->
+  let qty = L.assoc "eggs" in
+  let c0' = L.subst 1 ("eggs",qty+1) c0 in
+  sync () ~v:c0' >>=
+  fun c1 -> return ()
+\end{ocaml}
+&
+\begin{ocaml}
+let bob_f : L.t unit t =
+  get () >>= fun c0 ->
+  let c0' = L.insert ("candy",Qty.of_int 1) 1 @@
+            L.delete @@ L.lookup "milk" in
+  sync () ~v:c0' >>=
+  fun c1 -> return ()
+\end{ocaml}
+\\
+\end{tabular}
+\caption{Bob and Alice collaboratively build a grocery shopping list}
+\label{fig:shopping-list-code}
+\end{figure}
+
 A non-trivial demonstration illustrating the compositional power of
@@
 and Bob's quantity as per the definition of \C{Qty.merge}.
-The examples described above make certain assumptions about the
-concurrency model and the underlying system, such as the existence of
-a single lowest common ancestor (LCA) for any pair of versions, the
+
+The examples described in this section make certain assumptions about
+the concurrency model and the underlying system, such as the existence
+of a single lowest common ancestor (LCA) for any pair of versions, the
 ability to access a previous version on any branch, and mergeability
@@
 merges, are discussed in subsequent sections.
-
-\begin{figure}
-\centering
-\begin{tabular}{l||l}
-\begin{ocaml}
-let alice_f : L.t unit t =
-  fork bob_f >>= fun () -> (* Invite Bob *)
-  get () >>= fun c0 ->
-  let qty = L.assoc "eggs" in
-  let c0' = L.subst 1 ("eggs",qty+1) c0 in
-  sync () ~v:c0' >>=
-  fun c1 -> return ()
-\end{ocaml}
-&
-\begin{ocaml}
-let bob_f : L.t unit t =
-  get () >>= fun c0 ->
-  let c0' = L.insert ("candy",Qty.of_int 1) 1 @@
-            L.delete @@ L.lookup "milk" in
-  sync () ~v:c0' >>=
-  fun c1 -> return ()
-\end{ocaml}
-\\
-\end{tabular}
-\caption{Bob and Alice collaboratively build a grocery shopping list}
-\label{fig:shopping-list-code}
-\end{figure}

### 2017-03-15 paper/intro.tex
Fixed many errors.

@@
 To illustrate how the kinds of problems described above manifest in
-practice, consider how we might write a simple counter library (see
-Fig.~\ref{fig:counter-adt}).  A \C{Counter} supports two (update)
-operations - \C{add} and \C{mult} - that lets a non-negative integer
-value be added or multiplied to the counter, resp.  Observe that the
-library is written in an idiomatic functional style, with no special
-reasoning principles needed to realize desired functionality.  As long
-as applications use the library on a single machine, this
+practice, consider how we might write a simple monotonic counter
+library (see Fig.~\ref{fig:counter-adt}).  A \C{Counter} supports two
+(update) operations - \C{add} and \C{mult} - that lets a non-negative
+integer value be added or multiplied to the counter, resp.  Observe
+that the library is written in an idiomatic functional style, with no
+special reasoning principles needed to realize desired functionality.
+As long as applications use the library on a single machine, this
 implementation behaves as expected.  However, if the library is used
@@
 A method often adopted to address these concerns is to re-define a
-datatype's operations to return \emph{effects} instead of
+data type's operations to return \emph{effects} instead of
 values~\cite{SPB+11,Burckhardt2014}.  An \emph{effect} is a tag that
@@
 operation. Fig.~\ref{fig:counter-rdt} shows the \C{Counter} library
-with operations re-defined to returns effects.  Observe that the
+with operations re-defined to return effects.  Observe that the
 \C{Counter.add x v} operation now returns an \C{Add x} effect, which,
@@
 rather than file, coherence.  The \name (Versioned ML) programming
-model comprises the {\sc vst} monad which lets programmers write and
-compose concurrent computations around multiple (implicit) versions of
-a mergeable datatype, an ordinary ML datatype additionally equipped
-with a \emph{merge} function responsible for deriving a consistent
-global state from a collection of local versions of that state.  A
-computation progresses by forking (i.e., replicating) new versions of
-existing versions, allowing synchronization-free local computation to
-proceed on those nodes, creating new versions along an existing
-\emph{branch} that represents updated local instances, and by merging
-branches to realize global consistency.
+model comprises the \nameMonad (Versioned State) monad which lets
+programmers write and compose concurrent computations around multiple
+(implicit) versions of a mergeable datatype, an ordinary ML datatype
+additionally equipped with a \emph{merge} function responsible for
+deriving a consistent global state from a collection of local versions
+of that state.  A computation progresses by forking (i.e.,
+replicating) new versions of existing versions, allowing
+synchronization-free local computation to proceed on those nodes,
+creating new versions along an existing \emph{branch} that represents
+updated local instances, and by merging branches to realize global
+consistency.
 
@@
     \item We describe \name, a programming model that extends ML to
-      brings to bear the power and flexibility of a version control
+      bring to bear the power and flexibility of a version control
       system to the administration of replicated data. \name hides the
@@
     \item We present a formalization of the \name programming model,
-      and establish the conditions under which eventual convergence of
-      concurrent version state can be guaranteed.
+      and establish its soundess (consistency) and progress guarantees.
 
@@
 
-\noindent The remainder of the paper is organized as follows.  We present the
-\name\ programming model informally in the next section.
-Sec.~\ref{sec:mergeable_types} develops a series of examples that illustrate
-how merge operations can be defined and used. Sec.~\ref{sec:formalization}
-defines an operational semantics, and formalizes our intuition of a mergeable
-type in terms of morphisms over datatype operations.  We also present soundness
-(consistency) and progress guarantees enjoyed by the semantics.
-Sec.~\ref{sec:implementation} presents an instantiation of the model suitable
-for highly-available distributed environments with network partitions and
-failures. Experimental results and benchmarks are given in
-Sec.~\ref{sec:evaluation}.  Secs.~\ref{sec:related} and~\ref{sec:conclusions}
-describes related work and conclusions.
+\noindent The remainder of the paper is organized as follows.  We
+present the \name\ programming model informally in the next section.
+Sec.~\ref{sec:mergeable_types} develops a series of examples that
+illustrate how merge operations can be defined and composed for
+non-trivial data types.  Sec.~\ref{sec:formalization} defines an
+operational semantics, and formalizes our intuition of a mergeable
+type in terms of morphisms over datatype operations.  We also present
+soundness (consistency) and progress guarantees enjoyed by the
+semantics.  Sec.~\ref{sec:implementation} presents an instantiation of
+the model suitable for highly-available distributed environments with
+network partitions and failures. Experimental results and benchmarks
+are given in Sec.~\ref{sec:evaluation}.  Secs.~\ref{sec:related}
+and~\ref{sec:conclusions} describes related work and conclusions.
 

### 2017-03-15 paper/meta.tex
Fixed many errors.

@@
 without which \C{merge} may yield unexpected results. This is
-demonstrated for the grow-only counter in
+demonstrated for the monotonic counter in
 Fig.~\ref{fig:merge-needs-lca}, where an incorrect count is obtained if a
@@
 \includegraphics[scale=0.6]{Figures/merge-needs-lca}
-\caption{This example of a grow-only counter illustrate why \C{merge}
+\caption{This example of a monotonic counter illustrate why \C{merge}
 needs a lowest common ancestor, and not just a common ancestor. Both 0

### 2017-03-15 paper/model.tex
Fixed many errors.

@@
 \begin{figure}
-\begin{center}
+\centering
   \begin{ocaml}
@@
       | N of pixel
-      | B of {tl: tree; tr: tree; bl: tree; br: tree}
+      | B of {tl_t: tree; tr_t: tree; bl_t: tree; br_t: tree}
     type t = {max_x:int; max_y:int; canvas:tree}
@@
   \end{ocaml}
-\end{center}
 \caption{\drawsome: a sample \name application}
@@
 application. \drawsome represents a free-hand drawing canvas in terms
-of a tree of quadrants.  Each quadrant is simply a leaf node
-containing a single pixel (\C{r},\C{g}, or \C{b}) or a tree, if the
-quadrant contains multiple pixels of different colors. Quadrants are
+of a tree of quadrants.  A quadrant is either a leaf node containing a
+single pixel (an \C{r}-\C{g}-\C{b} tuple), or a tree of sub-quadrants,
+if the quadrant contains multiple pixels of different colors.
+Quadrants are
 \begin{wrapfigure}{l}{.4\textwidth}
@@
 around versioned data structures.  Fig.~\ref{fig:dali-monad} shows the
-signature of the \nameMonad module that implements the programming
+signature of the \nameMonad monad that implements the programming
 model along the lines of the well-known \C{State}
@@
   \begin{ocaml}
-  module type VML = sig
+  module type VST = sig
     type ('a, 'b) t
@@
   \end{ocaml}
-  \caption{Signature of the \name monad}
+  \caption{Signature of the Versioned State (\nameMonad) monad}
   \label{fig:dali-monad}
@@
 last merge or fork. Thus, \C{sync} effectively lets a computation
-synchronize with a subset of concurrent computations to obtain their
-latest updates.
+synchronize with concurrent computations to obtain their latest
+updates.
 
@@
 Cheryl's $C_0'$, and the LCA is the initial version ($C_0$). The
-result of this merge is installed as the next version ($\C{V_1}$) on
+result of this merge is installed as the next version ($V_1$) on
 Alice's branch. For the next merge, the two concurrent versions are

### 2017-03-15 paper/semantics.tex
Fixed many errors.

@@
 program state ($p;\,H$) consists of a program $p$ and a \emph{branch
-  history} $H$ that maps thread identifiers to corresponding branches;
-each thread is associated with a branch during evaluation. Evaluation
+history} $H$ that maps thread identifiers to corresponding branches.
+Each thread is associated with a branch during evaluation. Evaluation
 contexts have been defined separately for expressions ($E$) and
@@
 not lead to a case where a branch (i.e., its latest version) has
-multiple LCAs with remaining branches, andd hence gets
+multiple LCAs with remaining branches, and hence gets
 \emph{stuck}. Such nuances of LCA, and the guarantees provided by the

### 2017-04-19 paper/icfp17_rebuttal.md
ICFP Rebuttal.

We thank the reviewers for their extensive reviews and thoughtful
suggestions.

1. **Formalization** We agree with reviewer A in that disparate
   formalisms to capture various aspects of VML is far from ideal. Our
   aim in building these formalizations is three-fold: (a). To give
   semantics to the VML primitives introduced in Sec. 2 and 3, (b). To
   show that the side condition on merge makes three-way merges
   possible by enforcing the unique LCA property, and (c). To prove
   that the side condition does not affect the progress guarantee of
   the system, i.e, every pair of branches can eventually satisfy the
   side condition, allowing them to merge (progress theorem is the
   relation between the runs of the machine and structure of the
   history graphs). We appreciate reviewer's suggestion to drop the
   abstract machine. Perhaps an inductively defined graph structure
   should suffice to prove the properties of interest.

2. **Formal model vs the implementation** Formal model does have a bearing
   on the implementation. The ring deployment described in Fig. 21(a),
   where merges happen in a circular order, satisfies the side
   condition on merges by construction.

2. **Comparison with Concurrent Revisions (CRs)** Intuitively, the set
   of history graphs allowed under our model is a strict superset of
   revision diagrams. For instance, CR allows a branch to only merge
   with its parent, while VML allows merges between arbitrary
   branches. This generalization is useful in practice, since it
   admits a "server-less" fully decentralized system design required
   for highly-available web services. Both CR and VML admit three-way
   merges. CR gives a determinism guarantee, which is desirable in a
   concurrency model, but not so much in a distributed programming
   model since it is impossible to achieve without coordination
   (extensions of CR for distributed systems (ESOP'12, ECOOP'12)
   eschew determinism). Thus, CR for distributed systems is a
   specific instance of VML. We accept reviewer B's suggestion to
   include a performance comparison with CR.

### 2017-04-19 paper/icfp17_rebuttal.md
Edit.
EDits.

@@
 
-2. **Comparison with Concurrent Revisions (CRs)** Intuitively, the set
+3. **Comparison with Concurrent Revisions (CRs)** Intuitively, the set
    of history graphs allowed under our model is a strict superset of
@@
    include a performance comparison with CR.
+
+4. **Associativity** In the example with associative rope structures,
+   reviewer A is right in observing that the element `y` is duplicated
+   only if it is added independenty on two merging branches (can be
+   determined using the history). We understand reviewer C's intuition
+   about the need for associativity. However, associativity
+   requirement is waived in VML due to the system-level guarantees,
+   which only allow the latest states on the branches to be merged (no
+   criss-cross merges in the graph structure). Thus, if `a` is merged
+   onto `b`, then `b` can be merged onto `c` only if `a` is merged
+   onto `c`. The result of this merge (`a → b`, then `a → c`, then `b
+   → c`) is `a/2 + b/4 + c/4`. Subsequent merge from `c` to `b` would
+   be a fast-forward merge (no modification on `c`), replacing the
+   state on `c` with `a/2 + b/4 +c/4`.

### 2017-05-31 mlw2017/abstract.tex
ML Workshop submission

\documentclass[twocolumn,9pt]{extarticle}
\usepackage{fullpage}
\usepackage{xspace}
\usepackage{textcomp}
\usepackage{url,graphicx,color,subcaption,multirow}
\newcommand\cod[1]{{\small\sf\bfseries{#1}}}
\newcommand\xcod[1]{{\small\sf{#1}}}
\usepackage[lmargin=2cm,rmargin=2cm,tmargin=1cm,bmargin=2cm]{geometry}

\newenvironment{titemize}
{\begin{itemize}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
}
{\end{itemize}}

\newcommand{\etal}[1]{et al.\ #1}

\newcommand{\TODO}[1]{{\color{red}TODO:} #1}

\newcommand{\intboxskip}{\medskip}
\newcommand{\genfrag}[2]
{
        \intboxskip{}
        {
        \sffamily
        \small
        \input{code/#1-#2.tex}
        }
}

\newcommand{\genfragapp}[2]{
\begin{small}
{\sffamily \input{code/#1-#2.tex}}
\end{small}
}

\usepackage[usenames,dvipsnames]{xcolor}

\newcommand{\ocamlfrag}[1]{\genfrag{ml}{#1}\vspace{2mm}}

%\usepackage[compact]{titlesec}
%\titlespacing{\section}{0pt}{1ex}{1ex}
%\titlespacing{\subsection}{0pt}{1ex}{0.5ex}
%\titlespacing{\subsubsection}{0pt}{1ex}{0.5ex}
%\titlespacing{\paragraph}{1em}{0ex}{1ex}
%\setlength{\parskip}{0cm}
%\setlength{\parindent}{1em}

\newcommand{\tick}{\textquotesingle}

\usepackage{titling}

%\setlength{\droptitle}{-15mm}   % This is your set screw

\definecolor{mauve}{rgb}{0.58,0,0.82}

\usepackage{listings}

\lstdefinestyle{sOcaml}{language=[Objective]Caml,
  literate={+}{{$+\:$}}1 {/}{{$/$}}1 % { * }{{$*$}}1
           {=}{{$=$}}1
           {>}{{$>$}}1 {<}{{$<$}}1
           {<>}{$\not=\ $}1
           {->}{{$\rightarrow$} }2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
           {<=}{{$\leq$}}2
           {=>}{{$\Rightarrow$}}2
           {+->}{{$\hookrightarrow$}}2
           {==>}{{$\mapsto$}}2
           {fn}{$\lambda$}1
           {|}{{$\mid$}}1
           {'a}{$\alpha$ }1
           {+'a}{$\textrm{+}\alpha$}1
           {'b}{$\beta$}1
           {+'b}{$\textrm{+}\beta$}1
           {'c}{$\gamma$}1
           {'e}{$\epsilon$}1
           {'w}{$\omega$}1
           {'w.}{$\forall\omega.\ $}2
           {t1}{t$_1$}2
           {t2}{t$_2$}2
           {vO}{v$_o$}1
           {vV}{v$_v$}1
           {\ .\ }{$\;\circ\;$}1
           {TRB}{\mbox{\ensuremath\lceil}}1
           {TRE}{\mbox{\ensuremath\rceil}}1
           {...}{\ldots}2
           %% {\#\#+}{\color{dred}}1
           %% {\#\#*}{\color{dgreen}}1
           %% {\#\#-}{\color{black}}1
           {\#\#\#}{{$\leadsto$}}3
}

\definecolor{weborange}{RGB}{255,165,0}
\definecolor{frenchplum}{RGB}{129,20,83}
\definecolor{eminence}{RGB}{108,48,130}
\definecolor{commentgreen}{RGB}{2,112,10}
\definecolor{mauve}{rgb}{0.88,0.69,1.0}

\lstset{
	style=sOCaml,
  basicstyle=\small\ttfamily,
  commentstyle=\color{OliveGreen},
  escapeinside={(**}{)},
  keywordstyle=\color{blue},
  language=[Objective]Caml,
	emph={Async,Await,Yield,Done,Error,Waiting,Accept,Recv,Send,Delayed,Completed,Break,Signal_handle,Some,None,TimerTick},
	emphstyle={\color{weborange}},
  morekeywords={effect,continue,discontinue,perform,finally},
  stringstyle=\color{mauve},
  showstringspaces=false,
  breaklines=true,                    % allows line breaks
  tabsize=2,
}

% TODOs
% -----
\newcommand{\todo}[1]{{\par\noindent\small\color{red} \framebox{\parbox{\dimexpr\linewidth-2\fboxsep-2\fboxrule}{\textbf{TODO:} #1}}}}
\newcommand{\kc}[1]{{\par\noindent\small\color{red} \framebox{\parbox{\dimexpr\linewidth-2\fboxsep-2\fboxrule}{\textbf{KC:} #1}}}}

% Name
% ----
\newcommand{\name}{{\sc vml}\xspace}
\newcommand{\drawsome}{{\sf Canvas}\xspace}

\lstMakeShortInline[columns=fullflexible]|

\title{Mergeable Types}
\author{Gowtham Kaki \\
        Purdue University
        \and
				KC Sivaramakrishnan \\
        University of Cambridge
				\and
				Suresh Jagannathan \\
				Purdue University
				\and
				Samodya Abeysiriwardane \\
				Purdue University}
\date{~}

\newcommand{\cf}[1]{{\small\tt #1}}

\begin{document}

\maketitle

\emph{
	Distributed applications often eschew strong consistency and replicate data
	asynchronously to improve availability and fault tolerance. However,
	programming under eventual consistency is significantly more complex and
	often leads to onerous programming model where inconsistencies must be
	handled explicitly. We introduce \name, a programming model that extends ML
	datatypes with mergeability \`{a} la version control systems with the ability
	to define and compose distributed ML computations around such data. Our OCaml
	implementation instantiates mergeable types on Irmin, a distributed
	content-addressible store to enable composable and highly-available
	distributed applications.
}

\section{A Replicated Counter}

Consider a monotonic counter data type:

\begin{lstlisting}
	module Counter: sig
		type t
		val add: int -> t -> t
		val mult: int -> t -> t
		val read: t -> int
	end = struct
		type t = int
		let add x v = v + (abs x)
		let mult x v = v * (abs x)
		let read v = v
	end
\end{lstlisting}

Observe that the library is written in an idiomatic functional style, with no
special reasoning principles needed to realize desired functionality. As long
as applications use the library on a single machine, this implementation
behaves as expected.  However, if the library is used in the context of a more
sophisticated application, say one whose computation is distributed among a
collection of machines, its behavior can become significantly harder to
understand.  In particular, a distributed implementation might wish to
\emph{replicate} the counter state on each replica to improve response time or
fault tolerance.  Unfortunately, adding replication doesn't come for free.
Attempting to update every replicated copy atomically is problematic in the
absence of distributed transaction support, which impose significant
performance penalties.  But, without such heavyweight mechanisms, applying an
|Add| operation on one replica may not be instantaneously witnessed on another,
which may be in the process of simultaneously attempting to perform its own
|Add| or |Mult| action.

Rather than viewing each operation in terms its effect on the global state, can
we formulate a more declarative interpretation, directly in terms of the
counter value maintained by each replica? Since the counter is replicated, each
local operation can be thought of as yielding a new local version, collectively
producing a version tree, with one branch for each replica.  Every branch
represents different (immutable) versions maintained by different replicas, with
the state produced by the computation performed over a counter on a replica
recorded along the replica's local branch for the counter.  Now, to generate a
globally consistent view of a counter, we only need to define a merge operation
that explains how to combine two local versions to produce a new version that
reflects both their states. This operation is defined not in terms of replicas
or other system-specific artifacts, but in terms of the semantics of the
datatype itself.

Framing replication as merging leads to a counter implementation that bears
strong similarity to the original sequential one:
\begin{lstlisting}
	module Replicated_Counter = struct
		include Counter
		let merge lca v1 v2 =
				lca + (v1 - lca) + (v2 - lca)
	end
\end{lstlisting}
The role of |lca| (lowest common ancestor) here captures salient
history - the state resulting from the merge of two versions derived
from the same ancestor state should not unwittingly duplicate the
contributions of the ancestor. This interpretation of a replicated
datatype is thus given in terms of the evolution of a program state
implicitly associated with the different replicas that comprise a
distributed application with merge operations serving to communicate
and reconcile different local states.

\section{Collaborative drawing}

\name not only supports primitive data types but also algebraic data types.
This code snippet:

\begin{lstlisting}
module type CANVAS = sig
	type pixel = {r:char; g:char; b:char}
	type tree =
	| N of pixel
	| B of {tl_t:tree; tr_t:tree; bl_t:tree; br_t:tree}
	type t = {max_x:int; max_y:int; canvas:tree}
	type loc = {x:int; y:int}

	val new_canvas: int -> int -> t
	val set_px: t -> loc -> pixel -> t
	val get_px: t -> loc -> pixel
	val merge: (*lca*)t -> (*v1*)t -> (*v2*)t -> t
end
\end{lstlisting}

\noindent shows the signature of the \drawsome application. \drawsome
represents a free-hand drawing canvas in terms of a tree of quadrants.  A
quadrant is either a leaf replica containing a single pixel (an |r|-|g|-|b|
tuple), or a tree of sub-quadrants, if the quadrant contains multiple pixels of
different colors. Quadrants are expanded into a tree structures as and when
pixels are colored.  The representation is thus optimized for sparse canvases,
such as whiteboards. The application supports three simple operations: creating
a new canvas, setting the pixel at a specified coordinate, and returning the
pixel at a given coordinate.

\drawsome lets multiple users collaborate on a canvas that is conceptually
shared among them. Under a shared-memory abstraction, there would be a single
copy of the canvas that is updated concurrently by multiple clients; from the
perspective of any single client, the canvas could change without any explicit
intervention. \name ascribes functional semantics to sharing by letting each
client work on its own version of the state (the tree data structure in this
example), later merging concurrent versions on-demand.

\name requires a three-way |merge| function to merge concurrent versions of a
drawing canvas that includes two concurrent versions (|v1| and |v2|), and
their lowest common ancestor (|lca|) - the version from which the two
concurrent versions evolved independently.

\begin{lstlisting}
let color_mix px1 px2 : pixel =
	let f = Char.code in
	let h x y = Char.chr @@ (x + y)/ 2 in
	let (r1,g1,b1) = (f px1.r,f px1.g,f px1.b) in
	let (r2,g2,b2) = (f px2.r,f px2.g,f px2.b) in
	let (r,g,b) = (h r1 r2,h g1 g2,h b1 b2) in
	{r=r; g=g; b=b}

let b_of_n px =
	B {tl_t=N px; tr_t=N px; bl_t=N px; br_t=N px}

let rec merge lca v1 v2 =
	if v1=v2 then v1
	else if v1=lca then v2
	else if v2=lca then v1
	else match (lca,v1,v2) with
	| (_,B _,N px2) -> merge lca v1 @@ b_of_n px2
	| (_, N px1, B _) -> merge lca (b_of_n px1) v2
	| (N px, B _, B _) -> merge (b_of_n px) v1 v2
	| (B x, B x1, B x2) ->
		let tl_t = merge x.tl_t x1.tl_t x2.tl_t in
		let tr_t = merge x.tr_t x1.tr_t x2.tr_t in
		let bl_t = merge x.bl_t x1.bl_t x2.bl_t in
		let br_t = merge x.br_t x1.br_t x2.br_t in
		B {tl_t; tr_t; bl_t; br_t}
	| (_, N px1, N px2) ->
		(* pixels are merged by mixing colors *)
		let px' = color_mix px1 px2 in N px'
\end{lstlisting}

The merge function can make use of the pixel values of the common ancestor to
merge the pixel values on both the canvases. For instance, if the color of a
pixel in |v1| is white, and in |v2| it is green, and its color in |lca|
is white, then it means that only |v2| modified the color. Hence the pixel is
colored green in the merged canvas. On the other hand, if the pixel is red in
|v1|, then it means that both |v1| and |v2| have modified the color. In
such case, an appropriate color-mixing algorithm can be used to determine the
color of pixel.  For instance, the pixel can be colored yellow - an additive
combination of red and green. The logic is illustrated below.

\begin{center}
\includegraphics[scale=0.4]{Figures/canvas-merging}
\end{center}

We have built several mergeable datatypes including lists, ropes, etc, which
can be freely composed together. That is, a list of counters behaves like a
mergeable list for append and remove operations, with updates reconciled
through counter merge semantics.

\section{Distributed Instantiation}

The \name programming model is realized on top of Irmin~\cite{irmin}, an OCaml
library database implementation that is part of the MirageOS
project~\cite{mirage}. Irmin provides a persistent multi-versioned store with a
content-addressable heap abstraction. Simply put, content-addressability means
that the address of a data block is determined by its content. If the content
changes, then so does the address. Old content continues to be available from
the old address. Content-addressability also results in constant time
structural equality checks, which we exploit in our mergeable rope
implementation, among others.

Irmin provides support for distribution, fault-tolerance and concurrency
control by incorporating the Git distributed version control~\cite{git}
protocol over its object model. Indeed, Irmin is fully compatible with Git
command line tools. Distributed replicas in \name are created by cloning a
\name repository. Due to \name's support for mergeable types, each replica can
operate completely independently, accepting client requests, even when
disconnected from other replicas, resulting in a highly available distributed
system.

While Irmin's merge functions are defined over objects on Irmin's
content-addressable heap, \name's merge functions are defined over OCaml types.
We address this representational mismatch with the help of OCaml's PPX
metaprogramming support~\cite{ppx} to derive bi-directional transformations
between objects on OCaml and Irmin heaps. We also derive the various
serialization functions required by Irmin

\begin{figure}[t]
  \centering
	\begin{subfigure}{0.45\columnwidth}
		\includegraphics[width=0.9\textwidth]{Figures/cluster.pdf}
		\caption{{\small Our experimental configuration consists of an 8-node ring cluster
			executing on Google Cloud Platform. Edge labels are inter-node latencies in milliseconds.}}
		\label{fig:cluster}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.5\columnwidth}
		\includegraphics[width=\textwidth]{Figures/scalability.pdf}
		\caption{{\small Scalability: Overall throughput of the cluster and latency of each
		operation.}}
		\label{grf:scalability}
	\end{subfigure}
	\caption{\name performance evaluation.}
  \label{grf:evaluation}
	\vspace{-1em}
\end{figure}

We evaluated the performance of the system on a collaborative application that
simulates concurrent editing of the same document by several authors. The
benchmark itself was constructed with a list of ropes. The workload consists of
4000 edit operations at random indices with 85\% insertions and 15\% deletions.
We evaluate the scalability of concurrent editing application by increasing the
cluster size from 1 to 8 (the 4 node ring cluster consists of nodes numbered 0
to 3), with each node performing concurrent edits to the same document. In each
case, we measure the overall cluster throughput and latency of each operation.
The results are presented in Figure~\ref{grf:scalability}. The results show
that the cluster throughput increases linearly with the number of concurrent
editors, while the latency for each operation remains the same. This is because
each operation is performed locally and does not require synchronization with
other nodes. The nodes remain available to accept requests even if the node
gets disconnected. Since the document type is mergeable, eventually when the
node comes back online, the updates are synchronized with the cluster.

\clearpage

\bibliographystyle{abbrv}
\small
\bibliography{all}

\end{document}
